---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.width = Inf)
```

# Tibble now (tbl.now)

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/RodrigoZepeda/tbl.now/graph/badge.svg)](https://app.codecov.io/gh/RodrigoZepeda/tbl.now)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/tbl.now)](https://CRAN.R-project.org/package=tbl.now)
[![R-CMD-check](https://github.com/RodrigoZepeda/tbl.now/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RodrigoZepeda/tbl.now/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The **tbl.now** package contains a `data.frame` class that extends the [tibble](https://tibble.tidyverse.org/reference/tibble-package.html) (and, hence, the [tidyverse](https://tidyverse.org/)) to nowcasting data. The main purpose of the package is to provide a unified input within the [`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) framework. 

## Installation

You can install the development version of tbl.now from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("RodrigoZepeda/tbl.now")
```

And after installation: 

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(tbl.now)
library(almanac)    #Suggested for holiday effects
```

## Why tbl_now?

Traditionally in epidemiological nowcasting scenarios we have two dates: 

* `event_date`: When something happened (e.g. symptom onset or a test was taken).
* `report_date`: When it was reported (e.g. the patient went to the clinic or the test results were registered). 

The nowcasting problem is to estimate the total number of events **now** that have occurred at any past `event_date` given that not all of them have been reported yet. In the context of nowcasting, the **tbl_now** can be thought of as a specific [tibble()](https://tibble.tidyverse.org/reference/tibble.html) that **guarantees** an `event_date` and a `report_date`.


## Example

Consider the following `data.frame` representing the number of cases `n`  reported at `report_date` that happened at `event_date`:

```{r}
df <- tibble(
  event_date  = c(ymd("2023/12/25"), ymd("2023/12/26"), ymd("2023/12/25"), ymd("2023/12/26")),
  report_date = c(ymd("2023/12/26"), ymd("2023/12/26"), ymd("2023/12/27"), ymd("2023/12/27")),
  n = c(10, 2, 5, 11)
)

df
```

To convert to a `tbl_now` you can use the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) function: 

```{r}
df_now <- tbl_now(df, event_date = "event_date", report_date = "report_date")
df_now
```

The [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) automatically detects whether the **data-type** corresponds to _linelist_, _count-incidence_ or _count-cumulative_ data, the **date-units** for the event and report dates (frequency), and the **now** is given by the latest date (`r get_now(df_now)`). Additionally it transforms the `event_date` into numeric (`.event_num` column) as well as the `report_date` (`.report_num` column) and calculates the delay (`.delay` column). 

The [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) is compatible with the usual [dplyr](https://dplyr.tidyverse.org/) operations:

```{r}
df_now %>% 
  filter(n > 5)
```

If strata was given, the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) can easily tag the corresponding strata. 

```{r}
#Add the column using dplyr:
df_now <- df_now %>% 
  mutate(sex = c("M","M","F","M")) 

df_now
```

To specify strata you can use the `add_strata`:

```{r}
df_now %>% 
  add_strata("sex")
```

## Temporal effects

Temporal effects can be added as covariates of the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)  using the [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html). 

For example we can specify it includes the day of the week, the week of the year, and whether it is holiday in the US:

```{r}
t_eff <- temporal_effects(
  day_of_week  = TRUE,
  week_of_year = TRUE, 
  holidays     = cal_us_federal())
t_eff
```
Such effects can be added to the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) object with the [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html) function:

```{r}
df_now %>% 
  add_temporal_effects(t_eff)
```

Note that Christmas (`2023-12-25`) is marked as an `.event_holiday`, everything corresponds to the first epidemiological week of `2024`, and the days of the week correspond to Monday (`event_day_of_week = 2`) and Tuesday (`event_day_of_week = 3`). All of these effects can be used as covariates in the models. 


## Learn more about tbl.now 

+ Read the whitepaper on [tbl.now](https://rodrigozepeda.github.io/tbl.now/articles/Introduction.html)
+ Read an [example](https://rodrigozepeda.github.io/tbl.now/articles/Example.html) with real data. 


