---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Tibble now (tbl.now)

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/RodrigoZepeda/tbl.now/graph/badge.svg)](https://app.codecov.io/gh/RodrigoZepeda/tbl.now)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/tbl.now)](https://CRAN.R-project.org/package=tbl.now)
[![R-CMD-check](https://github.com/RodrigoZepeda/tbl.now/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RodrigoZepeda/tbl.now/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of tbl.now is to provide a unified `tibble`-like interface for nowcasting algorithms within the [`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) framework. 

## Installation

You can install the development version of tbl.now from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("RodrigoZepeda/tbl.now")
```

## Class

The `tbl.now` package contains the `tbl.now` class which extends the `tbl` class from ([`tibble()`](https://tibble.tidyverse.org/)). Nowcasts require an `event_date` and a `report_date` thus the`tbl.now` class can be thought of as a specific `tibble` that **guarantees** an `event_date` and a `report_date`. It also **tags** variables as: covariates, strata or as a batched indicator. Finally it converts event and report dates to numeric. 

Here is an example with a fake linelist data.frame where cases have happened at `symptom_onset` but have been reported at `report_time``. The tibble includes `sex` and `age_group` as covariate, `state` as strata and an unused variable called `comorbidity`. 

```{r}
library(tbl.now)
library(dplyr, quietly = TRUE)

#This is what the dataframe looks like
df <- tibble(
  patient_id  = 1:5,
  symptom_onset = c(as.Date("2020/12/01"), as.Date("2020/12/01"), as.Date("2020/12/01"), as.Date("2020/12/02"), as.Date("2020/12/02")),
  report_time   = c(as.Date("2020/12/01"), as.Date("2020/12/01"), as.Date("2020/12/02"), as.Date("2020/12/02"), as.Date("2020/12/03")),
  sex           = c("M", "F", "F", "M", "F"),
  age_group     = c("20-40","40-60","20-40","40-60","40-60"),
  state         = c("NY","NJ", "NY","NY","NY"),
  comorbidity   = c(T,F,F,F,T)
)

df
```


and transform it to `tbl_now` involves specifying the date of the event, the date of the report,
the strata and the covariates. 

```{r}
vdat <- tbl_now(df,
                event_date  = "symptom_onset",
                report_date = "report_time", 
                covariate   = c("sex", "age_group"),  
                strata      = "state")     

vdat
```

The algorithm automatically infers it corresponds to `linelist` data (vs count) and infers that the now is `r get_now(vdat)` (the last date). The usual `dplyr` operations can be performed on the dataset with it adjusting on the fly. For example, removing the `age_group` removes it from the covariate list at the bottom:

```{r}
vdat %>% select(-age_group)
```

While changing the structure in a way that removes the event and report dates returns a `tibble`:

```{r}
vdat %>% 
  group_by(sex) %>% 
  tally()
```








