---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.width = Inf)
```

# Tibble now (tbl.now)

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/RodrigoZepeda/tbl.now/graph/badge.svg)](https://app.codecov.io/gh/RodrigoZepeda/tbl.now)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/tbl.now)](https://CRAN.R-project.org/package=tbl.now)
[![R-CMD-check](https://github.com/RodrigoZepeda/tbl.now/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RodrigoZepeda/tbl.now/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

[`tbl.now`](https://rodrigozepeda.github.io/tbl.now/) provides an extension of the [`tibble()`](https://tibble.tidyverse.org/) for storing, validating, and manipulating epidemiological nowcasting data. It standardizes the representation of event dates, report dates, strata, temporal covariates, and data types (linelist and cumulative), ensuring that downstream models within the [`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) ecosystem can rely on a consistent interface.



## Installation

Install the development version from [GitHub](https://github.com/):

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("RodrigoZepeda/tbl.now")
```

Load the package:

```{r, message=FALSE, warning=FALSE}
library(dplyr, quietly = TRUE)
library(lubridate)
library(tbl.now)
library(almanac)    #Suggested for holiday effects
```

## A minimal example

Suppose you have a dataset where n cases reported on report_date belong to events occurring on event_date:

```{r}
df <- tibble(
  event_date  = c(ymd("2023-12-25"), ymd("2023-12-26"),
                  ymd("2023-12-25"), ymd("2023-12-26")),
  report_date = c(ymd("2023-12-26"), ymd("2023-12-26"),
                  ymd("2023-12-27"), ymd("2023-12-27")),
  n = c(10, 2, 5, 11)
)
```
Convert it to a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html): 

```{r}
df_now <- df %>% 
  tbl_now(event_date = event_date, report_date = report_date, case_count = n)

df_now
```

`tbl_now()` automatically:

+ infers that the **data type** is linelist (one row per event)

+ determines the **date units** (daily event and report frequencies)

+ computes numerical versions of the dates: `.event_num`, `.report_num`, and `.delay`

+ sets the **now** to the most recent reporting date

Use it like any tibble:

```{r}
df_now %>% 
  filter(n > 5)
```

## Adding strata

If strata was given, the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) can easily tag the corresponding strata. 

```{r}
#Add the column using dplyr:
df_now <- df_now %>% 
  mutate(sex = c("M","M","F","M")) 

df_now
```

Use the `add_strata` to specify the new column is a stratum:

```{r}
df_now %>% 
  add_strata("sex")
```

The object now records `"sex"` as a stratification variable, preserved through downstream operations.

## Adding temporal effects

Temporal covariates help nowcasting models incorporate weekly seasonality, holiday effects, etc. Define the effects:

```{r}
t_eff <- temporal_effects(
  day_of_week  = TRUE,
  week_of_year = TRUE, 
  holidays     = cal_us_federal())

t_eff
```

Attach them to the dataset:

```{r}
df_now <- df_now %>% 
  add_temporal_effects(t_eff)

df_now
```

This lazily adds to the table  `day_of_week`, `week_of_year`, and `holiday` related to `event_date`. However it does not compute. Use `compute_temporal_effects()` to add them as columns:

```{r}
df_now %>% 
  compute_temporal_effects()
```

You can also attach effects related to the `report_date`:

```{r}
r_eff <- temporal_effects(day_of_week = TRUE)

df_now %>% 
  add_temporal_effects(r_eff, date_type = "report_date")
```

## Working with the “now”

You may override the default now to perform historical evaluation:

```{r}
df_pruned <- df_now %>%
  filter(report_date <= ymd("2023-12-26")) %>%
  change_now(ymd("2023-12-26"))
```

Retrieve the current active nowcast horizon:

```{r}
get_now(df_pruned)
```


## Visualizing a `tbl_now`

The [autoplot()](https://rodrigozepeda.github.io/tbl.now/reference/autoplot.tbl_now.html) method gives a quick diagnostic overview of a `tbl_now`. 


```{r autoplot,fig.width=9, fig.height=8, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)
data("flusight")

flusight_now <- tbl_now(flusight,
                      event_date  = target_end_date,
                      report_date = as_of,
                      case_count = observation,
                      verbose     = FALSE)

autoplot(flusight_now, level = 1)
```


## Learning more

+ Introduction vignette: [https://rodrigozepeda.github.io/tbl.now/articles/Introduction.html](https://rodrigozepeda.github.io/tbl.now/articles/Introduction.html)
+ Full walk-through with real CDC Flusight data: [https://rodrigozepeda.github.io/tbl.now/articles/Example.html](https://rodrigozepeda.github.io/tbl.now/articles/Example.html) 
+ Package reference: [https://rodrigozepeda.github.io/tbl.now/reference/](https://rodrigozepeda.github.io/tbl.now/reference/)


