---
title: "Example analysis with Flusight"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Example analysis with Flusight}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "90%",
  dpi = 350,
  fig.align = "center"
)
```

In this vignette we demonstrate how to use the `tbl.now` framework with real data 
from the U.S. Centers for Disease Control and Prevention (CDC). 
Specifically, we work with the Flusight dataset, which contains weekly counts 
of hospital admissions for laboratory-confirmed influenza.

We begin by loading the required packages:

```{r}
library(dplyr)
library(lubridate)
library(tbl.now)
```

## Data

The Flusight dataset (`?flusight`) includes weekly influenza hospital 
admission counts. For each epidemiological week (`target_end_date`), the CDC
publishes revised counts across multiple future weeks (`as_of`). Thus, each event 
week is associated with multiple reporting dates reflecting updates or revisions.

```{r}
data(flusight)
```

```{r, echo=FALSE}
flusight %>% 
  arrange(location_name) %>% 
  head(10) %>% 
  knitr::kable(caption = "The Flusight dataset")
```

The columns are:

+ `target_end_date`: the epidemiological week when cases occurred

+ `as_of`: the week in which the CDC updated its estimate for that event week

+ `location_name`: the state or territory

+ `observation`: the reported number of cases for target_end_date as of as_of

The key feature of this dataset is that observation is cumulative: the value for a given `target_end_date` and `as_of` is the latest estimate of total cases up to that event week, not the incremental update for that reporting date. The following example illustrates the structure:


```{r}
flusight %>% 
  filter(location_name == "Puerto Rico" & target_end_date == ymd("2025/04/12")) 
```

Each unique pair of (`target_end_date`, `as_of`) therefore corresponds to a cumulative estimate.

## Creating the `tbl_now`

Creating the tbl_now Object

To construct a tbl_now object, we must indicate:

+ `event_date`: the date on which cases occurred

+ `report_date`: the date on which the estimate was released

+ `case_count`: the column containing case counts

+ `strata`: grouping variables that define separate strata (e.g., states)

A first attempt produces several warnings:

```{r}
df_wrong <- flusight %>% 
  tbl_now(event_date = target_end_date, 
          report_date = as_of, 
          case_count = observation, 
          strata = location_name)
```

These warnings arise because:

1. Duplicate rows exist for some event–report combinations.

2. Missing values appear in the case count column.

3. The data is incorrectly inferred to represent count-incidence rather than count-cumulative, because cumulative-type datasets often contain repeated records.

Inspecting a subset confirms duplicated rows:

```{r}
flusight[c(422146, 422147, 422148, 422149), ]
```

Using [`dplyr's distinct()](https://dplyr.tidyverse.org/reference/distinct.html) we remove the duplicates:

```{r}
flusight <- flusight %>% distinct()
```

Next, we remove observations with missing case counts:

```{r}
flusight <- flusight %>% filter(!is.na(observation))
```

However, reconstructing the object still yields a misclassified data type:

```{r}
df_still_wrong <- tbl_now(flusight, event_date = "target_end_date", report_date = "as_of", 
        case_count = "observation", strata = c("location_name"))
```

The function incorrectly infers incidence data (i.e., each row represents the incremental number reported on that report date). In contrast, the Flusight dataset contains cumulative values. We therefore explicitly declare the data type:


```{r}
df_flu <- tbl_now(flusight, event_date = "target_end_date", report_date = "as_of", 
        case_count = "observation", strata = c("location_name"), data_type = "count-cumulative")
```

This yields a correctly structured `tbl_now` object:

```{r}
df_flu
```

## Working with the `tbl_now` Object

`tbl_now` objects are fully compatible with `dplyr` verbs. For example, we may focus on Puerto Rico and observations after mid–2024:

```{r}
df_pr <- df_flu %>% 
  rename(latest_report = as_of) %>% 
  filter(location_name == "Puerto Rico") %>% 
  filter(target_end_date >= ymd("2024/07/01"))

df_pr
```

Because we are now working with a single geographic unit, the `location_name` variable is no longer meaningful as a stratum. We remove it from the strata definition (without removing the column itself):

```{r}
df_pr <- df_pr %>% 
  remove_strata("location_name")

df_pr
```

The `now` (the effective horizon for the nowcast) is:

```{r}
get_now(df_pr)
```

### Changing the “now” for Historical Backtesting

To perform retrospective analyses (backtesting), we can filter the dataset and explicitly set a historical reporting cutoff:

```{r}
df_pr_new_now <- df_pr %>% 
  filter(latest_report < ymd("2023/12/01")) %>% 
  change_now(ymd("2023/12/01"))

df_pr_new_now
```

The new now is:

```{r}
get_now(df_pr_new_now)
```

## Working with Initial and Latest Reports

Two helper functions extract initial and final reported values for each event date:

+ `get_initial_reported_cases()`: the earliest available report for each event

+ `get_latest_reported_cases()`: the most recent report relative to now

A simple plot highlights the differences between initial and final estimates:

```{r}
initial_reports <- get_initial_reported_cases(df_pr)
latest_reports  <- get_latest_reported_cases(df_pr)
```

A simple plot highlights the differences between initial and final estimates:

```{r, fig.width=5, fig.height=3}
plot(initial_reports$target_end_date, initial_reports$observation, 
     type = "p", col = "deepskyblue4",
     xlab = "Date of event", ylab = "Cases",
     main = "Cases in Puerto Rico")

lines(latest_reports$target_end_date, latest_reports$observation,
      col = "tomato4")

legend("right", legend = c("Initial report", "Final report"),
       fill = c("deepskyblue4", "tomato4"))

```

