---
title: "tbl.now"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{tbl.now}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(dplyr)
library(lubridate)
library(tbl.now)
```

## Introduction

In `R`, one of the main frameworks for data analysis is the [tidyverse](https://tidyverse.org/) [@tidyverse]. Within it, data is arranged with **rows** corresponding to single observations and **columns** representing different variables. This is referred to as [tidy data](https://tidyr.tidyverse.org/articles/tidy-data.html) [@wickham2014tidy].  

For time series, tidy extensions to the classical `data.frame` (or [`tibble`](https://tibble.tidyverse.org/reference/tibble.html)) have previously been proposed with the [tsibble](https://tsibble.tidyverts.org/), the [tibbletime](https://business-science.github.io/tibbletime/), and the [timetk](https://business-science.github.io/timetk/) packages [@wang2020new; @wang2019tidy; @timetk]. However, in the context of epidemiological nowcasting were two different time-indexes are on the same table, the previous methods are lacking. This is where `tbl.now` comes in. 

The tibble now (`tbl.now`) is an extension fo the [`tibble`](https://tibble.tidyverse.org/reference/tibble.html) that tags specific variables as indexes for nowcasting models in the context of [`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/). The main purpose of the `tbl.now` is to unify the data inputted to [`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/)'s models while allowing the user to keep a `tidy` structure on the data and perform the usual [`dplyr`] data cleaning operations. This allows the user to move easierly into one of the classical modeling workflow frameworks such as [@gelman2020bayesian; @wickham2023r]:

```
Data Cleaning -> Modeling -> New Data Cleaning -> New Modeling -> ...
```


## The `tbl.now` framework 

In epidemiological nowcasting scenarios we have at least two dates:

+ `event_date`: When something happened (_e.g._ symptom onset or a test was taken).

+ `report_date`: When it was reported (_e.g._ the patient visited a physician or the test results were registered).

The nowcasting problem is:

> To estimate the total number of events **now** that have occurred at any **past** `event_date` given that not all of them have been reported yet (i.e. some `report_date`(s) are potentially in the future). 

In the context of nowcasting, the `tbl_now` can be thought of as a specific `tibble()` that guarantees an `event_date` and a `report_date.` by tagging them as **attributes** and allows for additional operations such as delay calculation. 

### Example

The `tbl_now` works from a `data.frame` by specifying the `event_date` and `report_date` columns:

```{r}
df <- data.frame(
  symptom_onset = c(ymd("2023/12/25"), ymd("2023/12/26"), ymd("2023/12/25"), ymd("2023/12/26")),
  medical_visit = c(ymd("2023/12/26"), ymd("2023/12/26"), ymd("2023/12/27"), ymd("2023/12/27")),
  n = c(10, 2, 5, 11)
)

df
```

In the previous `data.frame`, column **symptom_onset** represents the `event` while **medical_visit** corresponds to the `report`. Finally `n` corresponds to the number of cases which can be specified in the `case_count` variable.

The previous options can be specified in the `tbl_now`:

```{r}
df %>% 
  tbl_now(event_date = symptom_onset, report_date = medical_visit, case_count = n)
```

Notice that the [`tbl_now()`](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) function does several things at once:

+ It auto-detects the **data-type** as `count-incidence`. Other data types such as `linelist` are available and are discussed in section \@ref(data-types).

+ It auto-detects the **units (frequency)** of the event and report dates as 'daily' for both. 

+ It **tags** `symptom_onset` as an `event_date`, `medical_visit` as a `report_date` and `n` as `cases`.

+ It generates the `.event_num` and `.report_num` columns which are numerical versions of the `event_date` and `report_date` columns. 

+ It generates the `.delay` column with the delay given by the difference between report and event:
`.delay = report_date - event_date`. 

+ It automatically detects the **now** for the nowcasting. 

The following sections explain each of these perks of the `tbl.now` framework as well as additional functions that can be applied to a `tbl.now`. 

## Attributes of a `tbl_now` 

A `tbl_now` contains saves several of its column names as `attributes()`. They are the following:

+ **Now** (`now`): Refers to the **now** of the nowcast. That is, the date for which the prediction from the nowcast should be made. Ideally it refers to the most recent `report_date`.  

+ **Event date** (`event_date`): Name of the column with the dates when the event of interest happened. The nowcast will predict how many observations will eventually be observed in `event_date`

+ **Report date** (`report_date`): Name of the column with the dates when the event of interest was reported. The nowcast will predict how many observations will eventually be observed for `event_date` across all `report_date`s.

+ **Event units** (`event_units`): The units (e.g. "days", "weeks", "numeric") in which the `event_date` is registered. This is used for computing the `delay`. 

+ **Report units** (`report_units`): The units (e.g. "days", "weeks", "numeric") in which the `report_date` is registered. This is used for computing the `delay`. 

+ **Data type** (`data_type`): Either `linelist` if each row corresponds to a different observation or one of the following counts if data is aggregated. In `count-incidence`, each row corresponds to the number of cases observed exactly in the `report_date`. The `count-cumulative` represents how many cases have been observed _up until (and including)_ the `report_date`. See the [data types](#data-types) section for more details. 

+ **Strata** (`strata`): [_optional_][^1] Name of the columns marking the strata for the nowcast. The strata should be those variables for which the nowcast should compute separate results. For example, if strata is age-group, the nowcast should compute one model for each age-group. 

+ **Covariates** (`covariates`): [_optional_]  Name of the columns marking the covariates for the nowcast. The covariates should be those variables that help improve the nowcast. For example, a covariate for a dengue nowcast can be precipitaion or humidity levels. 

+ **Batched indicator** (`is_batched`):  [_optional_] Name of the columns for the nowcast indicating whether the `report_date` is right-censored. As an example of cases when this happens is when some data-sources have a technical error and all reports are not registered until very late. The `report_date` would come in a **batch** in such cases and not represent the actual dynamics of the process but a right-censored version. 

+ **Cases** (`case_count`):  [_optional_] If the `data-type` is any type of **count** data (`count-incidence` or `count-prevalence`), the column where the counts of events is registered. 

+ **Temporal effects** (`temporal_effects`): [_optional_] Refer to the temporal effects in the data created by the `temporal_effects()` function. See the [temporal effects](#temporal-effects) section for more details. 

[^1]: Optional attributes don't exist unless they are set. Hence they are not accesible via the `attributes()` function unless they exist. 

Values for all attributes can be accessed via the `get_*` functions. Examples include: `get_event_date()`, and `get_strata()`. 

### Data types 

The following data-types are admitted at `tbl_now` objects:

+ **Linelist**: Each row is an observation that was reported at `report_date` as happening at `event_date`.

```{r, echo=FALSE, cache=FALSE}
df <- data.frame(
    patient     = 1:6,
   event_date  = c(rep(ymd("2020/09/12"), 3),
                   rep(ymd("2020/09/13"), 3)),
   report_date = c(ymd("2020/09/12"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/15")))
knitr::kable(df, caption = "Linelist data")
```

+ **Count-incidence**: Each `report_date`-`event_date` combination contains the total number of 
cases observed _exactly_ at `report_date` for `event_date`.

```{r, echo=FALSE, cache=FALSE}
df <- data.frame(
  n           = c(7, 1, 9, 5, 0, 2),
  event_date  = c(rep(ymd("2020/09/12"), 3),
                  rep(ymd("2020/09/13"), 3)),
  report_date = c(ymd("2020/09/12"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/15")))
knitr::kable(df, caption = "Count-incidence data")
```

+ **Count-cumulative** Each `report_date`-`event_date` combination contains the total number of
cases observed up until `report_date` for `event_date`. The most recent 
`report_date` contains the best estimation of cases happening at `event_date`.

```{r, echo=FALSE, cache=FALSE}
df <- data.frame(
  n           = c(1,5, 8, 2, 2, 4),
  event_date  = c(rep(ymd("2020/09/12"), 3),
                  rep(ymd("2020/09/13"), 3)),
  report_date = c(ymd("2020/09/12"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/15")))
knitr::kable(df, caption = "Count-cumulative data")
```

The `to_count` function allows you to easily convert between different 
data-types.


### Temporal effects

### Modifying the attributes of a `tbl_now`

A `tbl_now`'s attributes can be modified with the [`add_*`](https://rodrigozepeda.github.io/tbl.now/reference/add.html), [`change_*`](https://rodrigozepeda.github.io/tbl.now/reference/change.html) or [`remove_*`](https://rodrigozepeda.github.io/tbl.now/reference/remove.html)
functions. They follow sort of the same pattern. 

Here is an example of creating a `tbl_now` and adding strata and temporal effects, changing the strata later and removing the temporal effects. 

```{r}
data("mpoxdat")

df_now <- mpoxdat %>% 
  tbl_now(event_date = dx_date, report_date = dx_report_date,
          case_count = n, verbose = FALSE, strata = race)

df_now
```

We can `change` the strata to include a new `race` column which is in upper case:

```{r}
df_now <- df_now %>% 
  mutate(RACE_UPPER = toupper(race)) %>% 
  change_strata(RACE_UPPER)

df_now
```

Temporal effects can be added using `add_temporal_effects`: 

```{r}
df_now <- df_now %>% 
  add_temporal_effects(temporal_effects(week_of_year = TRUE))

df_now
```

The remove functions can be used in a similar way, for example, to remove the effects and strata we just added:

```{r}
df_now <- df_now %>% 
  remove_temporal_effects() %>% 
  remove_all_strata()

df_now
```

### Transforming a `tbl_now` to count data

### Modifying a `tbl_now` with `dplyr`

All `tbl_now` objects are extensions of the `tibble()`. Hence the classical `dplyr` verbs
can be used to operate on the `tbl_now` while it tries to automatically accomodate for the changes. As an example, renaming a strata also changes the strata attribute:

```{r, message=FALSE}
library(dplyr)
data(denguedat)
df_now <- tbl_now(denguedat, event_date = onset_week, 
                  report_date = report_week, strata = gender,
                  verbose = FALSE)
```

You can see that the strata is `gender`:

```{r}
get_strata(df_now)
```

However, if we rename the column, the strata will also be renamed:

```{r}
df_now <- df_now %>% 
  rename(male_or_female = gender)

get_strata(df_now)
```

If an object cannot keep its `tbl_now` structure it will collapse into a `tibble` which is what happens in this example where summarise collapses all dates:

```{r}
df_now %>% 
  summarise(male_prop = sum(male_or_female == "Male") / n())
```

### Updating a `tbl_now`

A `tbl_now` can be updated with the `update` function from another `data.frame`, `tibble` or `tbl_now`. As long as they have the same columns it will by default copy the strata, covariate and temporal effects from the first `tbl_now` and apply it to the update. It also updates the `now` estimation. 

Consider the following `data.frame`

```{r}
#Initial observations
df <- data.frame(
    patient     = 1:6,
    event_date  = c(rep(ymd("2020/09/12"), 3), rep(ymd("2020/09/13"), 3)),
    report_date = c(ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
                   ymd("2020/09/13"), ymd("2020/09/14"), ymd("2020/09/15")))

df_now <- tbl_now(df, event_date = event_date, 
                  report_date = report_date, verbose = FALSE)
```

and its updated version (new data):

```{r}
#Initial observations
df_new <- data.frame(
    patient     = 7:13,
    event_date  = c(ymd("2020/09/13"), rep(ymd("2020/09/14"), 3), 
                    rep(ymd("2020/09/15"), 3)),
    report_date = c(ymd("2020/09/14"), ymd("2020/09/14"), ymd("2020/09/15"), 
                    ymd("2020/09/16"),
                    ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17")))
```

The full `tbl_now` can be updated with the `update` function:

```{r}
df_updated <- update(df_now, new_data = df_new)

df_updated
```


## Other functions

### Reports

### Week alignment


## References

