---
title: "tbl.now"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{tbl.now}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(dplyr)
library(lubridate)
library(tbl.now)
```

## Introduction

In `R`, one of the main frameworks for data analysis is the [tidyverse](https://tidyverse.org/) [@tidyverse]. Within it, data is arranged with **rows** corresponding to single observations and **columns** representing different variables. This is referred to as [tidy data](https://tidyr.tidyverse.org/articles/tidy-data.html) [@wickham2014tidy].  

For time series, tidy extensions to the classical `data.frame` (or [`tibble`](https://tibble.tidyverse.org/reference/tibble.html)) have previously been proposed with the [tsibble](https://tsibble.tidyverts.org/), the [tibbletime](https://business-science.github.io/tibbletime/), and the [timetk](https://business-science.github.io/timetk/) packages [@wang2020new; @wang2019tidy; @timetk]. However, in the context of epidemiological nowcasting were two different time-indexes are on the same table, the previous methods are lacking. This is where `tbl.now` comes in. 

The tibble now (`tbl.now`) is an extension fo the [`tibble`](https://tibble.tidyverse.org/reference/tibble.html) that tags specific variables as indexes for nowcasting models in the context of [`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/). The main purpose of the `tbl.now` is to unify the data inputted to [`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/)'s models while allowing the user to keep a `tidy` structure on the data and perform the usual [`dplyr`] data cleaning operations. This allows the user to move easierly into one of the classical modeling workflow frameworks such as [@gelman2020bayesian; @wickham2023r]:

```
Data Cleaning -> Modeling -> New Data Cleaning -> New Modeling -> ...
```


## The `tbl.now` framework 

In epidemiological nowcasting scenarios we have at least two dates:

+ `event_date`: When something happened (_e.g._ symptom onset or a test was taken).

+ `report_date`: When it was reported (_e.g._ the patient visited a physician or the test results were registered).

The nowcasting problem is:

> To estimate the total number of events **now** that have occurred at any **past** `event_date` given that not all of them have been reported yet (i.e. some ``report_date`s are potentially in the future). 

In the context of nowcasting, the `tbl_now` can be thought of as a specific `tibble()` that guarantees an `event_date` and a `report_date.` by tagging them as **attributes** and allows for additional operations such as delay calculation. 

### Example

The `tbl_now` works from a `data.frame` by specifying the `event_date` and `report_date` columns:

```{r}
df <- data.frame(
  symptom_onset = c(ymd("2023/12/25"), ymd("2023/12/26"), ymd("2023/12/25"), ymd("2023/12/26")),
  medical_visit = c(ymd("2023/12/26"), ymd("2023/12/26"), ymd("2023/12/27"), ymd("2023/12/27")),
  n = c(10, 2, 5, 11)
)

df
```

In the previous `data.frame`, column **symptom_onset** represents the `event` while **medical_visit** corresponds to the `report`. Finally `n` corresponds to the number of cases which can be specified in the `case_count` variable.

The previous options can be specified in the `tbl_now`:

```{r}
df %>% 
  tbl_now(event_date = symptom_onset, report_date = medical_visit, case_count = n)
```

Notice that the [tbl_now()] function does several things at once:

+ It auto-detects the **data-type** as `count-incidence`. Other data types such as `linelist` are available and are discussed in section \@ref(data-types).

+ It auto-detects the **units (frequency)** of the event and report dates as 'daily' for both. 

+ It **tags** `symptom_onset` as an `event_date`, `medical_visit` as a `report_date` and `n` as `cases`.

+ It generates the `.event_num` and `.report_num` columns which are numerical versions of the `event_date` and `report_date` columns. 

+ It generates the `.delay` column with the delay given by the difference between report and event:
`.delay = report_date - event_date`. 

+ It automatically detects the **now** for the nowcasting. 

The following sections explain each of these perks of the `tbl.now` framework as well as additional functions that can be applied to a `tbl.now`. 


### Attributes of a `tbl_now` 

#### Data types {#data-types}

#### Temporal effects

#### Modifying the attributes of a `tbl_now`

#### Transforming a `tbl_now` to count data

#### Modifying a `tbl_now` with `dplyr`

#### Updating a `tbl_now`

## Other functions






## References

