---
title: "tbl.now"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{tbl.now}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(dplyr)
library(lubridate)
library(tbl.now)
```

## Introduction

In `R`, one of the main frameworks for data analysis is the [tidyverse](https://tidyverse.org/) [@tidyverse]. Within it, data is arranged with **rows** corresponding to single observations and **columns** representing different variables. This is referred to as [tidy data](https://tidyr.tidyverse.org/articles/tidy-data.html) [@wickham2014tidy].  

For time series, tidy extensions to the classical `data.frame` (or [`tibble`](https://tibble.tidyverse.org/reference/tibble.html)) have previously been proposed with the [tsibble](https://tsibble.tidyverts.org/), the [tibbletime](https://business-science.github.io/tibbletime/), and the [timetk](https://business-science.github.io/timetk/) packages [@wang2020new; @wang2019tidy; @timetk]. However, in the context of epidemiological nowcasting were two different time-indexes are on the same table, the previous methods are lacking. This is where `tbl.now` comes in. 

The tibble now (`tbl.now`) is an extension fo the [tibble()](https://tibble.tidyverse.org/)  that tags specific variables as indexes for nowcasting models in the context of [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/). The main purpose of the `tbl.now` is to unify the data inputted to [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/)'s models while allowing the user to keep a `tidy` structure on the data and perform the usual [dplyr](https://dplyr.tidyverse.org/) data cleaning operations. This allows the user to move easierly into one of the classical modeling workflow frameworks such as [@gelman2020bayesian; @wickham2023r]:

```
Data Cleaning -> Modeling -> New Data Cleaning -> New Modeling -> ...
```


## The `tbl.now` framework 

In epidemiological nowcasting scenarios we have at least two dates:

+ `event_date`: When something happened (_e.g._ symptom onset or a test was taken).

+ `report_date`: When it was reported (_e.g._ the patient visited a physician or the test results were registered).

The nowcasting problem is:

> To estimate the total number of events **now** that have occurred at any **past** `event_date` given that not all of them have been reported yet (i.e. some `report_date`(s) are potentially in the future). 

In the context of nowcasting, the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) can be thought of as a specific `tibble()` that guarantees an `event_date` and a `report_date.` by tagging them as **attributes** and allows for additional operations such as delay calculation. 

### Example

The [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) works from a `data.frame` by specifying the `event_date` and `report_date` columns:

```{r}
df <- data.frame(
  symptom_onset = c(ymd("2023/12/25"), ymd("2023/12/26"), ymd("2023/12/25"), ymd("2023/12/26")),
  medical_visit = c(ymd("2023/12/26"), ymd("2023/12/26"), ymd("2023/12/27"), ymd("2023/12/27")),
  n = c(10, 2, 5, 11)
)

df
```

In the previous `data.frame`, column **symptom_onset** represents the `event` while **medical_visit** corresponds to the `report`. Finally `n` corresponds to the number of cases which can be specified in the `case_count` variable.

The previous options can be specified in the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html):

```{r}
df %>% 
  tbl_now(event_date = symptom_onset, report_date = medical_visit, case_count = n)
```

Notice that the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) function does several things at once:

+ It auto-detects the **data-type** as `count-incidence`. Other data types such as `linelist` are available and are discussed in section \@ref(data-types).

+ It auto-detects the **units (frequency)** of the event and report dates as 'daily' for both. 

+ It **tags** `symptom_onset` as an `event_date`, `medical_visit` as a `report_date` and `n` as `cases`.

+ It generates the `.event_num` and `.report_num` columns which are numerical versions of the `event_date` and `report_date` columns. 

+ It generates the `.delay` column with the delay given by the difference between report and event:
`.delay = report_date - event_date`. 

+ It automatically detects the **now** for the nowcasting. 

The following sections explain each of these perks of the `tbl.now` framework as well as additional functions that can be applied to a `tbl.now`. 

## Attributes of a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) 

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) contains saves several of its column names as `attributes()`. They are the following:

+ **Now** (`now`): Refers to the **now** of the nowcast. That is, the date for which the prediction from the nowcast should be made. Ideally it refers to the most recent `report_date`.  

+ **Event date** (`event_date`): Name of the column with the dates when the event of interest happened. The nowcast will predict how many observations will eventually be observed in `event_date`

+ **Report date** (`report_date`): Name of the column with the dates when the event of interest was reported. The nowcast will predict how many observations will eventually be observed for `event_date` across all `report_date`s.

+ **Event units** (`event_units`): The units (e.g. "days", "weeks", "numeric") in which the `event_date` is registered. This is used for computing the `delay`. 

+ **Report units** (`report_units`): The units (e.g. "days", "weeks", "numeric") in which the `report_date` is registered. This is used for computing the `delay`. 

+ **Data type** (`data_type`): Either `linelist` if each row corresponds to a different observation or one of the following counts if data is aggregated. In `count-incidence`, each row corresponds to the number of cases observed exactly in the `report_date`. The `count-cumulative` represents how many cases have been observed _up until (and including)_ the `report_date`. See the [data types](#data-types) section for more details. 

+ **Strata** (`strata`): [_optional_][^1] Name of the columns marking the strata for the nowcast. The strata should be those variables for which the nowcast should compute separate results. For example, if strata is age-group, the nowcast should compute one model for each age-group. 

+ **Covariates** (`covariates`): [_optional_]  Name of the columns marking the covariates for the nowcast. The covariates should be those variables that help improve the nowcast. For example, a covariate for a dengue nowcast can be precipitaion or humidity levels. 

+ **Batched indicator** (`is_batched`):  [_optional_] Name of the columns for the nowcast indicating whether the `report_date` is right-censored. As an example of cases when this happens is when some data-sources have a technical error and all reports are not registered until very late. The `report_date` would come in a **batch** in such cases and not represent the actual dynamics of the process but a right-censored version. 

+ **Cases** (`case_count`):  [_optional_] If the `data-type` is any type of **count** data (`count-incidence` or `count-prevalence`), the column where the counts of events is registered. 

+ **Temporal effects** (`temporal_effects`): [_optional_] Refer to the temporal effects in the data created by the [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html) function. See the [temporal effects](#temporal-effects) section for more details. 



[^1]: Optional attributes don't exist unless they are set. Hence they are not accesible via the `attributes()` function unless they exist. 

Values for all attributes can be accessed via the `get_*` functions. Examples include: `get_event_date()`, and `get_strata()`. 

### Data types 

The following data-types are admitted at [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) objects:

+ **Linelist**: Each row is an observation that was reported at `report_date` as happening at `event_date`.

```{r, echo=FALSE, cache=FALSE}
df_linelist <- data.frame(
    patient     = 1:6,
   event_date  = c(rep(ymd("2020/09/12"), 3),
                   rep(ymd("2020/09/13"), 3)),
   report_date = c(ymd("2020/09/12"),
                  ymd("2020/09/12"),
                  ymd("2020/09/13"),
                  ymd("2020/09/13"),
                  ymd("2020/09/13"),
                  ymd("2020/09/13")))

df_linelist <- df_linelist %>% 
    tbl_now(event_date = event_date, report_date = report_date, 
            case_col = n, verbose = FALSE)

knitr::kable(df_linelist, caption = "Linelist data")
```

+ **Count-incidence**: Each `report_date`-`event_date` combination contains the total number of 
cases observed _exactly_ at `report_date` for `event_date`.

```{r, echo=FALSE, cache=FALSE}
df_count_inc <- data.frame(
  n           = c(7, 1, 9, 5, 0, 2),
  event_date  = c(rep(ymd("2020/09/12"), 3),
                  rep(ymd("2020/09/13"), 3)),
  report_date = c(ymd("2020/09/12"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/15"))) 

df_count_inc <- df_count_inc %>% 
  tbl_now(event_date = event_date, report_date = report_date, 
          case_count = n, verbose = FALSE)

knitr::kable(df_count_inc, caption = "Count-incidence data")
```

+ **Count-cumulative** Each `report_date`-`event_date` combination contains the total number of
cases observed up until `report_date` for `event_date`. The most recent 
`report_date` contains the best estimation of cases happening at `event_date`.

```{r, echo=FALSE, cache=FALSE}
df_count_cum <- data.frame(
  n           = c(1,5, 8, 2, 2, 4),
  event_date  = c(rep(ymd("2020/09/12"), 3),
                  rep(ymd("2020/09/13"), 3)),
  report_date = c(ymd("2020/09/12"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/15"))) 

df_count_cum <- df_count_cum %>% 
    tbl_now(event_date = event_date, report_date = report_date, 
            case_count = n, verbose = FALSE)

knitr::kable(df_count_cum, caption = "Count-cumulative data")
```

The `to_count()` function allows you to easily convert between different 
data-types.

### Transforming a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) to count data

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) can do the following transformations via the `to_count()` function between data types:

+ **linelist to count-incidence**: Aggregates each observation in the linelist by report and event date. The `n` column contains how many individuals were specifically observed at that `report_date` for the `event_date` in question. 

```{r}
df_linelist %>% 
  to_count(to = "count-incidence")
```

+ **linelist to count-cumulative**: Aggregates each observation in the linelist by report and event date. The `n` column contains how many individuals had been observed up untill that `report_date` (including the previous dates) for the `event_date` in question. 

```{r}
df_linelist %>% 
  to_count(to = "count-cumulative")
```

> **Note** In the previous example the `n` counts `3` as it is aggregating the `1` observed at `report_date = 2020-09-13` and the `2` observed at `report_date = 2020-09-12`. This is the difference between the **count-incidence** that specifies the ones observed **exactly** on that date and the **count-cumulative** that specifies the ones observed up **until and including** that date. 

+ **count-incidence to count-cumulative**: Aggregates each observation accumulating how many cases for that `event_date` had been observed up until that `report_date`:

```{r}
df_count_inc %>% 
  to_count(to = "count-cumulative")
```

+ **Aggregate repeated count-\* events**: The `to_count` function can also be used to aggregate data from one form to the same form. As an example of such a case consider the following  [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html):

```{r}
df_example <- data.frame(
  n           = c(8, 11, 0, 1, 1, 5, 2, 4, 1, 10, 9, 11, 3, 1),
  sex         = c(rep("M", 3), rep("F", 4), rep("M", 2), rep("F", 5)),
  event_date  = c(rep(ymd("2020/09/12"), 3),
                  rep(ymd("2020/09/12"), 4),
                  rep(ymd("2020/09/13"), 2),
                  rep(ymd("2020/09/13"), 5)),
  report_date = c(ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
                  ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
                  ymd("2020/09/15"), ymd("2020/09/13"), ymd("2020/09/14"),
                  ymd("2020/09/13"), ymd("2020/09/14"),
                  ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17"))) 

tbl_example <- df_example %>% 
  tbl_now(event_date = event_date, report_date = report_date, 
          case_count = n, strata = sex, verbose = FALSE)

tbl_example
```

Notice that the data is already in `count-incidence` format with `sex` being a strata. If we wanted the **overall** number of cases (not by strata) we can remove the strata and then use the `to_count` function to aggregate all the races while keeping the `count-incidence` structure:

1) Remove the strata

```{r}
tbl_example <- tbl_example %>% 
  remove_all_strata()

#This removes the strata but doesn't aggregate everything
tbl_example
```

2) And then aggregate to a `count-incidence`:

```{r}
tbl_example <- tbl_example %>% 
  to_count(to = "count-incidence")

#It summed all the `n` columns with one entry per observation 
tbl_example
```

### Temporal effects

Temporal effects can be added as a special type of covariate to the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)  using the [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html) function. 

For example, we can specify to include covariates for the day of the week, the week of the year, and whether it is holiday in the US:

```{r}
library(almanac)

t_eff <- temporal_effects(
  day_of_week  = TRUE,
  week_of_year = TRUE, 
  holidays     = cal_us_federal())
t_eff
```

Note that the holidays calendar is an [rcalendar](https://davisvaughan.github.io/almanac/reference/rcalendar.html) object from the [almanac](https://davisvaughan.github.io/almanac/articles/almanac.html) package. 

Temporal effects can be added to the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) object with the [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html) function:

```{r, R.options=list(pillar.width=250)}
data("denguedat")

#Create a tbl_now
df_now <- denguedat %>% 
  tbl_now(event_date = onset_week, report_date = report_week,
          verbose = FALSE, strata = gender)

#Add temporal effects (see them as . columns)
df_now %>% 
  add_temporal_effects(t_eff) 
```


### Modifying the attributes of a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)'s attributes can be modified with the [add_*](https://rodrigozepeda.github.io/tbl.now/reference/add.html), [change_*](https://rodrigozepeda.github.io/tbl.now/reference/change.html) or [remove_*](https://rodrigozepeda.github.io/tbl.now/reference/remove.html)
functions. They follow sort of the same pattern. 

Here is an example of creating a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) and adding strata and temporal effects, changing the strata later and removing the temporal effects. 

```{r}
data("mpoxdat")

df_now <- mpoxdat %>% 
  tbl_now(event_date = dx_date, report_date = dx_report_date,
          case_count = n, verbose = FALSE, strata = race)

df_now
```

We can `change` the strata to include a new `race` column which is in upper case:

```{r}
df_now <- df_now %>% 
  mutate(RACE_UPPER = toupper(race)) %>% 
  change_strata(RACE_UPPER)

df_now
```

Temporal effects can be added using [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html): 

```{r}
df_now <- df_now %>% 
  add_temporal_effects(temporal_effects(week_of_year = TRUE))

df_now
```

The remove functions can be used in a similar way, for example, to remove the effects and strata we just added:

```{r}
df_now <- df_now %>% 
  remove_temporal_effects() %>% 
  remove_all_strata()

df_now
```



### Modifying a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) with `dplyr`

All [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) objects are extensions of the [tibble()](https://tibble.tidyverse.org/). Hence the classical [dplyr](https://dplyr.tidyverse.org/) verbs
can be used to operate on the [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) while it tries to automatically accomodate for the changes. As an example, renaming a strata also changes the strata attribute:

```{r, message=FALSE}
library(dplyr)
data(denguedat)
df_now <- tbl_now(denguedat, event_date = onset_week, 
                  report_date = report_week, strata = gender,
                  verbose = FALSE)
```

You can see that the strata is `gender`:

```{r}
get_strata(df_now)
```

However, if we rename the column, the strata will also be renamed:

```{r}
df_now <- df_now %>% 
  rename(male_or_female = gender)

get_strata(df_now)
```

If an object cannot keep its [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) structure it will collapse into a [tibble()](https://tibble.tidyverse.org/) which is what happens in this example where summarise collapses all dates:

```{r}
df_now %>% 
  summarise(male_prop = sum(male_or_female == "Male") / n())
```

### Updating a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) can be updated with the `update` function from another `data.frame`, [tibble()](https://tibble.tidyverse.org/) or [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html). As long as they have the same columns it will by default copy the strata, covariate and temporal effects from the first [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) and apply it to the update. It also updates the `now` estimation. 

Consider the following `data.frame`

```{r}
#Initial observations
df <- data.frame(
    patient     = 1:6,
    event_date  = c(rep(ymd("2020/09/12"), 3), rep(ymd("2020/09/13"), 3)),
    report_date = c(ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
                   ymd("2020/09/13"), ymd("2020/09/14"), ymd("2020/09/15")))

df_now <- tbl_now(df, event_date = event_date, 
                  report_date = report_date, verbose = FALSE)
```

and its updated version (new data):

```{r}
#Initial observations
df_new <- data.frame(
    patient     = 7:13,
    event_date  = c(ymd("2020/09/13"), rep(ymd("2020/09/14"), 3), 
                    rep(ymd("2020/09/15"), 3)),
    report_date = c(ymd("2020/09/14"), ymd("2020/09/14"), ymd("2020/09/15"), 
                    ymd("2020/09/16"),
                    ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17")))
```

The full [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) can be updated with the [update()](https://rodrigozepeda.github.io/tbl.now/reference/update.tbl_now.html) function:

```{r}
df_updated <- update(df_now, new_data = df_new)

df_updated
```


## Other functions (utilities)

### Convert epiweeks to dates

The [week_2_date()](https://rodrigozepeda.github.io/tbl.now/reference/week_2_date.html) function allows you to convert from a `data.frame` with 
epidemiological weeks and years to specific dates:

```{r}
df <- data.frame(
  epidemiological_week = 1:5,
  epidemiological_year = rep(2024, 5)
)

df %>%
  week_2_date(
    week_col = epidemiological_week,
    year_col = epidemiological_year
  )
```

### Reports

The [get_*_reported_cases()](https://rodrigozepeda.github.io/tbl.now/reference/get_latest_first.html) functions get either the initial or the latest number of reported cases. This functions are useful to compare the initial number of cases believed to have happened against the latest number of cases. Consider the following example where initially 10 cases were suspected to have happened but in the latest reported number up to 15 (=10 + 1 + 1 + 3) cases happened for the same date:

```{r}
df_reports <- data.frame(
  n           = c(10, 1, 1, 0, 0, 3),
  event_date  = rep(ymd("2020/09/12"), 6),
  report_date = c(ymd("2020/09/12"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/15"),
                  ymd("2020/09/16"),
                  ymd("2020/09/17"))) 

tbl_reports <- df_reports %>% 
  tbl_now(event_date = event_date, report_date = report_date, 
          verbose = FALSE, case_count = n, report_units = "days", 
          event_units = "days")

tbl_reports
```

Initially in the first report this was the number of cases

```{r}
get_initial_reported_cases(tbl_reports)
```

But in the end 15 cases were observed:

```{r}
get_latest_reported_cases(tbl_reports)
```

### Week alignment

The [align_week()](https://rodrigozepeda.github.io/tbl.now/reference/align_week.html) function allows you to take dates from different days of the same epidemiological week
and set them all to the same date. This is useful for computing time differences between weekly reports (to avoid decimal times). 

```{r}
df <- data.frame(
  date = c(ymd("2022-10-31"), ymd("2022-11-07"), ymd("2022-11-13")),
  epiweek = c(44, 45, 46)
)

# Align to Sundays
df_aligned <- align_week(df, date_col = date)
df_aligned
```

we can check they are actually Sundays with the [wday()](https://lubridate.tidyverse.org/reference/day.html) function from the [lubridate](https://lubridate.tidyverse.org/index.html) package:

```{r}
df_aligned %>% 
  mutate(day_label = wday(date_aligned, label = TRUE, abbr = FALSE))
```





## References

