---
title: "tbl.now"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{tbl.now}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction 

The [tidyverse](https://tidyverse.org/) ecosystem [@tidyverse] provides a widely adopted framework for data analysis. Within this paradigm, data is structured such that rows represent individual observations and columns represent variables. This approach is commonly referred to as [tidy data](https://tidyr.tidyverse.org/articles/tidy-data.html) [@wickham2014tidy].

Several tidyverse extensions exist for working with time series including [tsibble](https://tsibble.tidyverts.org/), [tibbletime](https://business-science.github.io/tibbletime/), and [timetk](https://business-science.github.io/timetk/) [@wang2020new; @wang2019tidy; @timetk]. These packages provide time-aware tibbles and tools that integrate smoothly with the tidyverse. However, epidemiological nowcasting requires two time indices simultaneously: an **event time** and a **reporting time**. Classical time-series abstractions assume only a single index and therefore do not fully support this structure.

The `tbl.now` class and package addresses this gap. The `tbl.now` extends a regular  [tibble()](https://tibble.tidyverse.org/) to explicitly encode epidemiological event and report dates, allowing consistent data transformation, delay computation, and integration with the [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) modeling workflow.

More concretely, `tbl.now` was designed to:

+ Standardize the data inputs required by [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/)  models.

+ Preserve  [tidyverse](https://tidyverse.org/)  compatibility so users can continue to apply familiar [dplyr](https://dplyr.tidyverse.org/) operations.

+ [Diagnose reporting artefacts](https://rodrigozepeda.github.io/tbl.now/articles/batch-reporting.html) directly from the data (such as reporting-delay drift and change points) as well as batch (backlog) reporting.

+ Facilitate integration into iterative modeling workflows with [different nowcasting packages](https://rodrigozepeda.github.io/tbl.now/articles/nowcasting-models.html) [@gelman2020bayesian; @wickham2023r]:

```
Data Cleaning -> Modeling -> New Data Cleaning -> New Modeling -> ...
```

We'll begin by loading the required packages:

```{r setup, message=FALSE}
library(dplyr, quietly = TRUE)
library(lubridate)
library(tbl.now)
```


## How `tbl.now` works 

In an epidemiological nowcast, we typically observe at least two key dates[^1]:

[^1]: More key dates are possible such as a `confirmation_date`. For example in the case of influenza one might consider the `event_date` = symptom onset, the `report_date` = when the patient was first diagnosed by a medical professional, and `confirmation_date` = when the positive test's results for influenza were recorded. We will come back to these multiple dates later. 

+ `event_date`: when the underlying event occurred (e.g., symptom onset, exposure, sample collection).

+ `report_date`: when the event was recorded in the data system (e.g., lab result processed, clinical visit documented).

The nowcasting task is:

> To estimate, for each past `event_date`, how many events (e.g. cases) have already occurred
but have not yet been reported as of **now**. That is, the nowcast will predict how
many observations will eventually be observed for each (past or present) `event_date`.

Visually:

```{r nowcast-explainer, echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4.6, out.width = "100%", fig.alt = "Stacked bar chart of cases by event date. Green bars show cases already reported as of now; pale red bars stacked on top show the cases not yet reported, which grow rapidly over the most recent event dates. A black line traces the total that will eventually be reported, rising steadily, while the green observed bars turn downward near now."}
library(ggplot2)
library(dplyr)

now_date <- ymd("2024-03-31")
horizon <- 24

explain <- dplyr::tibble(event_date = now_date - (horizon:0))
explain$age <- as.numeric(now_date - explain$event_date)

# A rising epidemic: the eventual (final) count for each event date.
explain$eventual <- round(20 * exp(0.055 * (horizon - explain$age)))
# Reporting completeness grows with how old the event date is.
explain$completeness <- 1 - exp(-0.55 * (explain$age + 0.35))
explain$observed <- round(explain$eventual * explain$completeness)
explain$missing <- explain$eventual - explain$observed

lab_obs <- "Already reported as of now"
lab_mis <- "Not yet reported (what the nowcast estimates)"
lab_now <- "Nowcast: a model for what will eventually be reported"

bars <- tidyr::pivot_longer(
  explain[, c("event_date", "observed", "missing")],
  cols = c("observed", "missing"), names_to = "part", values_to = "cases"
)
bars$part <- factor(
  bars$part, levels = c("missing", "observed"), labels = c(lab_mis, lab_obs)
)

ggplot(bars, aes(x = event_date, y = cases)) +
  geom_col(aes(fill = part), width = 0.75) +
  geom_line(data = explain, aes(y = eventual, colour = lab_now), linewidth = 0.9) +
  geom_point(data = explain, aes(y = eventual, colour = lab_now), size = 1.5) +
  scale_fill_manual(
    name = NULL, breaks = c(lab_obs, lab_mis),
    values = stats::setNames(c("#5F7E62", "#e78b7f"), c(lab_obs, lab_mis))
  ) +
  scale_colour_manual(name = NULL, values = stats::setNames("#262626", lab_now)) +
  geom_vline(aes(xintercept = now_date), data = dplyr::tibble(now_date = now_date),
             linetype = "dashed", colour = "#262626", linewidth = 0.4) +
  annotate(
    "label", x = now_date, y = max(explain$eventual) * 1.10,
    label = "now", hjust = 0.5, vjust = 0, fontface = "bold",
    size = 3.4, colour = "#262626"
  ) +
  coord_cartesian(ylim = c(0, max(explain$eventual) * 1.15)) +
  labs(
    x = "Event date", y = "Cases",
    title = "What a nowcast estimates",
    subtitle = paste(
      "Recent event dates look like a downturn only because",
      "their reports have not arrived yet"
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    legend.box = "vertical",
    legend.margin = margin(b = -4),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", colour = "#262626")
  ) +
  guides(
    fill = guide_legend(order = 1, ncol = 2),
    colour = guide_legend(order = 2)
  )
```


>In the figure above, the green bars represent the number of cases (events)
that have been observed until **now**; the pale red segments are the 
reports still in transit. Because completeness decays sharply
over the most recent event dates, the observed counts bend *downwards* near the
right edge even though the epidemic is still growing. The **nowcast** is the red
line: an estimate of the height each bar will eventually reach.

A `tbl.now` object is a specialized [tibble()](https://tibble.tidyverse.org/) that:

+ Identifies the `event_date` and `report_date` columns.

+ Stores these as `attributes()` to enable consistent processing.

+ Automatically computes auxiliary fields such as delay, numerical indices, and frequency units.

+ Ensures the dataset is correctly formatted for [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) models.

## Example: A simple `tbl.now`

Consider the following dataset:

| symptom_onset | medical_visit |  n |
| ------------- | ------------- | -: |
| 2023-12-25    | 2023-12-26    | 10 |
| 2023-12-26    | 2023-12-26    |  2 |
| 2023-12-25    | 2023-12-27    |  5 |
| 2023-12-26    | 2023-12-27    | 11 |


Where:

+ `symptom_onset` is the `event_date`.

+ `medical_visit` is the `report_date`.

+ `n` is the number of reported cases for each event–report combination.

We can convert this into a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) by first creating a `data.frame` and then using the `tbl_now()` function:

```{r}
#Create a data.frame
df <- data.frame(
  symptom_onset = c(ymd("2023/12/25"), ymd("2023/12/26"), 
                    ymd("2023/12/25"), ymd("2023/12/26")),
  medical_visit = c(ymd("2023/12/26"), ymd("2023/12/26"), 
                    ymd("2023/12/27"), ymd("2023/12/27")),
  n = c(10, 2, 5, 11)
)

#Convert to tbl.now
df |>
  tbl_now(event_date = symptom_onset, report_date = medical_visit, case_count = n)
```


This performs several operations automatically:

+ Detects the **data type** (`count-incidence` in this example). See [below](#data-types) for all the data types available. 

+ Infers the **frequency units** of event and report dates (daily).

+ Tags the correct columns as `event_date`, `report_date`, and `case_count.`

+ Computes `.event_num`, `.report_num`, and `.delay` columns the numerical versions (indexed at 0) of event, report and `.delay = report_date - event_date` columns.

+ Identifies the appropriate **now** date (the most recent report date).

The remaining sections describe these features and the broader `tbl.now` toolkit.


## Attributes of a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) 

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) stores information about its structure using object attributes, ensuring consistent behavior across `dplyr` transformations. The primary attributes are:

+ **now**: the "current" reporting date used for nowcasting, typically the most recent `report_date`.

+ **event_date**: the column name storing event dates.

+ **report_date**: the column name storing report dates.

+ **event_units**: the temporal units for event dates (e.g., "days", "weeks", "numeric").

+ **report_units**: the temporal units for report dates.

+ **data_type**: one of the following (see [the data types section](#data-types)):
    + `"linelist"`: each row is an individual observation
    + `"count-incidence"`: each row contains counts. Those counts represent the
    exact number of events observed for each event-report date combination. 
    + `"count-cumulative"`:  each row contains cumulative counts. Those counts represent the
    cumulative number of events observed by each report date for that event. Later
    report dates accumulate the earlier ones into the count. 

+ **strata** (optional)[^2]: variables for which the nowcast should be 
computed separately (e.g., age group, sex).

+ **covariates** (optional): predictor variables that may improve the 
nowcast (e.g., weather covariates).

+ **is_censored** (optional): identifies cases where the report
date represents not the exact date it was reported but an upper limit
to that exact date (i.e. left-censored). As an example, one can consider 
a system error and reports from a lab are not registered until a week after. 

+ **case_count** (optional): the column storing case counts when the dataset 
is aggregated.

+ **temporal_effects** (optional): a lazy specification for temporal effects
such as day of the week, holiday, week of the year and other temporal effects. See the [temporal effects](#temporal-effects) section for more details.

You can access any attribute using the corresponding [getter](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html), e.g. [get_event_date()](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html) or [get_strata()](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html).

[^2]: Optional attributes are set to `NULL` by default. 


Below we provide more information on some of the attributes.

### Data types 

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)  can represent one of three data structures:


1. **Linelist**: Each row corresponds to a single reported observation.

```{r, echo=FALSE, cache=FALSE}
df_linelist <- data.frame(
  patient = 1:6,
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/13"), 3)
  ),
  report_date = c(
    ymd("2020/09/12"),
    ymd("2020/09/12"),
    ymd("2020/09/13"),
    ymd("2020/09/13"),
    ymd("2020/09/13"),
    ymd("2020/09/13")
  )
)

knitr::kable(df_linelist, caption = "Linelist data")
```

2. **Count-incidence**: Each row summarizes how many events with a given `event_date` were reported **exactly** on that `report_date.`

```{r, echo=FALSE, cache=FALSE}
df_count_inc <- data.frame(
  n = c(7, 1, 9, 5, 0, 2),
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/13"), 3)
  ),
  report_date = c(
    ymd("2020/09/12"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/15")
  )
)

knitr::kable(df_count_inc, caption = "Count-incidence data")
```

3. **Count-cumulative** Each row summarizes how many events with a given `event_date` had been reported up to and including that `report_date`. The distinction is crucial for nowcasting models that operate either on daily increments or cumulative totals.

```{r, echo=FALSE, cache=FALSE}
df_count_cum <- data.frame(
  n = c(1, 5, 8, 2, 2, 4),
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/13"), 3)
  ),
  report_date = c(
    ymd("2020/09/12"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/15")
  )
)

knitr::kable(df_count_cum, caption = "Count-cumulative data")
```

The `to_count()` function allows you to convert between different data-types as we see below:

### Converting Between Data Types

The [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html) function supports structured transformations. Here we start with linelist data as an example:

```{r}
#The original data.frame has one row per patient
df_linelist <- data.frame(
  patient     = 1:6,
  event_date  = c(rep(ymd("2020/09/12"), 3), rep(ymd("2020/09/13"), 3)),
  report_date = c(rep(ymd("2020/09/12"), 2), rep(ymd("2020/09/13"), 4))
)

#We can convert it to a tbl.now
df_linelist <- df_linelist |> 
  tbl_now(event_date = event_date, report_date = report_date, 
          data_type = "linelist")

#This is what it looks like
df_linelist
```

+ **Linelist → Count-Incidence**: Aggregates by event–report date, counting only cases reported on that date.

```{r}
df_linelist |>
  to_count(to = "count-incidence")
```

+ **Linelist → Count-Cumulative**: Aggregates by event–report date, producing cumulative counts up to each report date.

```{r}
df_linelist |>
  to_count(to = "count-cumulative")
```

> **Note** In the previous example the `n` counts `3` as it is aggregating the `1` observed at `report_date = 2020-09-13` and the `2` observed at `report_date = 2020-09-12`. This is the difference between the **count-incidence** that specifies the ones observed **exactly** on that date and the **count-cumulative** that specifies the ones observed up **until and including** that date. 

+ **Count-Incidence → Count-Cumulative**: Computes cumulative sums for each event date across report dates.

```{r}
df_count_inc <- df_linelist |>
  to_count(to = "count-incidence")

#This is count incidence:
df_count_inc

#Turns to count cumulative:
df_count_inc |>
  to_count(to = "count-cumulative")
```

+ **Aggregation within the same type**: The [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html) may also be used to re-aggregate datasets that contain duplicate event–report pairs. This is useful when raw surveillance feeds contain repeated entries such as in this case:

```{r}
tbl_example <- data.frame(
  n = c(8, 11, 0, 1, 1, 5, 2, 4, 1, 10, 9, 11, 3, 1),
  sex = c(rep("M", 3), rep("F", 4), rep("M", 2), rep("F", 5)),
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/12"), 4),
    rep(ymd("2020/09/13"), 2),
    rep(ymd("2020/09/13"), 5)
  ),
  report_date = c(
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/15"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17")
  )) |>
  tbl_now(
    event_date = event_date, report_date = report_date,
    data_type = "count-incidence", case_count = n, verbose = FALSE,
    warn_non_uniqueness = FALSE
  )

tbl_example
```

This dataset intentionally contains repeated event_date–report_date pairs for each `sex`. You can  aggregate redundant rows with the [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html)  function that collapses duplicates by summing the `case_count` column.


```{r}
tbl_example |>
  to_count(to = "count-incidence")
```

The function ensures that:

+ Rows are grouped by `event_date`, `report_date`, and any `strata`, and `is_censored`.

+ The `case_count` column is summed within each group.

+ Attributes are preserved so the resulting object remains a valid `tbl_now`.


### Temporal effects

Often, temporal covariates improve nowcasting performance by helping to adjust systematic changes within the calendar cycle (e.g., day-of-week effects, seasonal effects, or other reporting artefacts). The [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html) function creates a *specification* (recipe) of the features to compute:

```{r}
library(almanac)

t_eff <- temporal_effects(
  day_of_week  = TRUE,
  week_of_year = TRUE,
  holidays     = cal_us_federal()
)
t_eff
```

> Note that the holidays calendar is an [rcalendar](https://davisvaughan.github.io/almanac/reference/rcalendar.html) object from the [almanac](https://davisvaughan.github.io/almanac/articles/almanac.html) package.

#### How do thet work?

Temporal effects in `tbl.now` follow a **lazy evaluation** pattern:

1. **Add** with [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html) (or via the `t_effects` argument of `tbl_now()`). This records *what* should be computed but adds **no columns** yet.

2. **Materialise the columns** with [compute_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/compute_temporal_effects.html) when you are ready to use them in a model.

```{r, R.options=list(pillar.width=250)}
data("denguedat")

# Step 1 — create the tbl_now and attach the spec (no columns added yet)
df_now <- denguedat |>
  tbl_now(
    event_date = onset_week, report_date = report_week,
    verbose = FALSE, strata = gender
  )

df_now <- df_now |>
  add_temporal_effects(t_eff)

# The footer shows "T. effects (lazy): ..." — spec is recorded but not computed
df_now
```

The footer now shows the spec with `(lazy)` to signal that the columns have not been computed yet. No new columns appear in the tibble at this point. When we compute, they appear (scroll to the right):

```{r}
# Step 2 — materialise the columns when needed
df_computed <- compute_temporal_effects(df_now)

# Columns are now present and annotated [t_effect]
df_computed
```

After `compute_temporal_effects()`:

+ The effect columns (`.event_day_of_week`, `.event_week_of_year`, `.event_holiday`) are added.
+ The function `get_temporal_effect_cols()` lists the column names while
+ The original call remains accessible via `get_temporal_effects()`, so you always know which effects were requested even after further dplyr operations.

```{r}
get_temporal_effects(df_computed) # The spec (list of configs)
get_temporal_effect_cols(df_computed) # The computed column names
```

#### Around-holiday and around-weekend effects

Reporting often *rebounds* on the first working day(s) after a holiday or a
weekend. To capture that, `temporal_effects()` has `holiday_lags` and
`weekend_lags`: each takes a depth `N` and creates indicator columns
`..._holiday_lag_1 … ..._holiday_lag_N` (and `..._weekend_lag_k`) that flag dates
falling exactly `k` **working days** after a holiday / weekend. Working days skip
weekends and other holidays, so the effect lands on the first day back at work.

```{r}
# Flag the two working days after a holiday, and the working day after a weekend
after_eff <- temporal_effects(
  holidays     = cal_us_federal(),
  holiday_lags = 2,
  weekend_lags = 1
)
after_eff
```

The mirror image — a slowdown in the days *leading up to* a break — is a negative
depth. `..._holiday_lead_k` / `..._weekend_lead_k` then flag dates `k` working
days **before** a holiday / weekend, counting backwards from it, so `_lead_1` is
the working day closest to the break:

```{r}
# Flag Christmas Eve (and the eve of every other holiday), plus the Wednesday,
# Thursday and Friday before each weekend
before_eff <- temporal_effects(
  holidays     = cal_us_federal(),
  holiday_lags = -1,
  weekend_lags = -3
)
before_eff
```

To model both sides of the same break, attach one specification per direction:

```{r, eval=FALSE}
df_now |>
  add_temporal_effects(temporal_effects(weekend_lags = -1)) |> # the Friday before
  add_temporal_effects(temporal_effects(weekend_lags = 1)) # the Monday after
```

#### Event- vs report-date effects

By default effects are derived from the **event date** and named `.event_*`. Pass
`date_type = "report_date"` to `add_temporal_effects()` to derive them from the
**report date** instead (columns named `.report_*`). Both can coexist on the same
`tbl_now`, and every converter carries both sets of columns.

```{r, eval=FALSE}
df_now |>
  add_temporal_effects(temporal_effects(week_of_year = TRUE), date_type = "event_date") |>
  add_temporal_effects(temporal_effects(day_of_week = TRUE),  date_type = "report_date")
```


### The confirmation process

Some surveillance systems have a **third** date. A case is not only reported --
it is later *resolved*: a laboratory issues the result that confirms it, or
rules it out. Influenza is the standard picture: symptoms begin (the event), the
patient visits a doctor (the report), and days later a swab comes back positive
(the confirmation) or negative (a *retraction* -- the case was reported but is
not a case after all).

A `tbl_now` can carry this with `confirmation_date`, an optional
`confirmation_type`, and its own `confirmation_units`. The timeline it assumes
is

$$\text{event date} \le \text{report date} \le \text{confirmation date} \le \text{now}$$

`confirmation_type` takes the values `"confirmed"`, `"retracted"`, `"pending"`
or `NA`. **Pending** is the important one: it means the case has been reported
and is still waiting for a result, so it has *no* confirmation date. That is a
different thing from a case whose result you simply never recorded, which is
`NA`.

We will use `hai_bucaramanga`, an open extract of healthcare-associated
infections from Bucaramanga, Colombia, which records all three dates: when the
specimen was taken, when the laboratory received it, and when the laboratory
issued the result.

```{r, message=FALSE}
data("hai_bucaramanga")

hai <- hai_bucaramanga |>
  distinct() |>
  filter(
    !is.na(specimen_date), !is.na(received_date),
    received_date >= specimen_date
  ) |>
  mutate(
    # A result was issued: that is the confirmation. Everything else is still
    # waiting, and a case that is waiting has no confirmation date.
    result_date = if_else(
      !is.na(report_date) & report_date >= received_date,
      report_date, as.Date(NA)
    ),
    result = if_else(is.na(result_date), "pending", "confirmed")
  )

table(hai$result)
```

The three dates map onto the three roles directly:

```{r}
hai_now <- hai |>
  tbl_now(
    event_date        = specimen_date, # the specimen was taken
    report_date       = received_date, # the laboratory received it
    confirmation_date = result_date, # the laboratory issued a result
    confirmation_type = result,
    data_type         = "linelist",
    verbose           = FALSE
  )

hai_now
```

The footer now carries a confirmation line -- the column, its units, and how
many cases have actually been resolved. Two derived columns appear alongside
`.delay`: `.confirmation_num` (the confirmation date on the same numeric grid as
the other dates) and `.confirmation_delay`, the laboratory's **turnaround** --
the time from report to result, which is a different quantity from the reporting
delay.

```{r}
# Reporting delay: specimen to laboratory.
median(hai_now$.delay, na.rm = TRUE)

# Turnaround: laboratory to result.
median(hai_now$.confirmation_delay, na.rm = TRUE)
```

Confirmation also moves `now`. A result issued on a date means you were, by
definition, still observing the system on that date, so `now` is never earlier
than the last confirmation -- even when reporting stopped before it.

```{r}
get_now(hai_now)
get_confirmation_units(hai_now)
has_confirmation(hai_now)
```

#### Counting cases when some of them can be undone

Once cases can be retracted, "how many cases were there?" has more than one
answer, and the right one depends on the question:

```{r}
head(get_latest_confirmed(hai_now), 3) # cases the laboratory confirmed
head(get_net_confirmed(hai_now), 3) # confirmed minus retracted
```

`get_nth_confirmed(x, delay)` counts only the cases resolved *within* a given
number of periods -- what you would have known that soon after the report --
and `get_initial_confirmed()` is the same-period case. These mirror
`get_nth_reported_cases()` and `get_initial_reported_cases()` on the report
axis.

If your data records absurdly long confirmation delays -- a result "issued" two
years later is usually a data-entry artefact, not a laboratory --
`censor_confirmation_delays_above()` returns those cases to `"pending"`, which
is what they really were.

#### How much of the epidemic has been resolved?

`plot_confirmation_status()` shows the share of cases confirmed, retracted and
pending over time. A pending share that grows towards the right-hand edge is
normal -- recent cases have not had time to come back from the laboratory -- but
a pending share that grows in the *middle* of the series is a laboratory falling
behind.

```{r, fig.width=7, fig.height=4}
plot_confirmation_status(hai_now)
```

#### The confirmation axis

Every reporting diagnostic in the package asks one question: did an unusual
number of records arrive on this date, and with what delay? That question is
just as meaningful for results arriving from a laboratory, so the diagnostics
take an `axis` argument instead of being duplicated:

```{r, eval=FALSE}
# The same picture, drawn for the laboratory instead of the surveillance desk.
plot_reporting_triangle(hai_now, axis = "confirmation")
plot_delay_profiles(hai_now, axis = "confirmation")
plot_delay_drift(hai_now, axis = "confirmation")
diagnostic_plot(hai_now, axis = "confirmation")

# A laboratory clearing a backlog is a batch, exactly as a surveillance system
# clearing its inbox is.
batch_test(hai_now, axis = "confirmation")
```

Two notes on what `axis = "confirmation"` means. Delays are still measured
**from the event**, so the report and confirmation axes are directly comparable
and the gap between them is the time the laboratory adds. And cases that are
still `"pending"` are excluded: they have no confirmation date, so counting
them would invent an arrival on a date they do not have.

Finally, when a system produces both confirmations and retractions you can ask
whether they take equally long -- a laboratory that rules cases out faster than
it confirms them will bias any nowcast that treats the two alike.
`test_confirmation_delay()` tests exactly that (a Wilcoxon rank-sum test on the
two delay distributions) and `plot_confirmation_delay()` draws it. The
Bucaramanga extract records no retractions, so there is nothing to compare here.

### Getting, removing and changing attributes

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)’s attributes can be modified using the functions in [add_*](https://rodrigozepeda.github.io/tbl.now/reference/add.html), [change_*](https://rodrigozepeda.github.io/tbl.now/reference/change.html), or [remove_*](https://rodrigozepeda.github.io/tbl.now/reference/remove.html). These functions share a consistent interface that allows users to incrementally manipulate strata, covariates, and temporal effects.

The example below demonstrates how to create a  [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html), add strata and temporal effects, later modify the strata, and finally remove the temporal effects.

```{r, message=FALSE}
data("mpoxdat")

df_now <- mpoxdat |>
  tbl_now(
    event_date = dx_date, report_date = dx_report_date,
    case_count = n, verbose = FALSE, strata = race
  )

df_now
```

You can see that the strata is `race` with the corresponding [get_*](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html):

```{r}
get_strata(df_now)
```

Strata can be modified with the [change_*](https://rodrigozepeda.github.io/tbl.now/reference/change.html) family of functions. The following example adds a new column containing an uppercase version of the existing race variable and sets it as the new strata:

```{r}
df_now <- df_now |>
  mutate(RACE_UPPER = toupper(race)) |>
  change_strata(RACE_UPPER)

get_strata(df_now)
```

To attach a lazy temporal-effects spec, use [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html), then materialise with [compute_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/compute_temporal_effects.html):

```{r, R.options=list(pillar.width = 250)}
df_now <- df_now |>
  add_temporal_effects(temporal_effects(week_of_year = TRUE))

# Spec is stored (lazy):
get_temporal_effects(df_now)

# Compute to add columns:
df_now <- compute_temporal_effects(df_now)
get_temporal_effect_cols(df_now)
```

Attributes can be removed using the corresponding [remove_*](https://rodrigozepeda.github.io/tbl.now/reference/remove.html) functions. `remove_temporal_effects()` drops both the spec and any computed columns:

```{r}
df_now <- df_now |>
  remove_temporal_effects() |>
  remove_all_strata()

df_now
```

```{r}
get_temporal_effects(df_now) # Empty list — no spec
get_temporal_effect_cols(df_now) # character(0) — no computed cols
get_strata(df_now)
```

## Modifying a tbl_now() with dplyr

[tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) objects extend [tibble()](https://tibble.tidyverse.org/), and therefore support standard [dplyr](https://dplyr.tidyverse.org/)
verbs. The class attempts to preserve and adapt its internal attributes when operations are performed. For example, renaming a strata column will automatically update the stored strata attribute.

```{r}
library(dplyr, quietly = TRUE)
data(denguedat)

df_now <- tbl_now(denguedat,
  event_date = onset_week,
  report_date = report_week, strata = gender,
  verbose = FALSE
)

# Current strata
get_strata(df_now)
```


After renaming the column, the strata attribute updates accordingly:

```{r}
df_now <- df_now |>
  rename(male_or_female = gender)

get_strata(df_now)
```

Certain operations may cause a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) object to drop back to a standard tibble. This occurs when the operation removes necessary structure—for example, collapsing all data into a single row with [summarise()](https://dplyr.tidyverse.org/reference/summarise.html):

```{r}
df_now |>
  summarise(number_males = sum(male_or_female == "Male"))
```
 
 
Other operations like mutate and select work as long as the original columns and the required for the attributes are kept:

```{r}
df_now <- df_now |>
  mutate(GENDER = toupper(male_or_female)) |> 
  select(GENDER, everything())

df_now
```

 
### Updating a tbl_now()

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) object can be updated using the [update()](https://rodrigozepeda.github.io/tbl.now/reference/update.tbl_now.html) method with another data.frame, tibble, or [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) as input. When the column structure is compatible, the update process retains the strata, covariate, and temporal-effect attributes from the original object, and recalculates "now" estimates using the combined data.

Below is an example using an initial dataset:

```{r}
df <- data.frame(
  patient = 1:6,
  event_date = c(rep(ymd("2020/09/12"), 3), rep(ymd("2020/09/13"), 3)),
  report_date = c(
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/13"), ymd("2020/09/14"), ymd("2020/09/15")
  )
)

df_now <- tbl_now(df,
  event_date = event_date,
  report_date = report_date, verbose = FALSE
)
```

And a follow-up dataset containing newly reported cases:

```{r}
df_new <- data.frame(
  patient = 7:13,
  event_date = c(
    ymd("2020/09/13"),
    rep(ymd("2020/09/14"), 3),
    rep(ymd("2020/09/15"), 3)
  ),
  report_date = c(
    ymd("2020/09/14"), ymd("2020/09/14"), ymd("2020/09/15"),
    ymd("2020/09/16"), ymd("2020/09/15"), ymd("2020/09/16"),
    ymd("2020/09/17")
  )
)
```

We can update the original object by incorporating the new data:

```{r}
df_updated <- update(df_now, new_data = df_new)

df_updated
```

## Visualizing a `tbl_now`

<!-- The `eval=` conditions on the plotting chunks below spell the
requireNamespace() test out in full rather than reusing a helper variable. `R CMD
check` tangles this vignette with knitr::purl(), which evaluates chunk options in
a fresh environment: a variable defined by an earlier chunk's *code* does not
exist yet, so `eval=some_variable` silently drops the chunk, and a later chunk
relying on what it defined then fails with "object not found". -->


The [autoplot()](https://rodrigozepeda.github.io/tbl.now/reference/autoplot.tbl_now.html) method gives a quick diagnostic overview of a `tbl_now`:


```{r autoplot, eval=requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE), fig.width=9, fig.height=10, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)

dengue_now <- tbl_now(denguedat,
  event_date  = "onset_week",
  report_date = "report_week",
  verbose     = FALSE
)

autoplot(dengue_now)
```

We explore each of the panels below

### The delay distribution

```{r, fig.cap="Empirical delay distribution", echo = FALSE, eval=requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE)}
plot_delay_distribution(dengue_now)
```

The **empirical delay distribution** represents a case-count weighted histogram of the reporting delay.

### The observed epidemic process

```{r, fig.cap="Observed epidemic process", echo = FALSE, eval=requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE)}
plot_epidemic_process(dengue_now)
```

The **observed epidemic process** represents the latest reported counts per `event_date`.

### The calendar effects

```{r, fig.cap="Day of the week and week of the year effects", echo = FALSE, eval=requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE)}
plot_week_of_year_effects(dengue_now)
```

These are boxplots of the *percent* effect  by day of week / week of year / month. The plots shown are determined automatically by the units of the event and report dates. 

### The holiday effects 

If holidays are included in the `tbl_now()` they show the percent effects by *day type* (holiday or not) and by *position relative to the nearest holiday*. See [Holiday effects](#holiday-effects)

### The cycles

```{r, fig.cap="Periodogram showing the Fourier season's dominant peak", echo = FALSE, eval=requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE)}
plot_cycles(dengue_now)
```

The periodogram shows the dominant peak. This suggests a Fourier season length to pass to [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html). For the weekly dengue data for example, the periodogram peaks near 52 weeks, suggesting an annual cycle. We could capture this with a Fourier term of `temporal_effects(seasons = 52)`.


> The reporting delay panels reveal the same effects for the `report_delay`. 


### Additional arguments for `autoplot()`

Set `by_strata = TRUE` to split every panel by stratum: the boxplots become dodged boxes (one per stratum, side by side), the epidemic process and periodograms become one coloured line per stratum, and the delay distribution becomes dodged bars. The boxplots are then normalized *per stratum* (1 = that stratum's own average) so the calendar pattern is comparable across strata. By default the object's `strata` are used; pass `strata = "gender"` to group on a subset.

```{r, eval=requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE), fig.width=9, fig.height=10, message=FALSE, warning=FALSE}
autoplot(dengue_now, strata = "gender", by_strata = TRUE)
```

#### One panel at a time: the `plot_*()` functions

`autoplot()` draws the whole grid in one call. When you want a single effect on its own — in its own figure, at its own size — every panel also has a standalone `plot_*()` twin. They take the same object and return the identical plot, so `autoplot(x, panels = "calendar_weekday")` and `plot_day_of_week_effects(x)` are the same figure. Use `autoplot()` for the overview; reach for a `plot_*()` when you want to place one effect in a report.

Each calendar/holiday twin takes a `type` argument choosing the process: `type = "epidemic"` (the default, green — how the *cases* vary) or `type = "report"` (red — how the *reporting* does).

| Function | `autoplot()` panel |
|---|---|
| `plot_day_of_week_effects(x, type = )` | `calendar_weekday` / `delay_weekday` |
| `plot_week_of_year_effects(x, type = )` | `calendar_week` / `delay_week` |
| `plot_month_of_year_effects(x, type = )` | `calendar_month` / `delay_month` |
| `plot_holiday_effects(x, type = )` | `calendar_holiday` / `delay_holiday` |
| `plot_holiday_lag_effects(x, type = )` | `calendar_holiday_lag` / `delay_holiday_lag` |
| `plot_cycles(x, type = )` | `seasonality` / `delay_seasonality` |
| `plot_delay_distribution(x)` | `delay_distribution` |
| `plot_observed_cases(x)` | `epidemic` |

```{r plot-single, eval=FALSE}
# The same figure, two ways:
autoplot(dengue_now, panels = "calendar_week")

# The event effects:
plot_week_of_year_effects(dengue_now)

# The reporting twin (delay):
plot_week_of_year_effects(dengue_now, type = "report")
```


### Holiday effects {#holiday-effects}


The  holiday panels  describe the [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html) attached. For them to appear you need to describe a `holidays` calendar and/or `holiday_lags`. The temporal effect is read directly, so there is no need to call `compute_temporal_effects()` first.

Holidays are a daily phenomenon, so for this example we switch to daily data:

```{r holiday-panels-setup, include=FALSE}
library(almanac)
set.seed(42)

us_holidays <- cal_us_federal()
event_days <- seq(as.Date("2022-01-01"), as.Date("2023-12-31"), by = "day")

# Reporting takes longer when the desk is shut: +6 days on a holiday, +3 on a weekend.
reports <- lapply(event_days, function(day) {
  closed_for <- if (alma_in(day, us_holidays)) 6 else if (!is_weekday(day)) 3 else 0
  delays <- 0:rpois(1, 2 + closed_for)
  data.frame(onset = day, reported = day + delays, n = rpois(length(delays), 5))
})
reports <- do.call(rbind, reports)
```

```{r holiday-panels-build, eval=requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE) && requireNamespace("almanac", quietly = TRUE), message=FALSE}
holiday_now <- tbl_now(reports,
    event_date = onset, report_date = reported, case_count = n,
    data_type = "count-incidence", verbose = FALSE
  ) |>
  add_temporal_effects(
    temporal_effects(weekend = TRUE, holidays = cal_us_federal(), holiday_lags = 2)
  )
```

```{r holiday-panels, eval=requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE) && requireNamespace("almanac", quietly = TRUE), fig.width=9, fig.height=7, message=FALSE, warning=FALSE}
autoplot(holiday_now, panels = c("calendar_holiday", "calendar_holiday_lag",
                                 "delay_holiday", "delay_holiday_lag"))
```


The **holiday effect** panel splits the days by *type*. The categories you get follow the `temporal_effects()`: a calendar plus `weekend = TRUE` gives `Weekday` / `Weekend` / `Holiday`; a calendar alone gives `Non-holiday` / `Holiday`; a weekend effect alone gives `Weekday` / `Weekend`. A holiday that falls on a weekend counts as a **holiday**.

The **holiday lag effect** panel splits them by *position relative to the nearest holiday*, showing exactly the days that `holiday_lags` flags — `"1 after"`, `"2 after"`, and so on, with `"Other"` (every other day) as the reference to read them against. Negative depths add `"1 before"`, `"2 before"`, … to the left of the holiday, so the axis reads left-to-right as time does. Counting is in **working days**, so weekends and other holidays are skipped exactly as they are for the `..._holiday_lag_k` columns.


### Do delay distributions drift over time?

Reporting delays are not always stable: a surveillance system may speed up or slow down over a season or across years. [plot_delay_drift()](https://rodrigozepeda.github.io/tbl.now/reference/plot_delay_drift.html) draws a rolling *fan chart* of the delay distribution indexed by event date — a solid rolling median, a dashed rolling mean, and 25–75% / 10–90% quantile bands. A rising or falling centre line signals *location* drift; widening or narrowing bands signal *spread* drift.

```{r delay-drift, eval=requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE), fig.width=9, fig.height=4.5, message=FALSE, warning=FALSE}
plot_delay_drift(dengue_now)
```

Because the most recent event dates have not had time to be fully reported, their delays look artificially short; that immature region (after the completeness cutoff) is **shaded grey** and should not be read as drift.

For a formal answer, [test_delay_drift()](https://rodrigozepeda.github.io/tbl.now/reference/test_delay_drift.html) tests both a location statistic (the median) and a dispersion statistic (the 10–90 spread), on mature data only for any drift:

```{r delay-drift-test, eval=requireNamespace("modifiedmk", quietly = TRUE)}
test_delay_drift(dengue_now, stat = c("median", "spread"))
```

A significant `drift` for `spread` with a non-significant one for `median`, say, would mean the *typical* delay held steady while its *variability* changed.

The trend test looks for *gradual* change. If instead you suspect an **abrupt** shift — a reporting-system change on some date — use [test_delay_changepoint()](https://rodrigozepeda.github.io/tbl.now/reference/test_delay_changepoint.html), which applies Pettitt's nonparametric change-point test and returns the estimated change date together with the before/after delay level:

```{r delay-changepoint}
test_delay_changepoint(dengue_now, stat = c("median", "spread"))
```


You can get more information in the [corresponding article](https://rodrigozepeda.github.io/tbl.now/articles/batch-reporting.html).

### Detecting batch reporting

Reporting systems sometimes withhold results and then release a whole backlog at once — a **batch**. The key idea is that a batch *moves* reports along the report axis without *creating* them, so a window of report dates spanning both the lull and the spike has an unchanged total — whereas a genuine epidemic surge adds cases and inflates it. [batch_test()](https://rodrigozepeda.github.io/tbl.now/reference/batch_test.html) turns this into a per-report-date diagnostic that separates the two:

```{r batches, warning=FALSE}
batches <- batch_test(dengue_now, lookback = 2)

batches |>
  filter(batch) |>
  select(report_date, reported, baseline, everything())
```

Additional information on dealing with batches and other reporting delay artifacts can be found in the corresponding [article](https://rodrigozepeda.github.io/tbl.now/articles/batch-reporting.html). 

## Other functions (utilities)

### Convert epidemiological weeks to dates

The function [week_2_date()](https://rodrigozepeda.github.io/tbl.now/reference/week_2_date.html) converts epidemiological week/year combinations into a calendar date aligned on the first day of the week (Sunday).

```{r}
df <- data.frame(
  epidemiological_week = 1:5,
  epidemiological_year = rep(2024, 5)
)

df |>
  week_2_date(
    week_col = epidemiological_week,
    year_col = epidemiological_year
  )
```
 
### Reports

The functions [get_initial_reported_cases() and get_latest_reported_cases()](https://rodrigozepeda.github.io/tbl.now/reference/get_latest_first.html) extract the number of cases first reported for each event date and the most recently reported totals, respectively. These utilities allow users to quantify revisions between initial and final reports.
 
 
```{r}
df_reports <- data.frame(
  n = c(10, 1, 1, 0, 0, 3),
  event_date = rep(ymd("2020/09/12"), 6),
  report_date = c(
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17")
  )
)

tbl_reports <- df_reports |>
  tbl_now(
    event_date = event_date, report_date = report_date,
    verbose = FALSE, case_count = n, report_units = "days",
    event_units = "days"
  )

tbl_reports
```


The initial reported cases:

```{r}
get_initial_reported_cases(tbl_reports)
```

and the latest totals:

```{r}
get_latest_reported_cases(tbl_reports)
```

### Week alignment

The [align_weeks()](https://rodrigozepeda.github.io/tbl.now/reference/align_weeks.html) 
function standardizes dates within the same epidemiological week to a single reference day
(for example, the start of the week). This is helpful when computing differences across weekly reporting periods, avoiding fractional time intervals.

```{r}
df <- data.frame(
  date = c(ymd("2022-10-31"), ymd("2022-11-07"), ymd("2022-11-13")),
  epiweek = c(44, 45, 46)
)

# Align to Sundays
df_aligned <- align_weeks(df, date_col = date)
df_aligned
```

You can verify the resulting weekday using the [wday() function from the lubridate package](https://lubridate.tidyverse.org/reference/day.html):

```{r}
df_aligned |>
  mutate(day_label = wday(date_aligned, label = TRUE, abbr = FALSE))
```

### Complete zeroes

The [complete_zeroes()](https://rodrigozepeda.github.io/tbl.now/reference/complete_zeroes.html) function fills with zeroes those cases where the `event` or `report` weeks have not been observed. 

Consider for example the following data with just two observations per date:

```{r}
ndata <- tibble(
  event_date = c(as.Date("2021/01/12"), as.Date("2021/01/14"), as.Date("2021/01/14")),
  report_date = c(as.Date("2021/01/13"), as.Date("2021/01/15"), as.Date("2021/01/18")),
  case_count = c(10, 5, 1)
)

ndata <- tbl_now(ndata, event_date, report_date,
  verbose = FALSE, case_count = case_count, data_type = "count-incidence"
)

ndata
```

Notice that there are no observations for `2021/01/13`. Furthermore, if we assume that the maximum possible observed delay is of `4`,  we can fill the unobserved cases with:

```{r}
complete_zeroes(ndata, max_delay = 4)
```

Which looks at all the possible report dates and event dates and sets the counts to zero if they have not been observed.

### Censoring extreme delays

The function `censor_delays_above()` marks all delays above a threshold value (`max_delay`) as censored. This is useful to indicate extreme delays in some nowcast models:

```{r}
df <- data.frame(
  onset = as.Date("2020-01-01") + c(0, 0, 1, 2),
  reported = as.Date("2020-01-01") + c(1, 5, 2, 300)
)
tn <- tbl_now(df,
  event_date = onset, report_date = reported,
  data_type = "linelist", verbose = FALSE
)

# the 300-day report becomes censored (an upper bound on its delay)
censor_delays_above(tn, max_delay = 60)
```

### Converting to data formats from other packages

`tbl.now` ships converters that move data between a `tbl_now` and the data structures used by other nowcasting and delay-estimation packages. They all follow the same naming convention:

+ `tbl_now_from_*()` builds a `tbl_now` (it wraps [as_tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/as_tbl_now.html), so `...` is forwarded to [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)).

+ `tbl_now_to_*()` converts a `tbl_now` into that package's native object.

Each function accepts a `verbose` argument that reports the choices it made (the inferred `now`, the data type, For example here we can convert to `tsibble`:

```{r converters-roundtrip, eval=requireNamespace("tsibble", quietly = TRUE), message=FALSE}
dengue_now <- tbl_now(denguedat,
  event_date = "onset_week",
  report_date = "report_week", strata = "gender",
  verbose = FALSE
)

# tbl_now -> tsibble -> tbl_now
dengue_ts <- tbl_now_to_tsibble(dengue_now, verbose = FALSE)

#This returns a tsibble
dengue_ts

#Which can be converted back to tbl.now
as_tbl_now(dengue_ts, report_date = "report_week", verbose = FALSE)
```

See the [vignette on using different models](https://rodrigozepeda.github.io/tbl.now/articles/nowcasting-models.html) to see all conversion formats. 


## References

