---
title: "tbl.now"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{tbl.now}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Within `R`, the [tidyverse](https://tidyverse.org/) ecosystem [@tidyverse] provides a widely adopted framework for data analysis. In this paradigm, data is structured such that rows represent individual observations and columns represent variables. This approach is commonly referred to as tidy data [@wickham2014tidy].

Several tidyverse extensions exist for working with time series data, including [tsibble](https://tsibble.tidyverts.org/), [tibbletime](https://business-science.github.io/tibbletime/), and [timetk](https://business-science.github.io/timetk/) [@wang2020new; @wang2019tidy; @timetk]. These packages provide time-aware tibbles and tools that integrate smoothly with the tidyverse. However, epidemiological nowcasting requires two time indices simultaneously: an **event time** and a **reporting time**. Classical time-series abstractions assume only a single index and therefore do not fully support this structure.

The `tbl.now` class and package addresses this gap. The `tbl.now` extends a regular  [tibble()](https://tibble.tidyverse.org/) to explicitly encode epidemiological event and report dates, allowing consistent data transformation, delay computation, and integration with the [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) modeling workflow.

More concretely, tbl.now was designed to:

+ Standardize the data inputs required by [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/)  models.

+ Preserve  [tidyverse](https://tidyverse.org/)  compatibility so users can continue to apply familiar [dplyr](https://dplyr.tidyverse.org/) operations.

+ Facilitate seamless integration into iterative modeling workflows [@gelman2020bayesian; @wickham2023r]:

```
Data Cleaning -> Modeling -> New Data Cleaning -> New Modeling -> ...
```

We'll begin by loading the required packages:

```{r setup, message=FALSE}
library(dplyr, quietly = TRUE)
library(lubridate)
library(tbl.now)
```



## The `tbl.now` Framework 

In an epidemiological nowcast, we typically observe two key dates:

+ `event_date`: when the underlying event occurred (e.g., symptom onset, exposure, sample collection).

+ `report_date`: when the event was recorded in the data system (e.g., lab result processed, clinical visit documented).

The nowcasting task is:

> To estimate, for each past `event_date`, how many events have already occurred 
but have not yet been reported as of **now**. That is, the nowcast will predict how 
many observations will eventually be observed for the `event_date`s. 

A `tbl.now` object is therefore a specialized [tibble()](https://tibble.tidyverse.org/) that:

+ Identifies the `event_date` and `report_date` columns.

+ Stores these as attributes to enable consistent processing.

+ Automatically computes auxiliary fields such as delay, numerical indices, and frequency units.

+ Ensures the dataset is well-formed for downstream [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) models.

### Example

Consider the following dataset:

```{r}
df <- data.frame(
  symptom_onset = c(ymd("2023/12/25"), ymd("2023/12/26"), ymd("2023/12/25"), ymd("2023/12/26")),
  medical_visit = c(ymd("2023/12/26"), ymd("2023/12/26"), ymd("2023/12/27"), ymd("2023/12/27")),
  n = c(10, 2, 5, 11)
)

knitr::kable(df, caption = "Example dataset")
```

Here:

+ `symptom_onset` is the event date.

+ `medical_visit` is the report date.

+ `n` is the number of reported cases for each event–report combination.

We can convert this into a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html):

```{r}
df %>% 
  tbl_now(event_date = symptom_onset, report_date = medical_visit, case_count = n)
```


This performs several operations automatically:

+ Detects the **data type** (`count-incidence` in this example).

+ Infers the **frequency units** of event and report dates (daily).

+ Tags the correct columns as `event_date`, `report_date`, and `case_count.`

+ Computes `.event_num`, `.report_num`, and `.delay` columns.

+ Identifies the appropriate **now** date (the most recent report date).

The remaining sections describe these features and the broader `tbl.now` toolkit.


## Attributes of a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) 

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) stores information about its structure using object attributes, ensuring consistent behavior across transformations and modeling steps. The primary attributes are:

+ **now**: the "current" reporting date used for nowcasting, typically the most recent `report_date`.

+ **event_date**: the column name storing event dates.

+ **report_date**: the column name storing report dates.

+ **event_units**: the temporal units for event dates (e.g., "days", "weeks", "numeric").

+ **report_units**: the temporal units for report dates.

+ **data_type**: one of the following (see [the data types section](#data-types):
    + linelist
    + count-incidence
    + count-cumulative.

+ **strata** (optional)[^1]: variables for which the nowcast should be 
computed separately (e.g., age group, sex).

+ **covariates** (optional): predictor variables that may improve the 
nowcast (e.g., weather covariates).

+ **is_censored** (optional): identifies cases where some reports arrive 
in artificial "batches" representing right-censoring rather than true 
reporting dynamics (e.g. when there is a system error and reports from a lab 
are not registered until a week after)

+ **case_count** (optional): the column storing case counts when the dataset 
is aggregated.

+ **temporal_effects** (optional): a lazy specification (list of [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html) configs) stored via [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html). Columns are only materialised when [compute_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/compute_temporal_effects.html) is called. See the [temporal effects](#temporal-effects) section for more details.

You can access any attribute using the corresponding [getter](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html), e.g. [get_event_date()](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html) or [get_strata()](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html).

[^1]: Optional attributes are set to `NULL` by default. 


### Data types 

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)  can represent one of three data structures:


1. **Linelist**: Each row corresponds to a single reported observation.

```{r, echo=FALSE, cache=FALSE}
df_linelist <- data.frame(
    patient     = 1:6,
   event_date  = c(rep(ymd("2020/09/12"), 3),
                   rep(ymd("2020/09/13"), 3)),
   report_date = c(ymd("2020/09/12"),
                  ymd("2020/09/12"),
                  ymd("2020/09/13"),
                  ymd("2020/09/13"),
                  ymd("2020/09/13"),
                  ymd("2020/09/13")))

df_linelist <- df_linelist %>% 
    tbl_now(event_date = event_date, report_date = report_date, 
            case_col = n, verbose = FALSE)

knitr::kable(df_linelist, caption = "Linelist data")
```

2. **Count-incidence**: Each row summarizes how many events with a given `event_date` were reported exactly on that `report_date.`

```{r, echo=FALSE, cache=FALSE}
df_count_inc <- data.frame(
  n           = c(7, 1, 9, 5, 0, 2),
  event_date  = c(rep(ymd("2020/09/12"), 3),
                  rep(ymd("2020/09/13"), 3)),
  report_date = c(ymd("2020/09/12"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/15"))) 

df_count_inc <- df_count_inc %>% 
  tbl_now(event_date = event_date, report_date = report_date, 
          case_count = n, verbose = FALSE)

knitr::kable(df_count_inc, caption = "Count-incidence data")
```

3. **Count-cumulative** Each row summarizes how many events with a given `event_date` had been reported up to and including that `report_date`. The distinction is crucial for nowcasting models that operate either on daily increments or cumulative totals.

```{r, echo=FALSE, cache=FALSE}
df_count_cum <- data.frame(
  n           = c(1,5, 8, 2, 2, 4),
  event_date  = c(rep(ymd("2020/09/12"), 3),
                  rep(ymd("2020/09/13"), 3)),
  report_date = c(ymd("2020/09/12"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/13"),
                  ymd("2020/09/14"),
                  ymd("2020/09/15"))) 

df_count_cum <- df_count_cum %>% 
    tbl_now(event_date = event_date, report_date = report_date, 
            case_count = n, verbose = FALSE)

knitr::kable(df_count_cum, caption = "Count-cumulative data")
```

The `to_count()` function allows you to easily convert between different 
data-types.

### Converting Between Data Types

The [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html) function supports structured transformations:

+ **Linelist → Count-Incidence**: Aggregates by event–report date, counting only cases reported on that date.

```{r}
df_linelist %>% 
  to_count(to = "count-incidence")
```

+ **Linelist → Count-Cumulative**: Aggregates by event–report date, producing cumulative counts up to each report date.

```{r}
df_linelist %>% 
  to_count(to = "count-cumulative")
```

> **Note** In the previous example the `n` counts `3` as it is aggregating the `1` observed at `report_date = 2020-09-13` and the `2` observed at `report_date = 2020-09-12`. This is the difference between the **count-incidence** that specifies the ones observed **exactly** on that date and the **count-cumulative** that specifies the ones observed up **until and including** that date. 

+ **Count-Incidence → Count-Cumulative**: Computes cumulative sums for each event date across report dates.

```{r}
df_count_inc %>% 
  to_count(to = "count-cumulative")
```

+ **Aggregation within the same type**: The [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html) may also be used to re-aggregate datasets that contain duplicate event–report pairs. This is useful when raw surveillance feeds contain repeated entries such as in this case:

```{r}
df_example <- data.frame(
  n           = c(8, 11, 0, 1, 1, 5, 2, 4, 1, 10, 9, 11, 3, 1),
  sex         = c(rep("M", 3), rep("F", 4), rep("M", 2), rep("F", 5)),
  event_date  = c(rep(ymd("2020/09/12"), 3),
                  rep(ymd("2020/09/12"), 4),
                  rep(ymd("2020/09/13"), 2),
                  rep(ymd("2020/09/13"), 5)),
  report_date = c(ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
                  ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
                  ymd("2020/09/15"), ymd("2020/09/13"), ymd("2020/09/14"),
                  ymd("2020/09/13"), ymd("2020/09/14"),
                  ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17"))) 

tbl_example <- df_example %>% 
  tbl_now(event_date = event_date, report_date = report_date, 
          data_type = "count-incidence", case_count = n, verbose = FALSE,
          warn_non_uniqueness = FALSE)

tbl_example
```

This dataset intentionally contains repeated event_date–report_date pairs within each stratum. If the user wants to aggregate redundant rows, the [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html)  function can be used to collapse duplicates by summing the `case_count` column.


```{r}
tbl_example %>% 
  to_count(to = "count-incidence")
```

The function ensures that:

+ Rows are grouped by `event_date`, `report_date`, and any `strata`, and `is_censored`.

+ The `case_count` column is summed within each group.

+ Attributes are preserved so the resulting object remains a valid `tbl_now`.


### Temporal effects

Often, temporal covariates improve nowcasting performance by helping adjust for systematic changes occurring within the calendar cycle (e.g., day-of-week effects, seasonal effects, or reporting artefacts). The [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html) function creates a *specification* (recipe) of the features to compute:

```{r}
library(almanac)

t_eff <- temporal_effects(
  day_of_week  = TRUE,
  week_of_year = TRUE,
  holidays     = cal_us_federal())
t_eff
```

> Note that the holidays calendar is an [rcalendar](https://davisvaughan.github.io/almanac/reference/rcalendar.html) object from the [almanac](https://davisvaughan.github.io/almanac/articles/almanac.html) package.

#### Lazy evaluation: attach first, compute later

Temporal effects in `tbl.now` follow a **lazy evaluation** pattern:

1. **Attach the spec** with [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html) (or via the `t_effects` argument of `tbl_now()`). This records *what* should be computed but adds **no columns** yet.

2. **Materialise the columns** with [compute_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/compute_temporal_effects.html) when you are ready to use them in a model.

```{r, R.options=list(pillar.width=250)}
data("denguedat")

# Step 1 — create the tbl_now and attach the spec (no columns added yet)
df_now <- denguedat %>%
  tbl_now(event_date = onset_week, report_date = report_week,
          verbose = FALSE, strata = gender)

df_now <- df_now %>%
  add_temporal_effects(t_eff)

# The footer shows "T. effects (lazy): ..." — spec is recorded but not computed
df_now
```

The footer now shows the spec with `(lazy)` to signal that the columns have not been computed yet. No new columns appear in the tibble at this point.

```{r, R.options=list(pillar.width=250)}
# Step 2 — materialise the columns when needed
df_computed <- compute_temporal_effects(df_now)

# Columns are now present and annotated [t_effect]
df_computed
```

After `compute_temporal_effects()`:

+ The effect columns (`.event_day_of_week`, `.event_week_of_year`, `.event_holiday`) are added.
+ `get_temporal_effect_cols()` lists them.
+ The original spec remains accessible via `get_temporal_effects()`, so you always know which effects were requested even after further dplyr operations.

```{r}
get_temporal_effects(df_computed)   # The spec (list of configs)
get_temporal_effect_cols(df_computed) # The computed column names
```

#### dplyr operations preserve the spec

A key design property: **`filter()`, `select()`, `mutate()`, and all other dplyr verbs preserve the `temporal_effects` spec**. They do not trigger re-computation. Only `compute_temporal_effects()` computes the columns.

```{r}
# Filtering changes the rows but keeps the spec intact
df_filtered <- df_now %>%
  dplyr::filter(report_week <= as.Date("1991-06-01"))

identical(get_temporal_effects(df_filtered), get_temporal_effects(df_now))
```

#### You can also supply the spec directly to `tbl_now()`

```{r}
df_with_spec <- denguedat %>%
  tbl_now(event_date = onset_week, report_date = report_week,
          t_effects = temporal_effects(week_of_year = TRUE),
          verbose = FALSE)

# No effect columns yet — spec is stored lazily
".event_week_of_year" %in% names(df_with_spec)

# Compute on demand
df_with_spec <- compute_temporal_effects(df_with_spec)
".event_week_of_year" %in% names(df_with_spec)
```

### Modifying a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) with `dplyr`

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)’s attributes can be modified using the functions in [add_*](https://rodrigozepeda.github.io/tbl.now/reference/add.html), [change_*](https://rodrigozepeda.github.io/tbl.now/reference/change.html), or [remove_*](https://rodrigozepeda.github.io/tbl.now/reference/remove.html). These functions share a consistent interface that allows users to incrementally manipulate strata, covariates, and temporal effects.

The example below demonstrates how to create a  [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html), add strata and temporal effects, later modify the strata, and finally remove the temporal effects.

```{r, message=FALSE}
data("mpoxdat")

df_now <- mpoxdat %>%
  tbl_now(event_date = dx_date, report_date = dx_report_date,
          case_count = n, verbose = FALSE, strata = race)

df_now
```

You can see that the strata is `race` with the corresponding [get_*](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html):

```{r}
get_strata(df_now)
```

Strata can be modified with the [change_*](https://rodrigozepeda.github.io/tbl.now/reference/change.html) family of functions. The following example adds a new column containing an uppercase version of the existing race variable and sets it as the new strata:

```{r}
df_now <- df_now %>%
  mutate(RACE_UPPER = toupper(race)) %>%
  change_strata(RACE_UPPER)

get_strata(df_now)
```

To attach a lazy temporal-effects spec, use [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html), then materialise with [compute_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/compute_temporal_effects.html):

```{r, R.options=list(pillar.width = 250)}
df_now <- df_now %>%
  add_temporal_effects(temporal_effects(week_of_year = TRUE))

# Spec is stored (lazy):
get_temporal_effects(df_now)

# Compute to add columns:
df_now <- compute_temporal_effects(df_now)
get_temporal_effect_cols(df_now)
```

Attributes can be removed using the corresponding [remove_*](https://rodrigozepeda.github.io/tbl.now/reference/remove.html) functions. `remove_temporal_effects()` drops both the spec and any computed columns:

```{r, R.options=list(pillar.width = 250)}
df_now <- df_now %>%
  remove_temporal_effects() %>%
  remove_all_strata()

df_now
```

```{r}
get_temporal_effects(df_now)  # Empty list — no spec
get_temporal_effect_cols(df_now)  # character(0) — no computed cols
get_strata(df_now)
```

## Modifying a tbl_now() with dplyr

[tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) objects extend [tibble()](https://tibble.tidyverse.org/), and therefore support standard [dplyr](https://dplyr.tidyverse.org/)
verbs. The class attempts to preserve and adapt its internal attributes when operations are performed. For example, renaming a strata column will automatically update the stored strata attribute.

```{r}
library(dplyr, quietly = TRUE)
data(denguedat)

df_now <- tbl_now(denguedat, event_date = onset_week, 
                  report_date = report_week, strata = gender,
                  verbose = FALSE)

#Current strata
get_strata(df_now)
```


After renaming the column, the strata attribute updates accordingly:

```{r}
df_now <- df_now %>% 
  rename(male_or_female = gender)

get_strata(df_now)
```

Certain operations may cause a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) object to drop back to a standard tibble. This occurs when the operation removes necessary structure—for example, collapsing all data into a single row with [summarise()](https://dplyr.tidyverse.org/reference/summarise.html):

```{r}
df_now %>%
  summarise(number_males = sum(male_or_female == "Male"))
```
 
 
### Updating a tbl_now()

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) object can be updated using the [update()](https://rodrigozepeda.github.io/tbl.now/reference/update.tbl_now.html) method with another data.frame, tibble, or [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) as input. When the column structure is compatible, the update process retains the strata, covariate, and temporal-effect attributes from the original object, and recalculates "now" estimates using the combined data.

Below is an example using an initial dataset:

```{r}
df <- data.frame(
  patient = 1:6,
  event_date = c(rep(ymd("2020/09/12"), 3), rep(ymd("2020/09/13"), 3)),
  report_date = c(ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
                  ymd("2020/09/13"), ymd("2020/09/14"), ymd("2020/09/15")))

df_now <- tbl_now(df, event_date = event_date,
                  report_date = report_date, verbose = FALSE)
```

And a follow-up dataset containing newly reported cases:

```{r}
df_new <- data.frame(
  patient = 7:13,
  event_date = c(ymd("2020/09/13"), 
                 rep(ymd("2020/09/14"), 3), 
                 rep(ymd("2020/09/15"), 3)),
  report_date = c(ymd("2020/09/14"), ymd("2020/09/14"), ymd("2020/09/15"), 
                  ymd("2020/09/16"), ymd("2020/09/15"), ymd("2020/09/16"), 
                  ymd("2020/09/17")))
```

We can update the original object by incorporating the new data:

```{r}
df_updated <- update(df_now, new_data = df_new)

df_updated
```

## Visualizing a `tbl_now`

The [autoplot()](https://rodrigozepeda.github.io/tbl.now/reference/autoplot.tbl_now.html) method gives a quick four-panel diagnostic overview of a `tbl_now`. Using [ggplot2](https://ggplot2.tidyverse.org/) and [patchwork](https://patchwork.data-imaginist.com/), it shows:

1. **Empirical delay distribution**: a case-count weighted histogram of the reporting delay.
2. **Observed epidemic process**: the latest reported counts per `event_date`.
3. **Calendar effect**: boxplots of the *normalized* effect (each event date's cases divided by the overall mean, so 1 is average), built directly from the data.
4. **Seasonality** — a periodogram of the incidence series whose dominant peak suggests a Fourier season length to pass to [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html).

```{r autoplot-available, include=FALSE}
plotting_available <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("patchwork", quietly = TRUE)
```

```{r autoplot, eval=plotting_available, fig.width=9, fig.height=6.5, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)

dengue_now <- tbl_now(denguedat,
                      event_date  = "onset_week",
                      report_date = "report_week",
                      verbose     = FALSE)

autoplot(dengue_now)
```

For the weekly dengue data the periodogram peaks near 52 weeks, suggesting an annual cycle. We could capture this with a Fourier term of `temporal_effects(seasons = 52)`.

## Other functions (utilities)

### Convert epidemiological weeks to dates

The function [week_2_date()](https://rodrigozepeda.github.io/tbl.now/reference/week_2_date.html) converts epidemiological week/year combinations into an aligned calendar date.

```{r}
df <- data.frame(
  epidemiological_week = 1:5,
  epidemiological_year = rep(2024, 5)
  )

df %>%
  week_2_date(
    week_col = epidemiological_week,
    year_col = epidemiological_year
    )
```
 
### Reports

The functions [get_initial_reported_cases() and get_latest_reported_cases()](https://rodrigozepeda.github.io/tbl.now/reference/get_latest_first.html) extract the number of cases first reported for each event date and the most recently reported totals, respectively. These utilities allow users to quantify revisions between initial and final reports.
 
 
```{r}
df_reports <- data.frame(
    n           = c(10, 1, 1, 0, 0, 3),
    event_date  = rep(ymd("2020/09/12"), 6),
    report_date = c(ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
                    ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17")))

tbl_reports <- df_reports %>%
  tbl_now(event_date = event_date, report_date = report_date, 
          verbose = FALSE, case_count = n, report_units = "days",
          event_units = "days")

tbl_reports
```


The initial reported cases:

```{r}
get_initial_reported_cases(tbl_reports)
```

and the latest totals:

```{r}
get_latest_reported_cases(tbl_reports)
```

### Week alignment

The [align_weeks()](https://rodrigozepeda.github.io/tbl.now/reference/align_weeks.html) 
function standardizes dates within the same epidemiological week to a single reference day
(for example, the start of the week). This is helpful when computing differences across weekly reporting periods, avoiding fractional time intervals.

```{r}
df <- data.frame(
  date = c(ymd("2022-10-31"), ymd("2022-11-07"), ymd("2022-11-13")),
  epiweek = c(44, 45, 46)
)

# Align to Sundays
df_aligned <- align_weeks(df, date_col = date)
df_aligned
```

You can verify the resulting weekday using the [wday() function from the lubridate package](https://lubridate.tidyverse.org/reference/day.html):

```{r}
df_aligned %>%
  mutate(day_label = wday(date_aligned, label = TRUE, abbr = FALSE))
```

### Complete zeroes

The [complete_zeroes()](https://rodrigozepeda.github.io/tbl.now/reference/complete_zeroes.html) function fills with zeroes those cases where the `event` or `report` weeks have not been observed. 

Consider for example the following data with just two observations per date:

```{r}
ndata <- tibble(
  event_date  = c(as.Date("2021/01/12"), as.Date("2021/01/14"), as.Date("2021/01/14")),
  report_date = c(as.Date("2021/01/13"), as.Date("2021/01/15"), as.Date("2021/01/18")),
  case_count = c(10, 5, 1)
)

ndata <- tbl_now(ndata, event_date, report_date,
     verbose = FALSE, case_count = case_count, data_type = "count-incidence")

ndata
```

Notice that there are no observations for `2021/01/13`. Furthermore, if we assume that the maximum possible observed  We can fill the unobserved cases with:

```{r}
complete_zeroes(ndata)
```

Which looks at all the possible report dates and event dates and sets the counts to zero if they have not been observed.

### Censoring extreme delays

The function `censor_delays_above()` marks all delays above a threshold value (`max_delay`) as censored. This is useful to indicate extreme delays in some nowcast models:

```{r}
df <- data.frame(onset = as.Date("2020-01-01") + c(0, 0, 1, 2),
                 reported = as.Date("2020-01-01") + c(1, 5, 2, 300))
tn <- tbl_now(df, event_date = onset, report_date = reported,
              data_type = "linelist", verbose = FALSE)

# the 300-day report becomes censored (an upper bound on its delay)
censor_delays_above(tn, max_delay = 60)
```

### Converting to data formats from other packages

> **NOTE** This is still work in progress  `r lifecycle::badge("experimental")`

`tbl.now` ships converters that move data between a `tbl_now` and the data structures used by other nowcasting and delay-estimation packages. They all follow the same naming convention:

+ `tbl_now_from_*()` builds a `tbl_now` (it wraps [as_tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/as_tbl_now.html), so `...` is forwarded to [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)).

+ `tbl_now_to_*()` converts a `tbl_now` into that package's native object.

Each function accepts a `verbose` argument that reports the choices it made (the inferred `now`, the data type, the units, the column mapping, and so on). The supported packages — [epinowcast](https://package.epinowcast.org/), [baselinenowcast](https://baselinenowcast.epinowcast.org/), [EpiNow2](https://epiforecasts.io/EpiNow2/), [epidist](https://epidist.epinowcast.org/), [data.table](https://rdatatable.gitlab.io/data.table/) and [tsibble](https://tsibble.tidyverts.org/) — are all optional (`Suggests`).

| Package | `from` | `to` | Maps to |
|---|---|---|---|
| `epinowcast` | yes | yes | `count-cumulative` (`reference_date`, `report_date`, `confirm`) |
| `baselinenowcast` | yes | yes | `count-incidence` (long data frame or reporting-triangle matrix) |
| `EpiNow2` | — | yes | a single `date`/`confirm` incidence series |
| `data.table` | yes | yes | any data type |
| `epidist` | yes | yes | `linelist` (primary/secondary event dates) |
| `tsibble` | yes | yes | a `tbl_ts` (one date as index, the other date + strata as key) |

The object produced by any `tbl_now_to_*()` can be converted straight back with [as_tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/as_tbl_now.html), which has methods for each of these classes:

```{r converters-roundtrip, eval=requireNamespace("tsibble", quietly = TRUE), message=FALSE}
dengue_now <- tbl_now(denguedat, event_date = "onset_week",
                      report_date = "report_week", strata = "gender",
                      verbose = FALSE)

# tbl_now -> tsibble -> tbl_now
dengue_ts <- tbl_now_to_tsibble(dengue_now, verbose = FALSE)
as_tbl_now(dengue_ts, event_date = "onset_week", verbose = FALSE)
```

For example, the [epinowcast](https://package.epinowcast.org/) hospitalization data ships in long cumulative format and converts straight into a `count-cumulative` `tbl_now`:

```{r converters-available, include=FALSE}
epinowcast_available <- requireNamespace("epinowcast", quietly = TRUE)
```

```{r converters-from, eval=epinowcast_available, message=FALSE}
germany_hospitalizations <- head(epinowcast::germany_covid19_hosp, 1000)

hospitalizations_now <- tbl_now_from_epinowcast(
  germany_hospitalizations,
  strata = c("location", "age_group")
)

hospitalizations_now
```

Going the other way, `tbl_now_to_epinowcast()` returns an `enw_preprocess_data` object ready for [epinowcast::epinowcast()](https://package.epinowcast.org/reference/epinowcast.html):

```{r converters-to, eval=epinowcast_available, message=FALSE}
epinowcast_object <- tbl_now_to_epinowcast(hospitalizations_now, verbose = FALSE)
class(epinowcast_object)
```

The same pattern works for the other packages, for example `tbl_now_to_EpiNow2()` collapses the object into the single `date`/`confirm` time series that [EpiNow2](https://epiforecasts.io/EpiNow2/) expects.

## References

