---
title: "tbl.now"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{tbl.now}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Within `R`, the [tidyverse](https://tidyverse.org/) ecosystem [@tidyverse] provides a widely adopted framework for data analysis. In this paradigm, data is structured such that rows represent individual observations and columns represent variables. This approach is commonly referred to as tidy data [@wickham2014tidy].

Several tidyverse extensions exist for working with time series data, including [tsibble](https://tsibble.tidyverts.org/), [tibbletime](https://business-science.github.io/tibbletime/), and [timetk](https://business-science.github.io/timetk/) [@wang2020new; @wang2019tidy; @timetk]. These packages provide time-aware tibbles and tools that integrate smoothly with the tidyverse. However, epidemiological nowcasting requires two time indices simultaneously: an **event time** and a **reporting time**. Classical time-series abstractions assume only a single index and therefore do not fully support this structure.

The `tbl.now` class and package addresses this gap. The `tbl.now` extends a regular  [tibble()](https://tibble.tidyverse.org/) to explicitly encode epidemiological event and report dates, allowing consistent data transformation, delay computation, and integration with the [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) modeling workflow.

More concretely, tbl.now was designed to:

+ Standardize the data inputs required by [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/)  models.

+ Preserve  [tidyverse](https://tidyverse.org/)  compatibility so users can continue to apply familiar [dplyr](https://dplyr.tidyverse.org/) operations.

+ Diagnose reporting artefacts directly from the data — reporting-delay drift and change points ([test_delay_drift()](https://rodrigozepeda.github.io/tbl.now/reference/test_delay_drift.html), [test_delay_changepoint()](https://rodrigozepeda.github.io/tbl.now/reference/test_delay_changepoint.html)) and batch (backlog) reporting ([detect_report_batches()](https://rodrigozepeda.github.io/tbl.now/reference/detect_report_batches.html)).

+ Facilitate seamless integration into iterative modeling workflows [@gelman2020bayesian; @wickham2023r]:

```
Data Cleaning -> Modeling -> New Data Cleaning -> New Modeling -> ...
```

We'll begin by loading the required packages:

```{r setup, message=FALSE}
library(dplyr, quietly = TRUE)
library(lubridate)
library(tbl.now)
```



## The `tbl.now` Framework 

In an epidemiological nowcast, we typically observe two key dates:

+ `event_date`: when the underlying event occurred (e.g., symptom onset, exposure, sample collection).

+ `report_date`: when the event was recorded in the data system (e.g., lab result processed, clinical visit documented).

The nowcasting task is:

> To estimate, for each past `event_date`, how many events have already occurred 
but have not yet been reported as of **now**. That is, the nowcast will predict how 
many observations will eventually be observed for the `event_date`s. 

A `tbl.now` object is therefore a specialized [tibble()](https://tibble.tidyverse.org/) that:

+ Identifies the `event_date` and `report_date` columns.

+ Stores these as attributes to enable consistent processing.

+ Automatically computes auxiliary fields such as delay, numerical indices, and frequency units.

+ Ensures the dataset is well-formed for downstream [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) models.

### Example

Consider the following dataset:

```{r}
df <- data.frame(
  symptom_onset = c(ymd("2023/12/25"), ymd("2023/12/26"), ymd("2023/12/25"), ymd("2023/12/26")),
  medical_visit = c(ymd("2023/12/26"), ymd("2023/12/26"), ymd("2023/12/27"), ymd("2023/12/27")),
  n = c(10, 2, 5, 11)
)

knitr::kable(df, caption = "Example dataset")
```

Here:

+ `symptom_onset` is the event date.

+ `medical_visit` is the report date.

+ `n` is the number of reported cases for each event–report combination.

We can convert this into a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html):

```{r}
df |>
  tbl_now(event_date = symptom_onset, report_date = medical_visit, case_count = n)
```


This performs several operations automatically:

+ Detects the **data type** (`count-incidence` in this example).

+ Infers the **frequency units** of event and report dates (daily).

+ Tags the correct columns as `event_date`, `report_date`, and `case_count.`

+ Computes `.event_num`, `.report_num`, and `.delay` columns.

+ Identifies the appropriate **now** date (the most recent report date).

The remaining sections describe these features and the broader `tbl.now` toolkit.


## Attributes of a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) 

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) stores information about its structure using object attributes, ensuring consistent behavior across transformations and modeling steps. The primary attributes are:

+ **now**: the "current" reporting date used for nowcasting, typically the most recent `report_date`.

+ **event_date**: the column name storing event dates.

+ **report_date**: the column name storing report dates.

+ **event_units**: the temporal units for event dates (e.g., "days", "weeks", "numeric").

+ **report_units**: the temporal units for report dates.

+ **data_type**: one of the following (see [the data types section](#data-types)):
    + linelist
    + count-incidence
    + count-cumulative.

+ **strata** (optional)[^1]: variables for which the nowcast should be 
computed separately (e.g., age group, sex).

+ **covariates** (optional): predictor variables that may improve the 
nowcast (e.g., weather covariates).

+ **is_censored** (optional): identifies cases where some reports arrive 
in artificial "batches" representing right-censoring rather than true 
reporting dynamics (e.g. when there is a system error and reports from a lab 
are not registered until a week after)

+ **case_count** (optional): the column storing case counts when the dataset 
is aggregated.

+ **temporal_effects** (optional): a lazy specification (list of [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html) configs) stored via [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html). Columns are only materialised when [compute_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/compute_temporal_effects.html) is called. See the [temporal effects](#temporal-effects) section for more details.

You can access any attribute using the corresponding [getter](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html), e.g. [get_event_date()](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html) or [get_strata()](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html).

[^1]: Optional attributes are set to `NULL` by default. 


### Data types 

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)  can represent one of three data structures:


1. **Linelist**: Each row corresponds to a single reported observation.

```{r, echo=FALSE, cache=FALSE}
df_linelist <- data.frame(
  patient = 1:6,
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/13"), 3)
  ),
  report_date = c(
    ymd("2020/09/12"),
    ymd("2020/09/12"),
    ymd("2020/09/13"),
    ymd("2020/09/13"),
    ymd("2020/09/13"),
    ymd("2020/09/13")
  )
)

df_linelist <- df_linelist |>
  tbl_now(
    event_date = event_date, report_date = report_date,
    case_col = n, verbose = FALSE
  )

knitr::kable(df_linelist, caption = "Linelist data")
```

2. **Count-incidence**: Each row summarizes how many events with a given `event_date` were reported exactly on that `report_date.`

```{r, echo=FALSE, cache=FALSE}
df_count_inc <- data.frame(
  n = c(7, 1, 9, 5, 0, 2),
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/13"), 3)
  ),
  report_date = c(
    ymd("2020/09/12"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/15")
  )
)

df_count_inc <- df_count_inc |>
  tbl_now(
    event_date = event_date, report_date = report_date,
    case_count = n, verbose = FALSE
  )

knitr::kable(df_count_inc, caption = "Count-incidence data")
```

3. **Count-cumulative** Each row summarizes how many events with a given `event_date` had been reported up to and including that `report_date`. The distinction is crucial for nowcasting models that operate either on daily increments or cumulative totals.

```{r, echo=FALSE, cache=FALSE}
df_count_cum <- data.frame(
  n = c(1, 5, 8, 2, 2, 4),
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/13"), 3)
  ),
  report_date = c(
    ymd("2020/09/12"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/15")
  )
)

df_count_cum <- df_count_cum |>
  tbl_now(
    event_date = event_date, report_date = report_date,
    case_count = n, verbose = FALSE
  )

knitr::kable(df_count_cum, caption = "Count-cumulative data")
```

The `to_count()` function allows you to easily convert between different 
data-types.

### Converting Between Data Types

The [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html) function supports structured transformations:

+ **Linelist → Count-Incidence**: Aggregates by event–report date, counting only cases reported on that date.

```{r}
df_linelist |>
  to_count(to = "count-incidence")
```

+ **Linelist → Count-Cumulative**: Aggregates by event–report date, producing cumulative counts up to each report date.

```{r}
df_linelist |>
  to_count(to = "count-cumulative")
```

> **Note** In the previous example the `n` counts `3` as it is aggregating the `1` observed at `report_date = 2020-09-13` and the `2` observed at `report_date = 2020-09-12`. This is the difference between the **count-incidence** that specifies the ones observed **exactly** on that date and the **count-cumulative** that specifies the ones observed up **until and including** that date. 

+ **Count-Incidence → Count-Cumulative**: Computes cumulative sums for each event date across report dates.

```{r}
df_count_inc |>
  to_count(to = "count-cumulative")
```

+ **Aggregation within the same type**: The [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html) may also be used to re-aggregate datasets that contain duplicate event–report pairs. This is useful when raw surveillance feeds contain repeated entries such as in this case:

```{r}
df_example <- data.frame(
  n = c(8, 11, 0, 1, 1, 5, 2, 4, 1, 10, 9, 11, 3, 1),
  sex = c(rep("M", 3), rep("F", 4), rep("M", 2), rep("F", 5)),
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/12"), 4),
    rep(ymd("2020/09/13"), 2),
    rep(ymd("2020/09/13"), 5)
  ),
  report_date = c(
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/15"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17")
  )
)

tbl_example <- df_example |>
  tbl_now(
    event_date = event_date, report_date = report_date,
    data_type = "count-incidence", case_count = n, verbose = FALSE,
    warn_non_uniqueness = FALSE
  )

tbl_example
```

This dataset intentionally contains repeated event_date–report_date pairs within each stratum. If the user wants to aggregate redundant rows, the [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html)  function can be used to collapse duplicates by summing the `case_count` column.


```{r}
tbl_example |>
  to_count(to = "count-incidence")
```

The function ensures that:

+ Rows are grouped by `event_date`, `report_date`, and any `strata`, and `is_censored`.

+ The `case_count` column is summed within each group.

+ Attributes are preserved so the resulting object remains a valid `tbl_now`.


### Temporal effects

Often, temporal covariates improve nowcasting performance by helping adjust for systematic changes occurring within the calendar cycle (e.g., day-of-week effects, seasonal effects, or reporting artefacts). The [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html) function creates a *specification* (recipe) of the features to compute:

```{r}
library(almanac)

t_eff <- temporal_effects(
  day_of_week  = TRUE,
  week_of_year = TRUE,
  holidays     = cal_us_federal()
)
t_eff
```

> Note that the holidays calendar is an [rcalendar](https://davisvaughan.github.io/almanac/reference/rcalendar.html) object from the [almanac](https://davisvaughan.github.io/almanac/articles/almanac.html) package.

#### Lazy evaluation: attach first, compute later

Temporal effects in `tbl.now` follow a **lazy evaluation** pattern:

1. **Attach the spec** with [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html) (or via the `t_effects` argument of `tbl_now()`). This records *what* should be computed but adds **no columns** yet.

2. **Materialise the columns** with [compute_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/compute_temporal_effects.html) when you are ready to use them in a model.

```{r, R.options=list(pillar.width=250)}
data("denguedat")

# Step 1 — create the tbl_now and attach the spec (no columns added yet)
df_now <- denguedat |>
  tbl_now(
    event_date = onset_week, report_date = report_week,
    verbose = FALSE, strata = gender
  )

df_now <- df_now |>
  add_temporal_effects(t_eff)

# The footer shows "T. effects (lazy): ..." — spec is recorded but not computed
df_now
```

The footer now shows the spec with `(lazy)` to signal that the columns have not been computed yet. No new columns appear in the tibble at this point.

```{r, R.options=list(pillar.width=250)}
# Step 2 — materialise the columns when needed
df_computed <- compute_temporal_effects(df_now)

# Columns are now present and annotated [t_effect]
df_computed
```

After `compute_temporal_effects()`:

+ The effect columns (`.event_day_of_week`, `.event_week_of_year`, `.event_holiday`) are added.
+ `get_temporal_effect_cols()` lists them.
+ The original spec remains accessible via `get_temporal_effects()`, so you always know which effects were requested even after further dplyr operations.

```{r}
get_temporal_effects(df_computed) # The spec (list of configs)
get_temporal_effect_cols(df_computed) # The computed column names
```

#### Around-holiday and around-weekend effects

Reporting often *rebounds* on the first working day(s) after a holiday or a
weekend. To capture that, `temporal_effects()` has `holiday_lags` and
`weekend_lags`: each takes a depth `N` and creates indicator columns
`..._holiday_lag_1 … ..._holiday_lag_N` (and `..._weekend_lag_k`) that flag dates
falling exactly `k` **working days** after a holiday / weekend. Working days skip
weekends and other holidays, so the effect lands on the first day back at work.

```{r}
# Flag the two working days after a holiday, and the working day after a weekend
after_eff <- temporal_effects(
  holidays     = cal_us_federal(),
  holiday_lags = 2,
  weekend_lags = 1
)
after_eff
```

The mirror image — a slowdown in the days *leading up to* a break — is a negative
depth. `..._holiday_lead_k` / `..._weekend_lead_k` then flag dates `k` working
days **before** a holiday / weekend, counting backwards from it, so `_lead_1` is
the working day closest to the break:

```{r}
# Flag Christmas Eve (and the eve of every other holiday), plus the Wednesday,
# Thursday and Friday before each weekend
before_eff <- temporal_effects(
  holidays     = cal_us_federal(),
  holiday_lags = -1,
  weekend_lags = -3
)
before_eff
```

To model both sides of the same break, attach one specification per direction:

```{r, eval=FALSE}
df_now |>
  add_temporal_effects(temporal_effects(weekend_lags = -1)) |> # the Friday before
  add_temporal_effects(temporal_effects(weekend_lags = 1)) # the Monday after
```

#### Event- vs report-date effects

By default effects are derived from the **event date** and named `.event_*`. Pass
`date_type = "report_date"` to `add_temporal_effects()` to derive them from the
**report date** instead (columns named `.report_*`). Both can coexist on the same
`tbl_now`, and every converter carries both sets of columns.

```{r, eval=FALSE}
df_now |>
  add_temporal_effects(temporal_effects(week_of_year = TRUE), date_type = "event_date") |>
  add_temporal_effects(temporal_effects(day_of_week = TRUE),  date_type = "report_date")
```


### Modifying a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) with `dplyr`

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)’s attributes can be modified using the functions in [add_*](https://rodrigozepeda.github.io/tbl.now/reference/add.html), [change_*](https://rodrigozepeda.github.io/tbl.now/reference/change.html), or [remove_*](https://rodrigozepeda.github.io/tbl.now/reference/remove.html). These functions share a consistent interface that allows users to incrementally manipulate strata, covariates, and temporal effects.

The example below demonstrates how to create a  [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html), add strata and temporal effects, later modify the strata, and finally remove the temporal effects.

```{r, message=FALSE}
data("mpoxdat")

df_now <- mpoxdat |>
  tbl_now(
    event_date = dx_date, report_date = dx_report_date,
    case_count = n, verbose = FALSE, strata = race
  )

df_now
```

You can see that the strata is `race` with the corresponding [get_*](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html):

```{r}
get_strata(df_now)
```

Strata can be modified with the [change_*](https://rodrigozepeda.github.io/tbl.now/reference/change.html) family of functions. The following example adds a new column containing an uppercase version of the existing race variable and sets it as the new strata:

```{r}
df_now <- df_now |>
  mutate(RACE_UPPER = toupper(race)) |>
  change_strata(RACE_UPPER)

get_strata(df_now)
```

To attach a lazy temporal-effects spec, use [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html), then materialise with [compute_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/compute_temporal_effects.html):

```{r, R.options=list(pillar.width = 250)}
df_now <- df_now |>
  add_temporal_effects(temporal_effects(week_of_year = TRUE))

# Spec is stored (lazy):
get_temporal_effects(df_now)

# Compute to add columns:
df_now <- compute_temporal_effects(df_now)
get_temporal_effect_cols(df_now)
```

Attributes can be removed using the corresponding [remove_*](https://rodrigozepeda.github.io/tbl.now/reference/remove.html) functions. `remove_temporal_effects()` drops both the spec and any computed columns:

```{r, R.options=list(pillar.width = 250)}
df_now <- df_now |>
  remove_temporal_effects() |>
  remove_all_strata()

df_now
```

```{r}
get_temporal_effects(df_now) # Empty list — no spec
get_temporal_effect_cols(df_now) # character(0) — no computed cols
get_strata(df_now)
```

## Modifying a tbl_now() with dplyr

[tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) objects extend [tibble()](https://tibble.tidyverse.org/), and therefore support standard [dplyr](https://dplyr.tidyverse.org/)
verbs. The class attempts to preserve and adapt its internal attributes when operations are performed. For example, renaming a strata column will automatically update the stored strata attribute.

```{r}
library(dplyr, quietly = TRUE)
data(denguedat)

df_now <- tbl_now(denguedat,
  event_date = onset_week,
  report_date = report_week, strata = gender,
  verbose = FALSE
)

# Current strata
get_strata(df_now)
```


After renaming the column, the strata attribute updates accordingly:

```{r}
df_now <- df_now |>
  rename(male_or_female = gender)

get_strata(df_now)
```

Certain operations may cause a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) object to drop back to a standard tibble. This occurs when the operation removes necessary structure—for example, collapsing all data into a single row with [summarise()](https://dplyr.tidyverse.org/reference/summarise.html):

```{r}
df_now |>
  summarise(number_males = sum(male_or_female == "Male"))
```
 
 
### Updating a tbl_now()

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) object can be updated using the [update()](https://rodrigozepeda.github.io/tbl.now/reference/update.tbl_now.html) method with another data.frame, tibble, or [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) as input. When the column structure is compatible, the update process retains the strata, covariate, and temporal-effect attributes from the original object, and recalculates "now" estimates using the combined data.

Below is an example using an initial dataset:

```{r}
df <- data.frame(
  patient = 1:6,
  event_date = c(rep(ymd("2020/09/12"), 3), rep(ymd("2020/09/13"), 3)),
  report_date = c(
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/13"), ymd("2020/09/14"), ymd("2020/09/15")
  )
)

df_now <- tbl_now(df,
  event_date = event_date,
  report_date = report_date, verbose = FALSE
)
```

And a follow-up dataset containing newly reported cases:

```{r}
df_new <- data.frame(
  patient = 7:13,
  event_date = c(
    ymd("2020/09/13"),
    rep(ymd("2020/09/14"), 3),
    rep(ymd("2020/09/15"), 3)
  ),
  report_date = c(
    ymd("2020/09/14"), ymd("2020/09/14"), ymd("2020/09/15"),
    ymd("2020/09/16"), ymd("2020/09/15"), ymd("2020/09/16"),
    ymd("2020/09/17")
  )
)
```

We can update the original object by incorporating the new data:

```{r}
df_updated <- update(df_now, new_data = df_new)

df_updated
```

## Visualizing a `tbl_now`

The [autoplot()](https://rodrigozepeda.github.io/tbl.now/reference/autoplot.tbl_now.html) method gives a quick diagnostic overview of a `tbl_now`. Using [ggplot2](https://ggplot2.tidyverse.org/) and [patchwork](https://patchwork.data-imaginist.com/), it draws two families of panels. The **case-count** panels:

1. **Empirical delay distribution**: a case-count weighted histogram of the reporting delay.
2. **Observed epidemic process**: the latest reported counts per `event_date`.
3. **Calendar effect**: boxplots of the *normalized* effect (each event date's cases divided by the overall mean, so 1 is average) by day of week / week of year / month.
4. **Holiday effect** and **holiday lag effect**: the same normalized boxplots by *day type* and by *position relative to the nearest holiday* — see [Holiday effects](#holiday-effects) below.
5. **Seasonality** — a periodogram of the incidence series whose dominant peak suggests a Fourier season length to pass to [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html).

…and the **reporting-delay** panels, which reveal *delay effects* (whether the delay itself has a calendar or periodic pattern):

6. **Delay calendar effect**: boxplots of the mean reporting delay by day of week / week of year / month.
7. **Delay holiday effects**: the delay twins of panel 4.
8. **Delay periodicity**: a periodogram of the mean-delay series (e.g. a weekly reporting rhythm).

Which calendar/delay panels appear depends on the event unit (daily data gets day-of-week *and* week-of-year; weekly data week-of-year; monthly data month-of-year). Use the `panels` argument to pick a subset — for example `panels = "delay_calendar"` for just the delay boxplots, or a single key like `panels = "delay_week"` (a single panel is returned as a plain `ggplot`). The aliases are `"all"` (default), `"calendar"` and `"delay_calendar"`.

Set `by_strata = TRUE` to split every panel by stratum: the boxplots become dodged boxes (one per stratum, side by side), the epidemic process and periodograms become one coloured line per stratum, and the delay distribution becomes dodged bars. The boxplots are then normalized *per stratum* (1 = that stratum's own average) so the calendar pattern is comparable across strata. By default the object's `strata` are used; pass `strata = "gender"` to group on a subset.

The `strata` argument does **not** have to name a declared stratum — any column of the data works, and it is declared for you, so these two are the same plot:

```{r autoplot-strata-arg, eval=FALSE}
autoplot(mpox_now, strata = "race", by_strata = TRUE)
autoplot(mpox_now |> add_strata(race), by_strata = TRUE)
```

```{r autoplot-available, include=FALSE}
plotting_available <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("patchwork", quietly = TRUE)
```

```{r autoplot, eval=plotting_available, fig.width=9, fig.height=6.5, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)

dengue_now <- tbl_now(denguedat,
  event_date  = "onset_week",
  report_date = "report_week",
  verbose     = FALSE
)

autoplot(dengue_now)
```

For the weekly dengue data the periodogram peaks near 52 weeks, suggesting an annual cycle. We could capture this with a Fourier term of `temporal_effects(seasons = 52)`.

### Holiday effects {#holiday-effects}

The four holiday panels are different from the rest: they describe the [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html) spec you attached, so they appear only once there is one to describe. `calendar_holiday` / `delay_holiday` need a `holidays` calendar **or** `weekend = TRUE`; the two `_lag` panels also need a non-zero `holiday_lags`. The spec is read directly, so there is no need to call `compute_temporal_effects()` first.

Holidays are a daily phenomenon, so we switch to daily data — here a synthetic series whose reporting deliberately **stalls on weekends and holidays**:

```{r holiday-panels-setup, eval=plotting_available && requireNamespace("almanac", quietly = TRUE), message=FALSE}
library(almanac)
set.seed(42)

us_holidays <- cal_us_federal()
event_days <- seq(as.Date("2022-01-01"), as.Date("2023-12-31"), by = "day")

# Reporting takes longer when the desk is shut: +6 days on a holiday, +3 on a weekend.
reports <- lapply(event_days, function(day) {
  closed_for <- if (alma_in(day, us_holidays)) 6 else if (!is_weekday(day)) 3 else 0
  delays <- 0:rpois(1, 2 + closed_for)
  data.frame(onset = day, reported = day + delays, n = rpois(length(delays), 5))
})

holiday_now <- do.call(rbind, reports) |>
  tbl_now(
    event_date = onset, report_date = reported, case_count = n,
    data_type = "count-incidence", verbose = FALSE
  ) |>
  add_temporal_effects(
    temporal_effects(weekend = TRUE, holidays = us_holidays, holiday_lags = 2)
  )
```

The **holiday effect** panel splits the days by *type*. Which categories you get follows the spec: a calendar plus `weekend = TRUE` gives `Weekday` / `Weekend` / `Holiday`; a calendar alone gives `Non-holiday` / `Holiday`; a weekend effect alone gives `Weekday` / `Weekend`. A holiday that falls on a weekend counts as a **holiday**.

The **holiday lag effect** panel splits them by *position relative to the nearest holiday*, showing exactly the days that `holiday_lags` flags — `"1 after"`, `"2 after"`, and so on, with `"Other"` (every other day) as the reference to read them against. Negative depths add `"1 before"`, `"2 before"`, … to the left of the holiday, so the axis reads left-to-right as time does. Counting is in **working days**, so weekends and other holidays are skipped exactly as they are for the `..._holiday_lag_k` columns.

```{r holiday-panels, eval=plotting_available && requireNamespace("almanac", quietly = TRUE), fig.width=9, fig.height=7, message=FALSE, warning=FALSE}
autoplot(holiday_now, panels = c("calendar_holiday", "calendar_holiday_lag",
                                 "delay_holiday", "delay_holiday_lag"))
```

The **delay** twins (the green panels) are usually the more telling pair. A holiday rarely changes how many cases *occur* — but it very much changes how long they take to be *reported*, which is exactly what a nowcast has to model. Here the planted effect comes straight back out: the mean delay is well above average on holidays, above it at weekends, and below it on ordinary working days.

Use these panels to *check the spec before you model with it*. If the `"1 after"` box sits on the dashed line at 1, that lag is not doing anything and you can drop it; if `"2 before"` is the box that moves, your reporting slows down **ahead** of the break and you want a negative `holiday_lags` (see `?temporal_effects`).

### Do delay distributions drift over time?

Reporting delays are not always stable: a surveillance system may speed up or slow down over a season or across years. [plot_delay_drift()](https://rodrigozepeda.github.io/tbl.now/reference/plot_delay_drift.html) draws a rolling *fan chart* of the delay distribution indexed by event date — a solid rolling median, a dashed rolling mean, and 25–75% / 10–90% quantile bands. A rising or falling centre line signals *location* drift; widening or narrowing bands signal *spread* drift.

```{r delay-drift, eval=plotting_available, fig.width=9, fig.height=4.5, message=FALSE, warning=FALSE}
plot_delay_drift(dengue_now)
```

Because the most recent event dates have not had time to be fully reported, their delays look artificially short; that immature region (after the completeness cutoff) is **shaded grey** and should not be read as drift.

For a formal answer, [test_delay_drift()](https://rodrigozepeda.github.io/tbl.now/reference/test_delay_drift.html) runs an autocorrelation-robust monotonic-trend test (a modified Mann–Kendall test, via the `modifiedmk` package) on the per-period delay summaries — a delay series is correlated with itself, so an ordinary trend test would over-state significance. It tests both a location statistic (the median) and a dispersion statistic (the 10–90 spread), on mature data only:

```{r delay-drift-test, eval=requireNamespace("modifiedmk", quietly = TRUE)}
test_delay_drift(dengue_now, stat = c("median", "spread"))
```

A significant `drift` for `spread` with a non-significant one for `median`, say, would mean the *typical* delay held steady while its *variability* changed.

The trend test looks for *gradual* change. If instead you suspect an **abrupt** shift — a reporting-system change on some date — use [test_delay_changepoint()](https://rodrigozepeda.github.io/tbl.now/reference/test_delay_changepoint.html), which applies Pettitt's nonparametric change-point test and returns the estimated change date together with the before/after delay level:

```{r delay-changepoint}
test_delay_changepoint(dengue_now, stat = c("median", "spread"))
```

You can mark the detected change point on the fan chart with `plot_delay_drift(dengue_now, changepoint = TRUE)`.

### Detecting batch reporting

Laboratories sometimes withhold results and then release a whole backlog at once — a **batch**. The key idea is that a batch *moves* reports along the report axis without *creating* them, so a window of report dates spanning both the lull and the spike has an unchanged total — whereas a genuine epidemic surge adds cases and inflates it. [batch_test()](https://rodrigozepeda.github.io/tbl.now/reference/batch_test.html) turns this into a per-report-date diagnostic that separates the two:

```{r batches, warning=FALSE}
batches <- batch_test(dengue_now, lookback = 2)

batches |>
  filter(batch) |>
  select(report_date, reported, baseline, everything())
```

Additional information on dealing with batches and other reporting delay artifacts can be found in the [corresponding vignette](https://rodrigozepeda.github.io/tbl.now/reference/week_2_date.html). 

## Other functions (utilities)

### Convert epidemiological weeks to dates

The function [week_2_date()](https://rodrigozepeda.github.io/tbl.now/reference/week_2_date.html) converts epidemiological week/year combinations into an aligned calendar date.

```{r}
df <- data.frame(
  epidemiological_week = 1:5,
  epidemiological_year = rep(2024, 5)
)

df |>
  week_2_date(
    week_col = epidemiological_week,
    year_col = epidemiological_year
  )
```
 
### Reports

The functions [get_initial_reported_cases() and get_latest_reported_cases()](https://rodrigozepeda.github.io/tbl.now/reference/get_latest_first.html) extract the number of cases first reported for each event date and the most recently reported totals, respectively. These utilities allow users to quantify revisions between initial and final reports.
 
 
```{r}
df_reports <- data.frame(
  n = c(10, 1, 1, 0, 0, 3),
  event_date = rep(ymd("2020/09/12"), 6),
  report_date = c(
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17")
  )
)

tbl_reports <- df_reports |>
  tbl_now(
    event_date = event_date, report_date = report_date,
    verbose = FALSE, case_count = n, report_units = "days",
    event_units = "days"
  )

tbl_reports
```


The initial reported cases:

```{r}
get_initial_reported_cases(tbl_reports)
```

and the latest totals:

```{r}
get_latest_reported_cases(tbl_reports)
```

### Week alignment

The [align_weeks()](https://rodrigozepeda.github.io/tbl.now/reference/align_weeks.html) 
function standardizes dates within the same epidemiological week to a single reference day
(for example, the start of the week). This is helpful when computing differences across weekly reporting periods, avoiding fractional time intervals.

```{r}
df <- data.frame(
  date = c(ymd("2022-10-31"), ymd("2022-11-07"), ymd("2022-11-13")),
  epiweek = c(44, 45, 46)
)

# Align to Sundays
df_aligned <- align_weeks(df, date_col = date)
df_aligned
```

You can verify the resulting weekday using the [wday() function from the lubridate package](https://lubridate.tidyverse.org/reference/day.html):

```{r}
df_aligned |>
  mutate(day_label = wday(date_aligned, label = TRUE, abbr = FALSE))
```

### Complete zeroes

The [complete_zeroes()](https://rodrigozepeda.github.io/tbl.now/reference/complete_zeroes.html) function fills with zeroes those cases where the `event` or `report` weeks have not been observed. 

Consider for example the following data with just two observations per date:

```{r}
ndata <- tibble(
  event_date = c(as.Date("2021/01/12"), as.Date("2021/01/14"), as.Date("2021/01/14")),
  report_date = c(as.Date("2021/01/13"), as.Date("2021/01/15"), as.Date("2021/01/18")),
  case_count = c(10, 5, 1)
)

ndata <- tbl_now(ndata, event_date, report_date,
  verbose = FALSE, case_count = case_count, data_type = "count-incidence"
)

ndata
```

Notice that there are no observations for `2021/01/13`. Furthermore, if we assume that the maximum possible observed delay is of `4`,  we can fill the unobserved cases with:

```{r}
complete_zeroes(ndata, max_delay = 4)
```

Which looks at all the possible report dates and event dates and sets the counts to zero if they have not been observed.

### Censoring extreme delays

The function `censor_delays_above()` marks all delays above a threshold value (`max_delay`) as censored. This is useful to indicate extreme delays in some nowcast models:

```{r}
df <- data.frame(
  onset = as.Date("2020-01-01") + c(0, 0, 1, 2),
  reported = as.Date("2020-01-01") + c(1, 5, 2, 300)
)
tn <- tbl_now(df,
  event_date = onset, report_date = reported,
  data_type = "linelist", verbose = FALSE
)

# the 300-day report becomes censored (an upper bound on its delay)
censor_delays_above(tn, max_delay = 60)
```

### Converting to data formats from other packages

> **NOTE** This is still work in progress `r lifecycle::badge("experimental")`

`tbl.now` ships converters that move data between a `tbl_now` and the data structures used by other nowcasting and delay-estimation packages. They all follow the same naming convention:

+ `tbl_now_from_*()` builds a `tbl_now` (it wraps [as_tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/as_tbl_now.html), so `...` is forwarded to [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)).

+ `tbl_now_to_*()` converts a `tbl_now` into that package's native object.

Each function accepts a `verbose` argument that reports the choices it made (the inferred `now`, the data type, the units, the column mapping, and so on). The supported packages — [epinowcast](https://package.epinowcast.org/), [baselinenowcast](https://baselinenowcast.epinowcast.org/), [EpiNow2](https://epiforecasts.io/EpiNow2/), [epidist](https://epidist.epinowcast.org/), [data.table](https://rdatatable.gitlab.io/data.table/) and [tsibble](https://tsibble.tidyverts.org/) — are all optional (`Suggests`).

| Package | `from` | `to` | Maps to |
|---|---|---|---|
| `epinowcast` | yes | yes | `count-cumulative` (`reference_date`, `report_date`, `confirm`); `from` also accepts a preprocessed `enw_preprocess_data` or fitted object |
| `baselinenowcast` | yes | yes | `count-incidence` (long data frame or reporting-triangle matrix) |
| `EpiNow2` | — | yes | a single `date`/`confirm` series for `estimate_infections()`, or (with `model = "estimate_truncation"`) a list of report-date snapshots for `estimate_truncation()` |
| `data.table` | yes | yes | any data type |
| `epidist` | yes | yes | `linelist` (primary/secondary event dates) |
| `tsibble` | yes | yes | a `tbl_ts` (one date as index, the other date + strata as key) |

The object produced by any `tbl_now_to_*()` can be converted straight back with [as_tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/as_tbl_now.html), which has methods for each of these classes:

```{r converters-roundtrip, eval=requireNamespace("tsibble", quietly = TRUE), message=FALSE}
dengue_now <- tbl_now(denguedat,
  event_date = "onset_week",
  report_date = "report_week", strata = "gender",
  verbose = FALSE
)

# tbl_now -> tsibble -> tbl_now
dengue_ts <- tbl_now_to_tsibble(dengue_now, verbose = FALSE)
as_tbl_now(dengue_ts, report_date = "report_week", verbose = FALSE)
```

For example, the [epinowcast](https://package.epinowcast.org/) hospitalization data ships in long cumulative format and converts straight into a `count-cumulative` `tbl_now`:

```{r converters-available, include=FALSE}
epinowcast_available <- requireNamespace("epinowcast", quietly = TRUE)
```

```{r converters-from, eval=epinowcast_available, message=FALSE}
library(epinowcast)
data(germany_covid19_hosp)


#Read data from epinowcast
obs  <- germany_covid19_hosp[location == "DE"]

#Remove unused column
obs  <- obs[, location := NULL]

#Pre-process data
pobs <- epinowcast::enw_preprocess_data(obs, max_delay = 40, by = "age_group")

hospitalizations_now <- tbl_now_from_epinowcast(pobs)
  
hospitalizations_now
```

Going the other way, `tbl_now_to_epinowcast()` returns an `enw_preprocess_data` object ready for [epinowcast::epinowcast()](https://package.epinowcast.org/reference/epinowcast.html):

```{r converters-to, eval=epinowcast_available, message=FALSE}
epinowcast_object <- tbl_now_to_epinowcast(hospitalizations_now, verbose = FALSE)
epinowcast_object
```

The same pattern works for the other packages. <!---For example `tbl_now_to_EpiNow2()` collapses the object into the single `date`/`confirm` time series that [EpiNow2](https://epiforecasts.io/EpiNow2/)'s `estimate_infections()`/`epinow()` expect; passing `model = "estimate_truncation"` instead returns the list of report-date snapshots that `estimate_truncation()` consumes (the one EpiNow2 model that uses the report dimension).-->

## References

