---
title: "Diagnosing reporting batches and other reporting-delay artifacts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnosing reporting batches and other reporting-delay artifacts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 4.5,
  dpi = 120,
  cache = TRUE
)
library(tbl.now)
```


## Batch reporting

Surveillance data does not always arrive smoothly. Sometimes the reporting system
halts or reduces its output (e.g. a data-system outage, an overwhelmed jurisdiction) and the
backlog is released later all at once. That release is called a **batch**: a collection of 
reports from previous periods that were held and reported all at once during a different
reporting period. Intuivitively one can think of a batch as a collection of reports that 
--in an ideal scenario-- _should have been reported_ on a previous date but 
were actually released later. 

```{r, echo = FALSE}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

# --- 1. GENERATE SYNTHETIC SURVEILLANCE DATA ---
set.seed(42)
dates  <- seq(as.Date("2026-06-01"), as.Date("2026-06-25"), by = "day")
n_days <- length(dates)

# Ideal Scenario: A smooth, wave-like curve (e.g., normal distribution)
ideal_cases <- round(100 * dnorm(seq(-2, 2.5, length.out = n_days)) + 5)

# Create the data frame
data <- data.frame(
  date  = dates,
  ideal = ideal_cases
)

# Introduce a data-system freeze/backlog:
# - Days 12, 13, and 14: System breaks down. Reports drop to almost 0.
# - Day 15: The bottleneck is cleared, releasing all accumulated backlog as a "Batch".
backlog_days <- c(12, 13, 14)
batch_day <- 15

# Calculate how many cases were missed during the outage
backlog_volume <- sum(data$ideal[backlog_days])

# Build the "real" curve with the batch effect
data <- data %>%
  mutate(
    real = ideal,
    # Simulate the reporting drop
    real = ifelse(row_number() %in% backlog_days, 1, real), 
    # Add the accumulated backlog to the batch day
    real = ifelse(row_number() == batch_day, ideal + backlog_volume, real)
  )

# --- 2. DEFINE CUSTOM PALETTE (From your provided list) ---
# Main bars: #334335 (Dark green)
# Ideal curve: #5F7E62 (Muted green)
# Annotation accent: #B85348 (Coral red for warnings/anomalies)
# Texts & grid elements: #262626 (Off-black)
# Grid lines: #A8BFA9 (Soft pale green-gray)

# --- 3. CREATE THE PLOT ---
ggplot(data, aes(x = date)) +
  # Real reporting (bars)
  geom_col(aes(y = real, fill = "Real reporting with batch"), alpha = 0.9, width = 0.8) +
  # Ideal reporting (smooth line)
  geom_line(aes(y = ideal, color = "Ideal reporting scenario"), linewidth = 1, linetype = "solid") +
  
  # Set Colors
  scale_fill_manual(values = c("Real reporting with batch" = "#5F7E62")) +
  scale_color_manual(values = c("Ideal reporting scenario" = "#B85348")) +
  
  # --- ANNOTATION 1: The Valley (Where cases should have been reported) ---
  annotate(
    "text", 
    x = dates[mean(backlog_days)], 
    y = max(data$ideal) * 1.5, 
    label = "System Outage\n(Ideally, cases should\nhave been reported here)", 
    color = "black", fontface = "bold", size = 3.5, vjust = 0.5, hjust = 1
  ) +
  annotate(
    "segment", 
    x = dates[mean(backlog_days)], xend = dates[mean(backlog_days)], 
    y = max(data$ideal) * 1.35, yend = 5, 
    arrow = arrow(length = unit(0.2, "cm")), color = "black", linewidth = 0.8
  ) +
  
  # --- ANNOTATION 2: The Batch (The release day) ---
  annotate(
    "text", 
    x = dates[batch_day] + 3.5, 
    y = data$real[batch_day] - 10, 
    label = "The Batch!\nBacklog released\nall at once", 
    color = "black", fontface = "bold", size = 3.8, hjust = 0
  ) +
  annotate(
    "segment", 
    x = dates[batch_day] + 3.2, xend = dates[batch_day] + 0.3, 
    y = data$real[batch_day] - 10, yend = data$real[batch_day] - 5, 
    arrow = arrow(length = unit(0.2, "cm")), color = "black", linewidth = 0.8
  ) +
  
  # Theme and Formatting
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#262626"),
    plot.subtitle = element_text(size = 11, color = "#607060", margin = margin(b = 15)),
    axis.title = element_text(face = "bold", color = "#262626"),
    axis.text = element_text(color = "#262626"),
    panel.grid.major = element_line(color = "#A8BFA9", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 10, color = "#262626")
  ) +
  labs(
    title = "Understanding 'Batches' in Public Health Surveillance",
    x = "Reporting Date",
    y = "Number of Cases Reported"
  )
```


A batch is easy to confuse with an epidemic **surge**. The important part is that the batch happens
on the **reporting date** axis while a surge happens on the **event date** axis. A
batch just _moves_ reports to a later date not adding new cases while a real
epidemic surge _adds_ new cases. The tools here are designed to help you visualize that difference. 

> This article shows each plot **twice**: first on a clean, made-up
outbreak with one obvious batch (so you learn the signature), then on real,
COVID-19 data from the CDC (see [`covid_us`](https://rodrigozepeda.github.io/tbl.now/reference/covid_us.html)). 

## Two datasets to compare

### The made-up outbreak. 

This simulation consists of a bell-shaped curve over a hundred days, each
case reported within a few days. For one week near the peak, the reporting
system slows down with **half** of each day's reports being held back and released days after. 
That release is the **batch**. 

```{r ideal-data, warning = FALSE}
set.seed(82495)
#Simulate a curve
onset_days <- as.Date("2024-01-01") + 0:99
bell       <- dnorm(seq(-2.5, 2.5, length.out = 100))
per_day    <- round(400 * bell / max(bell)) + 8         
onset      <- rep(onset_days, per_day)
reported   <- onset + rpois(length(onset), 1.5)         

clean_tn <- tbl_now(tibble(onset = onset, reported = reported),
                    event_date = onset, report_date = reported,
                    data_type = "linelist", verbose = FALSE)

# Simulate a batch
ideal <- simulate_batch(clean_tn,
  closed_dates  = seq(as.Date("2024-02-19"), by = "day", length.out = 7),
  held_fraction = 0.5)
```

We can see the simulated data both from the event-date and the report-date perspectives:

```{r, eval = FALSE}
plot_epidemic_process(ideal)
plot_reporting_process(ideal)
```

```{r, echo = FALSE}
plot_epidemic_process(ideal) + plot_reporting_process(ideal)
```

### The real data

`covid_us` comes from the CDC's individual-level
[COVID-19 Case Surveillance Public Use Data](https://data.cdc.gov/Case-Surveillance/COVID-19-Case-Surveillance-Public-Use-Data/vbim-akqf/about_data).
These represent cases whose event date (`cdc_case_earliest_dt`)
*and* report date (`cdc_report_dt`) both fall before September 2020. The picture shows
what it would have looked back then.

```{r data}
data(covid_us)

covid_early <- covid_us %>% 
  filter(cdc_case_earliest_dt < as.Date("2020-09-01") &
           cdc_report_dt < as.Date("2020-09-01"))

tn <- tbl_now(covid_early, event_date = cdc_case_earliest_dt,
              report_date = cdc_report_dt, case_count = n,
              data_type = "count-incidence", verbose = FALSE)
```

Half of all cases were reported within a few days, but the tail is long: some cases take weeks or months to surface.

```{r delays}
stats::quantile(rep(tn$.delay, tn$n), c(0.5, 0.75, 0.9, 0.99))
```

We can see this dataset again from both the event-date and the report-date perspectives:

```{r, eval = FALSE}
plot_epidemic_process(tn)
plot_reporting_process(tn)
```

```{r, echo = FALSE}
plot_epidemic_process(tn) + plot_reporting_process(tn)
```


## The reporting process

This plot, which we have previously shown, shows how many reports arrived by date. Barches or surges might correspond
to spikes towering over their neighbours. 

```{r, fig.cap="Reporting process of the simulated data"}
plot_reporting_process(ideal)
```

On the real data the reporting is spikier; however, a couple of peaks in June 2020 stick up
where smooth epidemic reporting should be. Those are reporting artefacts either pure 
backlog releases, or a mix of backlog + a genuine surge (we'll come back to this characterization later). The tallest is a single 
day of about 160K reports, well above the 20-40K that arrive on a typical day that summer.

```{r proc-covid, fig.cap="Reporting process of the COVID-19 data"}
plot_reporting_process(tn)
```

## The reporting triangle

We provide two different visualizations of the reporting triangle. In both
we plot the three temporal dimensions involved in the process: the event date,
the reporting date and the delay. We cover the plots in tiles coloured by how many cases
were registered then. 


### The classical reporting triangle

The classical view is given by `plot_reporting_triangle()` where each
tile is described by _when it happened_ (across) and _how long they took to be 
reported_ (vertical). The diagonal shows reports that arrive on the same day.  
An indicator of a batch is **a bright diagonal reaching high up** which
reporesents many  delayed cases being all reported on the same day.

```{r proc-sim, eval = FALSE}
plot_reporting_triangle(ideal)
```

```{r, fig.cap="Classical reporting triangle of the simulated data", echo = FALSE}
plot_reporting_triangle(ideal) + (plot_reporting_process(ideal))
```

On COVID-19, the triangle is a broad blue-grey haze (most cases reported over many
months) crossed by two bright diagonals. They correspond to the same
spikes seen for June 2020:

```{r tri-covid, eval = FALSE}
plot_reporting_triangle(tn)
```

```{r, fig.cap="Classical reporting triangle of the COVID-19 data", echo = FALSE}
plot_reporting_triangle(tn) + (plot_reporting_process(tn))
```

### The reporting V

`plot_reporting_v()` shows *exactly the same data* but rotated 45 degrees 
so that the report date is the main vertical axis.  Here batches
that were diagonals in the previous interpretation are horizontal slices 
where the reporting dates can be seen more easily.

```{r v-sim, eval = FALSE}
plot_reporting_v(ideal)
```
```{r, echo = FALSE}
plot_reporting_v(ideal) + plot_reporting_process(ideal)
```

On covid the V opens into a full wedge; the potential batches are the faint horizontal
streaks in June 2020.

```{r v-covid, eval = FALSE}
plot_reporting_v(tn)
```

```{r v-covid2, echo = FALSE}
plot_reporting_v(tn) + plot_reporting_process(tn)
```

## Scalograms

::: {.alert .alert-warning}
The scalogram functions are **very experimental**. We have yet to confirm they work for all batch cases. Feel free to skip
to the @sec-transport
:::

Scalograms show reductions of cases. For this example, consider the following simulated reporting  process:

```{r, echo = FALSE}
data(denguedat)

ndata <- tibble(onset_week   = seq(as.Date("2000/01/01"), by = 7, length.out = 200),
                report_week  = seq(as.Date("2000/01/08"), by = 7, length.out = 200), 
                n = ceiling(sapply(1:200, \(x) 5*x + 1000*abs(sin(pi*x/52))))) |>
    tbl_now(
        event_date = onset_week, report_date = report_week, case_count = n, verbose = FALSE) 

#Smooth for the example
plot_reporting_process(ndata)

```


Its scalogram shows the decreases in the reporting cases as vertical streaks aligned with the minimal date of this decrease:

```{r, echo = FALSE}
#Smooth for the example
plot_scalogram(ndata)
```

One can see the same vertical streaks in the previous reporting process we had been working on corresponding to the dip before the batch:

```{r, eval = FALSE}
plot_scalogram(ideal)
```

```{r, echo = FALSE}
plot_scalogram(ideal, format = "%b") + plot_reporting_process(ideal)
```

and again the same June dates being the most identified with additional dates having less of a clear pattern:

```{r, eval = FALSE}
plot_scalogram(tn)
```


```{r, echo = FALSE}
plot_scalogram(tn, format = "%b") + plot_reporting_process(tn)
```

## Transport vs creation {#sec-transport} 

This is the main tool for detecting batches and surges. Before the plot, we
will explain the whole idea. Consider a daily outbreak with  reports incoming 
each day. Three things can change the number of reports:

- a **hold**  -- a reporting office falls behind and some days' reports are withheld;
- a **batch** -- the day the backlog (hold) is finally released, all at once;
- a **surge** -- an increase in the epidemic process: more people falling ill and being reported.

We simulate one of each in a clean epidemic and colour every day by which it is (grey
= an ordinary day):

```{r tut-data, echo = FALSE}
tut_cols <- c(Normal = "grey72", Hold = "#4E79A7",
              Batch = "#B85348", Surge = "#5F7E62")

set.seed(2024)
# a clean bell-shaped epidemic, reported within a day or two
tut_ev <- rep(as.Date("2024-01-01") + 0:99,
              times = round(300 * dnorm(seq(-2.5, 2.5, length.out = 100)) /
                              max(dnorm(seq(-2.5, 2.5, length.out = 100)))) + 6)
tut_rp <- tut_ev + rpois(length(tut_ev), 1.5)
# a SURGE: extra *real* cases in mid-March, reported promptly
surge_ev <- rep(as.Date("2024-03-11") + 0:2, each = 500)
tut <- rbind(
  data.frame(event_date = tut_ev,   report_date = tut_rp),
  data.frame(event_date = surge_ev, report_date = surge_ev + rpois(length(surge_ev), 1))
)
# a HOLD then BATCH: close reporting for a week near the peak; the backlog is
# released the day the desk catches up
closed <- seq(as.Date("2024-02-19"), by = "day", length.out = 3)
tut_tn <- simulate_batch(
  tbl_now(tut, event_date, report_date, data_type = "linelist", verbose = FALSE),
  closed_dates = closed, held_fraction = 0.7, verbose = FALSE)

# the two scores per report date, tagged by what we planted
scores <- transport_discriminant(tut_tn, lookback = 7)
scores$kind <- factor(dplyr::case_when(
  scores$report_date %in% closed                          ~ "Hold",
  scores$report_date == max(closed) + 1                   ~ "Batch",
  scores$report_date >= as.Date("2024-03-11") &
    scores$report_date <= as.Date("2024-03-13")           ~ "Surge",
  TRUE                                                     ~ "Normal"),
  levels = names(tut_cols))
ar <- arrow(length = unit(0.15, "cm"))

ggplot(scores, aes(report_date, reported, fill = kind)) +
  geom_col(width = 1) +
  scale_fill_manual(values = tut_cols, name = NULL) +
  annotate("segment", x = as.Date("2024-02-01"), xend = as.Date("2024-02-21"),
           y = 700, yend = 750, arrow = ar, colour = tut_cols[["Batch"]]) +
  annotate("text", x = as.Date("2024-01-31"), y = 700, hjust = 1, size = 3,
           colour = tut_cols[["Batch"]], label = "Batch: the backlog\nreleased in one lump") +
  annotate("segment", x = as.Date("2024-02-09"), xend = as.Date("2024-02-20"),
           y = 400, yend = 175, arrow = ar, colour = tut_cols[["Hold"]]) +
  annotate("text", x = as.Date("2024-02-08"), y = 400, hjust = 1, size = 3,
           colour = tut_cols[["Hold"]], label = "Hold: reports\nwithheld (a dip)") +
  annotate("segment", x = as.Date("2024-03-27"), xend = as.Date("2024-03-14"),
           y = 560, yend = 500, arrow = ar, colour = tut_cols[["Surge"]]) +
  annotate("text", x = as.Date("2024-03-28"), y = 590, hjust = 0, size = 3,
           colour = tut_cols[["Surge"]], label = "Surge: genuinely\nmore cases") +
  coord_cartesian(clip = "off") +
  labs(x = "Report date", y = "Reports",
       title = "1. The reporting timeline, coloured by what happened") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", plot.margin = margin(6, 80, 6, 6))
```

In the previous plot, every bar is one report date. The **batch** towers 
where the held reports land together; the **hold** is the small blue dip 
just before it; the **surge** is a genuine bump of new cases.

The transport discriminant turns each day into **two numbers**:

- A **creation score** -- did this stretch of days genuinely *gain* cases? A larger surge
implies a larger creation score.
- A **transport score** -- were the days just before *missing* reports? 
A backlog release after a hold pushes it up.

We plot every day by those two numbers and observe the directions of the three disturbances:

```{r tut-plane, fig.width = 8, fig.height = 6, warning = FALSE, echo = FALSE}
zc <- qnorm(0.975)
ty <- max(scores$transport_z, na.rm = TRUE)
rx <- max(scores$creation_z,  na.rm = TRUE)
lx <- min(scores$creation_z,  na.rm = TRUE)
ggplot(scores, aes(creation_z, transport_z, colour = kind)) +
  geom_hline(yintercept = zc, linetype = "dashed", colour = "grey85") +
  geom_vline(xintercept = zc, linetype = "dashed", colour = "grey85") +
  geom_point(data = ~subset(., kind == "Normal"), size = 1.7, alpha = 0.45, na.rm = TRUE) +
  geom_point(data = ~subset(., kind != "Normal"), size = 2.8, na.rm = TRUE) +
  scale_colour_manual(values = tut_cols, name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.14))) +
  scale_x_continuous(expand = expansion(mult = 0.09)) +
  annotate("text", x = 8, y = ty + 3, size = 3, fontface = "bold",
           colour = tut_cols[["Batch"]], label = "Backlog release: shoots UP and RIGHT") +
  annotate("text", x = rx * 0.72, y = 4, size = 3, fontface = "bold",
           colour = tut_cols[["Surge"]], label = "Real surge:\nshoots RIGHT") +
  annotate("text", x = lx * 0.55, y = 21, size = 3, fontface = "bold",
           colour = tut_cols[["Hold"]], label = "Hold:\ngoes UP and to the LEFT") +
  labs(x = "Creation score",
       y = "Transport score",
       title = "2. Where each day lands in the transport-vs-creation plane") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

Identifying any bar with its dot we can conclude:

- A **batch (backlog release, red)** shoots **up** -- the days before were
  depleted (high transport) but not as many new cases were created (right);
- A **surge (green)** shoots **right** -- cases genuinely appeared (high
  creation), with apparent preceding hole;
- A **hold (blue)** drifts **up and to the left** -- reports have apparently gone
missing (transport rising) while the window has *lost* cases (creation negative).

Ordinary days (grey) sit in the cloud through the middle. That is the whole idea
behind `transport_discriminant()` and `batch_test()`: **a batch is high transport
with little creation**. 

### The transport discriminant

The previous plot can be done with the `plot_transport_discriminant()` function: 

```{r disc-sim, eval = FALSE}
plot_transport_discriminant(ideal)
```

```{r disc-sim2, echo = FALSE}
plot_transport_discriminant(ideal) + plot_reporting_process(ideal)
```

Which also works to identify the COVID-19 cases:

```{r disc-covid, eval = FALSE}
plot_transport_discriminant(tn, period = 7)
```

```{r disc-covid2, echo = FALSE}
plot_transport_discriminant(tn, period = 7) +
  plot_reporting_process(tn)
```

### Recovering the data

The `batch_test()` function runs the transport test for the batch signature and
returns, for every report date, the `batch` flag -- a Benjamini-Hochberg-corrected
verdict that controls the false-discovery rate across all dates (see `?batch_test`
for the full column reference). Keeping the `batch` rows gives the confirmed
releases together with their `deficit` (how depleted the days just before were) and
`delta` (how little the window total actually changed).

```{r screen-sim, eval = FALSE}
batch_test(ideal) %>%
  filter(batch)
```

```{r screen-sim2, echo = FALSE}
batch_test(ideal) %>%
  filter(batch) %>%
  select(report_date, reported, baseline, deficit, delta, p_transport_bh, batch)
```

On covid we pass `period = 7` to divide out the weekly reporting cadence. The
confirmed batches (the Benjamini-Hochberg-corrected `batch` flag) are the
spring/summer-2020 releases -- each reported far more than its neighbours *and* was
preceded by a matching deficit. The clearest is **10 June 2020**, nearly ten times
its baseline:

```{r screen-covid, eval = FALSE}
batch_test(tn, period = 7) %>% 
  filter(batch)
```

```{r screen-covid2, echo = FALSE}
batch_test(tn, period = 7) %>%
  filter(batch) %>%
  select(report_date, reported, baseline, deficit, delta, p_transport_bh, batch) %>%
  arrange(desc(reported))
```


## Delay changes 

The reporting delay might change through time. Here we show two different plots 
for identifying delay problems. 

### Reporting-delay drift

The typical time from case to report, tracked over the outbreak. Shows the
overall trend of the delay. Normally it will be steady; a 
batch can be seen as a **sudden bump upward** on the release day.

```{r drift-sim}
plot_delay_drift(ideal)
```

On covid the delays show extreme variability in the beginning and a trend
that decreases the delay in time: 

```{r drift-covid}
plot_delay_drift(tn)
```

The functions `test_delay_drift()` and `test_delay_changepoint()` test 
for a gradual or abrupt changes in the delay. We can see, for example, that it correctly identifies the drift in the COVID-19 dataset that is not present in the ideal example:

```{r, warning=FALSE}
test_delay_drift(tn)
test_delay_drift(ideal)
```

while it also finds the June change-point in the COVID-19 delay distribution which fails to be sustained for long enought to be detected in the ideal example: 

```{r, warning=FALSE}
test_delay_changepoint(tn)
test_delay_changepoint(ideal)
```

### Delay profiles

Each faint line is one day's distirbution of reporting delays. Most days report quickly, so
their lines hug the left and concentrate around the same distribution:

```{r prof-sim}
plot_delay_profiles(ideal)
```

On COVID-19 a whole spray of lines reaches far to the right days that reported
cases months later.

```{r prof-covid}
plot_delay_profiles(tn)
```


## In summary

Here is the whole toolkit on one page. Every row is a different way of seeing the
same thing -- reports that were held back and then released together.

| Plot | What to look for |
|---|---|
| **Reporting process** | Shows how reports were registered. |
| **Reporting triangle** | Diagonals show cases with the same report date. |
| **The reporting V** | Horizontal slices show cases with the same report date. |
| **Wavelet scalogram** | Bright short-period ridges in the reporting series show _holds_ on the reporting. |
| **Transport discriminant** | A red dot up-and-left, in the "potential batch region" might indicate a batch. |
| **`batch_test()`** | Shows the flagged reports in the discriminant as a `data.frame` |
| **Delay profiles** | Show the delay distribution for each event date. |
| **Reporting-delay drift** | Shows how the delay and its variance change through time. |

Another practical notes from other data we have analyzed:

+ In general, **surges** (real new cases) seem harder to identify of that **batches** (moved reports).
+ Batches are very difficult to identify in low-incidence scenarios.  
+ Batches very close to the now are difficult to identify without additional modeling hypotheses.  
