---
title: "Adding your own nowcasting model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding your own nowcasting model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.width = 9,
  fig.height = 4.5,
  dpi = 120
)

options(tibble.width = Inf, width = 160, pillar.width = 160)
options(pillar.print_max = 5, pillar.print_min = 5,
        tibble.print_min = 5, dplyr.print_min = 5,
        tibble.print_max = 5, dplyr.print_max = 5)

```

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tbl.now)
```

`tbl.now` ships with back-ends for six nowcasting packages. The article [on ensemble nowcasting](https://rodrigozepeda.github.io/tbl.now/articles/ensemble-nowcasting.html), shows how to fit, score and combine them through one call. This article is about the model not included in the package: **yours**.

# 1. How does it work

The main function for running a nowcast is  `run_nowcast(x, engine("mymodel", ...))` which does exactly three things:

1. takes the **engine**: an object of class `c("mymodel", "nowcast_engine")`
   carrying your arguments;
2. calls **`nowcast_fit(method, x, ...)`** to run the model and return whatever it
   returns. The engine's arguments are passed in the dots (`...`);
3. utilizes **`nowcast_tidy(method, fit, x, ..., quantile_levels)`** to describe that
   result in the one shape the rest of the package understands.

and wraps the answer in a `tbl_nowcast`. In summary:

| you write | it receives | it must return |
|---|---|---|
| `nowcast_fit.mymodel()` | the `tbl_now`, your `...`, `quantile_levels`, `verbose` | anything at all — it is stored verbatim in `nowcast@fit` |
| `nowcast_tidy.mymodel()` | your fit, the `tbl_now`, `quantile_levels` | `list(predictions =, draws =)` |

`nowcast_tidy()` returns a list with two slots, and **only one of them may be
`NULL`**:

| slot | one row per | columns |
|---|---|---|
| `predictions` | (event date, stratum, quantile level) | `<event_date>`, the strata, `.quantile_level`, `.value` |
| `draws` | (event date, stratum, draw) | `<event_date>`, the strata, `.draw`, `.value` |

Return `draws` when your model has them. `tbl.now` derives the quantiles and can also
use them for the `linear_pool` ensemble. Return `predictions` when your model produces quantiles directly, which is
the case for the one we are about to write. You may (and preferably should) return both.


::: {.alert .alert-info}
If you are shipping the backend in a
package, also write an `engine_mymodel()` alongside it that names your arguments, the
way `engine_nobbs()` names `max_D` and `moving_window`: 

```r
engine_mymodel <- function(..., window = 12,
                           min_date = NULL,
                           quantile_levels = nowcast_quantile_levels(),
                           label = NULL) {
  engine("mymodel", window = window, ...,
         min_date = min_date, quantile_levels = quantile_levels, label = label)
}
```
:::


# 2. A worked example: the delay-ratio nowcast

## The model

> The main idea of our model is that we can multiply what has been observed until now by a constant factor *for each delay*. 

### Formalization (skip if you want)

Write $C(t, d)$ for the number of cases with event date $t$ that had been
reported within a delay of $d$ periods, and let $D$ be the delay past which we
are willing to call reporting "finished". Call an event date old enough that all $D$
periods have elapsed as **mature**. The proportion 
of the eventual total  visible at each delay is given by the ratio

$$r_t(d) \;=\; \frac{C(t, D)}{C(t, d)}$$

The nowcast for a young event date $t$, observed so
far at delay $d(t) = \mathrm{now} - t$, is then its current count scaled by that
distribution:

$$\hat q_\alpha(t) \;=\; C\big(t, d(t)\big) \cdot \mathrm{Quantile}_\alpha\Big(\big\{\, r_s(d(t)) : s \text{ mature} \,\big\}\Big).$$

The **median** multiplier gives the point estimate. The rest of the empirical
quantiles give the uncertainty.


## The data

For this example we'll use the `denguedat` dataset.  We nowcast as of
a date in the past so that there are later reports to score against.

```{r data}
data(denguedat)

dengue <- denguedat |>
  filter(onset_week >= as.Date("2008-01-01")) |>
  tbl_now(
    event_date  = onset_week,
    report_date = report_week,
    data_type   = "linelist",
    verbose     = FALSE
  ) 

now <- as.Date("2010-10-04")

snapshot <- dengue |>
  filter(report_week <= now) |>
  change_now(now = now)

get_now(snapshot)
```

`backtest_snapshot` is what the model may see; `dengue` still holds the reports that
arrived afterwards, and is the truth we will score against.

## `nowcast_fit.delay_ratio()`

The fitting step estimates the multiplier. Notice how much of it is spent
asking the object about itself rather than doing arithmetic.

```{r fit-method}
nowcast_fit.delay_ratio <- function(engine, x, ..., max_delay = NULL,
                                    quantile_levels, verbose = TRUE) {
  
  #We verify that x is count-incidence (in case people pass a linelist
  if (get_data_type(x) == "count-cumulative"){
    stop("Invalid data type this model doesn't work with cumulative data")
  } else {
    x <- x |> to_count(to = "count-incidence")
  }

  #It is important to use the getters and don't assume column names
  event_col <- get_event_date(x)
  strata    <- get_strata(x) %||% character(0)
  key       <- c(event_col, strata)

  # How far out we model. Past `max_delay` the model calls reporting finished.
  max_delay  <- if (is.null(max_delay)) max(x$.delay) else max_delay
  delay_grid <- seq(min(x$.delay), max_delay)

  # C(t, d) for every event date and every delay on the grid. This getter does
  # the cumulating, the strata and the three data types for us.
  snapshot_at <- function(d) {
    
    #We check the reported cases by each delay
    snapshot  <- get_nth_reported_cases(x, delay = d)
    
    #We get the case count column
    count_col <- get_case_count(snapshot)

    as_tibble(snapshot) |>
      summarise(.reported = sum(.data[[count_col]]), .by = all_of(key)) |>
      mutate(.delay_grid = d)
  }
  snapshots <- bind_rows(lapply(delay_grid, snapshot_at))

  # Work on the integer grid, not the calendar: `.event_num` and `.report_num`
  # are already in report units, so this is unit-agnostic.
  index     <- distinct(as_tibble(x), across(all_of(event_col)), .event_num)
  now_index <- max(x$.report_num)

  eventual <- snapshots |>
    filter(.delay_grid == max_delay) |>
    select(all_of(key), .eventual = ".reported")

  ratios <- snapshots |>
    filter(.delay_grid < max_delay) |>
    inner_join(eventual, by = key) |>
    inner_join(index, by = event_col) |>
    # Mature dates only: a date still filling in would teach the model that
    # reporting stops early. And a ratio needs a non-zero denominator.
    filter(.event_num + max_delay <= now_index, .reported > 0) |>
    mutate(.ratio = .eventual / .reported) |>
    select(all_of(strata), ".delay_grid", ".ratio")

  if (isTRUE(verbose)) {
    n_delays <- n_distinct(ratios$.delay_grid)
    cli::cli_alert_info(
      "Estimated {nrow(ratios)} multiplier{?s} over {n_delays} delay{?s}."
    )
  }

  list(
    ratios    = ratios,
    max_delay = max_delay,
    index     = index,
    now_index = now_index
  )
}
```

Two things of note:

1. The returned list is **the fit**. It is kept verbatim in
`nowcast@fit`, and it carries `index` and `now_index` because the tidying step
will need them. And `verbose` is used because a back-end that talks during `nowcast_backtest()` usually floods the console.

## `nowcast_tidy.delay_ratio()`

The tidying step applies the pool. This model produces quantiles directly, so it
fills `predictions` and leaves `draws` as `NULL`.

```{r tidy-method}
nowcast_tidy.delay_ratio <- function(engine, fit, x, ..., quantile_levels) {

  #It is important to use the getters and don't assume column names
  event_col <- get_event_date(x)
  strata    <- get_strata(x) %||% character(0)
  key       <- c(event_col, strata)

  # What each event date has reported as of `now`, and how old it is.
  latest    <- get_latest_reported_cases(x)
  count_col <- get_case_count(latest) %||% "n"

  current <- as_tibble(latest) |>
    summarise(.reported = sum(.data[[count_col]]), .by = all_of(key)) |>
    inner_join(fit$index, by = event_col) |>
    mutate(.delay_grid = pmin(fit$now_index - .event_num, fit$max_delay))

  # The empirical quantiles of the multiplier, per delay (and per stratum).
  # The 0.5 row is the median multiplier: the point estimate.
  multipliers <- fit$ratios |>
    reframe(
      .quantile_level = quantile_levels,
      .multiplier = quantile(.ratio, quantile_levels, names = FALSE),
      .by = all_of(c(strata, ".delay_grid"))
    )

  predictions <- current |>
    cross_join(tibble(.quantile_level = quantile_levels)) |>
    left_join(multipliers, by = c(strata, ".delay_grid", ".quantile_level")) |>
    # No multiplier means nothing left to correct: a mature date, or a delay the
    # history never showed us. Either way the honest factor is 1.
    mutate(
      .multiplier = coalesce(.multiplier, 1),
      .value      = .reported * .multiplier
    ) |>
    select(all_of(key), ".quantile_level", ".value")

  list(predictions = predictions, draws = NULL)
}
```

That is the entire back-end: two functions, no registration, no `tbl.now` change.

## Running it

```{r run}
nowcast <- run_nowcast(
  snapshot,
  engine("delay_ratio", max_delay = 8),
  verbose = FALSE
)

nowcast
```

The fit is kept in `nowcast@fit` if required. We can access the nowcasts via `tidy()`:

```{r multipliers, eval = FALSE}
tidy(nowcast)
```


```{r tidy, echo = FALSE}
tidy(nowcast) |> tail(5)
```

**`autoplot()`**  is also present:

```{r plot}
library(ggplot2)

autoplot(nowcast, date_lim = c(as.Date("2010/09/01"), as.Date("2010-10-04"))) 
```


**`nowcast_backtest()`** refits it at a series of past dates and scores each one:

```{r backtest}
backtest <- nowcast_backtest(
  dengue,
  engine("delay_ratio", max_delay = 8),
  now_dates = as.Date(c("2010-07-05", "2010-08-02", "2010-09-06")),
  seed      = 20260827,
  verbose   = FALSE
)

tidy(backtest) 
```

It can also use `scoringutils::as_forecast_quantile()` directly, `score_nowcast` and be used within a `nowcast_ensemble()`. All of the nowcast post-processing functions become available. 


# 3.  Do's and don'ts

Your `nowcast_fit()` method is handed the whole `tbl_now`. Almost every bug in a
back-end comes from assuming something about it that is not guaranteed.

**Never hard-code a column name.** The event-date column is called `onset_week`
in `denguedat`, `dx_date` in `mpoxdat` and `reference_date` in an `epinowcast`
import. Use the [getters](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html) instead:

```r
get_event_date(x)      # name of the event-date column
get_report_date(x)     # name of the report-date column
get_case_count(x)      # name of the counts column, or NULL for a line list
get_strata(x)          # character vector, or NULL
get_now(x)             # the as-of Date

              ...etcetera...
```

**Three protected numeric columns are always present**, and they are usually
what you actually want to compute on:

| column | meaning |
|---|---|
| `.event_num` | the event date as an integer number of report units from `min(event_date)` |
| `.report_num` | the report date, on the same scale |
| `.delay` | `.report_num - .event_num` — the reporting delay, in report units |

Working in `.event_num` / `.delay` means your model does not care whether the
data is daily, weekly or monthly, and does not have to do calendar arithmetic.

**Handle the three data types, or refuse one explicitly.** A line list has no
counts column and one row per case; `"count-incidence"` counts what was *newly*
reported in a cell; `"count-cumulative"` counts what had been reported *so far*.
Do not write three branches — call `to_count()` and be handed the one you want:

```r
x <- to_count(x, to = "count-incidence")
```

Remember that `cumulative → incidence` de-accumulates, so a downward revision
becomes a **negative** increment. If your model cannot represent that, say so
rather than silently taking a maximum. (`tbl_now_to_baselinenowcast()` refuses
cumulative input for exactly this reason.)

**A zero is not the same as a missing row.** A `tbl_now` only carries cells that
were reported, so an event date with no reports has no rows at all, and a model
that builds its time grid from the rows it was handed will quietly stop short of
`now` — which is precisely where a nowcast matters. `complete_zeroes()` fills the
grid out to `now` for count data:

```r
x <- complete_zeroes(x)
```

For a line list this cannot work — a zero-count row expands to zero rows — so the
grid has to come from somewhere else. The built-in `"surveillance"` back-end does
this by passing an explicit `control$dRange`.

**`now` is a declaration, not `max(report_date)`.** Use `get_now(x)`. An object
whose reporting has stalled has a `now` later than its last report, and that gap
is real information about the nowcast, not an error to round away.

## For developing in `tbl.now`:

If your model is a wrapper around another package, [create a converter](https://rodrigozepeda.github.io/tbl.now/reference/index.html#converters) that takes a `tbl.now` and returns the object required by the other package:

```r
tbl_now_to_baselinenowcast(x) 
tbl_now_to_epinowcast(x)      
tbl_now_to_EpiNow2(x, target = "estimate_infections")
tbl_now_to_epidist(x, format = "interval")        

              ...etcetera...
```

The converter should, when possible, also come back as a `tbl_now` with the  `as_tbl_now()` method:

```r
triangles <- tbl_now_to_baselinenowcast(x, format = "triangle_list")
as_tbl_now(triangles)      # back to a tbl_now, strata recoded
```

## Additional packaging requirements

Everything above worked from the global environment. To put it in a package:

* **Register both methods.** With roxygen, `@export` on `nowcast_fit.mymodel()`
  and `nowcast_tidy.mymodel()`.
* If developing your own package, **put `tbl.now` in `Imports`** and import the generics you extend
  (`@importFrom tbl.now nowcast_fit nowcast_tidy`), along with the getters you
  call.
* If developing for `tbl.now`, **put the modelling package in `Suggests`**, and check for it at fit time with `requireNamespace(..., quietly = TRUE)`
* **Give it an `engine_mymodel()`.** `engine("mymodel", ...)` already works; a
  constructor that names your arguments makes them discoverable.
* **Test the fit and the tidy separately.** `engine("mymodel")` builds the
  dispatch object, so `nowcast_tidy(engine("mymodel"), fit, x,
  quantile_levels = c(0.1, 0.5, 0.9))` can be tested against a stored fit without
  refitting anything.
* **Test against more than one shape.** A line list and a `count-cumulative`
  object with weekly dates and two strata will find the assumptions a daily
  incidence tibble never does.

A checklist for the method itself:

- [ ] No hard-coded column names — every one comes from a getter.
- [ ] All three `data_type`s handled, or one refused with a clear error.
- [ ] The time grid runs to `get_now(x)`, not to the last observed row.
- [ ] Strata honoured, or pooled with a warning.
- [ ] `quantile_levels` respected — or, if the model reports a fixed set, a
      warning saying so.
- [ ] `verbose = FALSE` really is silent.
- [ ] Everything the tidy step needs is inside the fit object.

# 4. Getting it into `tbl.now` itself

If you think a model belongs in the package, or you have written one and would like it maintained here:

> **Open an issue at
> [github.com/rodrigozepeda/tbl.now/issues](https://github.com/rodrigozepeda/tbl.now/issues)
> before writing a pull request.**

Please say in the issue:

* which modelling package or model the back-end wraps,
* what extras does it need at install time: a Stan toolchain, JAGS, nothing;
* how it handles **strata** — jointly, one fit per stratum, or not at all;
* whether it produces **draws** or only quantiles and a single interval;
* a minimal example that fits on one of the packaged datasets
  (`denguedat`, `mpoxdat`, `covid_colombia`, `covid_us`, `flusight`,
  `hai_bucaramanga`).


```{r learning-more, echo=FALSE, results="asis"}
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```
