---
title: "Adding your own nowcasting model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding your own nowcasting model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.width = 9,
  fig.height = 4.5,
  dpi = 120
)
options(tibble.print_min = 6, tibble.print_max = 6)
```

> This was written automatically by an AI model. A human has yet to review.

::: {.alert .alert-warning}
`r lifecycle::badge("experimental")`
`run_nowcast()`, `nowcast_fit()`, `nowcast_tidy()` and the `tbl_nowcast` class
are [experimental](https://lifecycle.r-lib.org/articles/stages.html#experimental).
They work and they are tested, but the extension contract is the youngest part
of the package and is the part most likely to change. Pin a version if you are
building on it.
:::

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tbl.now)
```

`tbl.now` ships back-ends for six nowcasting packages, and
`vignette("ensemble-nowcasting")` shows how to fit, score and combine them
through one call. This article is about the seventh model: **yours**.

Everything below really runs. The model we build needs no Stan, no JAGS and no
modelling package at all — it is a hundred lines of `dplyr` — which is the point:
the contract is small enough that a model you sketched on a napkin can be a
first-class citizen of the package by lunchtime.

# 1. The contract

`run_nowcast(x, engine("mymodel", ...))` does exactly three things:

1. takes the **engine** — an object of class `c("mymodel", "nowcast_engine")`
   carrying your arguments, built by `engine()`;
2. calls **`nowcast_fit(method, x, ...)`** — run the model, return whatever it
   returns. The engine's arguments arrive in that `...`;
3. calls **`nowcast_tidy(method, fit, x, ..., quantile_levels)`** — describe that
   result in the one shape the rest of the package understands.

and wraps the answer in a `tbl_nowcast`. That is the whole framework. Dispatch is
ordinary S3, so **the two methods can live anywhere** — your package, your
analysis script, or a chunk in a vignette like this one. Nothing inside `tbl.now`
changes, and nothing needs to be registered with it.

| you write | it receives | it must return |
|---|---|---|
| `nowcast_fit.mymodel()` | the `tbl_now`, your `...`, `quantile_levels`, `verbose` | anything at all — it is stored verbatim in `nowcast@fit` |
| `nowcast_tidy.mymodel()` | your fit, the `tbl_now`, `quantile_levels` | `list(predictions =, draws =)` |

`nowcast_tidy()` returns a list with two slots, and **exactly one of them may be
`NULL`**:

| slot | one row per | columns |
|---|---|---|
| `predictions` | (event date, stratum, quantile level) | `<event_date>`, the strata, `.quantile_level`, `.value` |
| `draws` | (event date, stratum, draw) | `<event_date>`, the strata, `.draw`, `.value` |

Return `draws` when your model has them: the quantiles are derived for you, and
draws are what `nowcast_ensemble(type = "linear_pool")` and `tidy(probs = )`
need. Return `predictions` when your model produces quantiles directly, which is
the case for the one we are about to write. You may return both.

::: {.alert .alert-info}
**The engine's arguments reach `nowcast_fit()`, not `nowcast_tidy()`.**
`run_nowcast()` splices them into the fitting step only; the tidying step is
called with the fit, the object and `quantile_levels`. Anything the tidy step
needs — a number of draws, a tuning parameter, a lookup table — must be **stored
in the object `nowcast_fit()` returns**. That is not a limitation so much as the
reason the fit is kept verbatim.
:::

::: {.alert .alert-info}
**Your backend gets `engine()` for free, and can have its own constructor.**
`engine("mymodel", window = 12)` works the moment `nowcast_fit.mymodel()` is
registered — there is nothing to declare. If you are shipping the backend in a
package, write an `engine_mymodel()` alongside it that names your arguments, the
way `engine_nobbs()` names `max_D` and `moving_window`: a named formal is
discoverable from the signature and a typo in one is an error, where a typo in
`...` is a default nobody notices.

```r
engine_mymodel <- function(..., window = 12,
                           min_date = NULL,
                           quantile_levels = nowcast_quantile_levels(),
                           label = NULL) {
  engine("mymodel", window = window, ...,
         min_date = min_date, quantile_levels = quantile_levels, label = label)
}
```
:::

# 2. Reading the `tbl_now` correctly

Your `nowcast_fit()` method is handed the whole `tbl_now`. Almost every bug in a
back-end comes from assuming something about it that is not guaranteed.

**Never hard-code a column name.** The event-date column is called `onset_week`
in `denguedat`, `dx_date` in `mpoxdat` and `reference_date` in an `epinowcast`
import. Ask:

```r
get_event_date(x)      # name of the event-date column
get_report_date(x)     # name of the report-date column
get_case_count(x)      # name of the counts column, or NULL for a line list
get_strata(x)          # character vector, or NULL
get_covariates(x)      # character vector, or NULL
get_now(x)             # the as-of Date
get_event_units(x)     # "days" | "weeks" | "months" | "years" | "numeric"
get_data_type(x)       # "linelist" | "count-incidence" | "count-cumulative"
```

**Three protected numeric columns are always present**, and they are usually
what you actually want to compute on:

| column | meaning |
|---|---|
| `.event_num` | the event date as an integer number of report units from `min(event_date)` |
| `.report_num` | the report date, on the same scale |
| `.delay` | `.report_num - .event_num` — the reporting delay, in report units |

Working in `.event_num` / `.delay` means your model does not care whether the
data is daily, weekly or monthly, and does not have to do calendar arithmetic.

**Handle the three data types, or refuse one explicitly.** A line list has no
counts column and one row per case; `"count-incidence"` counts what was *newly*
reported in a cell; `"count-cumulative"` counts what had been reported *so far*.
Do not write three branches — call `to_count()` and be handed the one you want:

```r
x <- to_count(x, to = "count-incidence")
```

Remember that `cumulative → incidence` de-accumulates, so a downward revision
becomes a **negative** increment. If your model cannot represent that, say so
rather than silently taking a maximum. (`tbl_now_to_baselinenowcast()` refuses
cumulative input for exactly this reason.)

**A zero is not the same as a missing row.** A `tbl_now` only carries cells that
were reported, so an event date with no reports has no rows at all, and a model
that builds its time grid from the rows it was handed will quietly stop short of
`now` — which is precisely where a nowcast matters. `complete_zeroes()` fills the
grid out to `now` for count data:

```r
x <- complete_zeroes(x)
```

For a line list this cannot work — a zero-count row expands to zero rows — so the
grid has to come from somewhere else. The built-in `"surveillance"` back-end does
this by passing an explicit `control$dRange`.

**`now` is a declaration, not `max(report_date)`.** Use `get_now(x)`. An object
whose reporting has stalled has a `now` later than its last report, and that gap
is real information about the nowcast, not an error to round away.

# 3. Don't reshape by hand: the converters

If your model is a wrapper around another package, or wants a reporting triangle,
a delay-interval frame or a `tsibble`, that work is already done and already
tested against all three data types, all four time units and the `now` edge:

```r
tbl_now_to_baselinenowcast(x, format = "matrix")   # reporting triangle
tbl_now_to_epinowcast(x)                           # long, cumulative, preprocessed
tbl_now_to_EpiNow2(x, target = "estimate_infections")
tbl_now_to_nobbs(x)                                # line list, one row per case
tbl_now_to_surveillance(x)
tbl_now_to_epidist(x, format = "interval")         # censored delay intervals
tbl_now_to_tsibble(x)
tbl_now_to_data_table(x)
```

They also pool the columns you never declared, so case totals survive the trip,
and they collapse a per-case censoring flag with a warning rather than putting
two rows in a triangle cell that has one slot.

The arrow points both ways. `as_tbl_now()` has a method for every one of those
outputs, plus the raw inputs of the packages themselves, so a foreign object
comes back into `tbl.now` without you writing the mapping:

```r
triangles <- tbl_now_to_baselinenowcast(x, format = "triangle_list")
as_tbl_now(triangles)      # back to a tbl_now, strata recoded
```

If you find yourself writing a `pivot_wider()` to build a triangle, stop: one of
the converters already did it, and it handled the case you were about to get
wrong.

# 4. A worked example: the delay-ratio nowcast

## The idea

Write $C(t, d)$ for the number of cases with event date $t$ that had been
reported within a delay of $d$ periods, and let $D$ be the delay past which we
are willing to call reporting finished. For an event date old enough that all $D$
periods have elapsed — call it **mature** — we can see the whole story: how much
of the eventual total was visible at each delay.

The ratio

$$r_t(d) \;=\; \frac{C(t, D)}{C(t, d)}$$

is the factor by which the count observed at delay $d$ eventually grew. Collect
it over every mature event date and you have an empirical distribution of
multipliers *for each delay*. The nowcast for a young event date $t$, observed so
far at delay $d(t) = \mathrm{now} - t$, is then its current count scaled by that
distribution:

$$\hat q_\alpha(t) \;=\; C\big(t, d(t)\big) \cdot \mathrm{Quantile}_\alpha\Big(\big\{\, r_s(d(t)) : s \text{ mature} \,\big\}\Big).$$

The **median** multiplier gives the point estimate. The rest of the empirical
quantiles give the uncertainty — no distribution is assumed anywhere, which is
the appeal: the spread of the intervals is the spread of what reporting has
actually done in the past.

This is the oldest idea in nowcasting (it is the chain-ladder of actuarial
reserving, wearing an epidemiological hat), and it is a genuinely useful
baseline. It also has two honest weaknesses that will show up in the output, and
we will come back to them.

## The data

Dengue in Puerto Rico: a weekly line list, aggregated to counts. We nowcast as of
a date in the past so that there are later reports to score against.

```{r data}
data(denguedat)

dengue <- denguedat |>
  filter(onset_week >= as.Date("2008-01-01")) |>
  count(onset_week, report_week, name = "n") |>
  tbl_now(
    event_date  = onset_week,
    report_date = report_week,
    case_count  = n,
    data_type   = "count-incidence",
    verbose     = FALSE
  )

now <- as.Date("2010-10-04")

snapshot <- dengue |>
  filter(report_week <= now) |>
  change_now(now = now)

get_now(snapshot)
```

`snapshot` is what the model may see; `dengue` still holds the reports that
arrived afterwards, and is the truth we will score against.

## `nowcast_fit.delay_ratio()`

The fitting step estimates the multiplier pool. Notice how much of it is spent
asking the object about itself rather than doing arithmetic.

```{r fit-method}
nowcast_fit.delay_ratio <- function(method, x, ..., max_delay = NULL,
                                    quantile_levels, verbose = TRUE) {

  event_col <- get_event_date(x)
  strata    <- get_strata(x)
  if (is.null(strata)) strata <- character(0)
  key       <- c(event_col, strata)

  # How far out we model. Past `max_delay` the model calls reporting finished.
  max_delay <- if (is.null(max_delay)) max(x$.delay) else max_delay
  delay_grid <- seq(min(x$.delay), max_delay)

  # C(t, d) for every event date and every delay on the grid. This getter does
  # the cumulating, the strata and the three data types for us.
  snapshot_at <- function(d) {
    snapshot <- get_nth_reported_cases(x, delay = d)
    count_col <- get_case_count(snapshot)
    if (is.null(count_col)) count_col <- "n"

    as_tibble(snapshot) |>
      summarise(.reported = sum(.data[[count_col]]), .by = all_of(key)) |>
      mutate(.delay_grid = d)
  }
  snapshots <- bind_rows(lapply(delay_grid, snapshot_at))

  # Work on the integer grid, not the calendar: `.event_num` and `.report_num`
  # are already in report units, so this is unit-agnostic.
  index     <- distinct(as_tibble(x), across(all_of(event_col)), .event_num)
  now_index <- max(x$.report_num)

  eventual <- snapshots |>
    filter(.delay_grid == max_delay) |>
    select(all_of(key), .eventual = ".reported")

  ratios <- snapshots |>
    filter(.delay_grid < max_delay) |>
    inner_join(eventual, by = key) |>
    inner_join(index, by = event_col) |>
    # Mature dates only: a date still filling in would teach the model that
    # reporting stops early. And a ratio needs a non-zero denominator.
    filter(.event_num + max_delay <= now_index, .reported > 0) |>
    mutate(.ratio = .eventual / .reported) |>
    select(all_of(strata), ".delay_grid", ".ratio")

  if (isTRUE(verbose)) {
    n_delays <- n_distinct(ratios$.delay_grid)
    cli::cli_alert_info(
      "Estimated {nrow(ratios)} multiplier{?s} over {n_delays} delay{?s}."
    )
  }

  list(
    ratios    = ratios,
    max_delay = max_delay,
    index     = index,
    now_index = now_index
  )
}
```

Two things to note. The returned list is **the fit** — it is kept verbatim in
`nowcast@fit`, and it carries `index` and `now_index` because the tidying step
will need them and `...` does not reach that far. And `verbose` is honoured,
because a back-end that talks when it is told to and is silent when it is not is
one that can be used inside `nowcast_backtest()` without flooding the console.

## `nowcast_tidy.delay_ratio()`

The tidying step applies the pool. This model produces quantiles directly, so it
fills `predictions` and leaves `draws` as `NULL`.

```{r tidy-method}
nowcast_tidy.delay_ratio <- function(method, fit, x, ..., quantile_levels) {

  event_col <- get_event_date(x)
  strata    <- get_strata(x)
  if (is.null(strata)) strata <- character(0)
  key       <- c(event_col, strata)

  # What each event date has reported as of `now`, and how old it is.
  latest    <- get_latest_reported_cases(x)
  count_col <- get_case_count(latest)
  if (is.null(count_col)) count_col <- "n"

  current <- as_tibble(latest) |>
    summarise(.reported = sum(.data[[count_col]]), .by = all_of(key)) |>
    inner_join(fit$index, by = event_col) |>
    mutate(.delay_grid = pmin(fit$now_index - .event_num, fit$max_delay))

  # The empirical quantiles of the multiplier, per delay (and per stratum).
  # The 0.5 row is the median multiplier: the point estimate.
  multipliers <- fit$ratios |>
    reframe(
      .quantile_level = quantile_levels,
      .multiplier = quantile(.ratio, quantile_levels, names = FALSE),
      .by = all_of(c(strata, ".delay_grid"))
    )

  predictions <- current |>
    cross_join(tibble(.quantile_level = quantile_levels)) |>
    left_join(multipliers, by = c(strata, ".delay_grid", ".quantile_level")) |>
    # No multiplier means nothing left to correct: a mature date, or a delay the
    # history never showed us. Either way the honest factor is 1.
    mutate(
      .multiplier = coalesce(.multiplier, 1),
      .value      = .reported * .multiplier
    ) |>
    select(all_of(key), ".quantile_level", ".value")

  list(predictions = predictions, draws = NULL)
}
```

That is the entire back-end: two functions, no registration, no `tbl.now` change.

## Running it

```{r run}
nowcast <- run_nowcast(
  snapshot,
  engine("delay_ratio", max_delay = 8),
  verbose = FALSE
)

nowcast
```

The fit is kept, so the multiplier pool is there to look at — and the median
multiplier by delay is the model, in one table:

```{r multipliers}
nowcast@fit$ratios |>
  summarise(
    n      = n(),
    median = round(median(.ratio), 2),
    q10    = round(quantile(.ratio, 0.1), 2),
    q90    = round(quantile(.ratio, 0.9), 2),
    .by    = .delay_grid
  ) |>
  arrange(.delay_grid)
```

Read it from the bottom up. By delay three a week's count is within a couple of
percent of its eventual total, and by delay four it is done. At delay one it
typically **doubles**. At delay zero it multiplies by thirty-five — and only 49
of the 136 mature weeks ever had a delay-zero report at all, so that row rests on
a third of the data and its 10–90% range runs from 8.7 to 106.

Those `q10`/`q90` columns are the entire uncertainty model: no distribution is
assumed anywhere, the intervals are just what reporting has actually done before.
And that top row is why the fan below opens so violently at the last week, which
is the model telling the truth rather than misbehaving.

# 5. What you get for free

The point of implementing the contract rather than writing a bespoke function is
everything in this section, none of which needed a line of extra code.

**`tidy()`** — the same table every other engine tidies into:

```{r tidy}
tidy(nowcast) |> tail(5)
```

**`autoplot()`** — the fan chart, with the counts reported by `now` drawn
underneath, so the gap is the correction the model is making:

```{r plot}
ggplot2::autoplot(nowcast) +
  ggplot2::coord_cartesian(
    xlim = as.Date(c("2010-04-01", "2010-10-11")),
    ylim = c(0, 800)
  )
```

**`score_nowcast()`** — the weighted interval score, the absolute error of the
median, and interval coverage, against the data the model never saw:

```{r score}
score_nowcast(nowcast, truth = dengue) |> tail(5)
```

This is where the model's two honest weaknesses appear. The fan widens sharply at
the `now` edge, because a multiplier estimated at delay 0 is estimated from very
few reports and is enormously variable — and the point estimate drifts high,
because a multiplier pooled over two and a half years remembers an epidemic that
was growing while these last weeks were shrinking. The intervals cover; the
median does not lead. That is a real property of this estimator, not a bug in the
example, and it is exactly what you would want a backtest to tell you.

**`list_nowcast_methods()`** finds it, because it scans the S3 methods registered
for `nowcast_fit()` in every loaded namespace:

```{r list}
list_nowcast_methods(installed_only = FALSE)
```

**`nowcast_backtest()`** refits it at a series of past dates and scores each one.
It takes engines, so `max_delay` travels with the model that needs it rather than
in an outer `...` that has to be routed to the right backend by name:

```{r backtest}
backtest <- nowcast_backtest(
  dengue,
  engine("delay_ratio", max_delay = 8),
  now_dates = as.Date(c("2010-07-05", "2010-08-02", "2010-09-06")),
  seed      = 20260827,
  verbose   = FALSE
)

tidy(backtest) |>
  summarise(mean_wis = round(mean(wis, na.rm = TRUE), 2), .by = now)
```

**`nowcast_ensemble()`** treats it exactly like a packaged method — which we can
show once there is a second member.

# 6. The other branch: returning draws

Our model has an empirical pool of multipliers, so it can just as well
*resample* it. That gives draws instead of quantiles, and draws unlock the two
things quantiles cannot do: `type = "linear_pool"` ensembles, and `tidy(probs =)`
at arbitrary levels.

The fitting step is unchanged, so reuse it — this is ordinary S3, and a method is
just a function:

```{r boot-fit}
nowcast_fit.delay_ratio_boot <- function(method, x, ..., n_draws = 1000,
                                         quantile_levels, verbose = TRUE) {
  fit <- nowcast_fit.delay_ratio(
    method, x, ..., quantile_levels = quantile_levels, verbose = verbose
  )
  # `...` never reaches `nowcast_tidy()`, so `n_draws` travels in the fit.
  fit$n_draws <- n_draws
  fit
}
```

```{r boot-tidy}
nowcast_tidy.delay_ratio_boot <- function(method, fit, x, ..., quantile_levels) {

  event_col <- get_event_date(x)
  strata    <- get_strata(x)
  if (is.null(strata)) strata <- character(0)
  key       <- c(event_col, strata)

  latest    <- get_latest_reported_cases(x)
  count_col <- get_case_count(latest)
  if (is.null(count_col)) count_col <- "n"

  current <- as_tibble(latest) |>
    summarise(.reported = sum(.data[[count_col]]), .by = all_of(key)) |>
    inner_join(fit$index, by = event_col) |>
    mutate(.delay_grid = pmin(fit$now_index - .event_num, fit$max_delay))

  pool <- fit$ratios

  draws <- current |>
    cross_join(tibble(.draw = seq_len(fit$n_draws))) |>
    mutate(
      .multiplier = vapply(.delay_grid, function(d) {
        available <- pool$.ratio[pool$.delay_grid == d]
        if (length(available) == 0) 1 else sample(available, 1)
      }, numeric(1)),
      .value = .reported * .multiplier
    ) |>
    select(all_of(key), ".draw", ".value")

  list(predictions = NULL, draws = draws)
}
```

`predictions` is `NULL`, and `run_nowcast()` derives the quantiles from the draws:

```{r boot-run}
set.seed(20260827)

bootstrapped <- run_nowcast(
  snapshot,
  engine("delay_ratio_boot", max_delay = 8, n_draws = 2000),
  verbose = FALSE
)

bootstrapped
```

Because there are draws, `tidy()` will compute any quantile you ask for:

```{r boot-probs}
tidy(bootstrapped, probs = c(0.1, 0.9)) |> tail(3)
```

On the quantile-only version the same call is an **error**, not an
approximation — which is the behaviour you want from a package that is about to
tell you how uncertain something is:

```{r boot-probs-fail, error=TRUE}
tidy(nowcast, probs = c(0.1, 0.9))
```

And the two versions ensemble with each other, or with any packaged method:

```{r ensemble}
ensemble <- nowcast_ensemble(
  median    = nowcast,
  bootstrap = bootstrapped,
  name      = "ratio-ensemble"
)

tidy(ensemble) |> tail(3)
```

They agree closely, which is the correct answer and a useful check: resampling
an empirical distribution and taking its quantiles are two routes to the same
estimator, and the two branches of the contract should not disagree about it.

# 7. Strata

`get_strata(x)` was threaded through both methods, so a stratified object needs
nothing new — one multiplier pool is estimated per stratum:

```{r strata}
dengue_by_sex <- denguedat |>
  filter(onset_week >= as.Date("2008-01-01")) |>
  count(onset_week, report_week, gender, name = "n") |>
  tbl_now(
    event_date  = onset_week,
    report_date = report_week,
    case_count  = n,
    strata      = gender,
    data_type   = "count-incidence",
    verbose     = FALSE
  ) |>
  filter(report_week <= now) |>
  change_now(now = now)

stratified <- run_nowcast(
  dengue_by_sex,
  engine("delay_ratio", max_delay = 8),
  verbose = FALSE
)

tidy(stratified) |> tail(4)
```

Whatever strata columns you put on the output are what the result reports as its
strata, and `score_nowcast()` and `nowcast_ensemble()` key on them.

**If your model cannot honour strata, warn and pool** — do not pretend. The
packaged back-ends show both ways out: `"baselinenowcast"` and `"surveillance"`
fit one model per stratum, while `"NobBS"` and `"EpiNow2"` are given the
*interaction* of the declared columns as a single pasted label, which is split
back into columns on the way out. If you paste labels yourself, refuse the paste
when a stratum value already contains your separator: splitting it back
ambiguously attaches a nowcast to the wrong stratum with no error at all.

# 8. Shipping it as a package

Everything above worked from the global environment. To put it in a package:

* **Register both methods.** With roxygen, `@export` on `nowcast_fit.mymodel()`
  and `nowcast_tidy.mymodel()`; by hand,
  `S3method(nowcast_fit, mymodel)` and `S3method(nowcast_tidy, mymodel)` in
  `NAMESPACE`. Because dispatch is ordinary S3, they are visible the moment your
  package is loaded, and `list_nowcast_methods()` picks them up.
* **Put `tbl.now` in `Imports`** and import the generics you extend
  (`@importFrom tbl.now nowcast_fit nowcast_tidy`), along with the getters you
  call.
* **Put the modelling package in `Suggests`**, and check for it at fit time with
  `requireNamespace(..., quietly = TRUE)`, so installing your package does not
  drag in a Stan or JAGS toolchain.
* **Reuse `tidy()` if that package already has a method for its own fits.** The
  built-in `"surveillance"` and `"EpiNow2"` back-ends do exactly this: `tidy()`
  already knows how to find their interval *and how wide it is*, and re-deriving
  that in a second place is how two real bugs got in.
* **Give it an `engine_mymodel()`.** `engine("mymodel", ...)` already works; a
  constructor that names your arguments makes them discoverable and turns a typo
  into an error. See section 1.
* **Test the fit and the tidy separately.** `engine("mymodel")` builds the
  dispatch object, so `nowcast_tidy(engine("mymodel"), fit, x,
  quantile_levels = c(0.1, 0.5, 0.9))` can be tested against a stored fit without
  refitting anything.
* **Test against more than one shape.** A line list and a `count-cumulative`
  object with weekly dates and two strata will find the assumptions a daily
  incidence tibble never does.

A checklist for the method itself:

- [ ] No hard-coded column names — every one comes from a getter.
- [ ] All three `data_type`s handled, or one refused with a clear error.
- [ ] The time grid runs to `get_now(x)`, not to the last observed row.
- [ ] Strata honoured, or pooled with a warning.
- [ ] `quantile_levels` respected — or, if the model reports a fixed set, a
      warning saying so rather than interpolation.
- [ ] `verbose = FALSE` really is silent.
- [ ] Everything the tidy step needs is inside the fit object.

# 9. Getting it into `tbl.now` itself

A back-end does **not** have to live in `tbl.now` to be a first-class citizen —
that is the whole design. But if you think a model belongs in the package, or you
have written one and would like it maintained here:

> **Open an issue at
> [github.com/rodrigozepeda/tbl.now/issues](https://github.com/rodrigozepeda/tbl.now/issues)
> before writing a pull request.**

Please say in the issue:

* which modelling package the back-end wraps, and where it lives (CRAN, GitHub,
  r-universe) — this decides whether it can be a `Suggests` at all;
* what it needs at install time: a Stan toolchain, JAGS, or nothing;
* how it handles **strata** — jointly, one fit per stratum, or not at all;
* whether it produces **draws** or only quantiles and a single interval;
* a minimal example that fits on one of the packaged datasets
  (`denguedat`, `mpoxdat`, `covid_colombia`, `covid_us`, `flusight`,
  `hai_bucaramanga`).

An issue first is not bureaucracy: a new back-end also needs a converter, a
section in `vignette("nowcasting-models")` showing both the plain and the
stratified fit, a row in the converter matrix, and an entry in the model
comparison. It is much easier to agree on the shape of all that before the code
is written than after.

Bug reports and back-ends that should stay in your own package are equally
welcome as issues — if a getter you needed does not exist, or a converter got
your data type wrong, that is the same address.

# See also

* `vignette("ensemble-nowcasting")` — `run_nowcast()`, scoring, backtesting and
  ensembles across the packaged methods.
* `vignette("nowcasting-models")` — the same six packages driven by hand through
  the converters.
* `?nowcast_fit` and `?nowcast_tidy` — the contract in reference form.
* `?tbl_nowcast` — the object your two methods are turned into.
* `?tidy.tbl_nowcast`, `?score_nowcast`, `?nowcast_ensemble`.
