---
title: "Describing and diagnosing a tbl_now"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Describing and diagnosing a tbl_now}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
options(tibble.width = Inf, width = 160, pillar.width = 160)
```

```{r setup, message=FALSE, warning=FALSE}
library(dplyr, quietly = TRUE)
library(tbl.now)
```

Two questions come up with every new dataset, and they are different questions:

* **What is in it?** — how many cases, over what period, arriving how late, how
  sparse, how concentrated in one stratum. `summary()` answers this.
* **What is wrong with it?** — missing dates, impossible orderings, repeated
  rows, units that do not line up, data that stops before its own `now`.
  `diagnose()` answers this.

Both return a **tibble**, not printed text. That is the design decision
everything else follows from: an answer you can `filter()`, `join()`, plot, or
assert on in a test is worth far more than one you can only read.

```{r data}
data(denguedat)

dengue_now <- tbl_now(denguedat,
  event_date  = "onset_week",
  report_date = "report_week",
  strata      = "gender",
  verbose     = FALSE
)
```

# Part 1 — `summary()`

## One table, several blocks

`summary()` stacks blocks of related statistics into a single table. The
`component` column says which block a row belongs to:

```{r summary-blocks}
summary(dengue_now) |>
  count(component)
```

Every row is one quantity, described by up to eighteen columns. Not every column
applies to every row — a delay distribution has a `mean` but no `prop`, a
proportion has a `prop` but no `q90` — so the table is deliberately sparse:

```{r summary-shape}
summary(dengue_now) |>
  filter(component == "cases") |>
  select(component, quantity, stratum, n, total, mean, sd, min, q50, q90, max, prop_zero)
```

`stratum` is always "which subset of the data does this row describe": `"all"`
for the pooled rows, or a stratum label. The *category* of a compositional row
goes in `quantity` instead, which keeps the two ideas from colliding:

```{r summary-composition}
summary(dengue_now) |>
  filter(component == "composition") |>
  select(quantity, stratum, n, total, prop)
```

## The blocks worth knowing

**`delay`** — the case-weighted reporting delay. Weighted means "equal to what
you would get by expanding the counts to one row per case", so it needs no
separate explanation:

```{r summary-delay}
delay_summary(dengue_now) |>
  select(quantity, stratum, n, total, mean, sd, q50, q90, max)
```

**`completeness`** — the share of each event date's eventual total that had
arrived by delay `d`. This is usually the most decision-relevant block in the
table, because it says how far back a nowcast has anything left to estimate:

```{r summary-completeness}
reporting_completeness(dengue_now, delays = 0:4) |>
  filter(stratum == "all") |>
  select(quantity, n, mean, q50, prop)
```

Four percent of a week's cases are in by the end of that week, half by the end
of the next, and 94% by three weeks. Beyond about three weeks there is very
little missing to reconstruct.

**`zero_run`** — how sparse the series is, measured as runs of consecutive
zero dates rather than as a simple proportion. A series that is 30% zeros
scattered at random is a very different modelling problem from one that is 30%
zeros in three long gaps:

```{r summary-zeroruns}
zero_run_summary(dengue_now, axis = "event") |>
  select(quantity, stratum, n, total, mean, q50, max)
```

**`coverage`** — the reach of the object, including how stale it is:

```{r summary-coverage}
triangle_occupancy(dengue_now) |>
  filter(stratum == "all") |>
  select(quantity, n, value)
```

## Every block is also a function

`summary()` is exactly the `bind_rows()` of its components, and each one is
exported, so you can ask a single question without computing the rest:

```{r summary-components}
prop_strata(dengue_now) |>
  select(quantity, total, prop)
```

The full set is `cases_per_date()`, `delay_summary()`, `zero_run_summary()`,
`prop_censored()`, `prop_confirmation_type()`, `prop_strata()`,
`prop_covariate_levels()`, `case_autocorrelation()`, `date_ranges()`,
`triangle_occupancy()`, `reporting_completeness()` and `cumulative_growth()`.

<div class="alert alert-info">
**Quantiles are inverse-ECDF (type 1).** `q50` is the smallest value whose
cumulative weight reaches 0.5 — for an even number of observations, the upper of
the two middle values rather than their average. This is the same estimator
`autoplot()` and `plot_delay_drift()` use, so the table and the figures always
agree, and for count data it always returns a value that was actually observed.
A half-case delay is not.
</div>

# Part 2 — `diagnose()`

## Findings, worst first

```{r diagnose-all}
diagnose(dengue_now) |>
  count(status)
```

`status` is an **ordered factor** — `error` > `warning` > `note` > `ok` >
`not_run` > `skipped` — so the table sorts itself and filtering to what needs
acting on is a comparison rather than a set membership test:

```{r diagnose-actionable}
diagnose(dengue_now) |>
  filter(status <= "note") |>
  select(check, scope, stratum, status, n_affected, n_total, message)
```

Each finding also carries a `hint` saying what to do about it, and a `rows`
list-column of the offending row indices, so you can go straight to them:

```{r diagnose-rows}
finding <- diagnose(dengue_now) |> filter(check == "declarations", status <= "note")

finding$hint
```

## The six statuses, and why `skipped` is not `ok`

The distinction that matters most is between a check that **ran and found
nothing** and one that **could not run**:

```{r diagnose-skipped}
diagnose(dengue_now) |>
  filter(status == "skipped") |>
  select(check, scope, message)
```

`dengue_now` is a line list with no confirmation process, so four checks have
nothing to work on. None of them is a pass. Reporting them as `ok` would be a
quiet lie, and this is the single most common way a health check misleads:
silence that reads as approval.

Note the first row in particular. **`diagnose()` will not look for duplicate
records in a line list**, because a line list is one row per case — two
identical rows are two infections, not a repeat. Deduplicating a line list needs
a key the object does not have (a record id, say), so that check stays yours.

## Structural, never statistical

`diagnose()` runs no statistical tests at all. It is fast, and its answer never
depends on a random seed or an optional package. The questions that *do* need a
test come back as `not_run` signposts naming the call that answers each one:

```{r diagnose-signposts}
diagnose_signposts(dengue_now) |>
  select(scope, status, message)
```

Those calls — `diagnose_drift()`, `diagnose_changepoint()`,
`diagnose_batches()`, `diagnose_batch_shape()` — are the statistical
diagnostics, documented in the
[batch-reporting article](https://rodrigozepeda.github.io/tbl.now/articles/batch-reporting.html).
They return their own shapes rather than the findings schema, because a
hypothesis test has a p-value and an effect size that a findings row has nowhere
to put.

## Cumulative data: the revisions

Cumulative counts get revised downwards, and a downward revision becomes a
**negative increment** the moment the series is de-accumulated. Anything
consuming incidence has to cope with that, so `diagnose()` counts them.
FluSight is the standard example:

```{r flusight}
data(flusight)

flu_now <- flusight |>
  filter(location_name %in% c("Alabama", "Alaska", "Arizona"),
         target_end_date >= as.Date("2023-01-01")) |>
  tbl_now(
    event_date  = target_end_date,
    report_date = as_of,
    case_count  = observation,
    strata      = location_name,
    data_type   = "count-cumulative",
    verbose     = FALSE
  )

diagnose_negatives(flu_now) |>
  select(scope, stratum, status, n_affected, n_total, message)
```

For cumulative data `summary()` also swaps its `delay` block for a `growth`
one — the ratio of one delay's running total to the previous one's — because a
cumulative total is not additive across delays and a case-weighted delay
distribution would be meaningless:

```{r flusight-growth}
cumulative_growth(flu_now, k = 3) |>
  filter(stratum == "all") |>
  select(quantity, n, total, mean, q50, max)
```

The mean ratio falls below 1 by the third delay, and the totals go negative:
these series shrink on revision more often than they grow. `delay_summary()`
refuses to run on this object at all, and says why.

# The same findings, two presentations

`validate_tbl_now()` runs the **same engine** as `diagnose()`. It does not
return the findings; it re-emits the `error` rows as an abort and the `warning`
rows as warnings. That is why a malformed object complains the moment you build
it, and why the two can never drift apart:

```{r validate, error=TRUE}
bad <- denguedat |>
  mutate(report_week = report_week - 400) |>
  tbl_now(event_date = "onset_week", report_date = "report_week", verbose = FALSE)
```

The warning above and the `ordering` row of `diagnose(bad)` are the same
finding, formatted for two different audiences: one for a person who is about to
make a mistake, one for a program that needs to decide what to do.

`note` rows are deliberately never promoted to warnings. `validate_tbl_now()`
runs on every `dplyr` verb, so a new warning there would turn a quiet
construction into a noisy one for data that has always been accepted.

# Where to go next

* `?tbl_now_summary` and `?nowcast_summary_components` — every column, every block.
* `?diagnose` and `?nowcast_diagnose_components` — every check, every status.
* The [worked example](https://rodrigozepeda.github.io/tbl.now/articles/example.html)
  applies both to a real, deliberately unpolished dataset, and cleans it on the
  strength of what `diagnose()` reports.
