---
title: "Diagnosing a tbl_now"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnosing a tbl_now}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 4.5,
  dpi = 120,
  cache = TRUE
)
# The summary and diagnose tables are wide, and wrapping them costs more than
# the horizontal scroll bar does.
options(tibble.width = Inf, width = 160, pillar.width = 160)
```

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(tbl.now)
```

Three questions come up with every new surveillance dataset, and they are
different questions that want different tools:

1. **What is in it?** -- how many cases, over what period, arriving how late,
   how sparse, how concentrated in one stratum. `summary()` answers this.
2. **What is structurally wrong with it?** -- missing dates, impossible
   orderings, repeated rows, units that do not line up, data that stops before
   its own `now`. `diagnose()` answers this, deterministically and with no
   model.
3. **What needs a statistical test?** -- has the reporting delay drifted, did
   reports arrive in batches, is a spike real cases or released backlog.
   `diagnose()` deliberately refuses these -- answering them means choosing a
   method, a window and a multiplicity correction -- and leaves them to
   `diagnose_drift()` and `diagnose_batches()`.

This article walks the three in order. The first two return a **tibble**, not
printed text, which is the design decision everything else follows from: an
answer you can `filter()`, `join()`, plot or assert on in a test is worth far
more than one you can only read. The third returns test results and figures,
because a hypothesis test has a p-value and an effect size that a findings row
has nowhere to put.

```{r dengue-data}
data(denguedat)

dengue_now <- tbl_now(denguedat,
  event_date  = "onset_week",
  report_date = "report_week",
  strata      = "gender",
  verbose     = FALSE
)
```

# Part 1 — What is in the data: `summary()`

## One table, several blocks

`summary()` returns a table, and prints it one **component** at a time with the
columns that component does not populate dropped — the schema is wide because it
has to hold every block at once, and no single block fills more than a handful
of it:

```{r summary-print}
summary(dengue_now)
```

Underneath it is an ordinary tibble, so every `dplyr` verb works on it and
`tibble::as_tibble()` gives the full schema back. The `component` column says
which block a row belongs to:

```{r summary-blocks}
summary(dengue_now) |>
  count(component)
```

Every row is one quantity, described by up to eighteen columns. Not every column
applies to every row — a delay distribution has a `mean` but no `prop`, a
proportion has a `prop` but no `q90` — so the table is deliberately sparse:

```{r summary-shape}
summary(dengue_now) |>
  filter(component == "cases") |>
  select(component, quantity, stratum, n, total, mean, sd, min, q50, q90, max, prop_zero)
```

`stratum` is always "which subset of the data does this row describe": `"all"`
for the pooled rows, or a stratum label. The *category* of a compositional row
goes in `quantity` instead, which keeps the two ideas from colliding:

```{r summary-composition}
summary(dengue_now) |>
  filter(component == "composition") |>
  select(quantity, stratum, n, total, prop)
```

## The blocks worth knowing

**`delay`** — the case-weighted reporting delay. Weighted means "equal to what
you would get by expanding the counts to one row per case", so it needs no
separate explanation:

```{r summary-delay}
delay_summary(dengue_now) |>
  select(quantity, stratum, n, total, mean, sd, q50, q90, max)
```

**`reporting_completeness()`** — the share of each event date's eventual total
that had arrived by delay `d`. It says how far back a nowcast has anything left
to estimate, and it is the one number here that most often changes a decision.
It is **not** part of `summary()`, and it warns on every call: it was written by
an AI and has not yet been checked by a human, so read it as a starting point
rather than as a verified statistic.

```{r summary-completeness}
reporting_completeness(dengue_now, delays = 0:4) |>
  filter(stratum == "all") |>
  select(quantity, n, mean, q50, prop)
```

Four percent of a week's cases are in by the end of that week, half by the end
of the next, and 94% by three weeks. Beyond about three weeks there is very
little missing to reconstruct.

**`zero_run`** — how sparse the series is, measured as runs of consecutive
zero dates rather than as a simple proportion. A series that is 30% zeros
scattered at random is a very different modelling problem from one that is 30%
zeros in three long gaps:

```{r summary-zeroruns}
zero_run_summary(dengue_now, axis = "event") |>
  select(quantity, stratum, n, total, mean, q50, max)
```

**`coverage`** — the reach of the object, including how stale it is:

```{r summary-coverage}
triangle_occupancy(dengue_now) |>
  filter(stratum == "all") |>
  select(quantity, n, value)
```

## Every block is also a function

`summary()` is exactly the `bind_rows()` of its components, and each one is
exported, so you can ask a single question without computing the rest:

```{r summary-components}
prop_strata(dengue_now) |>
  select(quantity, total, prop)
```

The full set is `cases_per_date()`, `delay_summary()`, `zero_run_summary()`,
`prop_censored()`, `prop_validation_type()`, `prop_strata()`,
`prop_covariate_levels()`, `date_ranges()`, `triangle_occupancy()` and
`cumulative_growth()`, plus the two unreviewed ones -- `case_autocorrelation()`
and `reporting_completeness()` -- which share the schema but are not part of
`summary()`.

<div class="alert alert-info">
**Quantiles are inverse-ECDF (type 1).** `q50` is the smallest value whose
cumulative weight reaches 0.5 — for an even number of observations, the upper of
the two middle values rather than their average. This is the same estimator
`autoplot()` and `plot_delay_drift()` use, so the table and the figures always
agree, and for count data it always returns a value that was actually observed.
A half-case delay is not.
</div>

# Part 2 — What is structurally wrong: `diagnose()`

## Findings, worst first

`diagnose()` prints as a report: the errors, warnings and notes in full, each
with its hint, and one line each for the checks that passed, that were
deliberately not run, and that could not be assessed. `print(x, all = TRUE)`
spells those out too.

```{r diagnose-print}
diagnose(dengue_now)
```

It is a tibble underneath, in the schema the rest of this section reads:

```{r diagnose-all}
diagnose(dengue_now) |>
  count(status)
```

`status` is an **ordered factor** — `error` > `warning` > `note` > `ok` >
`skipped` — so the table sorts itself and filtering to what needs
acting on is a comparison rather than a set membership test:

```{r diagnose-actionable}
diagnose(dengue_now) |>
  filter(status <= "note") |>
  select(check, scope, stratum, status, n_affected, n_total, message)
```

Each finding also carries a `hint` saying what to do about it, and a `rows`
list-column of the offending row indices, so you can go straight to them:

```{r diagnose-rows}
finding <- diagnose(dengue_now) |> filter(check == "declarations", status <= "note")

finding$hint
```

## The five statuses, and why `skipped` is not `ok`

The distinction that matters most is between a check that **ran and found
nothing** and one that **could not run**:

```{r diagnose-skipped}
diagnose(dengue_now) |>
  filter(status == "skipped") |>
  select(check, scope, message)
```

`dengue_now` is a line list with no validation process, so four checks have
nothing to work on. None of them is a pass. Reporting them as `ok` would be a
quiet lie, and this is the single most common way a health check misleads:
silence that reads as approval.

Note the first row in particular. **`diagnose()` will not look for duplicate
records in a line list**, because a line list is one row per case — two
identical rows are two infections, not a repeat. Deduplicating a line list needs
a key the object does not have (a record id, say), so that check stays yours.

## Structural, never statistical

`diagnose()` runs no statistical tests at all. It is fast, and its answer never
depends on a random seed or an optional package. The questions that *do* need a
test are simply not in it: they have their own functions, and you call one when
you want its answer.

Those calls — `diagnose_drift()`, `diagnose_changepoint()`,
`diagnose_batches()`, `diagnose_batches2()` — are Part 3 of this article.
They return their own shapes rather than the findings schema, because a
hypothesis test has a p-value and an effect size that a findings row has nowhere
to put.

## Cumulative data: the revisions

Cumulative counts get revised downwards, and a downward revision becomes a
**negative increment** the moment the series is de-accumulated. Anything
consuming incidence has to cope with that, so `diagnose()` counts them.
FluSight is the standard example:

```{r flusight}
data(flusight)

flu_now <- flusight |>
  filter(location_name %in% c("Alabama", "Alaska", "Arizona"),
         target_end_date >= as.Date("2023-01-01")) |>
  tbl_now(
    event_date  = target_end_date,
    report_date = as_of,
    case_count  = observation,
    strata      = location_name,
    data_type   = "count-cumulative",
    verbose     = FALSE
  )

diagnose_negatives(flu_now) |>
  select(scope, stratum, status, n_affected, n_total, message)
```

For cumulative data `summary()` also swaps its `delay` block for a `growth`
one — the ratio of one delay's running total to the previous one's — because a
cumulative total is not additive across delays and a case-weighted delay
distribution would be meaningless:

```{r flusight-growth}
cumulative_growth(flu_now, k = 3) |>
  filter(stratum == "all") |>
  select(quantity, n, total, mean, q50, max)
```

The mean ratio falls below 1 by the third delay, and the totals go negative:
these series shrink on revision more often than they grow. `delay_summary()`
refuses to run on this object at all, and says why.

## The same findings, two presentations

`validate_tbl_now()` runs the **same engine** as `diagnose()`. It does not
return the findings; it re-emits the `error` rows as an abort and the `warning`
rows as warnings. That is why a malformed object complains the moment you build
it, and why the two can never drift apart:

```{r validate, error=TRUE}
bad <- denguedat |>
  mutate(report_week = report_week - 400) |>
  tbl_now(event_date = "onset_week", report_date = "report_week", verbose = FALSE)
```

The warning above and the `ordering` row of `diagnose(bad)` are the same
finding, formatted for two different audiences: one for a person who is about to
make a mistake, one for a program that needs to decide what to do.

`note` rows are deliberately never promoted to warnings. `validate_tbl_now()`
runs on every `dplyr` verb, so a new warning there would turn a quiet
construction into a noisy one for data that has always been accepted.


# Part 3 — What needs a statistical test: reporting artefacts

Everything above is structural: it can be decided by looking at the object, it
gives the same answer every time, and it costs nothing. The questions Part 2
would not touch are where that stops. Whether the reporting delay has
drifted, and whether a spike is released backlog or genuine new cases, are
statements about a **distribution** — and answering them means picking a method,
a window and a multiplicity correction, which is not a decision a health check
should make on your behalf.

So the rest of this article is the other half of the toolkit: the tests and the
pictures for the questions `diagnose()` refuses to answer.

## Batch reporting

Surveillance data does not always arrive smoothly. Sometimes the reporting system
halts or reduces its output (e.g. a data-system outage, an overwhelmed jurisdiction) and the
backlog is released later all at once. That release is called a **batch**: a collection of 
reports from previous periods that were held and reported all at once during a different
reporting period. Intuivitively one can think of a batch as a collection of reports that 
--in an ideal scenario-- _should have been reported_ on a previous date but 
were actually released later. 

```{r, echo = FALSE}
# --- 1. GENERATE SYNTHETIC SURVEILLANCE DATA ---
set.seed(42)
dates  <- seq(as.Date("2026-06-01"), as.Date("2026-06-25"), by = "day")
n_days <- length(dates)

# Ideal Scenario: A smooth, wave-like curve (e.g., normal distribution)
ideal_cases <- round(100 * dnorm(seq(-2, 2.5, length.out = n_days)) + 5)

# Create the data frame
data <- data.frame(
  date  = dates,
  ideal = ideal_cases
)

# Introduce a data-system freeze/backlog:
# - Days 12, 13, and 14: System breaks down. Reports drop to almost 0.
# - Day 15: The bottleneck is cleared, releasing all accumulated backlog as a "Batch".
backlog_days <- c(12, 13, 14)
batch_day <- 15

# Calculate how many cases were missed during the outage
backlog_volume <- sum(data$ideal[backlog_days])

# Build the "real" curve with the batch effect
data <- data |>
  mutate(
    real = ideal,
    # Simulate the reporting drop
    real = ifelse(row_number() %in% backlog_days, 1, real), 
    # Add the accumulated backlog to the batch day
    real = ifelse(row_number() == batch_day, ideal + backlog_volume, real)
  )

# --- 2. DEFINE CUSTOM PALETTE (From your provided list) ---
# Main bars: #334335 (Dark green)
# Ideal curve: #5F7E62 (Muted green)
# Annotation accent: #B85348 (Coral red for warnings/anomalies)
# Texts & grid elements: #262626 (Off-black)
# Grid lines: #A8BFA9 (Soft pale green-gray)

# --- 3. CREATE THE PLOT ---
ggplot(data, aes(x = date)) +
  # Real reporting (bars)
  geom_col(aes(y = real, fill = "Real reporting with batch"), alpha = 0.9, width = 0.8) +
  # Ideal reporting (smooth line)
  geom_line(aes(y = ideal, color = "Ideal reporting scenario"), linewidth = 1, linetype = "solid") +
  
  # Set Colors
  scale_fill_manual(values = c("Real reporting with batch" = "#5F7E62")) +
  scale_color_manual(values = c("Ideal reporting scenario" = "#B85348")) +
  
  # --- ANNOTATION 1: The Valley (Where cases should have been reported) ---
  annotate(
    "text", 
    x = dates[mean(backlog_days)], 
    y = max(data$ideal) * 1.5, 
    label = "System Outage\n(Ideally, cases should\nhave been reported here)", 
    color = "black", fontface = "bold", size = 3.5, vjust = 0.5, hjust = 1
  ) +
  annotate(
    "segment", 
    x = dates[mean(backlog_days)], xend = dates[mean(backlog_days)], 
    y = max(data$ideal) * 1.35, yend = 5, 
    arrow = arrow(length = unit(0.2, "cm")), color = "black", linewidth = 0.8
  ) +
  
  # --- ANNOTATION 2: The Batch (The release day) ---
  annotate(
    "text", 
    x = dates[batch_day] + 3.5, 
    y = data$real[batch_day] - 10, 
    label = "The Batch!\nBacklog released\nall at once", 
    color = "black", fontface = "bold", size = 3.8, hjust = 0
  ) +
  annotate(
    "segment", 
    x = dates[batch_day] + 3.2, xend = dates[batch_day] + 0.3, 
    y = data$real[batch_day] - 10, yend = data$real[batch_day] - 5, 
    arrow = arrow(length = unit(0.2, "cm")), color = "black", linewidth = 0.8
  ) +
  
  # Theme and Formatting
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#262626"),
    plot.subtitle = element_text(size = 11, color = "#607060", margin = margin(b = 15)),
    axis.title = element_text(face = "bold", color = "#262626"),
    axis.text = element_text(color = "#262626"),
    panel.grid.major = element_line(color = "#A8BFA9", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 10, color = "#262626")
  ) +
  labs(
    title = "Understanding 'Batches' in Public Health Surveillance",
    x = "Reporting Date",
    y = "Number of Cases Reported"
  )
```


A batch is easy to confuse with an epidemic **surge**. The important part is that the batch happens
on the **reporting date** axis while a surge happens on the **event date** axis. A
batch just _moves_ reports to a later date not adding new cases while a real
epidemic surge _adds_ new cases. The tools here are designed to help you visualize that difference. 

> This article shows each plot **twice**: first on a clean, made-up
outbreak with one obvious batch (so you learn the signature), then on real,
COVID-19 data from the CDC (see [`covid_us`](https://rodrigozepeda.github.io/tbl.now/reference/covid_us.html)). 

## Two datasets to compare

### The made-up outbreak. 

This simulation consists of a bell-shaped curve over a hundred days, each
case reported within a few days. For one week near the peak, the reporting
system slows down with **half** of each day's reports being held back and released days after. 
That release is the **batch**. 

```{r ideal-data, warning = FALSE}
set.seed(82495)
#Simulate a curve
onset_days <- as.Date("2024-01-01") + 0:99
bell       <- dnorm(seq(-2.5, 2.5, length.out = 100))
per_day    <- round(400 * bell / max(bell)) + 8         
onset      <- rep(onset_days, per_day)
reported   <- onset + rpois(length(onset), 1.5)         

clean_tn <- tbl_now(tibble(onset = onset, reported = reported),
                    event_date = onset, report_date = reported,
                    data_type = "linelist", verbose = FALSE)

# Simulate a batch
ideal <- simulate_batch(clean_tn,
  closed_dates  = seq(as.Date("2024-02-19"), by = "day", length.out = 7),
  held_fraction = 0.5)
```

We can see the simulated data both from the event-date and the report-date perspectives:

```{r, eval = FALSE}
plot_epidemic_process(ideal)
plot_reporting_process(ideal)
```

```{r, echo = FALSE}
plot_epidemic_process(ideal) + plot_reporting_process(ideal)
```

### The real data

`covid_us` comes from the CDC's individual-level
[COVID-19 Case Surveillance Public Use Data](https://data.cdc.gov/Case-Surveillance/COVID-19-Case-Surveillance-Public-Use-Data/vbim-akqf/about_data).
It carries three dates. Here we use the first and the last: symptom onset
(`onset_dt`) as the event, and registration at CDC (`cdc_report_dt`) as the
report. The middle one, `pos_spec_dt`, is the specimen collection, and it is
pooled away along with `current_status` and `sex` -- the batch question is about
when reports *arrived*, not about who they were.

```{r data}
data(covid_us)

covid_early <- covid_us |>
  summarise(n = sum(n), .by = c(onset_dt, cdc_report_dt))

tn <- tbl_now(covid_early, event_date = onset_dt,
              report_date = cdc_report_dt, case_count = n,
              data_type = "count-incidence", verbose = FALSE)
```

Half of all cases were reported within a few days, but the tail is long: some cases take weeks or months to surface.

```{r delays}
stats::quantile(rep(tn$.delay, tn$n), c(0.5, 0.75, 0.9, 0.99))
```

We can see this dataset again from both the event-date and the report-date perspectives:

```{r, eval = FALSE}
plot_epidemic_process(tn)
plot_reporting_process(tn)
```

```{r, echo = FALSE}
plot_epidemic_process(tn) + plot_reporting_process(tn)
```


## The reporting process

This plot, which we have previously shown, shows how many reports arrived by date. Batches or surges might correspond to spikes towering over their neighbours. 

```{r, fig.cap="Reporting process of the simulated data"}
plot_reporting_process(ideal)
```

On the real data the reporting is spikier; a handful of peaks stick up where smooth
epidemic reporting should be. Those are reporting artefacts either pure
backlog releases, or a mix of backlog + a genuine surge (we'll come back to this characterization later). The tallest is a single
day of about 50K reports on 12 December 2020, sixteen times the 3K that arrive on a
typical day; 10 June and 5 September are the other conspicuous ones.

```{r proc-covid, fig.cap="Reporting process of the COVID-19 data"}
plot_reporting_process(tn)
```

## The reporting triangle

We provide two different visualizations of the reporting triangle. In both
we plot the three temporal dimensions involved in the process: the event date,
the reporting date and the delay. We cover the plots in tiles coloured by how many cases
were registered then. 


### The classical reporting triangle

The classical view is given by `plot_reporting_triangle()` where each
tile is described by _when it happened_ (across) and _how long they took to be 
reported_ (vertical). The diagonal shows reports that arrive on the same day.  
An indicator of a batch is **a bright diagonal reaching high up** which
reporesents many  delayed cases being all reported on the same day.

```{r proc-sim, eval = FALSE}
plot_reporting_triangle(ideal)
```

```{r, fig.cap="Classical reporting triangle of the simulated data", echo = FALSE}
plot_reporting_triangle(ideal) + (plot_reporting_process(ideal))
```

On COVID-19, the triangle is a broad blue-grey haze (most cases reported over many
months) crossed by bright diagonals. They correspond to the same
spikes seen in the reporting process:

```{r tri-covid, eval = FALSE}
plot_reporting_triangle(tn)
```

```{r, fig.cap="Classical reporting triangle of the COVID-19 data", echo = FALSE}
plot_reporting_triangle(tn) + (plot_reporting_process(tn))
```

### The reporting hexamap

Event date, report date and reporting delay can be seen as an **age-period-cohort** triple
(`report = event + delay`, exactly `period = cohort + age`), so the reporting
triangle can be drawn as a hexamap in the style of [Jalal and Burke (2020)](https://doi.org/10.1097/EDE.0000000000001236): each
`(event, delay)` cell is a point on a hexagonal lattice, coloured by its report
count, with event date,
report date and delay running along the three 60-degree axes. Because a batch is a
happens in the **report date**, it shows up as a  **vertical stripe**; the fast
reporting bulk sits along the short-delay bottom edge.

```{r hex-sim, eval = TRUE, fig.width = 8, fig.height = 8}
plot_reporting_hexamap(ideal)
```

The marks are sized in millimetres while the lattice is sized in data units, so
no default can suit every combination of cell count and figure size. `size`
is the knob: raise it until the points nearly touch at the size you are actually
drawing, and `shape = 15` swaps the circles for squares, which tile more closely.

```{r hex-sim-big, eval = TRUE, fig.width = 8, fig.height = 8}
plot_reporting_hexamap(ideal, size = 3, shape = 15)
```

On covid the vertical stripes are the 2020 backlog releases. The
delay axis is capped with `max_delay` to keep the map to where the reports are.

```{r hex-covid, eval = TRUE, fig.width = 8.5, fig.height = 8}
plot_reporting_hexamap(tn, max_delay = 60)
```


## Scalograms

::: {.alert .alert-warning}
The scalogram functions are **very experimental**. We have yet to confirm they work for all batch cases. Feel free to skip to the section on [transport](https://rodrigozepeda.github.io/tbl.now/articles/diagnosing-a-tbl-now.html#sec-transport)
:::

Scalograms show reductions of cases. For this example, consider the following simulated reporting  process:

```{r, echo = FALSE}
data(denguedat)

ndata <- tibble(onset_week   = seq(as.Date("2000/01/01"), by = 7, length.out = 200),
                report_week  = seq(as.Date("2000/01/08"), by = 7, length.out = 200), 
                n = ceiling(sapply(1:200, \(x) 5*x + 1000*abs(sin(pi*x/52))))) |>
    tbl_now(
        event_date = onset_week, report_date = report_week, case_count = n, verbose = FALSE) 

#Smooth for the example
plot_reporting_process(ndata)

```


Its scalogram shows the decreases in the reporting cases as vertical streaks aligned with the minimal date of this decrease:

```{r, echo = FALSE}
#Smooth for the example
plot_scalogram(ndata)
```

One can see the same vertical streaks in the previous reporting process we had been working on corresponding to the dip before the batch:

```{r, eval = FALSE}
plot_scalogram(ideal)
```

```{r, echo = FALSE}
plot_scalogram(ideal, format = "%b") + plot_reporting_process(ideal)
```

and again the same release dates being the most identified, with additional dates having less of a clear pattern:

```{r, eval = FALSE}
plot_scalogram(tn)
```


```{r, echo = FALSE}
plot_scalogram(tn, format = "%b") + plot_reporting_process(tn)
```

## Transport vs creation {#sec-transport} 

This is the main tool for detecting batches and surges. Before the plot, we
will explain the whole idea. Consider a daily outbreak with  reports incoming 
each day. Three things can change the number of reports:

- a **hold**  -- a reporting office falls behind and some days' reports are withheld;
- a **batch** -- the day the backlog (hold) is finally released, all at once;
- a **surge** -- an increase in the epidemic process: more people falling ill and being reported.

We simulate one of each in a clean epidemic and colour every day by its type (grey
= ordinary day):

```{r tut-data, echo = FALSE}
tut_cols <- c(Normal = "grey72", Hold = "#4E79A7",
              Batch = "#B85348", Surge = "#5F7E62")

set.seed(2024)
# a clean bell-shaped epidemic, reported within a day or two
tut_ev <- rep(as.Date("2024-01-01") + 0:99,
              times = round(300 * dnorm(seq(-2.5, 2.5, length.out = 100)) /
                              max(dnorm(seq(-2.5, 2.5, length.out = 100)))) + 6)
tut_rp <- tut_ev + rpois(length(tut_ev), 1.5)
# a SURGE: extra *real* cases in mid-March, reported promptly
surge_ev <- rep(as.Date("2024-03-11") + 0:2, each = 500)
tut <- rbind(
  data.frame(event_date = tut_ev,   report_date = tut_rp),
  data.frame(event_date = surge_ev, report_date = surge_ev + rpois(length(surge_ev), 1))
)
# a HOLD then BATCH: close reporting for a week near the peak; the backlog is
# released the day the desk catches up
closed <- seq(as.Date("2024-02-19"), by = "day", length.out = 3)
tut_tn <- simulate_batch(
  tbl_now(tut, event_date, report_date, data_type = "linelist", verbose = FALSE),
  closed_dates = closed, held_fraction = 0.7, verbose = FALSE)

# the two scores per report date, tagged by what we planted
scores <- transport_discriminant(tut_tn, lookback = 7)
scores$kind <- factor(dplyr::case_when(
  scores$report_date %in% closed                          ~ "Hold",
  scores$report_date == max(closed) + 1                   ~ "Batch",
  scores$report_date >= as.Date("2024-03-11") &
    scores$report_date <= as.Date("2024-03-13")           ~ "Surge",
  TRUE                                                     ~ "Normal"),
  levels = names(tut_cols))
ar <- arrow(length = unit(0.15, "cm"))

ggplot(scores, aes(report_date, reported, fill = kind)) +
  geom_col(width = 1) +
  scale_fill_manual(values = tut_cols, name = NULL) +
  annotate("segment", x = as.Date("2024-02-01"), xend = as.Date("2024-02-21"),
           y = 700, yend = 750, arrow = ar, colour = tut_cols[["Batch"]]) +
  annotate("text", x = as.Date("2024-01-31"), y = 700, hjust = 1, size = 3,
           colour = tut_cols[["Batch"]], label = "Batch: the backlog\nreleased in one lump") +
  annotate("segment", x = as.Date("2024-02-09"), xend = as.Date("2024-02-20"),
           y = 400, yend = 175, arrow = ar, colour = tut_cols[["Hold"]]) +
  annotate("text", x = as.Date("2024-02-08"), y = 400, hjust = 1, size = 3,
           colour = tut_cols[["Hold"]], label = "Hold: reports\nwithheld (a dip)") +
  annotate("segment", x = as.Date("2024-03-27"), xend = as.Date("2024-03-14"),
           y = 560, yend = 500, arrow = ar, colour = tut_cols[["Surge"]]) +
  annotate("text", x = as.Date("2024-03-28"), y = 590, hjust = 0, size = 3,
           colour = tut_cols[["Surge"]], label = "Surge: genuinely\nmore cases") +
  coord_cartesian(clip = "off") +
  labs(x = "Report date", y = "Reports",
       title = "1. The reporting timeline, coloured by what happened") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", plot.margin = margin(6, 80, 6, 6))
```

In the previous plot, every bar is one report date. The **batch** towers 
where the held reports land together; the **hold** is the small blue dip 
just before it; the **surge** is a genuine bump of new cases.

The transport discriminant turns each day into **two numbers**:

- A **creation score** -- did this stretch of days genuinely *gain* cases? A larger surge
implies a larger creation score.
- A **transport score** -- were the days just before *missing* reports? 
A backlog release after a hold pushes it up.

We plot every day by those two numbers and observe the directions of the three disturbances:

```{r tut-plane, fig.width = 8, fig.height = 6, warning = FALSE, echo = FALSE}
zc <- qnorm(0.975)
ty <- max(scores$transport_z, na.rm = TRUE)
rx <- max(scores$creation_z,  na.rm = TRUE)
lx <- min(scores$creation_z,  na.rm = TRUE)
ggplot(scores, aes(creation_z, transport_z, colour = kind)) +
  geom_hline(yintercept = zc, linetype = "dashed", colour = "grey85") +
  geom_vline(xintercept = zc, linetype = "dashed", colour = "grey85") +
  geom_point(data = ~subset(., kind == "Normal"), size = 1.7, alpha = 0.45, na.rm = TRUE) +
  geom_point(data = ~subset(., kind != "Normal"), size = 2.8, na.rm = TRUE) +
  scale_colour_manual(values = tut_cols, name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.14))) +
  scale_x_continuous(expand = expansion(mult = 0.09)) +
  annotate("text", x = 8, y = ty + 3, size = 3, fontface = "bold",
           colour = tut_cols[["Batch"]], label = "Batch: shoots UP and RIGHT") +
  annotate("text", x = 20, y = -5, size = 3, fontface = "bold",
           colour = tut_cols[["Surge"]], label = "Surge:\nshoots RIGHT") +
  annotate("text", x = lx * 0.55, y = -10, size = 3, fontface = "bold",
           colour = tut_cols[["Hold"]], label = "Hold:\ngoes goes DOWN") +
  labs(x = "Creation score",
       y = "Transport score",
       title = "2. Where each day lands in the transport-vs-creation plane") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

Identifying any bar with its dot we can conclude:

- A **batch (backlog release, red)** shoots **up and right** -- the days before were
  depleted (high transport) but not as many new cases were created (right);
- A **surge (green)** shoots **right** -- cases genuinely appeared (high
  creation), with apparent preceding hole;
- A **hold (blue)** drifts **left** -- reports have apparently gone
missing (transport rising) while the window has *lost* cases (creation negative).

Ordinary days (grey) sit in the cloud through the middle. That is the whole idea
behind `transport_discriminant()` and `diagnose_batches()`: **a batch is high transport
with little creation**. 

### The transport discriminant

The previous plot can be done with the `plot_transport_discriminant()` function: 

```{r disc-sim, eval = FALSE}
plot_transport_discriminant(ideal)
```

```{r disc-sim2, echo = FALSE}
plot_transport_discriminant(ideal) + plot_reporting_process(ideal)
```

Which also works to identify the COVID-19 cases:

```{r disc-covid, eval = FALSE}
plot_transport_discriminant(tn, period = 7)
```

```{r disc-covid2, echo = FALSE}
plot_transport_discriminant(tn, period = 7) +
  plot_reporting_process(tn)
```

### Recovering the data

The `diagnose_batches()` function runs the transport test for the batch signature and
returns, for every report date, the `batch` flag -- a Benjamini-Hochberg-corrected
verdict that controls the false-discovery rate across all dates (see `?diagnose_batches`
for the full column reference). Keeping the `batch` rows gives the confirmed
releases together with their `deficit` (how depleted the days just before were) and
`delta` (how little the window total actually changed).

```{r screen-sim, eval = FALSE}
diagnose_batches(ideal) |>
  filter(batch)
```

```{r screen-sim2, echo = FALSE}
diagnose_batches(ideal) |>
  filter(batch) |>
  select(report_date, reported, baseline, deficit, delta, p_transport_bh, batch)
```

On covid we pass `period = 7` to divide out the weekly reporting cadence. Only
one date survives the Benjamini-Hochberg-corrected `batch` flag: **7 November
2020**, which reported about twice its baseline *and* was preceded by a matching
deficit of roughly the same size. That pairing is the whole test -- the taller
spikes of 12 December and 10 June are not flagged, because nothing was withheld
beforehand to release, which makes them surges rather than batches:

```{r screen-covid, eval = FALSE}
diagnose_batches(tn, period = 7) |>
  filter(batch)
```

```{r screen-covid2, echo = FALSE}
diagnose_batches(tn, period = 7) |>
  filter(batch) |>
  select(report_date, reported, baseline, deficit, delta, p_transport_bh, batch) |>
  arrange(desc(reported))
```

The sensitivity of the batch flag can be adapted with `alpha`. 


## Delay changes 

The reporting delay might change through time. Here we show two different plots 
for identifying delay problems. 

### Reporting-delay drift

The typical time from case to report, tracked over the outbreak. Shows the
overall trend of the delay. Normally it will be steady; a 
batch can be seen as a **sudden bump upward** on the release day.

```{r drift-sim}
plot_delay_drift(ideal)
```

On covid the delays show extreme variability in the beginning and a trend
that decreases the delay in time: 

```{r drift-covid}
plot_delay_drift(tn)
```

The functions `diagnose_drift()` and `diagnose_changepoint()` test 
for a gradual or abrupt changes in the delay. We can see, for example, that it correctly identifies the drift in the COVID-19 dataset both for the median (trend) and the spread (see the quantiles getting tighter). On the ideal example this doesn't happen so it does not detect a drift:

```{r, warning=FALSE}
diagnose_drift(tn)
diagnose_drift(ideal)
```

The change-point function also detects that by early June the COVID-19 delay distribution has completely changed from before. In the case of the ideal example the change is not  long enough to be detected: 

```{r, warning=FALSE}
diagnose_changepoint(tn)
diagnose_changepoint(ideal)
```

### Delay profiles

Each faint line is one day's distirbution of reporting delays. Most days report quickly, so
their lines hug the left and concentrate around the same distribution:

```{r prof-sim}
plot_delay_profiles(ideal)
```

On COVID-19 a whole spray of lines reaches far to the right days that reported
cases months later.

```{r prof-covid}
plot_delay_profiles(tn)
```


# The whole toolkit, on one page

The structural tools come first because they are cheap and never wrong; the
statistical ones come second because they cost a choice.

| Tool | Question it answers |
|---|---|
| **`summary()`** | What is in the data: counts, delays, sparsity, composition, reach. |
| **`diagnose()`** | What is structurally wrong: ordering, missingness, duplicates, units, right-truncation. |
| **`validate_tbl_now()`** | The same findings, raised as errors and warnings instead of returned. |

Everything below needs a statistical choice, and every row is a different way of
seeing the same thing -- reports that were held back and then released together.

| Plot or test | What to look for |
|---|---|
| **Reporting process** | Shows how reports were registered. |
| **Reporting triangle** | Diagonals show cases with the same report date. |
| **The reporting V** | Horizontal slices show cases with the same report date. |
| **Wavelet scalogram** | Bright short-period ridges in the reporting series show _holds_ on the reporting. |
| **Transport discriminant** | A red dot up-and-left, in the "potential batch region" might indicate a batch. |
| **`diagnose_batches()`** | Shows the flagged reports in the discriminant as a `data.frame` |
| **Delay profiles** | Show the delay distribution for each event date. |
| **Reporting-delay drift** | Shows how the delay and its variance change through time. |

Another practical notes from other data we have analyzed:

+ In general, **surges** (real new cases) seem harder to identify of that **batches** (moved reports).
+ Batches are very difficult to identify in low-incidence scenarios.  
+ Batches very close to the now are difficult to identify without additional modeling hypotheses.  


# Where to go next

* `?tbl_now_summary` and `?nowcast_summary_components` — every column, every
  block of Part 1.
* `?diagnose` and `?nowcast_diagnose_components` — every check, every status of
  Part 2.
* `?diagnose_drift`, `?diagnose_batches` and `?diagnostic_plot` — the tests and
  figures of Part 3.

```{r learning-more, echo=FALSE, results="asis"}
# The fragment is looked up by ABSOLUTE path via system.file(), never by a
# relative one. pkgdown/rmarkdown render into an intermediates directory under
# tempdir() and copy relative resources alongside; a path containing "../.."
# escapes that directory ("cannot create file '/tmp/RtmpXXX/../../man/...'"),
# which fails wherever tempdir() sits near the filesystem root -- as it does on
# CI. `system.file()` also works under pkgload::load_all(), which shims it to
# find inst/ in the source tree.
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```
