---
title: "Diagnosing a tbl_now"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnosing a tbl_now}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 4.5,
  dpi = 120,
  cache = TRUE
)
options(tibble.width = Inf, width = 160, pillar.width = 160)
options(pillar.print_max = 5, pillar.print_min = 5,
        tibble.print_min = 5, dplyr.print_min = 5,
        tibble.print_max = 5, dplyr.print_max = 5)
```

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(tbl.now)
```

Three questions frequently appear with every surveillance dataset:

1. **What is structurally wrong with it?**: missing and impossible dates, 
    data that stops before its own `now`. `diagnose()` answers this.
    
2. **What is in it?**: how many cases, over what period, arriving how late,
   how sparse, how do they vary among strata. `summary()` answers these.
   
3. **Are there any changes in the epidemic or the reporting mechanism?** To answer
  these we look at batches, drifts as well as visualizations of the processes. 

This article follows the `sari_bh` dataset, a linelist of severe acute respiratory
illness (SARI) cases from Belo Horizonte (Brazil) from 2020 to 2022:


```{r dengue-data, message=FALSE,warning=FALSE}
data(sari_bh)

#Create an age category for strata
sari_bh <- sari_bh |> 
  mutate(age_cat = cut(age_yrs, breaks = c(0, 20, 40, 60, Inf), include.lowest = TRUE))

sari <- tbl_now(sari_bh,
  event_date  = symptom_onset_date,
  report_date = record_date,
  strata      = age_cat
)

sari
```


# 1. Diagnose what is structurally wrong

## Linelist data

Once the `tbl_now` is specified, `diagnose()` can be used to print a report: with errors, warnings and notes:

```{r diagnose-print}
diagnose(sari)
```

in our case we have rows with missing report dates which we'll input via censoring (we'll utilize the last value observed):

```{r, message=FALSE, warning=FALSE}
sari <- sari |> censor_reports(is.na(record_date))
```

while for the 30 rows where the report date was set before the event we'll drop them as there is no principled way of handling these cases:

```{r}
sari <- sari |> filter(record_date >= symptom_onset_date)
```

The notes can also be reduced. `tbl_now()` aims for the bare minimum for a nowcast hence it complains about the extra columns that are neither covariates nor strata. We can remove them to silence the note:

```{r}
sari <- sari |> select(-age_yrs, -final_classification, -case_evolution)
```

This now leads to a cleaner diagnose which can be called the same way or printed as a tibble by assigning it to an object:

```{r}
sari_diagnostic <- diagnose(sari) |> filter(status != "ok" & status != "skipped")
```

<div class="small-table">
```{r, echo=FALSE}
sari_diagnostic |> select(-status, -scope, -check, -hint) |> knitr::kable()
```
</div>

## Cumulative data: the revisions

Cumulative counts represent at each entry the best estimate of the running total of cases by
that event-report date. As such, they can get revised downwards, and a downward revision can become a
**negative 'increment'** the moment the series is de-accumulated. We can see this
example with the `flusight` dataset: 

```{r flusight, message=FALSE, warning=FALSE}
data(flusight)

flu_now <- flusight |>
  filter(location_name == "Alabama", target_end_date >= as.Date("2023-01-01")) |>
  tbl_now(
    event_date  = target_end_date,
    report_date = as_of,
    case_count  = observation,
    data_type   = "count-cumulative"
  )

diagnose(flu_now)
```

The first thing to note here is that data is weekly though sometimes reported
on a Saturday and sometimes on a Wednesday. We can use the `align_weeks()` function
to set all the days to the same day of the week so that the delays are integer numbers:

```{r}
#Day 4 is a wednesday as per wday()
flu_now <- flu_now |> align_weeks(align_on_day = 4)
```

The remaining notes refer to the distance between today and the now as well as the fact
that this dataset sometimes gets revised in its number of cases (hence the negative 'increments')

# 2. Summarise what is in the data

The `summary()` function returns a table, and prints it one **component** at a time:

```{r summary-print}
summary(sari)
```

We'll discuss each of the blocks individually: 

### Delay block

Summarises the distribution of the reporting delay. Here we can see that on average it takes 29 days to
get reported but some cases took more than 500 days!

You can also see the reporting delay via its plot:

```{r, fig.width=9, fig.height=4.5}
plot_delay_distribution(sari)
```

Or get that specific table with:

```{r}
delay_summary(sari)
```

## Composition


The strata composition (i.e. the proportion of cases by strata) can be accessed with the 
`prop_strata()` function:

```{r}
prop_strata(sari)
```

Here we see that half of the cases occurred in people 60 years or older. 

## Sparcity

The `zero_run_summary` explains  how sparse the series is. It measures a zero-run: a collection of consecutive dates with no cases and quantifies how those runs behave. Here we can see that there is no sparcity (the longest zero runs are of 7 days with most of them lasting 1 day or less). The data is not at all sparse as we can see that if we see a zero, it is very likely the next day there will be cases (in contrast with another 0). 

```{r summary-zeroruns}
zero_run_summary(sari, axis = "event")
```

## Coverage

The `date_ranges` show the minimum and maximum for each date per strata to see if there was a lapse in coverage for one: 

```{r summary-coverage}
date_ranges(sari)
```

## The epidemic process

The `cases_per_date()` function recovers the description of the number of cases
with the mean number of cases reported per day and its distribution

```{r}
cases_per_date(sari)
```

The epidemic process can be seen with a plot too:S

```{r, fig.width=9, fig.height=4.5}
plot_epidemic_process(sari)
```

(to plot the unstratified one use `remove_strata` on the `tbl_now` before `plot_epidemic_process`)

We can see, for example, the flusight dataset too where we can see that there were no reports for the off-season in 2024. 

```{r, fig.width=9, fig.height=4.5}
plot_epidemic_process(flu_now)
``` 

We can use the `complete_zeroes()` function to substitute those values for zeroes and complete the observations:

```{r}
flu_now <- flu_now |> complete_zeroes(max_delay = 10)
```

# 3. Visualizing a `tbl.now` 

One can use `autoplot()` to glance at the main properties of a `tbl.now` where we can see that there is an effect of the weekend on the reporting:

```{r, fig.width=11, fig.height=13}
autoplot(sari)
```

with less cases getting reported on Saturday/Sunday. Additional plots help you observe how the reports have changed over time where you can clearly see the emptyness of the weekends:

```{r, fig.width=9, fig.height=4.5}
plot_reporting_process(sari)
```

The reporting triangle shows in the same plot the delays, reports and  event dates

```{r, fig.width=9, fig.height=4.5}
sari |> remove_all_strata() |> plot_reporting_triangle()
```

where we can again see the weekends leaving empty streaks. 

# 4. Identifying delay changes

The reporting delay might change through time. Here we show two different plots 
for identifying delay problems. 

### Reporting-delay drift

This shows the typical time from event to report as it moved through the outbreak. It shows the
overall trend of the delay and its quantiles. 

```{r drift-covid, fig.width=9, fig.height=4.5}
plot_delay_drift(sari)
```

Here we can see that the first cases took so long to get reported. However, we can also see that after the initial chaos int he first months of 2020 the reporting delay became pretty stable.  The functions `diagnose_drift()` and `diagnose_changepoint()` test 
for a gradual or abrupt changes in the delay. 

Here we can see that they identify a slight reduction (the delay reduces 0.0105 days per day) and no changes in the intervals

```{r, warning=FALSE}
diagnose_drift(sari)
```

The change-point function on the other hand looks for changes in the distribution of the delay. The function however has to be used carfully as the changepoint function will **always** look for a changepoint in the data. So one has to first see the delay distribution to detect a possible changepoint before calling the diagnostic. The diagnostic will always find a changepoint no matter what. So here although it gives a very small p value we have decided to conclude there is no changepoint based on the delay plot:

```{r, warning=FALSE}
diagnose_changepoint(sari)
```

# Batches

Reporting batches are an important hurdle in surveillance systems. See [this vignette on batches](https://rodrigozepeda.github.io/tbl.now/articles/batches.html) for how we represent them and how to diagnose them. For the purpose of this tutorial we present two diagnosing functions. The `diagnose_batches()` identifies potential dates that might carry reporting batches:

```{r}
sari |> remove_all_strata() |> diagnose_batches()
```

while `diagnose_batches2()` requires you to identify a date to test for batch-reporting:

```{r}
sari |> remove_all_strata() |> diagnose_batches2(at = as.Date("2021/11/11"))
```




# Summary

Here we have shown how to diagnose, summarise and visualize a `tbl.now` to characterize its data. 



```{r learning-more, echo=FALSE, results="asis"}
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```
