---
title: "One call, many models: `run_nowcast()`, custom backends and ensembles"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{One call, many models: run_nowcast(), custom backends and ensembles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The Bayesian fits (Stan / JAGS / INLA) are shown as illustrative code but are
# not run when the vignette is built. Flip `run_models` to TRUE (and make sure
# the corresponding packages are installed) to actually fit them.
run_models <- FALSE

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)

# Several of the modelling packages route their data through data.table; pinning
# it to a single thread avoids an OpenMP crash seen when knitr forks the build.
if (has("data.table")) data.table::setDTthreads(1)
```

> This was written automatically by an AI model. A human has yet to review.

## Why this vignette?

`vignette("nowcasting-models")` shows how one `tbl_now` can be handed to five
different modelling packages. That still leaves you writing five different
calls, reading five different output formats, and doing the bookkeeping yourself
if you want to compare or combine them.

This vignette is about the layer that removes that bookkeeping:

* **`run_nowcast(x, method = ...)`** fits *any* supported package and always
  returns the same kind of object,
* **`nowcast_ensemble()`** combines several of those objects into an ensemble,
* **`nowcast_backtest()`** and **`nowcast_weights()`** decide, from data, how
  much each member should count,
* and **`nowcast_fit()` / `nowcast_tidy()`** let you plug in a model that
  `tbl.now` has never heard of.

```{r setup, message=FALSE}
library(dplyr)
library(tbl.now)

data(denguedat)

dengue <- tbl_now(denguedat,
  event_date  = onset_week,
  report_date = report_week,
  verbose     = FALSE
)
```

A compact recent window keeps everything on this page fast:

```{r window}
dengue <- dengue |>
  filter(onset_week >= as.Date("2009-01-01"))

get_now(dengue)
```

## 1. One call per model

`run_nowcast()` takes the `tbl_now`, the name of a modelling package, and
anything that package accepts:

```{r bln-run, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
baseline <- run_nowcast(dengue, method = "baselinenowcast", draws = 1000, verbose = FALSE)

baseline
```

The `...` goes straight through to the backend, so switching models mostly means
switching the arguments you already know:

```{r other-methods, eval=FALSE}
# Bayesian structural time series (diseasenowcasting)
dnc <- run_nowcast(dengue, "diseasenowcasting", type = "one_stage", n_draws = 2000)

# Bayesian delay + reference modules (epinowcast)
enw <- run_nowcast(dengue, "epinowcast",
  preprocess_args = list(max_delay = 10),
  fit = epinowcast::enw_fit_opts(chains = 2, iter_sampling = 500)
)

# Nowcasting by Bayesian Smoothing (NobBS)
nobbs <- run_nowcast(dengue, "NobBS", max_D = 8, moving_window = 30)

# INLA-based nowcast (nowcaster); needs the INLA package
inla <- run_nowcast(dengue, "nowcaster", Dmax = 10)
```

`list_nowcast_methods()` tells you what is available in your session:

```{r methods}
list_nowcast_methods()
```

### What you get back

Every method returns a `tbl_nowcast`. Whatever the package produced — a Stan
fit, a matrix of samples, a data frame of quantiles — the predictions are always
in the same tidy shape:

```{r bln-tidy, eval=has("baselinenowcast")}
as_tibble(baseline)
```

and, when the backend is sample-based, the draws are there too:

```{r bln-draws, eval=has("baselinenowcast")}
as_tibble(baseline, type = "draws")
```

Nothing is thrown away. The `fit` property still holds the backend's own object,
so you can keep using that package's diagnostics:

```{r bln-fit, eval=has("baselinenowcast")}
class(baseline@fit)
```

There is an `autoplot()` method for a quick look:

```{r bln-plot, eval=has("baselinenowcast") && has("ggplot2"), fig.width=7, fig.height=4}
autoplot(baseline)
```

### Strata are handled for you

When the `tbl_now` declares strata, `run_nowcast()` does whatever that backend
needs to honour them — one reporting triangle per stratum for `baselinenowcast`,
`NobBS.strat()` instead of `NobBS()`, the grouping variable for `epinowcast` —
and the stratum ends up as an ordinary column of the output:

```{r strata, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
dengue_by_sex <- dengue |>
  add_strata(gender)

by_sex <- run_nowcast(dengue_by_sex, "baselinenowcast", draws = 500, verbose = FALSE)

as_tibble(by_sex)
```

If a backend genuinely cannot stratify, it warns and pools rather than
pretending.

## 2. Scoring a nowcast

`score_nowcast()` compares the predictive quantiles with what was eventually
observed, using the weighted interval score (WIS) of @bracher2021.

Scoring only makes sense against data the model had not seen, so the usual
pattern is to nowcast at a date in the past and score against the full series:

```{r score, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
retrospective <- dengue |>
  filter(report_week <= as.Date("2010-09-01")) |>
  change_now(now = as.Date("2010-09-01"))

past_nowcast <- run_nowcast(retrospective, "baselinenowcast", draws = 500, verbose = FALSE)

# Scored against everything that was eventually reported
score_nowcast(past_nowcast, truth = nowcast_truth(dengue)) |>
  tail()
```

If you would rather use the full \pkg{scoringutils} score suite,
`as_scoringutils()` reshapes the same object into the format it expects:

```{r scoringutils, eval=has("baselinenowcast") && has("scoringutils"), message=FALSE}
past_nowcast |>
  as_scoringutils(truth = nowcast_truth(dengue)) |>
  as.data.frame() |>
  scoringutils::as_forecast_quantile() |>
  scoringutils::score() |>
  head()
```

## 3. Ensembles

An ensemble of several nowcasts is usually better calibrated than any of its
members: models fail in different directions, and averaging them cancels part of
that. Because every `run_nowcast()` result has the same shape, combining them is
one call.

```{r members, eval=FALSE}
members <- list(
  baseline = run_nowcast(dengue, "baselinenowcast", draws = 1000),
  dnc      = run_nowcast(dengue, "diseasenowcasting", type = "one_stage"),
  nobbs    = run_nowcast(dengue, "NobBS", max_D = 8, moving_window = 30)
)

ensemble <- nowcast_ensemble(members)
```

### Two ways of combining

`type = "quantile"` (the default) averages the members' quantiles level by
level. It is the workhorse of the forecast hubs, it always applies, and it tends
to produce *narrower* intervals than the members.

`type = "linear_pool"` instead pools the members' draws into a mixture
distribution. It needs draws from every member and generally produces *wider*
intervals, because disagreement between models becomes extra spread rather than
being averaged away.

```{r pool, eval=FALSE}
sharp <- nowcast_ensemble(members, type = "quantile")
wide  <- nowcast_ensemble(members, type = "linear_pool", n_draws = 4000)
```

Which one you want depends on whether you read between-model disagreement as
noise to be averaged out (quantile) or as genuine uncertainty to be propagated
(linear pool).

### Weights

By default every member counts the same. You can also give weights directly:

```{r weights-manual, eval=FALSE}
nowcast_ensemble(members, weights = c(baseline = 0.5, dnc = 0.3, nobbs = 0.2))
```

or *learn* them from how the members actually performed. `nowcast_backtest()`
walks back through time: at each retrospective date it truncates the data to the
reports available then, refits every method, and scores the result against what
was eventually observed.

```{r backtest, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
backtest <- nowcast_backtest(
  dengue,
  methods   = "baselinenowcast",
  now_dates = as.Date(c("2010-08-01", "2010-09-01", "2010-10-01")),
  draws     = 500,
  verbose   = FALSE
)

backtest$scores |>
  group_by(.method, .now) |>
  summarise(mean_wis = mean(wis, na.rm = TRUE), .groups = "drop")
```

This is the expensive part of the workflow — it is `length(methods) ×
length(now_dates)` model fits — so keep `now_dates` short when the members are
Bayesian.

From the backtest, `nowcast_weights()` offers two rules:

```{r weights-fitted, eval=FALSE}
nowcast_weights(backtest, type = "inverse_score") # w proportional to 1 / mean WIS
nowcast_weights(backtest, type = "optim")         # w minimising the ensemble's WIS
```

`"inverse_score"` is the safe default: it is monotone in performance, never
collapses onto a single model, and cannot overfit. `"optim"` searches the
simplex for the weights that would have minimised the ensemble's WIS over the
training window; it is better in principle, but with only a handful of
retrospective dates it happily overfits, so prefer it when the training window
is long.

Either rule plugs straight into the ensemble:

```{r ensemble-weighted, eval=FALSE}
nowcast_ensemble(members, weights = "inverse_score", backtest = backtest)
```

## 4. Does the ensemble help? Three epidemics

> This was written automatically by an AI model. A human has yet to review.

Talk of robustness is cheap; let us measure it. We take three very different
outbreaks that ship with `tbl.now` — **dengue** (weekly, Puerto Rico),
**mpox** (daily, US) and **covid** (weekly, Colombia) — and on each one we
nowcast at a fixed `now`, then score every model, and the ensemble, against the
counts that were *eventually* reported.

The recipe is the same for all three:

```{r comparison-recipe, eval=FALSE}
# For one epidemic (see below for how each dataset is prepared):
#   1. snapshot the data to the reports available at `now`
#   2. fit each member on that snapshot
#   3. build the ensemble
#   4. score every model AND the ensemble against the eventual truth

evaluate <- function(x_full, now, members, eval_dates, method_args = list()) {
  snapshot <- x_full |>
    filter(.data[[get_report_date(x_full)]] <= now) |>
    change_now(now = now)

  fits <- lapply(members, function(m) {
    do.call(run_nowcast, c(list(snapshot, m, verbose = FALSE), method_args[[m]]))
  })
  names(fits) <- members

  ensemble <- nowcast_ensemble(fits, type = "quantile", verbose = FALSE)

  # scoringutils scores the members and the ensemble side by side
  truth <- nowcast_truth(x_full)
  c(fits, list(ensemble = ensemble)) |>
    lapply(as_scoringutils, truth = truth) |>
    bind_rows(.id = "model")
}
```

The members are the three methods that nowcast on the object's **own**
event-date grid, so their predictions line up with the truth date for date: the
baseline reporting-triangle model, `diseasenowcasting` and `NobBS`. (The counts
for mpox and covid are expanded to a line list first so `NobBS` can read them,
and covid — with its long, heavy reporting tail — is nowcast weekly with delays
capped at eight weeks.) `nowcaster` and `epinowcast` plug in exactly the same
way; `nowcaster` is left out of the *scoring* only because it rebins to
epiweeks, which would not share the daily grid the other members use.

Running all of this takes a few minutes of Bayesian fitting, so the results are
pre-computed and shipped with the article; the block above is the shape of what
produced them. Here we just load them and let \pkg{scoringutils} do the scoring:

```{r comparison-score, eval=has("scoringutils"), message=FALSE, warning=FALSE}
library(scoringutils)

forecasts <- readRDS("ensemble-comparison.rds")

wis <- forecasts |>
  split(~epidemic) |>
  lapply(function(df) {
    df |>
      as.data.frame() |>
      as_forecast_quantile(forecast_unit = c("target_date", "model", "epidemic")) |>
      score() |>
      summarise(mean_wis = mean(wis), .by = c(model, epidemic))
  }) |>
  bind_rows()

wis |>
  mutate(mean_wis = round(mean_wis, 1)) |>
  tidyr::pivot_wider(names_from = epidemic, values_from = mean_wis) |>
  arrange(model)
```

Lower WIS is better. Two things stand out.

**No single model wins everywhere.** `diseasenowcasting` is best on dengue and
mpox but *worst* on covid; `baselinenowcast` is the reverse. If you had to commit
to one model up front, you would have to be right about which epidemic you were
about to face.

**The ensemble is never best, but never worst.** It sits in the upper middle of
every column — on mpox it is within a whisker of the best model, on covid and
dengue it comfortably beats the weaker members. That is the whole point: the
ensemble buys you insurance against picking the wrong model, at the cost of
rarely being the single best.

```{r comparison-plot, eval=has("scoringutils") && has("ggplot2"), message=FALSE, warning=FALSE, fig.width=7, fig.height=3.2}
library(ggplot2)

wis |>
  mutate(
    is_ensemble = grepl("ensemble", model),
    model = reorder(model, mean_wis)
  ) |>
  ggplot(aes(mean_wis, model, fill = is_ensemble)) +
  geom_col() +
  facet_wrap(~epidemic, scales = "free") +
  scale_fill_manual(
    values = c(`TRUE` = "#0072B2", `FALSE` = "grey70"),
    labels = c(`TRUE` = "ensemble", `FALSE` = "single model"),
    name = NULL
  ) +
  labs(x = "Mean WIS (lower is better)", y = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

Where one member is much weaker than the others — `diseasenowcasting` on covid,
`NobBS` on dengue — the equal-weight ensemble is dragged towards it. That is
exactly what the performance weighting of the previous section is for: feeding a
`nowcast_backtest()` into `nowcast_ensemble(weights = "inverse_score")` down-
weights the members that have been missing, and pulls the ensemble back towards
the leaders.

## 5. Adding your own model

This is the part that makes the whole thing worth building. A backend is
**two S3 methods**, and they can live in your own package, your analysis script,
or a one-off chunk like this one. Nothing inside `tbl.now` needs to change.

The dispatch works like this: `run_nowcast(x, method = "mymodel")` builds a
little object of class `c("mymodel", "nowcast_method")` and calls

* `nowcast_fit(method, x, ...)` — run the model, return whatever it returns;
* `nowcast_tidy(method, fit, x, ..., quantile_levels)` — describe the result in
  the tidy format.

### A worked example

Let us register a genuinely simple model: multiply the counts reported so far by
a fixed inflation factor, with lognormal noise around it. It is not a good
nowcast, but it is a complete one.

```{r custom-fit}
nowcast_fit.inflate <- function(method, x, ..., factor = 1.2, sigma = 0.15,
                                quantile_levels, verbose = TRUE) {
  reported <- nowcast_truth(x)

  list(
    dates  = reported[[get_event_date(x)]],
    counts = reported[[".observed"]],
    factor = factor,
    sigma  = sigma
  )
}
```

`nowcast_fit()` returns whatever is convenient — here a plain list, elsewhere a
Stan fit. It is stored verbatim in the result's `fit` property.

The second method turns that into predictions. You may return **draws**
(`tbl.now` derives the quantiles for you), **quantiles**, or both:

```{r custom-tidy}
nowcast_tidy.inflate <- function(method, fit, x, ..., quantile_levels) {
  n_draws <- 1000

  draws <- tidyr::expand_grid(
    .index = seq_along(fit$dates),
    .draw  = seq_len(n_draws)
  ) |>
    dplyr::mutate(
      !!get_event_date(x) := fit$dates[.index],
      .value = fit$counts[.index] * fit$factor *
        stats::rlnorm(dplyr::n(), meanlog = 0, sdlog = fit$sigma)
    ) |>
    dplyr::select(-".index")

  list(predictions = NULL, draws = draws)
}
```

The contract for `nowcast_tidy()` is short:

| element | shape |
|---|---|
| `predictions` | one row per (event date, stratum, quantile level), columns `<event_date>`, the strata, `.quantile_level`, `.value` |
| `draws` | one row per (event date, stratum, draw), columns `<event_date>`, the strata, `.draw`, `.value` |

Either may be `NULL`, but not both. Returning draws is preferable: it is what
`type = "linear_pool"` needs, and the quantiles follow from them automatically.

That is the whole backend. It now behaves like any other method:

```{r custom-use}
inflated <- run_nowcast(dengue, method = "inflate", factor = 1.15, verbose = FALSE)

inflated

as_tibble(inflated)
```

It shows up in the catalogue,

```{r custom-list}
list_nowcast_methods(installed_only = FALSE)
```

it can be scored (here against the same data it was fitted on, which flatters
it enormously — see the backtest below for the honest version),

```{r custom-score}
score_nowcast(inflated) |> tail(3)
```

it can be backtested,

```{r custom-backtest}
custom_backtest <- nowcast_backtest(
  dengue,
  methods   = "inflate",
  now_dates = as.Date(c("2010-09-01", "2010-10-01")),
  verbose   = FALSE
)

custom_backtest$scores |>
  group_by(.now) |>
  summarise(mean_wis = mean(wis, na.rm = TRUE), .groups = "drop")
```

and it can be ensembled with the packaged methods:

```{r custom-ensemble, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
mixed <- nowcast_ensemble(
  baselinenowcast = baseline,
  inflate         = inflated,
  weights         = c(baselinenowcast = 0.7, inflate = 0.3),
  verbose         = FALSE
)

as_tibble(mixed)
```

### Notes for backend authors

A few things worth knowing when the backend lives in a package rather than a
chunk:

* **Register the methods in your `NAMESPACE`** with
  `S3method(nowcast_fit, mymodel)` and `S3method(nowcast_tidy, mymodel)` (or
  `@export` on `nowcast_fit.mymodel` if you use roxygen), and put `tbl.now` in
  `Imports`. Because dispatch is ordinary S3, the methods are found as soon as
  your package is loaded.
* **Use the event-date column name from the object**, `get_event_date(x)`, not a
  hard-coded `"event_date"`. The same goes for `get_strata(x)`,
  `get_case_count(x)` and `get_now(x)`.
* **Honour the strata if you can**, and warn and pool if you cannot. Whatever
  strata columns you put on the output are what the result reports as its
  strata.
* **Convert the data with the existing bridges** where one fits your model's
  input: `tbl_now_to_baselinenowcast()`, `tbl_now_to_epinowcast()`,
  `tbl_now_to_epidist()`, `tbl_now_to_tsibble()`, `tbl_now_to_data_table()`.
* **`quantile_levels` reaches both methods.** Most backends can ignore it at fit
  time; the ones that must be told up front which quantiles to report (as
  `NobBS` is) use it in `nowcast_fit()`.
* **Declare the modelling package in `Suggests`**, and check for it at fit time,
  so that installing your package does not drag in a Stan toolchain.

## See also

* `vignette("nowcasting-models")` — the same five packages, driven by hand.
* `?run_nowcast`, `?nowcast_ensemble`, `?nowcast_backtest`.
* `?nowcast_fit` and `?nowcast_tidy` — the extension contract in reference form.

## References
