---
title: "Fitting models from different nowcasting frameworks, backtesting and building an ensemble"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Fitting models from different nowcasting frameworks, backtesting and building an ensemble}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)

options(tibble.width = Inf, width = 160, pillar.width = 160)
options(pillar.print_max = 5, pillar.print_min = 5, 
        tibble.print_min = 5, dplyr.print_min = 5,
        tibble.print_max = 5, dplyr.print_max = 5)



# Several of the modelling packages route their data through data.table; pinning
# it to a single thread avoids an OpenMP crash seen when knitr forks the build.
if (has("data.table")) data.table::setDTthreads(1)
```


## Why this vignette?

This vignette is about how to run nowcasts from different `R` packages
all within the same `tbl_now` framework as well as on how to backtest
and do ensembles. 

Here we describe how to:

* Use **`engine_*()`** to setup the model and its arguments,
* Fit any model with **`run_nowcast(x, engine)`**. 
* Build backtest with **`nowcast_backtest()`** to evaluate your models.
* Use **`nowcast_ensemble()`** to combines several models into an ensemble model.

::: {.alert .alert-info}
**There are two ways to do the same.** You can either use
`run_nowcast(x, engine_epinowcast())` to use the same call for any
nowcast (this is what this vignette explains). Alternatively you can
use the converters to use the original method from the specific package:

```r
tbl_now_to_epinowcast(x) |> epinowcast::epinowcast() 
```

Ideally you should:

1.  Use the converter (`tbl_now_to_*`) when you want to pass that package's
own arguments, inspect what it was handed, or do something 
the `tbl.now` backend does not.

2. Use `run_nowcast()` when you want several models that can be compared
via `nowcast_backtest()` and `nowcast_ensemble()`.

:::

```{r setup, message=FALSE}
library(dplyr)
library(tbl.now)

data(denguedat)
```

The examples below use `denguedat`:  a **weekly line list** of dengue cases in Puerto Rico:

```{r window}
dengue <- denguedat |>
  tbl_now(
    event_date  = onset_week,  #symptom onset
    report_date = report_week, #when it was reported
    verbose     = FALSE
  )

dengue
```


::: {.alert .alert-info}
This tutorial requires installation of baselinenowcast, NobBS and surveillance. You can do it as:

```{r, eval = FALSE}
install.packages(c("baselinenowcast","surveillance","NobBS"))
```
:::


## 1. Fitting a nowcast

The `run_nowcast()` function takes two arguments: the `tbl_now`, and an `engine()`:

```{r bln-run, eval=has("baselinenowcast"), warning=FALSE, message=FALSE}
#Here we run a baselinenowcast as an example with very few draws
#because its a tutorial
baseline <- dengue |> 
  run_nowcast(engine_baselinenowcast(draws = 100))

baseline
```


An engine includes  *the model and all of its arguments*. For example,
`engine_baselinenowcast()` names `baselinenowcast::baselinenowcast()`'s own arguments. 
For example in the previous case we modified the number of draws which is
an argument from `baselinenowcast::baselinenowcast()`:

```{r engine-print}
engine_baselinenowcast(draws = 1000)
```

You can call `list_nowcast_methods()` to see all the methods available in this package

```{r methods}
list_nowcast_methods()
```

In general each of the methods for the `engine()` requires you to install the corresponding package. You can only use those you have installed. 

### How much history to fit on: `min_date`

In general not all nowcasts can fit all of the same data fast enough to be useful. As a rule of thumb:

1. `diseasenowcasting` can take the **whole series** and produce a nowcast in less than a minute.
2. `baselinenowcast` can take all of the event dates however one usually has to truncate the
number of delays in the delay distribution. 
3. `epinowcast`, `EpiNow2`, `surveillance` and `NobBS` all require an event-window so that for a long time
series not all event dates are passed. You can use the `min_date` argument to select the size of the window
for the engine. 

```{r min-date-demo, eval=FALSE}
run_nowcast(dengue, engine_epinowcast(min_date = 20))
```

::: {.alert .alert-info}
For backtesting and ensembles it is important to utilize the `min_date` in the engine so that the backtest
knows to truncate the data to the last `min_date` periods for every single test. 
:::

## 2. Cleaning and evaluating the nowcast

Every method called by `run_engine` returns a `tbl_nowcast` which can be 
cleaned with `tidy`:


```{r bln-tidy-1}
tidy(baseline)
```

or visualized with `autoplot()`: 

```{r, fig.width=9, fig.height=4.5}
autoplot(baseline)
```

If enough cases have been observed so that the "truth" is settled, one can use the  `score_nowcast()` function which compares the predictive quantiles with a resolved truth table using the weighted interval score (WIS) of @bracher2021, plus the absolute error of the median and the 50% and 90% interval coverage:

```{r score}
score_nowcast(baseline, truth = dengue)
```

or more directly one can use the \pkg{scoringutils} package:

```{r scoringutils, eval=has("scoringutils")}
baseline |>
  scoringutils::as_forecast_quantile(truth = dengue) |>
  scoringutils::score()
```

::: {.alert .alert-warning}
Because the dataset ends at `r `get_now(dengue)` and we nowcasted for that date the
scores are not very useful here (we don't know how many cases _arrived_ eventually). A
better option is to perform a backtest to evaluate historical performance as the dataset knows
how many cases have already settled from dates in the past. 
:::



## 3. Backtests and ensembles

Models fail in different directions and combining them cancels part of that. An
ensemble combines multiple nowcasting for robustness. To build an ensemble
one has to specify different engines, backtest them and then build the ensemble. 

Here for example we will create an ensemble from `NobBS`,  and
two `baselinenowcast` options. It is important to utilize the `label` argument
when using the same engine with different parameters:



```{r}
#NobBS
model1 <- engine_nobbs(min_date = 52, max_D = 15, label = "NobBS")

#baselinenowcast
model2 <- engine_baselinenowcast(max_delay = 15, label = "baselinenowcast")

#a baselinenowcast model with a different specification
model3 <- engine_baselinenowcast(max_delay = 10, prop_delay = 0.75, label = "baselinenowcast 2")
```

Once the models are specified we can backtest through several past dates:

```{r members-out, warning=FALSE}
#We evaluate the performance of our models in 2 past dates
#in real life change n_dates to a bigger number (we use 2 for the tutorial)
models_backtest <- dengue |> 
  nowcast_backtest(model1, model2, model3, n_dates = 2)
```

The backtest now gives us enough information to evaluate each of the models:

```{r, eval=has("scoringutils")}
models_backtest |> 
  scoringutils::as_forecast_quantile(truth = dengue) |>
  scoringutils::score() |> 
  scoringutils::summarise_scores() 
```

Once the historical evaluation is performed one needs to fit each of the models individually and then can put them together into a `nowcast_ensemble`  along its backtest. The ensemble will then average all the models in a way that optimizes the WIS (if `weights = "optim"` or proportional to the WIS if `weights = "inverse_score"`).  

```{r, warning=FALSE, message=FALSE}
#We fit each of the individual models
nowcast1 <- dengue |> run_nowcast(model1)
nowcast2 <- dengue |> run_nowcast(model2)
nowcast3 <- dengue |> run_nowcast(model3)

#And then ensemble to optimize for the best WIS
ensemble_nowcast <-  
  nowcast_ensemble(nowcast1, nowcast2, nowcast3, 
                   weights = "optim", backtest = models_backtest)
```
One can visualize the ensemble with `autoplot()` or get its results with `tidy()` as in the case of a regular nowcast. 

```{r, fig.width=9, fig.height=4.5}
autoplot(ensemble_nowcast, date_lim = c(as.Date("2010/10/01"), as.Date("2010/12/20")))
```

## 4. Adding your own model

You can add any model built by yourself or from any other package to the fitting so that it has its own engine, its own tidy to clean and can be used in conjunction with `tidy()`, `autoplot()` as well as the backtesting and ensemble methods. You can read more about it in [the corresponding article](https://rodrigozepeda.github.io/tbl.now/articles/custom-nowcast-models.html). 

```{r learning-more, echo=FALSE, results="asis"}
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```

## References
