---
title: "One call, many models: run_nowcast(), custom backends and ensembles"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{One call, many models: run_nowcast(), custom backends and ensembles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The model fits (Stan / JAGS) are shown as illustrative code but are NOT run
# when the article is built: everything displayed below is precomputed by
# `data-raw/ensemble_comparison.R` and read from `ensemble-comparison.rds`.
# Flip `run_models` to TRUE, with the modelling packages installed, to fit them.
run_models <- FALSE

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)

# Several of the modelling packages route their data through data.table; pinning
# it to a single thread avoids an OpenMP crash seen when knitr forks the build.
if (has("data.table")) data.table::setDTthreads(1)
```

> This was written automatically by an AI model. A human has yet to review.

::: {.alert .alert-warning}
`r lifecycle::badge("experimental")`
**Everything on this page is experimental.** `run_nowcast()`,
`nowcast_ensemble()`, `nowcast_backtest()`, `nowcast_weights()`,
`score_nowcast()` and the `tbl_nowcast` class are all marked
[experimental](https://lifecycle.r-lib.org/articles/stages.html#experimental):
they work, they are tested, and their **interfaces may still change** — argument
names, defaults and the shape of what comes back. The extension contract
(`nowcast_fit()` / `nowcast_tidy()`) is the part most likely to move, because it
is the youngest.

Pin a version if you are building something on top of this. The stable, settled
half of the package is the `tbl_now` class and the `tbl_now_to_*()` converters
in `vignette("nowcasting-models")` — you can always drive a model by hand
through those, and `tidy()` will normalise the result the same way.
:::

## Why this vignette?

`vignette("nowcasting-models")` shows how one `tbl_now` can be handed to six
different modelling packages. That still leaves you writing six different calls,
reading six different output formats, and doing the bookkeeping yourself if you
want to compare or combine them.

This vignette is about the layer that removes that bookkeeping:

* **`run_nowcast(x, method = ...)`** fits *any* supported package and always
  returns the same kind of object,
* **`nowcast_ensemble()`** combines several of those objects into an ensemble,
* **`nowcast_backtest()`** and **`nowcast_weights()`** decide, from data, how
  much each member should count,
* and **`nowcast_fit()` / `nowcast_tidy()`** let you plug in a model that
  `tbl.now` has never heard of.

::: {.alert .alert-info}
**Two front doors, and both stay open.** `run_nowcast(x, "epinowcast")` and the
hand-written `tbl_now_to_epinowcast(x) |> epinowcast::epinowcast() |> tidy()`
do the same work. Reach for the converter when you want to pass that package's
own arguments, inspect what it was handed, or do something the backend does not
expose. Reach for `run_nowcast()` when you want several models on one footing —
which is the only way `nowcast_ensemble()` and `score_nowcast()` can compare
them at all.
:::

```{r setup, message=FALSE}
library(dplyr)
library(tbl.now)

data(denguedat)
```

The examples below use dengue in Puerto Rico: a **weekly line list**, one row per
case, with an onset week and a report week.

```{r window}
dengue <- denguedat |>
  filter(onset_week >= as.Date("2008-01-01")) |>
  count(onset_week, report_week, name = "n") |>
  tbl_now(
    event_date  = onset_week,
    report_date = report_week,
    case_count  = n,
    data_type   = "count-incidence",
    verbose     = FALSE
  )

# Nowcast as of a date in the past, so that later reports exist to score against
now <- as.Date("2010-10-04")   # a Monday: these weeks start on Mondays

snapshot <- dengue |>
  filter(report_week <= now) |>
  change_now(now = now)

get_now(snapshot)
```

```{r cache-load, echo=FALSE, message=FALSE}
cached <- readRDS("ensemble-comparison.rds")
display <- cached$display

# The article shows code and prints the REAL output of that code, captured by
# the precompute. `replay()` renders it as if the visible chunk above had run.
replay <- function(x) cat(x, sep = "\n")
```

## 1. One call per model

`run_nowcast()` takes the `tbl_now`, the name of a modelling package, and
anything that package accepts:

```{r bln-run, eval=run_models && has("baselinenowcast")}
baseline <- run_nowcast(
  snapshot,
  method  = "baselinenowcast",
  draws   = 1000,
  verbose = FALSE
)

baseline
```

```{r bln-run-out, echo=FALSE, eval=!run_models}
replay(display$baseline_print)
```

The `...` goes straight through to the backend, so switching models mostly means
switching the arguments you already know:

```{r other-methods, eval=FALSE}
# Bayesian structural time series (diseasenowcasting)
dnc <- run_nowcast(snapshot, "diseasenowcasting")

# Bayesian delay + reference modules (epinowcast)
enw <- run_nowcast(snapshot, "epinowcast",
  preprocess_args = list(max_delay = 10),
  fit = epinowcast::enw_fit_opts(chains = 2, iter_sampling = 500)
)

# Nowcasting by Bayesian Smoothing (NobBS); needs JAGS
nobbs <- run_nowcast(snapshot, "NobBS", max_D = 10, moving_window = 64)

# Hoehle & an der Heiden (surveillance)
sur <- run_nowcast(snapshot, "surveillance", D = 10)

# EpiNow2's estimate_infections()
en2 <- run_nowcast(snapshot, "EpiNow2")
```

Each of those routes through the same `tbl_now_to_*()` converter documented in
`vignette("nowcasting-models")` — so `"NobBS"` expands your counts to the one row
per case that `NobBS()` counts, and `"surveillance"` hands over a grid that
reaches the `now` even when the newest days carry no reports yet.

`list_nowcast_methods()` tells you what is available in your session:

```{r methods}
list_nowcast_methods()
```

### What you get back

Every method returns a `tbl_nowcast`. Whatever the package produced — a Stan
fit, a matrix of samples, a data frame of quantiles — the predictions are always
in the same tidy shape:

```{r bln-quantiles-1, eval=FALSE}
as_tibble(baseline)
```

```{r bln-quantiles, echo=FALSE}
display$baseline_quantiles
```

and, when the backend is sample-based, the draws are there too:

```{r bln-draws-1, eval=FALSE}
as_tibble(baseline, type = "draws")
```

```{r bln-draws, echo=FALSE}
display$baseline_draws
```

### `tidy()` works here too

`as_tibble()` gives you the quantiles in full. `tidy()` gives you the **same
summary table every other engine in this package produces** — the one documented
at `?tidy.nowcast` — so a nowcast fitted through `run_nowcast()` and one fitted
by calling a package by hand are read the same way:

```{r bln-tidy-1, eval=FALSE}
tidy(baseline)
```

```{r bln-tidy, echo=FALSE}
display$baseline_tidy
```

Those first weeks look odd until you notice what they are: `estimate`,
`conf.low` and `conf.high` are identical because early 2008 was **settled** long
before this `now`. Every case had been reported, so there is nothing left to
nowcast and no uncertainty to report. The interesting rows are at the `now` end.

Two columns there are read off the object rather than assumed. `engine` is the
method that produced it. `level` is the width of the **widest symmetric pair of
quantiles the nowcast actually carries** — `0.95` for the default
`nowcast_quantile_levels()`, `0.8` for a nowcast summarised at
`c(0.1, 0.5, 0.9)`, and `NA` when no symmetric pair exists at all. A guessed
width would defeat the point of the column, which is to stop a 90% band being
compared with a 95% one as though they were the same thing.

Nothing is thrown away. The `fit` property still holds the backend's own object,
so you can keep using that package's diagnostics:

```{r bln-fit, eval=FALSE}
class(baseline@fit)
#> [1] "baselinenowcast_df" "data.frame"
```

And there is an `autoplot()` method for a quick look. It is drawn in **green**,
the colour this package reserves for the epidemic process — a nowcast is an
estimate of *what happened*, not of *when we found out*:

```{r bln-plot-1, eval=FALSE}
autoplot(baseline)
```

```{r bln-plot, echo=FALSE, eval=has("ggplot2"), message=FALSE, warning=FALSE, fig.width=7, fig.height=4, fig.alt="A green fan chart of the dengue nowcast: nested prediction intervals around a median line, with the counts reported so far as points."}
library(ggplot2)

# Redrawn from the cached predictions, so this is a picture of the fit the code
# above produces rather than of a different one.
autoplot(tbl_nowcast(
  predictions = display$ensemble_predictions,
  method      = "baselinenowcast",
  event_date  = display$event_date,
  now         = display$now
)) +
  coord_cartesian(xlim = c(display$now - 180, display$now))
```

### Strata are handled for you

When the `tbl_now` declares strata, `run_nowcast()` does whatever that backend
needs to honour them — one reporting triangle per stratum for `baselinenowcast`,
`NobBS.strat()` instead of `NobBS()`, one fit per stratum for `surveillance`,
`regional_epinow()` instead of `estimate_infections()` — and the stratum ends up
as an ordinary column of the output:

```{r strata, eval=run_models && has("baselinenowcast")}
dengue_by_sex <- denguedat |>
  filter(onset_week >= as.Date("2008-01-01")) |>
  count(onset_week, report_week, gender, name = "n") |>
  tbl_now(
    event_date = onset_week, report_date = report_week, case_count = n,
    strata = gender, data_type = "count-incidence", verbose = FALSE
  ) |>
  filter(report_week <= now) |>
  change_now(now = now)

by_sex <- run_nowcast(dengue_by_sex, "baselinenowcast", draws = 1000, verbose = FALSE)

as_tibble(by_sex)
```

```{r strata-out, echo=FALSE, eval=!run_models}
display$stratified_quantiles
```

`tidy()` labels each block with its stratum, so `(stratum, event_date)` is a
unique key and the two series can never be confused for one another:

```{r strata-tidy-1, eval=FALSE}
tidy(by_sex)
```

```{r strata-tidy, echo=FALSE}
display$stratified_tidy
```

If a backend genuinely cannot stratify, it warns and pools rather than
pretending.

## 2. Scoring a nowcast

`score_nowcast()` compares the predictive quantiles with what was eventually
observed, using the weighted interval score (WIS) of @bracher2021, plus the
absolute error of the median and the 50% and 90% interval coverage.

Scoring only makes sense against data the model had not seen, which is why
`snapshot` was truncated at `now` and the truth comes from the full series:

```{r score, eval=FALSE}
score_nowcast(baseline, truth = dengue)
```

```{r score-out, echo=FALSE}
score_nowcast(
  tbl_nowcast(
    predictions = display$ensemble_predictions,
    method = "baselinenowcast", event_date = display$event_date
  ),
  truth = display$truth, observed_col = ".observed"
) |>
  filter(.data$onset_week > display$now - 56)
```

Lower WIS is better; `coverage_90` should be `TRUE` about nine times in ten if
the intervals are honest.

If you would rather use the full \pkg{scoringutils} score suite,
`as_scoringutils()` reshapes the same object into the format it expects:

```{r scoringutils, eval=FALSE}
baseline |>
  as_scoringutils(truth = dengue) |>
  as.data.frame() |>
  scoringutils::as_forecast_quantile() |>
  scoringutils::score()
```

`tbl.now`'s own `wis` and the one \pkg{scoringutils} computes agree to machine
precision — the package's test suite checks exactly that, on the same numbers,
rather than trusting either implementation on its own.

## 3. Ensembles

Models fail in different directions, and combining them cancels part of that: an
ensemble is rarely the single best model, but it is also rarely the worst, which
is worth a great deal when you have to commit to something before you know which
epidemic you are facing. *How* you combine them matters more than most write-ups
admit, and section 4 measures it rather than asserting it. Because every
`run_nowcast()` result has the same shape, combining them is one call.

```{r members, eval=FALSE}
members <- list(
  baselinenowcast   = run_nowcast(snapshot, "baselinenowcast", draws = 1000),
  diseasenowcasting = run_nowcast(snapshot, "diseasenowcasting"),
  NobBS             = run_nowcast(snapshot, "NobBS", max_D = 10, moving_window = 64)
)

ensemble <- nowcast_ensemble(members)
```

```{r ensemble-out, echo=FALSE, eval=!run_models}
replay(display$ensemble_print)
```

An ensemble is a `tbl_nowcast` like any other, so it scores, plots and tidies
the same way. Its `engine` is the ensemble's name rather than a package:

```{r ensemble-tidy-1, eval=FALSE}
tidy(ensemble)
```

```{r ensemble-tidy, echo=FALSE}
display$ensemble_tidy
```

### Two ways of combining

`type = "quantile"` (the default) averages the members' quantiles level by
level — *vincentization*. It is the workhorse of the forecast hubs, it always
applies because every backend produces quantiles, and it tends to produce
**narrower** intervals than the members.

`type = "linear_pool"` instead pools the members' draws into a mixture
distribution and re-summarises it. It needs draws from every member and
generally produces **wider** intervals, because disagreement between models
becomes extra spread rather than being averaged away.

```{r pool, eval=FALSE}
sharp <- nowcast_ensemble(members, type = "quantile")
wide  <- nowcast_ensemble(members, type = "linear_pool", n_draws = 4000)
```

Which one you want depends on whether you read between-model disagreement as
noise to be averaged out (quantile) or as genuine uncertainty to be propagated
(linear pool).

::: {.alert .alert-warning}
**The linear pool needs draws, and not every backend has them.** `NobBS` reports
quantiles only — its `nowcast.post.samps` cover the `now` date rather than every
event date — so `type = "linear_pool"` refuses a set of members that includes
it, rather than silently dropping it and returning a differently-composed
ensemble under the same name.
:::

### Weights

By default every member counts the same. You can also give weights directly:

```{r weights-manual, eval=FALSE}
nowcast_ensemble(members, weights = c(baselinenowcast = 0.5,
                                      diseasenowcasting = 0.3, NobBS = 0.2))
```

or *learn* them from how the members actually performed. `nowcast_backtest()`
walks back through time: at each retrospective date it truncates the data to the
reports available then, refits every method, and scores the result against what
was eventually observed.

```{r backtest, eval=FALSE}
backtest <- nowcast_backtest(
  dengue,
  methods   = c("baselinenowcast", "diseasenowcasting", "NobBS"),
  now_dates = now - 7 * 8 * (3:1),
  seed      = 20260824
)

tidy(backtest)
```

```{r backtest-out, echo=FALSE}
cached$backtests |>
  filter(.data$epidemic == "dengue") |>
  select(-"epidemic") |>
  head()
```

`tidy()` on a backtest gives one row per (method, `now` date, target), which goes
straight into `dplyr` or `ggplot2`:

```{r backtest-summary, echo=FALSE}
cached$backtests |>
  filter(.data$epidemic == "dengue") |>
  summarise(
    mean_wis = round(mean(wis, na.rm = TRUE), 1),
    coverage_90 = round(mean(coverage_90, na.rm = TRUE), 2),
    .by = c(method, now)
  ) |>
  arrange(now, mean_wis)
```

This is the expensive part of the workflow — it is `length(methods) ×
length(now_dates)` model fits — so keep `now_dates` short when the members are
Bayesian. Pass `seed` so the backtest is reproducible: one `set.seed()` before
the whole thing only pins anything if every method draws the same random numbers
in the same order, which stops being true the moment you drop a method.

From the backtest, `nowcast_weights()` offers two rules:

```{r weights-fitted, eval=FALSE}
nowcast_weights(backtest, type = "inverse_score") # w proportional to 1 / mean WIS
nowcast_weights(backtest, type = "optim")         # w minimising the ensemble's WIS
```

```{r weights-out, echo=FALSE}
round(cached$weights$dengue, 3)
```

`"inverse_score"` is the safe default: it is monotone in performance, never
collapses onto a single model, and cannot overfit. `"optim"` searches the simplex
for the weights that would have minimised the ensemble's WIS over the training
window; it is better in principle, but with only a handful of retrospective dates
it happily overfits, so prefer it when the training window is long.

Either rule plugs straight into the ensemble:

```{r ensemble-weighted, eval=FALSE}
nowcast_ensemble(members, weights = "inverse_score", backtest = backtest)
```

## 4. Does the ensemble help? Three epidemics

Talk of robustness is cheap; let us measure it. We take three very different
outbreaks that ship with `tbl.now` — **dengue** (weekly, Puerto Rico), **mpox**
(daily, US) and **covid** (daily, Colombia) — and on each one nowcast at a fixed
`now`, then score every model, and every ensemble, against the counts that were
*eventually* reported.

The recipe is the same for all three:

```{r comparison-recipe, eval=FALSE}
# For one epidemic:
#   1. snapshot the data to the reports available at `now`
#   2. fit each member on that snapshot
#   3. build the ensembles
#   4. score every model AND every ensemble against the eventual truth

evaluate <- function(x_full, now, members, method_args = list()) {
  snapshot <- x_full |>
    filter(.data[[get_report_date(x_full)]] <= now) |>
    change_now(now = now)

  fits <- lapply(members, function(m) {
    do.call(run_nowcast, c(list(snapshot, m, verbose = FALSE), method_args[[m]]))
  })
  names(fits) <- members

  everything <- c(fits, list(
    `ensemble (quantile)` = nowcast_ensemble(fits, type = "quantile")
  ))

  everything |>
    lapply(as_scoringutils, truth = x_full) |>
    bind_rows(.id = "model")
}
```

The members are the three methods that nowcast on the object's **own** event-date
grid, so their predictions line up with the truth date for date:
`baselinenowcast`, `diseasenowcasting` and `NobBS`. The full script is
`data-raw/ensemble_comparison.R`; it takes minutes of Bayesian fitting, so the
results are precomputed and shipped with the article and read back here.

```{r comparison-score, eval=has("scoringutils"), message=FALSE, warning=FALSE}
library(scoringutils)

wis <- cached$forecasts |>
  split(~epidemic) |>
  lapply(function(df) {
    df |>
      as.data.frame() |>
      as_forecast_quantile(
        forecast_unit = c("target_date", "model", "epidemic")
      ) |>
      score() |>
      summarise(mean_wis = mean(wis), .by = c(model, epidemic))
  }) |>
  bind_rows()

wis |>
  mutate(mean_wis = round(mean_wis, 1)) |>
  tidyr::pivot_wider(names_from = epidemic, values_from = mean_wis) |>
  arrange(model)
```

Lower WIS is better.

```{r comparison-plot, eval=has("scoringutils") && has("ggplot2"), message=FALSE, warning=FALSE, fig.width=8, fig.height=3.2, fig.alt="Three panels, one per epidemic, with a horizontal bar per model showing its mean weighted interval score. Ensembles are drawn in green and single models in grey."}
# The tbl.now palette: green is the epidemic process, which is what every one of
# these models is estimating. Single models are greyed so the ensembles read as
# the subject of the figure.
wis |>
  mutate(
    is_ensemble = grepl("ensemble", model),
    model = reorder(model, mean_wis)
  ) |>
  ggplot(aes(mean_wis, model, fill = is_ensemble)) +
  geom_col() +
  facet_wrap(~epidemic, scales = "free") +
  scale_fill_manual(
    values = c(`TRUE` = "#5F7E62", `FALSE` = "#c9cec9"),
    labels = c(`TRUE` = "ensemble", `FALSE` = "single model"),
    name = NULL
  ) +
  labs(x = "Mean WIS (lower is better)", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", colour = "#334335")
  )
```

### What that looks like day by day

The table says the ensemble is mid-pack. This is what "mid-pack" actually means:

```{r comparison-lines, eval=has("ggplot2"), message=FALSE, warning=FALSE, fig.width=9, fig.height=3.4, fig.alt="Three panels, one per epidemic. Thin coloured lines show each member model's median nowcast, a thick green line the ensemble, and a black dashed line what those dates eventually reached. The members disagree and the ensemble runs between them."}
members <- cached$members

medians <- cached$forecasts |>
  filter(quantile_level == 0.5, model %in% c(members, "ensemble (quantile)"))

truth_line <- cached$forecasts |>
  distinct(epidemic, target_date, observed)

# Okabe-Ito for the members: they need to be told APART, which is a categorical
# job. The ensemble gets the tbl.now palette's green, the colour this package
# reserves for the epidemic process.
member_colours <- c(
  baselinenowcast   = "#D55E00",
  diseasenowcasting = "#0072B2",
  NobBS             = "#CC79A7"
)

ggplot(mapping = aes(target_date)) +
  geom_line(
    data = filter(medians, model %in% members),
    aes(y = predicted, colour = model), linewidth = 0.5, alpha = 0.9
  ) +
  # The ensemble in the palette's green: it is an estimate of the epidemic
  # process, like every other line here, but it is the subject of the figure.
  geom_line(
    data = filter(medians, model == "ensemble (quantile)"),
    aes(y = predicted, colour = "ensemble (quantile)"), linewidth = 1.2
  ) +
  geom_line(
    data = truth_line, aes(y = observed, linetype = "eventual truth"),
    colour = "#262626", linewidth = 0.6
  ) +
  facet_wrap(~epidemic, scales = "free", nrow = 1) +
  scale_colour_manual(
    name = NULL,
    values = c(member_colours, `ensemble (quantile)` = "#5F7E62"),
    breaks = c(members, "ensemble (quantile)")
  ) +
  scale_linetype_manual(name = NULL, values = c(`eventual truth` = "dashed")) +
  labs(x = NULL, y = "Cases") +
  theme_minimal(base_size = 9) +
  theme(
    legend.position = "top", legend.box = "horizontal",
    panel.grid.minor = element_blank()
  ) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE))
```

The members disagree — on dengue by a wide margin — and the ensemble runs
between them. That is the whole mechanism: it cannot be the best line on the
page, because it is an average of lines, but it also cannot be the worst.

### Is it calibrated?

A sharp ensemble with dishonest intervals is worse than no ensemble, so WIS
alone is not enough. The 90% intervals should contain the truth about nine times
in ten:

```{r comparison-coverage, eval=has("scoringutils"), message=FALSE, warning=FALSE}
coverage <- cached$forecasts |>
  split(~epidemic) |>
  lapply(function(df) {
    df |>
      as.data.frame() |>
      as_forecast_quantile(
        forecast_unit = c("target_date", "model", "epidemic")
      ) |>
      score() |>
      summarise(cov90 = mean(interval_coverage_90), .by = c(model, epidemic))
  }) |>
  bind_rows()

coverage |>
  mutate(cov90 = round(cov90, 2)) |>
  tidyr::pivot_wider(names_from = epidemic, values_from = cov90) |>
  arrange(model)
```

The quantile ensemble **under**-covers on all three; the linear pool
**over**-covers. That is not a coincidence, and it is worth seeing why, because
it is arithmetic rather than luck. Here is the width of the same 90% interval:

```{r comparison-width, eval=has("scoringutils"), message=FALSE, warning=FALSE}
cached$forecasts |>
  filter(quantile_level %in% c(0.05, 0.95)) |>
  tidyr::pivot_wider(
    names_from = quantile_level, values_from = predicted, names_prefix = "q"
  ) |>
  summarise(width = round(mean(q0.95 - q0.05), 1), .by = c(model, epidemic)) |>
  tidyr::pivot_wider(names_from = epidemic, values_from = width) |>
  arrange(model)
```

Averaging quantiles averages *widths*: the ensemble's 90% interval at a given
date is the mean of the members' 90% intervals, exactly. So it can never be wider
than its widest member, and a single overconfident member pulls the whole
ensemble in — on dengue the mean member width is 103, against `baselinenowcast`'s
own 185.

Pooling draws does the opposite. A mixture is at least as dispersed as the
average of its components, plus whatever spread comes from the components
disagreeing about where the centre is — so the pool comes out wider than that
mean every time, and wider than *any* member on covid and mpox.

Neither is free. Too wide costs you sharpness; too narrow tells you the answer is
more certain than it is. For a nowcast feeding a decision that asymmetry usually
favours the linear pool — and it needs draws from every member, so plan the
members accordingly.

### Does the ensemble beat its best member?

```{r comparison-best, eval=has("scoringutils")}
wis |>
  mutate(kind = ifelse(grepl("ensemble", model), "ensemble", "single model")) |>
  summarise(best = min(mean_wis), .by = c(epidemic, kind)) |>
  tidyr::pivot_wider(names_from = kind, values_from = best) |>
  mutate(
    ratio = round(ensemble / `single model`, 2),
    verdict = ifelse(ratio < 1, "ensemble wins", "best single model wins")
  )
```

### Does performance weighting beat equal weighting?

```{r comparison-weighted, eval=has("scoringutils")}
wis |>
  filter(model %in% c("ensemble (quantile)", "ensemble (weighted)")) |>
  tidyr::pivot_wider(names_from = model, values_from = mean_wis) |>
  mutate(
    ratio = round(`ensemble (weighted)` / `ensemble (quantile)`, 3),
    verdict = ifelse(ratio < 1, "weighting helps", "weighting does not help")
  )
```

The two tables above are computed from the cache rather than asserted in prose,
so they say whatever the data say. Read them together with the caveat that this
is **one `now` per epidemic**: a single retrospective date is an anecdote, not an
evaluation. What it can show is the shape of the answer — whether an ensemble is
in the running at all, and whether learning weights from three backtest dates
moves it.

The general finding in the forecasting literature [@bracher2021] is the one worth
carrying away: an ensemble is rarely the single best model on any one dataset,
but it is also rarely the worst, and *which* model is best changes between
outbreaks. If you have to commit to one model before you know what epidemic you
are facing, that insurance is usually worth more than the extra sharpness of a
model that might turn out to be the wrong one.

## 5. Adding your own model

This is the part that makes the whole thing worth building. A backend is
**two S3 methods**, and they can live in your own package, your analysis script,
or a one-off chunk like this one. Nothing inside `tbl.now` needs to change.

The dispatch works like this: `run_nowcast(x, method = "mymodel")` builds a
little object of class `c("mymodel", "nowcast_method")` and calls

* `nowcast_fit(method, x, ...)` — run the model, return whatever it returns;
* `nowcast_tidy(method, fit, x, ..., quantile_levels)` — describe the result in
  the tidy format.

### A worked example

Let us register a genuinely simple model: multiply the counts reported so far by
a fixed inflation factor, with lognormal noise around it. It is not a good
nowcast, but it is a complete one.

```{r custom-fit}
nowcast_fit.inflate <- function(method, x, ..., factor = 1.2, sigma = 0.15,
                                quantile_levels, verbose = TRUE) {
  reported <- get_latest_reported_cases(x)
  count_col <- get_case_count(reported)

  list(
    dates  = reported[[get_event_date(x)]],
    counts = reported[[count_col]],
    factor = factor,
    sigma  = sigma
  )
}
```

`nowcast_fit()` returns whatever is convenient — here a plain list, elsewhere a
Stan fit. It is stored verbatim in the result's `fit` property.

The second method turns that into predictions. You may return **draws**
(`tbl.now` derives the quantiles for you), **quantiles**, or both:

```{r custom-tidy}
nowcast_tidy.inflate <- function(method, fit, x, ..., quantile_levels) {
  n_draws <- 1000

  draws <- tidyr::expand_grid(
    .index = seq_along(fit$dates),
    .draw  = seq_len(n_draws)
  ) |>
    dplyr::mutate(
      !!get_event_date(x) := fit$dates[.index],
      .value = fit$counts[.index] * fit$factor *
        stats::rlnorm(dplyr::n(), meanlog = 0, sdlog = fit$sigma)
    ) |>
    dplyr::select(-".index")

  list(predictions = NULL, draws = draws)
}
```

The contract for `nowcast_tidy()` is short:

| element | shape |
|---|---|
| `predictions` | one row per (event date, stratum, quantile level), columns `<event_date>`, the strata, `.quantile_level`, `.value` |
| `draws` | one row per (event date, stratum, draw), columns `<event_date>`, the strata, `.draw`, `.value` |

Either may be `NULL`, but not both. Returning draws is preferable: it is what
`type = "linear_pool"` needs, and the quantiles follow from them automatically.

That is the whole backend. It now behaves like any other method — no modelling
package required, so this chunk really does run:

```{r custom-use}
set.seed(20260824)
inflated <- run_nowcast(snapshot, method = "inflate", factor = 1.15, verbose = FALSE)

inflated
```

It tidies like everything else,

```{r custom-tidy-out}
tidy(inflated) |> tail(3)
```

it shows up in the catalogue,

```{r custom-list}
list_nowcast_methods(installed_only = FALSE)
```

it can be scored (here against the same data it was fitted on, which flatters it
enormously — the backtest below is the honest version),

```{r custom-score}
score_nowcast(inflated) |> tail(3)
```

and it can be backtested:

```{r custom-backtest}
custom_backtest <- nowcast_backtest(
  dengue,
  methods   = "inflate",
  now_dates = as.Date(c("2010-08-01", "2010-09-01")),
  seed      = 20260824,
  verbose   = FALSE
)

tidy(custom_backtest) |>
  summarise(mean_wis = round(mean(wis, na.rm = TRUE), 1), .by = now)
```

It also ensembles with the packaged methods, exactly like one of them:

```{r custom-ensemble, eval=FALSE}
mixed <- nowcast_ensemble(
  baselinenowcast = baseline,
  inflate         = inflated,
  weights         = c(baselinenowcast = 0.7, inflate = 0.3)
)
```

### Notes for backend authors

A few things worth knowing when the backend lives in a package rather than a
chunk:

* **Register the methods in your `NAMESPACE`** with
  `S3method(nowcast_fit, mymodel)` and `S3method(nowcast_tidy, mymodel)` (or
  `@export` on `nowcast_fit.mymodel` if you use roxygen), and put `tbl.now` in
  `Imports`. Because dispatch is ordinary S3, the methods are found as soon as
  your package is loaded.
* **Use the event-date column name from the object**, `get_event_date(x)`, not a
  hard-coded `"event_date"`. The same goes for `get_strata(x)`,
  `get_case_count(x)` and `get_now(x)`.
* **Convert the data with the existing bridges** rather than by hand:
  `tbl_now_to_baselinenowcast()`, `tbl_now_to_epinowcast()`,
  `tbl_now_to_nobbs()`, `tbl_now_to_surveillance()`, `tbl_now_to_EpiNow2()`,
  `tbl_now_to_epidist()`, `tbl_now_to_tsibble()`, `tbl_now_to_data_table()`.
  They handle the three data types, the time units and the `now` edge; an inline
  version handles whichever one you were thinking about that day.
* **Reuse `tidy()` if your package already has a method for its own fits.** The
  built-in `surveillance` and `EpiNow2` backends do exactly that: `tidy()`
  already knows how to find their interval *and how wide it is*, and re-deriving
  that in a second place is how two real bugs got in.
* **Honour the strata if you can**, and warn and pool if you cannot. Whatever
  strata columns you put on the output are what the result reports as its strata.
* **`quantile_levels` reaches both methods.** Most backends can ignore it at fit
  time; the ones that must be told up front which quantiles to report (as `NobBS`
  is) use it in `nowcast_fit()`.
* **Declare the modelling package in `Suggests`**, and check for it at fit time,
  so that installing your package does not drag in a Stan toolchain.

## See also

* `vignette("nowcasting-models")` — the same packages, driven by hand through
  the converters.
* `?run_nowcast`, `?nowcast_ensemble`, `?nowcast_backtest`.
* `?tidy.tbl_nowcast` — the table every nowcast tidies into.
* `?nowcast_fit` and `?nowcast_tidy` — the extension contract in reference form.

## References
