---
title: "One call, many models: run_nowcast(), custom backends and ensembles"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{One call, many models: run_nowcast(), custom backends and ensembles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The model fits (Stan / JAGS) are shown as illustrative code but are NOT run
# when the article is built: everything displayed below is precomputed by
# `data-raw/ensemble_comparison.R` and read from `ensemble-comparison.rds`.
# Flip `run_models` to TRUE, with the modelling packages installed, to fit them.
run_models <- FALSE

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)

# Several of the modelling packages route their data through data.table; pinning
# it to a single thread avoids an OpenMP crash seen when knitr forks the build.
if (has("data.table")) data.table::setDTthreads(1)
```


::: {.alert .alert-warning}
`r lifecycle::badge("experimental")`
**Everything on this page is experimental.** `run_nowcast()`,
`nowcast_ensemble()`, `nowcast_backtest()`, `nowcast_weights()`,
`score_nowcast()` and the `tbl_nowcast` class are all marked
[experimental](https://lifecycle.r-lib.org/articles/stages.html#experimental):
they work, they are tested, and their **interfaces may still change** — argument
names, defaults and the shape of what comes back. The extension 
(`nowcast_fit()` / `nowcast_tidy()`) is the part most likely to move, because it
is the most recent.
:::

## Why this vignette?

This vignette is about how to run nowcasts from different `R` packages
all within the same `tbl_now` framework as well as on how to backtest
and do ensembles. 

Here we describe how:

* **`engine_*()`** says which model to fit and with what arguments,
* **`run_nowcast(x, engine)`** fits *any* supported package and always
  returns the same kind of object,
* **`nowcast_ensemble()`** combines several of those objects into an ensemble,
* **`nowcast_backtest()`** and **`nowcast_weights()`** decide, from data, how
  much each member should count,
* and **`nowcast_fit()` / `nowcast_tidy()`** let you add a nowcasting 
  framework that has not been built into `tbl.now`.

::: {.alert .alert-info}
**Two ways to do the same.** You can either use
`run_nowcast(x, engine_epinowcast())` as explained in this vignette or step by
step:
```
tbl_now_to_epinowcast(x) |> epinowcast::epinowcast() |> tidy()
```
Ideally you should:

1.  Use the converter when you want to pass that package's own arguments, inspect 
what it was handed, or do something the `tbl.now` backend does not.

2. Use `run_nowcast()` when you want several models that can be compared
via `nowcast_ensemble()` and `score_nowcast()`.

:::

```{r setup, message=FALSE}
library(dplyr)
library(tbl.now)

data(denguedat)
```

The examples below use dengue in Puerto Rico: a **weekly line list**: one row per
case, with an onset and a report week.

```{r window}
dengue <- denguedat |>
  tbl_now(
    event_date  = onset_week,
    report_date = report_week,
    verbose     = FALSE
  )

# Nowcast as of a date in the past, so that later reports exist to score against
now <- as.Date("2010-10-04")   # a Monday: these weeks start on Mondays

snapshot <- dengue |>
  filter(report_week <= now) |>
  change_now(now = now)

get_now(snapshot)
```

```{r cache-load, echo=FALSE, message=FALSE, purl=FALSE}
cached <- readRDS("ensemble-comparison.rds")
display <- cached$display
if (is.null(cached$article_weights) && !is.null(cached$weights)) {
  cached$article_weights <- cached$weights
}
if (is.null(display$member_predictions) && !is.null(display$baseline_predictions)) {
  display$member_predictions <- display$baseline_predictions |>
    dplyr::mutate(model = "baselinenowcast", .before = 1)
}

# The article shows code and prints the REAL output of that code, captured by
# the precompute. `replay()` renders it as if the visible chunk above had run.
replay <- function(x) cat(x, sep = "\n")
```

## 1. One call per model

`run_nowcast()` takes two things: the `tbl_now`, and an **engine**.

```{r bln-run, eval=run_models && has("baselinenowcast"), warning=FALSE}
baseline <- run_nowcast(
  snapshot,
  engine_baselinenowcast(draws = 1000),
  verbose = FALSE
)

baseline
```

```{r bln-run-out, echo=FALSE, eval=!run_models, purl=FALSE}
replay(display$baseline_print)
```

An engine is *the model and everything it needs*: `engine_baselinenowcast()`
names `baselinenowcast`'s own arguments, and there is one such constructor per
supported package. The data and `verbose` are the only things that sit outside
it.

That is not decoration. Before engines, arguments travelled in a `...` on
`run_nowcast()` and in a `method_args` list of lists on `nowcast_backtest()`,
and both failed the same silent way: an argument that missed its backend simply
vanished, and you got a fitted model at its default with nothing to say so. A
named formal turns that into an error at the call, where you can see it.

```{r engine-print}
engine_baselinenowcast(draws = 1000)
```

`list_nowcast_methods()` tells you what is available in your session:

```{r methods}
list_nowcast_methods()
```

### One card per engine

Each of these routes through the `tbl_now_to_*()` converter documented in
`vignette("nowcasting-models")`, which is where you go to learn a package's own
API. What follows is the other half: the call through `run_nowcast()`, and the
**one thing about that engine most likely to catch you out**.

#### baselinenowcast

Fast, assumption-light, and the only engine here that needs no external
toolchain. It works from a reporting triangle, so the delay axis is a modelling
choice you make.

```{r card-bln, eval=FALSE}
baseline <- run_nowcast(snapshot, engine_baselinenowcast(draws = 1000))  # fitted above
```

**Watch for:** `max_delay` caps the triangle's *width*. Left unset it is inferred
from the longest delay present, so a single straggler can give the triangle
hundreds of near-empty columns and turn a fast fit into a slow one. It keeps
draws, so it can join a `type = "linear_pool"` ensemble.

#### diseasenowcasting

The one engine that takes the `tbl_now` **directly**: it reads the strata and the
temporal-effect columns off the object, so there is nothing to pass.

```{r card-dnc, eval=run_models && has("diseasenowcasting")}
dnc <- run_nowcast(snapshot, engine_diseasenowcasting())

dnc
```

```{r card-dnc-out, echo=FALSE, eval=!run_models, purl=FALSE}
replay(display$card_dnc)
```

**Watch for:** it is also the one engine whose *model* you can swap without
changing packages — see the ensemble below, which uses two of them. On
`count-cumulative` data that revises downwards, pass a
`model(confirmation = confirmation_process())`, or the de-accumulated negatives
have nowhere to go.

#### epinowcast

A flexible Bayesian model with separate modules for the reporting delay and the
reference process. Preprocessing arguments go through `preprocess_args`;
everything else goes to `epinowcast()` itself.

```{r card-enw, eval=run_models && has("epinowcast")}
enw <- run_nowcast(snapshot, engine_epinowcast(
  preprocess_args = list(max_delay = 10),
  fit = epinowcast::enw_fit_opts(
    sampler = epinowcast::enw_pathfinder, draws = 1000, seed = 20260824
  ),
  # The slowest engine here, and it scales with the number of REFERENCE dates,
  # so it gets a window where the others get the whole series. See below.
  min_date = 96
))

enw
```

```{r card-enw-out, echo=FALSE, eval=!run_models, purl=FALSE}
replay(display$card_enw)
```

**Watch for:** it is **unseeded** unless you say otherwise. `enw_fit_opts()`
passes `...` to the sampler, so `seed =` reaches Stan — without it, the same fit
can take forty minutes on one run and six hours on the next, and neither
reproduces. It handles a weekly object natively (`timestep = "week"`), so its
reference dates line up with the object's grid.

#### NobBS

Nowcasting by Bayesian Smoothing. Needs **JAGS** installed as a separate program.

```{r card-nobbs, eval=run_models && has("NobBS")}
nobbs <- run_nowcast(snapshot, engine_nobbs(max_D = 10, moving_window = 64))

nobbs
```

```{r card-nobbs-out, echo=FALSE, eval=!run_models, purl=FALSE}
replay(display$card_nobbs)
```

**Watch for:** it keeps **no draws per event date**, so it cannot join a
`type = "linear_pool"` ensemble, and it can only report quantiles it was asked
for *at fit time*: pass `specs = list(quantiles = ...)` and `tidy(fit, probs =)`
will return them, but a level it never computed is an error rather than an
approximation. It also counts **rows**, so the converter expands your counts to
one row per case — trim before fitting on a long series.

#### surveillance

The classic Höhle & an der Heiden nowcast. No external toolchain for the method
used here.

```{r card-sur, eval=run_models && has("surveillance")}
sur <- run_nowcast(snapshot, engine_surveillance(D = 10))

sur
```

```{r card-sur-out, echo=FALSE, eval=!run_models, purl=FALSE}
replay(display$card_sur)
```

**Watch for:** `surveillance::nowcast()` has **no strata argument at all**, so
`run_nowcast()` fits one model per stratum and labels the blocks for you. Both
date grids it needs come from the object via `get_surveillance_when()` and
`get_surveillance_range()` — the second matters because a line list cannot
express a zero, so the quiet days at the `now` edge would otherwise fall off the
grid entirely.

#### EpiNow2

A renewal-equation model of the *infection* process, with the reporting
correction supplied separately.

```{r card-en2, eval=run_models && has("EpiNow2")}
en2 <- run_nowcast(snapshot, engine_epinow2(
  generation_time = EpiNow2::gt_opts(EpiNow2::example_generation_time),
  delays          = EpiNow2::delay_opts(EpiNow2::example_incubation_period),
  stan            = EpiNow2::stan_opts(
    method = "pathfinder", backend = "cmdstanr", samples = 500
  ),
  min_date        = 96
))

en2
```

```{r card-en2-out, echo=FALSE, eval=!run_models, purl=FALSE}
replay(display$card_en2)
```

**Watch for:** EpiNow2 has no `timestep` — it always models a *daily* process. The
converter therefore lays a weekly object onto EpiNow2's daily grid, putting each
week's count on the week-ending day and marking the rest `accumulate = TRUE`.
EpiNow2 honours that on the way out as well as in the likelihood, so predictions
come back on **your** grid at **your** scale; you do not have to undo anything.
It is also the one engine here that *forecasts* — it returns one period past the
end of the data — and `run_nowcast()` drops that, because a nowcast estimates
what has already happened.

::: {.alert .alert-info}
**Its priors are not fitted from your data unless you say so.** The generation
time and incubation period default to EpiNow2's shipped examples. Those are
properties of transmission rather than of reporting, and no amount of reporting
data identifies them — but they are still an assumption you are making, not a
detail. See `vignette("nowcasting-models")` for fitting the truncation from the
report dimension, which *is* what a `tbl_now` measures.
:::

### How much history to fit on: `min_date`

The engines above are not shown the same data, and that is deliberate.

`baselinenowcast` and `diseasenowcasting` take the **whole series** in their
stride: one estimates a delay distribution from the reporting triangle, the other
fits a state-space model whose cost grows gently with the number of periods. More
history is more information about the delay, and there is no reason to throw it
away.

`epinowcast` and `EpiNow2` are different. Both scale with the number of
**reference dates** they are given -- `epinowcast` carries a parameter block per
reference date, `EpiNow2` models a latent infection curve over every day of it --
so a series that costs `baselinenowcast` a second costs them an afternoon.
Trimming them is not a workaround; it is a modelling decision about how much
history the reporting process is assumed to be stable over.

`min_date` puts that decision **on the engine**, so each model gets the window it
needs and no global `filter()` has to be applied to all of them at once:

```{r min-date-demo, eval=FALSE}
run_nowcast(snapshot, engine_epinowcast(min_date = 96))          # last 96 weeks
run_nowcast(snapshot, engine_epinowcast(min_date = as.Date("2009-01-05")))
run_nowcast(snapshot, engine_baselinenowcast())                  # whole series
```

It takes either shape:

| `min_date` | means |
|---|---|
| `NULL` (default) | the whole series |
| a `Date` | keep event dates on or after it |
| a number | keep the last *n* periods before `now`, in the object's **own units** |

The number is usually what you want, and in a `nowcast_backtest()` it is the only
one that behaves: `now` moves between fits, so a fixed calendar cut makes the
fitted window **grow** as the backtest walks forward, and the last fit is trained
on more data than the first. `min_date = 96` on this weekly object is ninety-six
weeks at every retrospective date.

::: {.alert .alert-info}
**`min_date` trims the event axis, not `now`.** The nowcast is still made as of
the same date, and the trimmed object is what the result carries -- so
`score_nowcast()` and `autoplot()`'s reported counts describe the series the
model was actually shown.
:::

### What you get back

Every method returns a `tbl_nowcast`. Whatever the package produced — a Stan
fit, a matrix of samples, a data frame of quantiles — the predictions are always
in the same tidy shape:

```{r bln-quantiles-1, eval=FALSE}
as_tibble(baseline)
```

```{r bln-quantiles, echo=FALSE, purl=FALSE}
display$baseline_quantiles
```

and, when the backend is sample-based, the draws are there too:

```{r bln-draws-1, eval=FALSE}
as_tibble(baseline, type = "draws")
```

```{r bln-draws, echo=FALSE, purl=FALSE}
display$baseline_draws
```

### `tidy()` works here too

`as_tibble()` gives you the quantiles in full. `tidy()` gives you the **same
summary table every other engine in this package produces** — the one documented
at `?tidy.nowcast` — so a nowcast fitted through `run_nowcast()` and one fitted
by calling a package by hand are read the same way:

```{r bln-tidy-1, eval=FALSE}
tidy(baseline)
```

```{r bln-tidy, echo=FALSE, purl=FALSE}
display$baseline_tidy
```

Those first weeks look odd until you notice what they are: `estimate`,
`conf.low` and `conf.high` are identical because early 2008 was **settled** long
before this `now`. Every case had been reported, so there is nothing left to
nowcast and no uncertainty to report. The interesting rows are at the `now` end.

Two columns there are read off the object rather than assumed. `engine` is the
method that produced it. `level` is the width of the **widest symmetric pair of
quantiles the nowcast actually carries** — `0.95` for the default
`nowcast_quantile_levels()`, `0.8` for a nowcast summarised at
`c(0.1, 0.5, 0.9)`, and `NA` when no symmetric pair exists at all. A guessed
width would defeat the point of the column, which is to stop a 90% band being
compared with a 95% one as though they were the same thing.

Nothing is thrown away. The `fit` property still holds the backend's own object,
so you can keep using that package's diagnostics:

```{r bln-fit, eval=FALSE}
class(baseline@fit)
#> [1] "baselinenowcast_df" "data.frame"
```

And there is an `autoplot()` method for a quick look. It is drawn in **green**,
the colour this package reserves for the epidemic process — a nowcast is an
estimate of *what happened*, not of *when we found out*:

```{r bln-plot-1, eval=FALSE}
autoplot(baseline)
```

```{r bln-plot, echo=FALSE, purl=FALSE, eval=has("ggplot2"), message=FALSE, warning=FALSE, fig.width=7, fig.height=4, fig.alt="A green fan chart of the dengue nowcast: nested prediction intervals around a median line, with the counts reported so far as points."}
library(ggplot2)

# Redrawn from the cached predictions of THIS fit. It used to be drawn from the
# ensemble's predictions under the label "baselinenowcast", which is a picture of
# a different object than the code above produces.
autoplot(tbl_nowcast(
  predictions = display$baseline_predictions,
  method      = "baselinenowcast",
  event_date  = display$event_date,
  now         = display$now
)) +
  coord_cartesian(xlim = c(display$now - 180, display$now))
```

### Strata are handled for you

When the `tbl_now` declares strata, `run_nowcast()` does whatever that backend
needs to honour them — one reporting triangle per stratum for `baselinenowcast`,
`NobBS.strat()` instead of `NobBS()`, one fit per stratum for `surveillance`,
`regional_epinow()` instead of `estimate_infections()` — and the stratum ends up
as an ordinary column of the output:

```{r strata, eval=run_models && has("baselinenowcast")}
dengue_by_sex <- denguedat |>
  filter(onset_week >= as.Date("2008-01-01")) |>
  count(onset_week, report_week, gender, name = "n") |>
  tbl_now(
    event_date = onset_week, report_date = report_week, case_count = n,
    strata = gender, data_type = "count-incidence", verbose = FALSE
  ) |>
  filter(report_week <= now) |>
  change_now(now = now)

by_sex <- run_nowcast(dengue_by_sex, engine = engine_baselinenowcast(draws = 1000))

as_tibble(by_sex)
```

```{r strata-out, echo=FALSE, eval=!run_models, purl=FALSE}
display$stratified_quantiles
```

`tidy()` labels each block with its stratum, so `(stratum, event_date)` is a
unique key and the two series can never be confused for one another:

```{r strata-tidy-1, eval=FALSE}
tidy(by_sex)
```

```{r strata-tidy, echo=FALSE, purl=FALSE}
display$stratified_tidy
```

If a backend genuinely cannot stratify, it warns and pools rather than
pretending.

## 2. Scoring a nowcast

`score_nowcast()` compares the predictive quantiles with what was eventually
observed, using the weighted interval score (WIS) of @bracher2021, plus the
absolute error of the median and the 50% and 90% interval coverage.

Scoring only makes sense against data the model had not seen, which is why
`snapshot` was truncated at `now` and the truth comes from the full series:

```{r score, eval=FALSE}
score_nowcast(baseline, truth = dengue)
```

```{r score-out, echo=FALSE, purl=FALSE}
display$baseline_score
```

Lower WIS is better; `coverage_90` should be `TRUE` about nine times in ten if
the intervals are honest.

If you would rather use the full \pkg{scoringutils} score suite,
its coercion generic accepts the nowcast directly. Supply the later, full
`tbl_now` as `truth`:

```{r scoringutils, eval=FALSE}
baseline |>
  scoringutils::as_forecast_quantile(truth = dengue) |>
  scoringutils::score()
```

`as_scoringutils(baseline, truth = dengue)` is the long tibble underneath that
object when you want to inspect or alter the forecast units first.

When the nowcast retains posterior draws, it can instead become a sample
forecast. This follows the same adapter pattern used by \pkg{epinowcast}:

```{r scoringutils-sample, eval=FALSE, purl=FALSE}
baseline |>
  scoringutils::as_forecast_sample(truth = dengue) |>
  scoringutils::score()
```

Sample conversion is deliberately unavailable for a quantile-only nowcast:
samples cannot be reconstructed from a handful of quantiles.

`tbl.now`'s own `wis` and the one \pkg{scoringutils} computes agree to machine
precision — the package's test suite checks exactly that, on the same numbers,
rather than trusting either implementation on its own.

## 3. Ensembles

Models fail in different directions, and combining them cancels part of that: an
ensemble is rarely the single best model, but it is also rarely the worst, which
is worth a great deal when you have to commit to something before you know which
epidemic you are facing. *How* you combine them matters more than most write-ups
admit, and the two rules below behave differently enough that the choice is
worth making deliberately. Because every `run_nowcast()` result has the same
shape, combining them is one call.

The members below are the fits from the cards above, so nothing is refitted:

```{r members, eval=run_models}
members <- list(
  baselinenowcast   = baseline,
  diseasenowcasting = dnc,
  # The SAME package, a different epidemic process. An ensemble is not only a
  # hedge across packages -- two structurally different models from one package
  # disagree in their own way, and that disagreement is worth combining too.
  dnc_ar1           = run_nowcast(
    snapshot,
    engine_diseasenowcasting(
      model = diseasenowcasting::model(
        epidemic = diseasenowcasting::ar1_epidemic()
      ),
      label = "dnc_ar1"
    )
  ),
  epinowcast        = enw,
  NobBS             = nobbs,
  EpiNow2           = en2
)

ensemble <- nowcast_ensemble(members)
```

```{r members-out, echo=FALSE, eval=!run_models, purl=FALSE}
replay(display$ensemble_print)
```

### The ensemble against its parts

The reason to build one is easier to see than to describe. Each thin line below
is one member's median, the green fan is the ensemble, and the dashed line is
what those weeks eventually reached. The window is the ensemble's own: it keeps
only the dates **every** member covers, and `epinowcast` reports just the
reference dates it is nowcasting.

```{r ensemble-vs-members, echo=FALSE, purl=FALSE, eval=has("ggplot2"), message=FALSE, warning=FALSE, fig.width=8, fig.height=4.6, out.width="100%", fig.cap="The quantile ensemble against each of its members over the last weeks before `now`.", fig.alt="Coloured lines, one per member, fan out over the final weeks; the ensemble's shaded band sits among them and its median tracks the eventual truth more closely than most single members."}
library(ggplot2)

event_col <- display$event_date
# The ensemble's OWN targets, not a hard-coded window: it keeps only the dates
# EVERY member covers, and `epinowcast` reports just the reference dates it is
# actually nowcasting. A fixed twelve-week window would have drawn empty space.
window <- sort(unique(display$ensemble_predictions[[event_col]]))
in_window <- function(d) dplyr::filter(d, .data[[event_col]] %in% window)

# The fan comes from `autoplot()` itself rather than from a second
# implementation of the band logic here. The nowcast is rebuilt from the cached
# predictions, so the bands are the ones the ensemble actually carries -- the
# members below report different quantile levels, and the ensemble keeps only
# the ones they share, which is not a set this chunk should be guessing at.
ensemble_cached <- tbl_nowcast(
  predictions = in_window(display$ensemble_predictions),
  method      = "ensemble",
  event_date  = event_col,
  now         = display$now
)

autoplot(ensemble_cached) +
  geom_line(
    data = display$member_predictions |>
      dplyr::filter(abs(.quantile_level - 0.5) < 1e-8) |>
      in_window(),
    aes(x = .data[[event_col]], y = .value, colour = model),
    linewidth = 0.5
  ) +
  geom_line(
    data = in_window(display$truth),
    aes(x = .data[[event_col]], y = .observed, linetype = "Eventual truth"),
    colour = "#262626", linewidth = 0.8
  ) +
  scale_colour_viridis_d(name = "Member median", option = "turbo", end = 0.9) +
  # DASHED, because `autoplot()` draws the ensemble's median as a solid line in
  # the package green: a second solid dark line beside it reads as a third model
  # rather than as the answer.
  scale_linetype_manual(values = c("Eventual truth" = "22"), name = NULL) +
  labs(
    title = "The ensemble and its members",
    subtitle = paste0("Now = ", format(display$now),
                      "; the green fan and its median line are the ensemble"),
    x = "Onset week", y = "Cases"
  )
```

Three things are worth reading off it.

Most members **fan out at the `now` edge** and lie on top of each other before
it. That is the shape of the problem: the settled weeks have nothing left to
nowcast, so any sane model reproduces them, and the disagreement concentrates
exactly where the answer is not yet known. The spread at the right is what an
ensemble exists to absorb.

`EpiNow2` is the exception, and it disagrees **everywhere**. It is not estimating
the same thing as the others: it fits a latent *infection* curve and reports the
smooth implied by it, rather than correcting each week's reported count for its
outstanding delay. A member that is off by a constant factor on settled weeks is
usually a sign of that kind of mismatch rather than of a bad fit.

And the ensemble's median sits **inside** the spread rather than at either edge
of it. That is the entire trade: it is rarely the best line on any given week,
and it is rarely the worst either — which is worth a great deal when you have to
commit before knowing which member was right.

::: {.alert .alert-warning}
**An ensemble reports only what every member can.** The fan above has as many
bands as the levels its members share, and here that is all nine of
`nowcast_quantile_levels()` — every member either keeps draws or was asked for
those levels at fit time. Add `surveillance`, which keeps none and reports a
fixed `{0.025, 0.5, 0.975}`, and the fan collapses to a single 95% band. That is
why it is a card above but not a member here — see the note below.
:::

::: {.alert .alert-warning}
**These fits use approximate inference.** `epinowcast` runs through
`enw_pathfinder()` and `EpiNow2` through `stan_opts(method = "pathfinder")` so this
article rebuilds in minutes rather than overnight. That is fine for showing how
the ensemble *machinery* behaves and is **not** a tuned posterior for either
model — do not read any single member's band as that package's considered answer.
:::

An ensemble is a `tbl_nowcast` like any other, so it scores, plots and tidies
the same way. Its `engine` is the ensemble's name rather than a package:

```{r ensemble-tidy-1, eval=FALSE}
tidy(ensemble)
```

```{r ensemble-tidy, echo=FALSE, purl=FALSE}
display$ensemble_tidy
```

It can also be a member of another ensemble. This makes it possible, for
example, to combine a `diseasenowcasting` HSGP/AR ensemble with an `epinowcast`
random walk and a baseline. Quantile ensembles can always be nested; a nested
member of a `linear_pool` must itself have draws, so its inner ensemble must
also use `type = "linear_pool"`.

### Two ways of combining

`type = "quantile"` (the default) averages the members' quantiles level by
level — *vincentization*. It is the workhorse of the forecast hubs, it always
applies because every backend produces quantiles, and it tends to produce
**narrower** intervals than the members.

`type = "linear_pool"` instead pools the members' draws into a mixture
distribution and re-summarises it. It needs draws from every member and
generally produces **wider** intervals, because disagreement between models
becomes extra spread rather than being averaged away.

```{r pool, eval=run_models}
sharp <- nowcast_ensemble(members, type = "quantile")

# The pool needs draws, so it takes the members that have them.
with_draws <- members[c("baselinenowcast", "diseasenowcasting",
                        "dnc_ar1", "epinowcast")]
wide <- nowcast_ensemble(with_draws, type = "linear_pool", n_draws = 4000)
```

Which one you want depends on whether you read between-model disagreement as
noise to be averaged out (quantile) or as genuine uncertainty to be propagated
(linear pool).

::: {.alert .alert-warning}
**The linear pool needs draws, and not every backend has them.** Of the six
engines above, `baselinenowcast`, `diseasenowcasting`, `epinowcast` and
`EpiNow2` keep per-event-date draws; `NobBS` and `surveillance` report summaries
only. (`NobBS`'s `nowcast.post.samps` cover the `now` date rather than every
event date, which is not the same thing.) `type = "linear_pool"` **refuses** a
set of members that includes one of the two, rather than silently dropping it
and returning a differently-composed ensemble under the same name — which is why
the call above selects its members explicitly.
:::

::: {.alert .alert-info}
**Members must agree on what they are predicting.** An ensemble combines members
target by target, so a date one member covers and another does not cannot be
combined: averaging over whoever happens to be present would report a single
member's own value as the ensemble's. Those targets are dropped, with a warning
saying how many. This is not hypothetical — `EpiNow2` forecasts one period past
the end of the data, and before `run_nowcast()` trimmed it that lone extra week
landed exactly at the `now` edge, which is the part of the picture people read.
:::

::: {.alert .alert-info}
**Why `surveillance` is not a member here.** There *is* a principled reason, and
it is arithmetic. `surveillance` keeps no per-date draws and reports a **fixed**
`{0.025, 0.5, 0.975}`, so it cannot be asked for any other level after the fit.
An ensemble reports only the levels **every** member carries, so adding it drops
the ensemble from the nine hub levels to those three — one 95% band instead of a
fan — and every other member's extra resolution is thrown away to accommodate
one that has none.

`nowcast_ensemble()` warns when this happens rather than pretending. The fix is
at fit time: leave it out, or widen what it reports where the package lets you
(`surveillance`'s `control$alpha`). Nothing in `nowcast_ensemble()` treats it
specially.
:::

### Weights

By default every member counts the same. You can also give weights directly:

```{r weights-manual, eval=FALSE}
nowcast_ensemble(members, weights = c(
  baselinenowcast = 0.3, diseasenowcasting = 0.2, dnc_ar1 = 0.1,
  epinowcast = 0.2, NobBS = 0.1, EpiNow2 = 0.1
))
```

or *learn* them from how the members actually performed. `nowcast_backtest()`
walks back through time: at each retrospective date it truncates the data to the
reports available then, refits every method, and scores the result against what
was eventually observed.

It takes the **same engines** you fitted with, so there is no second place to
keep the arguments in step. An engine's `label` becomes its name in the result —
which is how one package appears twice, so the two `diseasenowcasting` models
above can be weighted separately rather than sharing one weight:

```{r backtest, eval=run_models}
backtest <- nowcast_backtest(
  dengue,
  engine_baselinenowcast(draws = 1000),
  engine_diseasenowcasting(),
  engine_diseasenowcasting(
    model = diseasenowcasting::model(
      epidemic = diseasenowcasting::ar1_epidemic()
    ),
    label = "dnc_ar1"
  ),
  engine_nobbs(max_D = 10, moving_window = 64),
  now_dates = now - 7 * 8 * (3:1),
  seed      = 20260824
)

tidy(backtest)
```

```{r backtest-out, echo=FALSE, eval=!run_models, purl=FALSE}
head(cached$backtest_tidy)
```

::: {.alert .alert-info}
**Why not every ensemble member is backtested here.** A backtest is
`length(methods) × length(now_dates)` model fits. `epinowcast` and `EpiNow2` are
both Stan models and each would add three more, so they are left out of *this*
backtest to keep the article buildable — not because anything stops them. If you
weight an ensemble from a backtest that does not cover every member, give the
uncovered ones weights yourself; `nowcast_ensemble()` will not invent them.
:::

`tidy()` on a backtest gives one row per (method, `now` date, target), carrying
the retrospective prediction (`estimate`, `conf.low`, `conf.high`) next to what
was eventually `observed` and the scores that comparison earned. It goes
straight into `dplyr` or `ggplot2`:

```{r backtest-summary, echo=FALSE, purl=FALSE}
cached$backtest_summary
```

The backtest also converts directly to \pkg{scoringutils}; it already stores
the truth, and `now` remains a forecast unit so repeated retrospective
predictions of one target stay distinct:

```{r backtest-scoringutils, eval=FALSE, purl=FALSE}
backtest |>
  scoringutils::as_forecast_quantile() |>
  scoringutils::score() |>
  scoringutils::add_relative_skill()
```

Quantile conversion above needs no extra storage. To use
`as_forecast_sample()` instead, build the backtest with `keep_draws = TRUE`.
This can make the object much larger, and every included engine must return
draws; a mixed backtest containing a quantile-only engine is refused rather
than scored on only a subset of its methods.

This is the expensive part of the workflow — it is `length(methods) ×
length(now_dates)` model fits — so keep `now_dates` short when the members are
Bayesian. Pass `seed` so the backtest is reproducible: one `set.seed()` before
the whole thing only pins anything if every method draws the same random numbers
in the same order, which stops being true the moment you drop a method.

From the backtest, `nowcast_weights()` offers two rules:

```{r weights-fitted, eval=run_models}
fitted_weights <- nowcast_weights(backtest, type = "inverse_score")
fitted_weights                                    # w proportional to 1 / mean WIS

nowcast_weights(backtest, type = "optim")         # w minimising the ensemble's WIS
```

```{r weights-out, echo=FALSE, purl=FALSE}
round(cached$article_weights, 3)
```

`"inverse_score"` is the safe default: it is monotone in performance, never
collapses onto a single model, and cannot overfit. `"optim"` searches the simplex
for the weights that would have minimised the ensemble's WIS over the training
window; it is better in principle, but with only a handful of retrospective dates
it happily overfits, so prefer it when the training window is long.

Either rule plugs straight into the ensemble:

```{r ensemble-weighted, eval=FALSE}
nowcast_ensemble(members, weights = "inverse_score", backtest = backtest)
```

## 4. Adding your own model

This is the part that makes the whole thing worth building. A back-end is **two
S3 methods**, and they can live in your own package, your analysis script, or a
one-off chunk. Nothing inside `tbl.now` needs to change.

`run_nowcast(x, engine = engine("mymodel"))` builds a little object of class
`c("mymodel", "nowcast_method")` and calls

* `nowcast_fit(method, x, ...)` — run the model, return whatever it returns;
* `nowcast_tidy(method, fit, x, ..., quantile_levels)` — describe the result as
  `predictions` (one row per event date, stratum and quantile level) or `draws`
  (one row per event date, stratum and draw). Either may be `NULL`, not both.

`vignette("custom-nowcast-models")` is the full account: what your method may
assume about the `tbl_now` it is handed, how to reuse the converters instead of
reshaping by hand, a complete worked back-end that needs no modelling package,
and what shipping one in a package involves.

```{r learning-more, echo=FALSE, results="asis"}
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```

## References
