---
title: "The nowcasting workflow: hospital-acquired infections in Bucaramanga, Colombia"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{The nowcasting workflow: hospital-acquired infections in Bucaramanga, Colombia}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 150,
  fig.align = "center"
)
options(tibble.width = Inf, width = 160, pillar.width = 160)

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)

# epinowcast compiles a Stan model, which takes minutes. Flip to TRUE to fit it.
run_stan <- FALSE
```

This article is an end-to-end walk-through of the `tbl.now` workflow on a real,
**deliberately unpolished** dataset. We will:

1. **Build** a `tbl_now` from a messy surveillance extract and show
   how to `diagnose()` possible data errors.
2. **Clean** what it reported: duplicate records, missing dates, etc. 
3. **Look** at the data with `autoplot()`, and the  `summary()` functions.
4. **Describe** the behaviour of the reporting delay. 
5. **Attach temporal effects** to a model using the data.
6. **Nowcast** with two different engines using everything we found.

We'll start the process by nowcasting with two dates (event and report dates)
and then we'll focus on nowcasting with three dates (event, report, and validation dates). 

Let's start by calling the libraries:

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(tbl.now)
```

## The data

`hai_bucaramanga` is a raw line list of **healthcare-associated infections** (IAAS,
*Infecciones Asociadas a la Atención en Salud*) notified in the county of
Bucaramanga, Sandander, Colombia, as published by the county. Each row is
one infection: a specimen taken from a hospitalised patient, the laboratory
result, and the microorganism isolated.

```{r load-data}
data(hai_bucaramanga)
```

```{r}
hai_bucaramanga
```

For our example, six columns matter:

* `id`            — a unique patient identifier. 
* `specimen_date` — the **event date**, when the sample was taken.
* `report_date`   — the **report date**, when the laboratory issued the result.
* `received_date` — the **validation date**, when the report was received by INS.
* `microorganism` - the species of the organism isolated in the culture. 
* `sex`           - the sex of the individual. 

For the tutorial, we will do perform a nowcast assuming we are standing on July 18th 2022. For
that purpose we filter our data to what it would have looked like:


```{r}
hai_bucaramanga <- hai_bucaramanga |>
  filter((is.na(specimen_date) | specimen_date <= ymd("2022/07/18") & specimen_date >= ymd("2020/01/01")), 
         (is.na(report_date)   | report_date   <= ymd("2022/07/18")),
         (is.na(received_date) | received_date <= ymd("2022/07/18")))
```

::: {.alert .alert-warning}
For this tutorial  we are keeping those events that have missing dates. That 
is what the `is.na()` section on the left side of the filter is doing as 
we keep either cases by October 18th 2022 or missing. 
:::

## 1. Initial data cleaning

In this tutorial we show how to discover several anomalies in the data
as well as several modeling indicators. We suggest, however, that before
transforming the dataset into a `tbl_now` you use traditional data 
analysis tools to clean the data first. Here we'll just select
the columns that matter to us and move onto the `tbl_now` capabilities. However
if you already have a pipeline for cleaning and preparing your data
this is where you would do it as we'll see the `tbl_now` diagnostics
don't cover everything. 

```{r}
#We keep only the columns of interest
hai_bucaramanga <- hai_bucaramanga |> 
  select(id, specimen_date, report_date, received_date, microorganism, sex)
```

<!--
## 0. Initial data transformation

The column `microorganism` contains all the isolated organism. We will
consider the following as our microorgainsms of interest and
recode them as `"confirmed"` (they will be confirmed cases of our disease).
Those that were classified as something else will be recoded
as `"retracted"` (they were suspected Klebsiella but not anymore). Finally
those with `NA` will be recoded as `"pending"` as we still don't know their result.

We note that for our analysis we will only focus on Klebsiella pneumoniae hence
solely code as **confirmed** the following microorganism entries: `Klebsiella pneumoniae`, 
`Klebsiella pneumoniae subsp. pneumoniae`, and `Klebsiella pneumoniae (multidrug-resistant)`.
The entry `Klebsiella pneumoniae/oxytoca` will be coded as **pending**. The remaining
microorganisms will be considered **retracted**:

```{r}
#We list those that we will consider "confirmed"
valid_organisms <- c("Klebsiella pneumoniae",  
                     "Klebsiella pneumoniae subsp. pneumoniae",  
                     "Klebsiella pneumoniae (multidrug-resistant)")

hai_bucaramanga <- hai_bucaramanga |> 
  mutate(validation = case_when(
    microorganism %in% valid_organisms               ~ "confirmed",
    is.na(microorganism)                             ~ "pending",
    microorganism == "Klebsiella pneumoniae/oxytoca" ~ "pending",
    .default = "retracted"
  ))
```
-->

## 2. Building the `tbl_now`

We'll create the `tbl_now()` object. For the beginning of the example, 
we will declare two dates (`specimen_date` for the event and `report_date`) for
the report. We'll further specify that the  data is in line-list format 
(_i.e._ each row is a case), and stratify by `sex`. 


```{r build-tbl-now-raw}
hai_bucaramanga <- hai_bucaramanga |>
  tbl_now(
    event_date   = specimen_date,
    report_date  = report_date,
    strata       = sex,
    data_type    = "linelist",
    event_units  = "days",
    report_units = "days"
  )
```

The `tbl.now()` automatically fires several warnings. Let's take a look!

## 3. What is wrong?

These same warnings can also be accessed with the `diagnose()` function which returns
the warnings as well as additional notes:

```{r diagnose-raw}
diagnose(hai_bucaramanga) 
```

Let's see what they identify. 

### The warnings


In general the warnings will look for missing data within the dates
and the strata or for violations of the nowcasting hypotheses (for example
reports that "see the future" and identify an event before it happens!). In
our case, specifically we identify missing dates in both the
report and the event as well as some reports that have incorrect dates.

There is no solution that works in all cases for these problems and oftentimes
the ideal solution is to identify the reason for the misingness. Here we posit 
some approaches for the warnings:

1. **For simultaneosuly missing event and report dates** there is nothing one can do.
In those cases we don't know when something happened (to add to the epidemic process)
or when it was reported (to add to the delay process). Hence we remove those rows.

```{r}
#Remove those that have both dates missing
hai_bucaramanga <- hai_bucaramanga |> 
  filter(!(is.na(report_date) & is.na(specimen_date)))
```

2. **For the rows with missing reports, we can censor them above**. This will tell the models in [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) that the
measurement is not exact; instead, that we know the report arrived before
a certain date though we don't know that _exact_ date in particular. We
can do this because if we are seeing the reports **today** then we know that at most
they took until **today** to arrive. 

This censoring will add a column `.is_censored_report` which will indicate 
that that report is censored. 

When calling the function we specify that the ones we are censoring are the missing 
values and that the date they should input is "2022/10/18" which corresponds
to the date of the nowcast in this example. 

```{r, warning=FALSE, message=FALSE}
hai_bucaramanga <- hai_bucaramanga |> 
  censor_reports(is.na(report_date), to_report = as.Date("2022/07/18"))
```

2. **For the delays before the reports we visualize the data**. Doing a quick analysis
we see that its a date formatting issue. You can see in the first entries that
the specimen date is in format `year-day-month` (read with `lubridate::ydm()`) 
while the `report_date` and `received_date` is in format `year-month-day`
(read with `lubridate::ymd()`) . 

```{r, eval=TRUE}
hai_bucaramanga |> filter(report_date < specimen_date) 
```


We can correct the date format to fix this issue: 

```{r, warning=FALSE}
hai_bucaramanga <- hai_bucaramanga |> 
  mutate(specimen_date = if_else(specimen_date > report_date, 
                                 ydm(as.character(specimen_date)), 
                                 ymd(as.character(specimen_date))))
```


3. Currently there is **nothing we can do for the cases were the event date is unknown**. Hence we just document them
and we'll remove from the dataset given that they represent such a small percent (`r 100*(hai_bucaramanga |> filter(is.na(specimen_date)) |> nrow())/nrow(hai_bucaramanga)`%). 

```{r, warning=FALSE}
hai_bucaramanga <- hai_bucaramanga |> filter(!is.na(specimen_date))
```

Re-running `diagnose()` confirms we solved this issues:

```{r}
diagnose(hai_bucaramanga)
```

### The notes

The notes convey additinal information that, at a glance, can tell you whether something
needs to be corrected in your data. 

It helps you know that you have columns that are not strata or covariates (in case you forgot to declare them!):

```
declarations/undeclared: 3 columns "id", "received_date", and "microorganism" are not declared as strata or covariates.
```

It quantifies how much time has happened between the last observation and the now for each strata:

```
now/now_gap_event [Female]: The last event date is 50 days before now ("2022-10-18").
now/now_gap_event [Male]: The last event date is 35 days before now ("2022-10-18").
```

It also tells you how what is the smallest stratum (in case you have strata with so few
observations you need to collapse them into another one):

```
strata/size [Female]: The smallest stratum is "Female" with 420 cases, 42.9% of the total.
```

And how sparse your data is: 

```
strata/sparsity [Female]: The sparsest stratum is "Female": 86% of the event dates on the grid carry no cases at all.
```

This is particularly useful for triangle-based methods that require to fill with zeroes the whole reporting triangle. 


## 4. More cleaning

Besides the `diagnose()` one always has to do some additional cleaning of datasets.
One particular issue is for repeated rows; that is when an individual is
registered twice in the dataset for the same disease at the same exact times.

Here we can see that the same individuals `id` are sometimes repeated
for the same event reporting the same microorganism:

```{r dupes}
hai_bucaramanga |>
  count(id, microorganism, report_date, specimen_date) |>
  filter(n > 1) |> 
  as_tibble()
```

We can use `dplyr::distinct()` to remove them without losing information:

```{r clean-distinct}
hai_bucaramanga <- hai_bucaramanga |> 
  distinct()
```


## 5. Looking at the data

`autoplot()` draws a diagnostic grid with the panels in green describing
the **epidemic** process and the red ones, the **reporting** process.

```{r autoplot, fig.width=11, fig.height=13, message=FALSE, warning=FALSE, fig.alt="A six-panel diagnostic grid: empirical delay distribution, observed epidemic curve, month-of-year effects for cases and for reporting, and periodograms for both processes."}
autoplot(hai_bucaramanga)
```

Four things stand out, and each of them changes a modelling decision later.

**The epidemic curve is extremely sparse.** There are only a handful of cases a year
with most of the times 0 or 1 cases being reported. 

```{r delay-dist, fig.width=9, fig.height=4.5, fig.alt="Epidemic process."}
plot_epidemic_process(hai_bucaramanga)
```

> **Note** This is the same sparcity the `note` in the `diagnostic()`
function gave us. 


**The delay distribution has some extreme delays.** This is the most consequential
feature of the dataset, so it is worth its own plot:

```{r delay-dist2, fig.width=9, fig.height=4.5, fig.alt="Histogram of the reporting delay."}
plot_delay_distribution(hai_bucaramanga)
```

A lot of results are reported way after 500 days!

**There doesn't seem to be a day-of-the-week effect** In both reports and events
there is no aparent day of the week effect.

**One week had a lot of reports** Epiweek 29 is an anomaly in the amount of reports
it seems to have. This would warrant further investigation.

### The same picture, as numbers

`autoplot()` is for looking; `summary()` is for reading those values. You can either call `summary()` all at once:

```{r summary-delay, eval = FALSE}
summary(hai_bucaramanga) 
```

or as we do here call each of the summaries we are going to discuss. We can start with the delays
where we can see that up until the 3rd quartile delays are less than 100 days but then they explode by the 90th quantile:

```{r}
delay_summary(hai_bucaramanga)
```
We can see the same idea in the completeness summary which quantifies thr proportion of cases that had arrived.

The `zero_run_summary()` quantifies the sparcity by counting how many days with continuous zeroes happened for either
the event or the report dates. That is, it measures the distribution of days of continuous zeroes. You can see that usually there are on average `r `zero_run_summary(hai_bucaramanga)$mean[1]` continuous zeroes before a case. However there have been runs of `r zero_run_summary(hai_bucaramanga)$max[1]` continuous days with no cases (continuous zeroes) until one case:

```{r}
zero_run_summary(hai_bucaramanga)
```

The composition block shows that there are way less females than males. Which, again was already observed in the notes as the smaller stratum:

```{r summary-strata}
prop_strata(hai_bucaramanga)
```

Finally the cases per date shows how many cases on average occur per date-axis. That is on average how many
cases there are per event date or per report date. In this case we can see that at least half of the dates have zero cases:

```{r}
cases_per_date(hai_bucaramanga)
```

Given the sparcity here we can still nowcast using the diseasenowcasting package:

```{r}
hai_fit <- hai_bucaramanga |> 
  run_nowcast(engine = engine_diseasenowcasting(temporal_effects = "none"))
``` 

Given the overall sparcity we have observed the nowcast will predict almost no cases at any
time with just maybe a second case at the now for males:

```{r, eval=FALSE}
autoplot(hai_fit)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5, fig.alt="A sparse nowcast"}
autoplot(hai_fit) +
  ggplot2::scale_y_continuous(breaks = seq(0, 3, by = 0.5))
```
 
## 6. Changing to weekly data

Though nowcasting is possible with sparse data, oftentimes people prefer to rescale data to a coarser time frame. Here
we'll rescale the dataset to a weekly reporting time frame with `aggregate_time_units()`:

```{r}
hai_bucaramanga <- hai_bucaramanga |> 
  aggregate_time_units(to = "weeks")
```

This aggregation of the epidemic process allows the nowcasting models to infer more from
it as you can see a more clear epidemic shape in the autoplot:

```{r, fig.width=11, fig.height=13, message=FALSE, warning=FALSE, fig.alt="The now auto plot with weekly data"}
autoplot(hai_bucaramanga)
```

One should run again the `summary()` in the aggregated to draw conclusions from this data.
For us, however it will suffice to see that the sparcity has reduced to usually just 1 week
with zero cases before seeing a week with cases:

```{r}
zero_run_summary(hai_bucaramanga)
```

## 7. Is the reporting delay stable?

A nowcast might assume that  the delay distribution is the same throughout. 
Here we check those assumptions: 

First, we just look at the delay aggregated through time

```{r drift-plot, fig.width=9, fig.height=5.5, message=FALSE, warning=FALSE, fig.alt="Fan chart of the reporting delay over time."}
plot_delay_drift(hai_bucaramanga)
```

The median moves from an extremely wide distribution in 2020 to a shorter delay
after 2021. We can confirm that split of two delay behaviours with the following tests: 

### Testing for the delay's drift

The `diagnose_drift()` function runs an autocorrelation-robust Mann-Kendall test for a
*monotonic* trend in the delay:

```{r drift-test, warning=FALSE}
diagnose_drift(hai_bucaramanga)
```

Here the median shows a reduction (`sens_slope` of `r diagnose_drift(hai_bucaramanga)$sens_slope[1]`) which is consistent with the plot. The `-1` can be interpreted as the decrease in the median per unit of time. In this case, the median delay reduced 1 week per week.   

### Is there an abrupt shift?

A trend test cannot see a steep change like a jump. The `diagnose_changepoint()` 
function uses Pettitt's test to look for exactly that. 

```{r changepoint-test, warning=FALSE}
diagnose_changepoint(hai_bucaramanga)
```

Here neither the median nor the spread have a changepoint. The change
was a drift in the trend as we saw in the plot and not an abrupt shift. You can see the 
section on [delay changes](https://rodrigozepeda.github.io/tbl.now/articles/diagnosing-a-tbl-now.html#part-3-what-needs-a-statistical-test-reporting-artefacts) of the [Diagnosing a tbl_now](https://rodrigozepeda.github.io/tbl.now/articles/diagnosing-a-tbl-now.html) article for more information. 

## 8. Are there any reporting batches?

A reporting batch consists on several reports that were all held for any reason by the reporter (e.g. a laboratory)
and then reported all together as a group so that there is a giant signal in the reporting process. To see them we can start by plotting against the reporting axis:

```{r,  fig.width=9, fig.height=4.8}
plot_reporting_process(hai_bucaramanga)
```

We can see that there were several reports that dropped near the end of 2022. The reporting hexamap allows us to visualize the date of the report at the same time as the day of the event and the corresponding delay

```{r,  fig.width=9, fig.height=4.8}
plot_reporting_hexamap(hai_bucaramanga)
```

Here lies a figure with 3 axis. We can see for example following the axis of event date `19/Apr/20` all the way to the
intersection with the report date of `19/Jun/22` that there are many cases with event dates in 2020 that have delays of 100-128 weeks and lie in between December 2019 and April 2020. These reports will certainly influence our estimation of the delay distribution. 

::: {.alert .alert-info}
You can also use 

```{r, eval = FALSE}
plot_reporting_triangle(hai_bucaramanga)
```

to visualize the batches; however for this specific case they are easier to visualize in the hexamap. 
:::

Either way we can see that `2022-07-17` is a candidate for possible batches. In this package, the 
`diagnose_batches()` function helps identify potential dates with batches. We recommend using it with or without the strata depending on whether the context warrants the possibility that batching could differ by strata (e.g. if strata correspond to healthcare facilities maybe yes; if they correspond to gender of the patient maybe no). 

Here we run it without the strata which we remove first. We'll keep only those observations that have been identified as potential batches: 

```{r}
hai_bucaramanga |> 
  remove_all_strata() |> 
  diagnose_batches() |> 
  filter(batch)
```

This identifies some dates that might be batches due to their high reported numbers. 
A second batch method is the `diagnose_batch_shape()` function. This one takes a specific date and checks whether 
the reports that arrived on a date came from systematically older dates. We can combine the findings from the previous run to test some of them and add evidence to which are batches:

```{r}
hai_bucaramanga |> 
  remove_all_strata() |> 
  diagnose_batch_shape(at = ymd("2021-11-28"))
```
Where we further identify the date of `2021-11-28` as a potential batch with its `p_value < 0.003`. 


::: {.alert .alert-info}
The  batch at `2022-07-17` was actually introduced by us by censoring all the cases at that date in section 3. 
:::


<!--
## 7. Temporal effects

`temporal_effects()` attaches calendar covariates a model can use. The
temptation is to add everything available; the discipline is to add only what
the data supports.

**Day-of-week effects are not justified here.** They are the usual first choice
for surveillance data — laboratories are quieter at weekends — but this dataset
does not behave that way:

```{r dow-check}
hai |>
  count(weekday = wday(specimen_date, label = TRUE, week_start = 1)) |>
  mutate(percent = round(100 * n / sum(n), 1))
```

Specimens are spread almost evenly across all seven days, weekends included. And
in any case we aggregated to weeks, so a day-of-week term has nothing to attach
to. Adding one would be modelling noise.

**An annual cycle is justified**, on the reporting side, by the periodogram in
section 5 and by the bimodal delay. Since the data are weekly, that is a Fourier
season of period 52:

```{r temporal-effects}
hai_effects <- temporal_effects(
  week_of_year = TRUE,   # position within the year
  seasons      = 52      # an annual Fourier cycle on weekly data
)

hai_effects
```

Attach them to the report date, because the annual rhythm we saw is a property
of *reporting*, not of infection:

```{r attach-effects}
hai_seasonal <- hai_now |>
  add_temporal_effects(hai_effects, date_type = "report_date")

hai_seasonal
```

Effects are attached **lazily** — the object records the specification but does
not materialise the columns until a model asks for them (or you call
`compute_temporal_effects()`):

```{r compute-effects}
hai_seasonal |>
  compute_temporal_effects() |>
  select(event_week, report_week, starts_with(".report_")) |>
  head(4)
```

## 8. Nowcasting with `diseasenowcasting`

Now we let the diagnostics decide the model. Three findings carry over:

* **The delay distribution changed in March 2020.** Fitting across that boundary
  would average two reporting regimes. We restrict to the post-change-point
  period.
* **The delay reaches 98% by about 20 weeks.** That sets the maximum delay: long
  enough to capture the second reporting stream, short enough not to carry a
  mostly-empty tail.
* **No monotonic drift within that period**, so a single delay distribution for
  the restricted window is defensible.

```{r informed-window}
changepoint_date <- as.Date("2020-03-09")

hai_model <- hai_weekly |>
  filter(event_week >= changepoint_date) |>
  tbl_now(
    event_date  = event_week,
    report_date = report_week,
    strata      = sex,
    data_type   = "linelist",
    verbose     = FALSE
  ) |>
  censor_reporting_delays_above(max_delay = 20)

hai_model
```

`censor_reporting_delays_above()` is doing real work: it caps the delay dimension at 20
weeks so the model is not asked to estimate probabilities for delays that occur
a handful of times in the whole series.

`diseasenowcasting` consumes a `tbl_now` **directly** — no converter, no
restating of which column is which:

```{r dnc-fit, eval=has("diseasenowcasting"), message=FALSE, warning=FALSE}
# `diseasenowcasting::` rather than `library()` -- it is not on CRAN, and
# attaching it would leave `R CMD check` with an undeclared dependency.
hai_fit <- diseasenowcasting::nowcast(hai_model, type = "one_stage", n_draws = 2000)

hai_fit
```

The posterior draws of the eventual counts come from `predict()`:

```{r dnc-predict, eval=has("diseasenowcasting"), message=FALSE, warning=FALSE}
hai_prediction <- predict(hai_fit)

draws       <- S7::prop(hai_prediction, "draws")        # draws x event weeks
event_weeks <- S7::prop(hai_prediction, "event_dates")
observed    <- S7::prop(hai_prediction, "observed_series")

nowcast_summary <- tibble::tibble(
  event_week = event_weeks,
  observed   = observed,
  median     = apply(draws, 2, median),
  lower      = apply(draws, 2, quantile, 0.025),
  upper      = apply(draws, 2, quantile, 0.975)
)

tail(nowcast_summary, 8)
```

```{r dnc-plot, eval=has("diseasenowcasting"), fig.width=9, fig.height=4.8, message=FALSE, warning=FALSE, fig.alt="Nowcast of weekly hospital-acquired infections. Grey bars show cases reported so far, a red line and ribbon show the nowcast median and 95 percent interval, which rises above the bars over the most recent weeks."}
recent <- nowcast_summary |> filter(event_week >= max(event_week) - 7 * 30)

ggplot(recent, aes(x = event_week)) +
  geom_col(aes(y = observed, fill = "Reported so far"), width = 5.5) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#e78b7f", alpha = 0.35) +
  geom_line(aes(y = median, colour = "Nowcast"), linewidth = 0.9) +
  scale_fill_manual(name = NULL, values = c("Reported so far" = "#dfe5e0")) +
  scale_colour_manual(name = NULL, values = c("Nowcast" = "#B85348")) +
  labs(
    x = "Event week (specimen taken)", y = "Infections",
    title = "Nowcast of hospital-acquired infections",
    subtitle = paste0("Now = ", format(get_now(hai_model)),
                      "; the gap at the right edge is what has not been reported yet")
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335"))
```

The bars are what had been reported by the `now`; the ribbon is the model's
estimate of where those weeks will end up.

Look closely at what the nowcast actually does here, because it is not the
textbook picture. The **median barely moves**: it exceeds the observed count in
only one of the 152 modelled weeks. What changes dramatically is the
**uncertainty** — the mean width of the 95% interval is about 0.1 cases over the
settled part of the series and about 2 cases over the final ten weeks, a
twenty-fold widening.

That is the correct behaviour for data this sparse, and it is worth
understanding rather than treating as a disappointment. We know from the delay
distribution that roughly a quarter of results arrive more than a week late, so
a week sitting five weeks from `now` really is missing about a quarter of its
eventual cases. But a quarter of *two cases* is half a case, and the median of a
posterior over counts is an integer. The correction is real; it is simply
smaller than the resolution of the answer. With one to three infections a week,
a nowcast's honest output is a statement about what it does not know, not a
large upward revision.

This is also why the point of the earlier diagnostics was never to produce a
better point estimate. It was to make sure the interval is trustworthy — that it
comes from one reporting regime rather than two averaged together.

### Carrying the strata and effects through

The enriched object needs no extra arguments: `diseasenowcasting` reads the
`sex` stratum and the seasonal columns off the object itself.

```{r dnc-seasonal, eval=run_stan && has("diseasenowcasting"), message=FALSE}
hai_model_seasonal <- hai_model |>
  add_temporal_effects(hai_effects, date_type = "report_date")

diseasenowcasting::nowcast(hai_model_seasonal, type = "one_stage", n_draws = 2000)
```


## Summary

The workflow, start to finish:

```{r recap, eval=FALSE}
# 1. Choose a resolution the data (and your engine) can support.
hai_weekly_raw <- hai_bucaramanga |>
  mutate(event_week  = floor_date(specimen_date, "week", week_start = 1),
         report_week = floor_date(report_date,   "week", week_start = 1))

# 2. Describe it once -- before cleaning, so the object can tell you what is wrong.
hai_raw_now <- tbl_now(hai_weekly_raw, event_date = event_week,
                       report_date = report_week, strata = sex,
                       data_type = "linelist")

# 3. Ask what is wrong with it, instead of guessing.
diagnose(hai_raw_now) |> filter(status <= "note")

# 4. Fix what it reported, and confirm the fix.
hai <- hai_bucaramanga |>
  distinct() |>                                          # duplicates: diagnose() skips
  filter(!is.na(specimen_date), !is.na(report_date)) |>  #   these two it found
  filter(report_date >= specimen_date)                   #   and this one

hai_weekly <- hai |>
  mutate(event_week  = floor_date(specimen_date, "week", week_start = 1),
         report_week = floor_date(report_date,   "week", week_start = 1))

hai_now <- tbl_now(hai_weekly, event_date = event_week, report_date = report_week,
                   strata = sex, data_type = "linelist")

diagnose(hai_now) |> filter(status <= "note")

# 5. Look at it, and read the numbers off it.
autoplot(hai_now)
summary(hai_now)

# 6. Run the statistical tests diagnose() deliberately leaves to you.
diagnose_drift(hai_now)
diagnose_changepoint(hai_now)

# 7. Let those findings set the model window and the maximum delay.
hai_model <- hai_now |> censor_reporting_delays_above(max_delay = 20)

# 8. Hand it to whichever engine you like.
diseasenowcasting::nowcast(hai_model)         # diseasenowcasting
tbl_now_to_baselinenowcast(hai_model)         # baselinenowcast
tbl_now_to_epinowcast(hai_model)              # epinowcast
tbl_now_to_surveillance(hai_model)            # surveillance
as.data.frame(hai_model)                      # NobBS
```

Two closing points worth carrying to your own data.

**The diagnostics were not decoration.** The change-point test found a real
break in March 2020 and that set the model window; the delay quantiles set
`max_delay`; the day-of-week check *stopped* us adding a term the data did not
support. Had we skipped straight to `nowcast()`, all three would have been
guesses.

**Half of this dataset never reached the model.** Duplicate records, missing
dates and impossible delays removed 750 of 1,423 rows. A nowcast built on the
remainder is only as good as the assumption that what was dropped resembles what
was kept — which for the missing dates, concentrated in the early years, is
worth stating out loud rather than burying.
-->

```{r learning-more, echo=FALSE, results="asis"}
# The fragment is looked up by ABSOLUTE path via system.file(), never by a
# relative one. pkgdown/rmarkdown render into an intermediates directory under
# tempdir() and copy relative resources alongside; a path containing "../.."
# escapes that directory ("cannot create file '/tmp/RtmpXXX/../../man/...'"),
# which fails wherever tempdir() sits near the filesystem root -- as it does on
# CI. `system.file()` also works under pkgload::load_all(), which shims it to
# find inst/ in the source tree.
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```
