---
title: "The nowcasting workflow 2: COVID-19 case surveillance with revisions in the United States"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{The nowcasting workflow 2: COVID-19 case surveillance with revisions in the United States}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 150,
  fig.align = "center"
)
options(tibble.width = Inf, width = 160, pillar.width = 160)
options(pillar.print_max = 5, pillar.print_min = 5,
        tibble.print_min = 5, dplyr.print_min = 5,
        tibble.print_max = 5, dplyr.print_max = 5)

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)
```

This article is the sequel to [the nowcasting
workflow](https://rodrigozepeda.github.io/tbl.now/articles/example.html).
That one used **two** dates: the `event_date` for when a case happened, and the
`report_date` for when it was registered in the system. This one adds the **third**
`revision_date` for cases that can be revised to be either confirmed or rejected.

In surveillance contexts, a record might arrive before anybody knows whether it is
real case. A specimen might be collected, a case registered in the system, and only later
does a laboratory settle it. Maybe additional information comes along that causes the patient
to be de-registered (for example the case was registered twice or it did not belong to the population of interest). 
`tbl.now` calls that third date the **revision date**, and models the following process

$$
\texttt{event} \to \texttt{report} \to \texttt{revision}
$$

A case that has not been settled yet is called `"pending"`: it has been reported,
but may still need to be confirmed (and/or retracted).  **Pending cases are the reason the revision axis exists.** as it allows for a nowcast to _decrease_ based on how many we expect to be retracted. 

For this tutorial, we assume you have read the [previous one](https://rodrigozepeda.github.io/tbl.now/articles/example.html) which included how to
build a `tbl_now()`, `diagnose()`, and `summarise()`. We'll move through those sections quickly here
while focusing only on specific methods for the third date. 

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(tbl.now)
```

## 1. The data

`covid_us` is an aggregation of the U.S. CDC's individual-level COVID-19
case surveillance database for 2020. Each row is a unique combination of
three dates, a status and a sex, with the number of cases `n`.

```{r load-data}
data(covid_us)
```

```{r, eval = FALSE}
covid_us
```

```{r, echo = FALSE}
covid_us |> tail()
```

The three dates are the following:

-   `onset_dt` — the **`event_date`**: symptoms begin.
-   `pos_spec_dt` — the **`report_date`**: when the surveillance system first saw the case.
-   `cdc_report_dt` — the **`revision_date``**: when the status as probable or confirmed
was sent to the Center for Disease Control (CDC).

The data also includes a `current_status` as the outcome that revision carries:

```{r}
covid_us |> count(current_status)
```

For our nowcast we will model as if we were standing on the first day of October 2020: `2020-10-01`. To do so we get all of those with reports before that date and we recode the cases that were confirmed after then as `Probable Case` (the confirmation would have not yet arrived by that date if they were confirmed later):s

```{r}
covid <- covid_us |>
  filter(cdc_report_dt <= ymd("2020-10-01") & 
         onset_dt      <= ymd("2020-10-01")) |> 
  mutate(current_status = if_else(
    cdc_report_dt >= ymd("2020-10-01"), "Probable Case", current_status))
```

## 2. Pending means *no settled revision*

Here is the one rule that has no counterpart in the two-date workflow. 

::: {.alert .alert-warning}
A row is `"pending"` **because its revision has not yet given a confirmed or retracted result**, so a
pending row must have a **missing** revision date.
:::

CDC's `"Probable Case"` is a case that met the clinical and
epidemiological criteria without ever meeting the laboratory-confirmed
definition. In this package's vocabulary those will be the `"pending"`. 

Because the pending corresponds to the final confirmation and the probable
cases have yet to be confirmed we will remove their report dates as the **final**
confirmation has not happened for them and they have not settled yet as **Laboratory-confirmed** or not. 



```{r}
covid <- covid |>
  mutate(cdc_report_dt = 
           if_else(current_status == "Probable Case", NA_Date_, cdc_report_dt)) 
```


## 3. Building the `tbl_now`

In the `tbl_now` we add the  `revision_date` and `revision_type`. Furthermore
we will utilize `revision_levels` to translate the source's vocabulary into this package's:

```{r build}
covid <- covid |>
  tbl_now(
    event_date      = onset_dt,
    report_date     = pos_spec_dt,
    revision_date   = cdc_report_dt,
    revision_type   = current_status,
    revision_levels = c(
      "Laboratory-confirmed case" = "confirmed",
      "Probable Case"             = "pending"
    ),
    case_count = n,
    strata     = sex,
    data_type  = "count-incidence"
  )
```

```{r}
covid
```

The `tbl_now` now also has a `.revision_num` and `.revision_delay`.


3. ## Diagnosing

The `diagnose()` function runs the same structural checks as before with some additional checks
for the third date.

```{r diagnose}
diagnose(covid)
```

Specifically we can see that Other, Missing and Unknown aren't very frequent strata. Hence we'll collapse them into a strata call "Other". To do so we use `mutate` to generate the new `sex_category` variable and then `change_strata()` to that so that `tbl.now` recognizes the new one:

```{r}
#You can see that the strata is sex
get_strata(covid)

#Create the new sex category column
covid <- covid |> 
  mutate(sex_category = if_else(sex %in% c("Female","Male"), sex, "Other"))

#Substitute as the strata
covid <- covid |> 
  change_strata(sex_category)

#and now is sex category
get_strata(covid)
```

As we get no warnings from `diagnose()` we can conclude that the data is in good shape. 

## 4. Describing the revision process

Here we analyze some of the decision-process specific summaries. The other  diagnostic summaries are [already in the previous example](https://rodrigozepeda.github.io/tbl.now/articles/example.html) 

### How much is still pending?

The `prop_revision_type()` function splits the cases by outcome, overall and by stratum:

```{r prop-revision}
prop_revision_type(covid)
```

About 3% of cases are still pending, and the share barely moves between
the two strata (which is a good sign, we don't want the confirmation to vary by sex!). 

This same idea can be visualized with  `plot_revision_status()` which 
shows how the confirmation has varied by time:

```{r revision-status, fig.width=9, fig.height=4.5, fig.alt="Share of cases confirmed versus pending by report date, showing pending cases concentrating in the most recent weeks."}
plot_revision_status(covid, by = "sex_category")
```

### How long does a case take to resolve?

The `plot_delay_profiles()` draws the distributions of the delays between revision and report:

```{r revision-delay, fig.width=9, fig.height=4.5, fig.alt="Distribution of the report-to-revision delay."}
plot_delay_distribution(covid, axis = "revision")
```

### Does the outcome change the delay?

The `diagnose_revision_delay()` function checks whether the revision-delay
changes between "confirmed" and "retracted" cases. In our case because 
we don't have retracted cases this function returns `NA`. However, its
usefulness is to be able to diangose whether our models require just one 
revision-delay distirbution or two (one for confirmed and one for retracted cases).

```{r revision-delay-test, warning=FALSE}
diagnose_revision_delay(covid)
```


### Getting the cases that survived the revision

The `*_revised_cases()` getters give you a slice of
the latest confirmed cases:

```{r revised-cases}
get_latest_revised_cases(covid, type = "confirmed") 
```


## 6. Looking at the data

`autoplot()` draws the usual grid, in green for the **epidemic** process
and red for the **reporting** one and yellow for the **revision**:

```{r autoplot, fig.width=11, fig.height=13, message=FALSE, warning=FALSE, fig.alt="A diagnostic grid of the weekly COVID-19 data: delay distribution, epidemic curve, calendar effects and periodograms."}
autoplot(covid)
```


## 7. Nowcasting

The 
[`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) package
includes functions for revised data with its `revision_process()`

```{r engine, eval = has("diseasenowcasting"), message=FALSE}
library(diseasenowcasting)

covid_fit <- covid |> 
  run_nowcast(engine = engine_diseasenowcasting(label = "hsgp"))
```

We can see the results with `autoplot()` which shows a more
conservative nowcast of the overall confirmed vs the total number of observed:

```{r fit-plot, eval = has("diseasenowcasting"), message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5, fig.alt="Nowcast by sex: grey reported columns with a green prediction fan over the final weeks."}
autoplot(covid_fit, date_lim = c(ymd("2020-08-01"), ymd("2020-10-05")))
```

`tidy()` gives the same thing as a table:

```{r, eval=FALSE}
tidy(covid_fit)
```

```{r fit-tidy, echo = FALSE, eval = has("diseasenowcasting")}
tidy(covid_fit) |> arrange(event_date) |> tail()
```

## 8. Backtesting

`nowcast_backtest()` walks back through time: for each past date it
truncates the data to what was known **then**, refits every engine, and
scores the result against what the series eventually reached.

We will compare our model against a second `diseasenowcasting` configuration.
Both carry the same confirmation-only revision process; they differ in the
**epidemic** components  with the new one being an autorregresive model:

```{r engine2, eval = has("diseasenowcasting")}
engine_ar <- engine_diseasenowcasting(
  model = model(
    epidemic = ar1_epidemic(),
  ),
  label = "ar1" #We add the label for the backtest
) 
```


```{r fit2, eval = has("diseasenowcasting"), message=FALSE, warning=FALSE}
covid_fit_ar <- covid |> run_nowcast(engine = engine_ar)
```

```{r backtest, eval = has("diseasenowcasting"), message=FALSE, warning=FALSE}
covid_backtest <- covid |>
  nowcast_backtest(
    engine_diseasenowcasting(label = "hsgp"),
    engine_ar,
    now_dates = ymd(c("2020-08-16", "2020-08-30", "2020-09-13")),
    seed      = 20260909
  )
```

Printing it gives the weighted interval score (`wis`), the median absolute
error (`ae_median`) and the coverage of the nominal intervals:

```{r backtest-print, eval = has("diseasenowcasting")}
covid_backtest
```


## 9. Ensembling

An ensemble averages models, on the theory that no single one is right.
The members here are the two fits, and the backtest supplies the weights:

```{r ensemble, eval = has("diseasenowcasting"), message=FALSE, warning=FALSE}
covid_ensemble <- nowcast_ensemble(
  covid_fit,
  covid_fit_ar,
  weights  = "inverse_score",
  backtest = covid_backtest
)
```

`weights = "inverse_score"` gives each member a weight inversely
proportional to its backtest score, so the members are matched to their
scores:

```{r ensemble-plot, eval = has("diseasenowcasting"), message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5, fig.alt="The ensemble nowcast: grey reported columns with a green prediction fan."}
autoplot(covid_ensemble, date_lim = c(ymd("2020-08-01"), ymd("2020-10-05")))
```

For ensembles of more than two models, and for what the different
weighting schemes do, see the article on [ensemble
nowcasting](https://rodrigozepeda.github.io/tbl.now/articles/ensemble-nowcasting.html).

## Summary

What the third date added to the workflow:

1.  **A declaration**: `revision_date`, `revision_type` and
    `revision_levels` in the `tbl_now()` call.

2.  **A rule**: a `"pending"` row carries **no** revision date. Anything
    that does carry one is settled, and its outcome must be `"confirmed"`
    or `"retracted"`.

3.  **Descriptions**: `prop_revision_type()`, `plot_revision_status()`,
    `plot_delay_distribution(axis = "revision")` and the `*_revised_cases()`
    getters.

4.  **A test**: `diagnose_revision_delay()`, which asks whether the
    outcome changes how long a case takes to settle.

Everything else is the same as it was with two dates.

```{r learning-more, echo=FALSE, results="asis"}
# The fragment is looked up by ABSOLUTE path via system.file(), never by a
# relative one. See the note in `example.Rmd` for why.
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```
