---
title: "The nowcasting workflow: hospital-acquired infections in Bucaramanga, Colombia"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{The nowcasting workflow: hospital-acquired infections in Bucaramanga, Colombia}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 150,
  fig.align = "center"
)
options(tibble.width = Inf, width = 160, pillar.width = 160)

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)

# epinowcast compiles a Stan model, which takes minutes. Flip to TRUE to fit it.
run_stan <- FALSE
```

This article is an end-to-end walk-through of the `tbl.now` workflow on
a real, **deliberately unpolished** dataset. We will:

1.  **Build** a `tbl_now` from a messy surveillance extract and show how
    to `diagnose()` possible data errors.
2.  **Clean** what it reported: duplicate records, missing dates, etc.
3.  **Look** at the data with `autoplot()`, and the `summary()`
    functions.
4.  **Describe** the behaviour of the reporting delay.
5.  **Attach temporal effects** to a model using the data.
6.  **Nowcast** with two different engines using everything we found.

We'll start the process by nowcasting with two dates (event and report
dates) and then we'll focus on nowcasting with three dates (event,
report, and validation dates).

Let's start by calling the libraries:

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(tbl.now)
```

## The data

`hai_bucaramanga` is a raw line list of **healthcare-associated
infections** (IAAS, *Infecciones Asociadas a la Atención en Salud*)
notified in the county of Bucaramanga, Sandander, Colombia, as published
by the county. Each row is one infection: a specimen taken from a
hospitalised patient, the laboratory result, and the microorganism
isolated.

```{r load-data}
data(hai_bucaramanga)
```

```{r}
hai_bucaramanga
```

For our example, six columns matter:

-   `id` — a unique patient identifier.
-   `specimen_date` — the **event date**, when the sample was taken.
-   `report_date` — the **report date**, when the laboratory issued the
    result.
-   `received_date` — the **validation date**, when the report was
    received by INS.
-   `microorganism` - the species of the organism isolated in the
    culture.
-   `sex` - the sex of the individual.

For the tutorial, we will do perform a nowcast assuming we are standing
on July 18th 2022. For that purpose we filter our data to what it would
have looked like:

```{r}
hai_bucaramanga <- hai_bucaramanga |>
  filter((is.na(specimen_date) | specimen_date <= ymd("2022/07/18")), 
         (is.na(report_date)   | report_date   <= ymd("2022/07/18")),
         (is.na(received_date) | received_date <= ymd("2022/07/18")))
```

::: {.alert .alert-warning}
For this tutorial we are keeping those events that have missing dates.
That is what the `is.na()` section on the left side of the filter is
doing as we keep either cases by July 18th 2022 or missing.
:::

## 1. Initial data cleaning

In this tutorial we show how to discover several anomalies in the data
as well as several modeling indicators. We suggest, however, that before
transforming the dataset into a `tbl_now` you use traditional data
analysis tools to clean the data first. Here we'll just select the
columns that matter to us and move onto the `tbl_now` capabilities.
However if you already have a pipeline for cleaning and preparing your
data this is where you would do it as we'll see the `tbl_now`
diagnostics don't cover everything.

```{r}
#We keep only the columns of interest
hai_bucaramanga <- hai_bucaramanga |> 
  select(id, specimen_date, report_date, received_date, microorganism, sex)
```

````{=html}
<!--
## 0. Initial data transformation

The column `microorganism` contains all the isolated organism. We will
consider the following as our microorgainsms of interest and
recode them as `"confirmed"` (they will be confirmed cases of our disease).
Those that were classified as something else will be recoded
as `"retracted"` (they were suspected Klebsiella but not anymore). Finally
those with `NA` will be recoded as `"pending"` as we still don't know their result.

We note that for our analysis we will only focus on Klebsiella pneumoniae hence
solely code as **confirmed** the following microorganism entries: `Klebsiella pneumoniae`, 
`Klebsiella pneumoniae subsp. pneumoniae`, and `Klebsiella pneumoniae (multidrug-resistant)`.
The entry `Klebsiella pneumoniae/oxytoca` will be coded as **pending**. The remaining
microorganisms will be considered **retracted**:

```{r}
#We list those that we will consider "confirmed"
valid_organisms <- c("Klebsiella pneumoniae",  
                     "Klebsiella pneumoniae subsp. pneumoniae",  
                     "Klebsiella pneumoniae (multidrug-resistant)")

hai_bucaramanga <- hai_bucaramanga |> 
  mutate(validation = case_when(
    microorganism %in% valid_organisms               ~ "confirmed",
    is.na(microorganism)                             ~ "pending",
    microorganism == "Klebsiella pneumoniae/oxytoca" ~ "pending",
    .default = "retracted"
  ))
```
-->
````

## 2. Building the `tbl_now`

We'll create the `tbl_now()` object. For the beginning of the example,
we will declare two dates (`specimen_date` for the event and
`report_date`) for the report. We'll further specify that the data is in
line-list format (*i.e.* each row is a case), and stratify by `sex`.

```{r build-tbl-now-raw}
hai_bucaramanga <- hai_bucaramanga |>
  tbl_now(
    event_date   = specimen_date,
    report_date  = report_date,
    strata       = sex,
    data_type    = "linelist",
    event_units  = "days",
    report_units = "days"
  )
```

The `tbl.now()` automatically fires several warnings. Let's take a look!

## 3. What is wrong?

These same warnings can also be accessed with the `diagnose()` function
which returns the warnings as well as additional notes:

```{r diagnose-raw}
diagnose(hai_bucaramanga) 
```

Let's see what they identify.

### The warnings

In general the warnings will look for missing data within the dates and
the strata or for violations of the nowcasting hypotheses (for example
reports that "see the future" and identify an event before it happens!).
In our case, specifically we identify missing dates in both the report
and the event as well as some reports that have incorrect dates.

There is no solution that works in all cases for these problems and
oftentimes the ideal solution is to identify the reason for the
misingness. Here we posit some approaches for the warnings:

1.  **For simultaneosuly missing event and report dates** there is
    nothing one can do. In those cases we don't know when something
    happened (to add to the epidemic process) or when it was reported
    (to add to the delay process). Hence we remove those rows.

```{r}
#Remove those that have both dates missing
hai_bucaramanga <- hai_bucaramanga |> 
  filter(!(is.na(report_date) & is.na(specimen_date)))
```

2.  **For the rows with missing reports, we can censor them above**.
    This will tell the models in
    [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/)
    that the measurement is not exact; instead, that we know the report
    arrived before a certain date though we don't know that *exact* date
    in particular. We can do this because if we are seeing the reports
    **today** then we know that at most they took until **today** to
    arrive. This censoring will add a column `.is_censored_report` which
    will indicate that that report is censored.

When calling the function we specify that the ones we are censoring are
the missing values and that the date they should input is "2022/10/18"
which corresponds to the date of the nowcast in this example.

```{r, warning=FALSE, message=FALSE}
hai_bucaramanga <- hai_bucaramanga |> 
  censor_reports(is.na(report_date), to_report = as.Date("2022/07/18"))
```

3.  **For the delays before the reports we visualize the data**. Doing a
    quick analysis we see that its a date formatting issue. You can see
    in the first entries that the specimen date is in format
    `year-day-month` (read with `lubridate::ydm()`) while the
    `report_date` and `received_date` is in format `year-month-day`
    (read with `lubridate::ymd()`) .

```{r, eval=TRUE}
hai_bucaramanga |> filter(report_date < specimen_date) 
```

We can correct the date format to fix this issue:

```{r, warning=FALSE}
hai_bucaramanga <- hai_bucaramanga |> 
  mutate(specimen_date = if_else(specimen_date > report_date, 
                                 ydm(as.character(specimen_date)), 
                                 ymd(as.character(specimen_date))))
```

3.  Currently there is **nothing we can do for the cases were the event
    date is unknown**. Hence we just document them and we'll remove from
    the dataset given that they represent such a small percent
    (`r 100*(hai_bucaramanga |> filter(is.na(specimen_date)) |> nrow())/nrow(hai_bucaramanga)`%).

```{r, warning=FALSE}
hai_bucaramanga <- hai_bucaramanga |> filter(!is.na(specimen_date))
```

Re-running `diagnose()` confirms we solved this issues:

```{r}
diagnose(hai_bucaramanga)
```

### The notes

The notes convey additinal information that, at a glance, can tell you
whether something needs to be corrected in your data.

It helps you know that you have columns that are not strata or
covariates (in case you forgot to declare them!):

```         
declarations/undeclared: 3 columns "id", "received_date", and "microorganism" are not declared as strata or covariates.
```

It quantifies how much time has happened between the last observation
and the now for each strata:

```         
now/now_gap_event [Female]: The last event date is 32 days before now ("2022-07-18").
```

It also tells you how what is the smallest stratum (in case you have
strata with so few observations you need to collapse them into another
one):

```         
strata/size [Female]: The smallest stratum is "Female" with 195 cases, 37.7% of the total.
```

And how sparse your data is:

```         
strata/sparsity [Female]: The sparsest stratum is "Female": 777 of the 929 "days" between the minimum event (2020-01-02) and the now (2022-07-18) carry no cases at all (83.6%, against 64.5% pooled over every stratum).
```

This is particularly useful for triangle-based methods that require to
fill with zeroes the whole reporting triangle.

## 4. More cleaning

Besides the `diagnose()` one always has to do some additional cleaning
of datasets. One particular issue is for repeated rows; that is when an
individual is registered twice in the dataset for the same disease at
the same exact times.

Here we can see that the same individuals `id` are sometimes repeated
for the same event reporting the same microorganism:

```{r dupes}
hai_bucaramanga |>
  count(id, microorganism, report_date, specimen_date) |>
  filter(n > 1) |> 
  as_tibble()
```

We can use `dplyr::distinct()` to remove them without losing
information:

```{r clean-distinct}
hai_bucaramanga <- hai_bucaramanga |> distinct()
```

## 5. Temporal effects

Temporal covariates can be added to a `tbl_now` so that we can describe
their effect or use them for nowcasting. They can be specified via the
`temporal_effects()` function. It can add holidays, weekends, day of the week 
effects and more. For the purpose of this tutorial we will add the weekends
and check whether there is a weekend effect or not:

```{r}
temporal_effects(weekend = TRUE)
```

You can add the effects to the `tbl_now()` with the `add_temporal_effects()` function:

```{r}
hai_bucaramanga <- hai_bucaramanga |> 
  add_temporal_effects(t_effects = temporal_effects(weekend = TRUE))

hai_bucaramanga
```

The temporal effect gets added lazily. You can use
`compute_temporal_effects()` to show it as a column in case you require it. 

```{r}
hai_bucaramanga
```


## 6. Looking at the data

`autoplot()` draws a diagnostic grid with the panels in green describing
the **epidemic** process and the red ones, the **reporting** process.

```{r autoplot, fig.width=11, fig.height=13, message=FALSE, warning=FALSE, fig.alt="A six-panel diagnostic grid: empirical delay distribution, observed epidemic curve, month-of-year effects for cases and for reporting, and periodograms for both processes."}
autoplot(hai_bucaramanga)
```

Four things stand out, and each of them changes a modelling decision
later.

**The epidemic curve is extremely sparse.** There are only a handful of
cases a year with most of the times 0 or 1 cases being reported.

```{r delay-dist, fig.width=9, fig.height=4.5, fig.alt="Epidemic process."}
plot_epidemic_process(hai_bucaramanga)
```

> **Note** This is the same sparcity the `note` in the `diagnostic()`
> function gave us.

**The delay distribution has some extreme delays.** This is the most
consequential feature of the dataset, so it is worth its own plot:

```{r delay-dist2, fig.width=9, fig.height=4.5, fig.alt="Histogram of the reporting delay."}
plot_delay_distribution(hai_bucaramanga)
```

A lot of results are reported way after 500 days!

**There doesn't seem to be a day-of-the-week effect** In both reports
and events there is no aparent day of the week effect.

```{r delay-dist3, fig.width=9, fig.height=4.5, fig.alt="Effects in the week"}
plot_day_of_week_effects(hai_bucaramanga)
```

**One week had a lot of reports** Epiweek 29 is an anomaly in the amount
of reports it seems to have. This would warrant further investigation.

```{r delay-dist4, fig.width=9, fig.height=4.5, fig.alt="Effects on the report"}
plot_week_of_year_effects(hai_bucaramanga, type = "report")
```

**There doesn't seem to be a weekend effect either** In both reports
and events there is no aparent effect of a weekend.

```{r delay-dist5, fig.width=9, fig.height=4.5, fig.alt="Effects in the week"}
plot_weekend_effects(hai_bucaramanga)
```

## 7. Summarising the data

The `summary()` of a `tbl_now` quantifies what we see in the plots. You
can either call `summary()` all at once:

```{r summary-delay, eval = FALSE}
summary(hai_bucaramanga) 
```

or as we do here call each of the summaries we are going to discuss (see
`?nowcast_summary_components` for the whole list).

We can start with the delays where we can see that up until the 3rd
quartile delays are less than 100 days but then they explode by the 90th
quantile:

```{r}
delay_summary(hai_bucaramanga)
```

The `zero_run_summary()` quantifies the sparcity by counting how many
days with continuous zeroes happened for either the event or the report
dates. That is, it measures the distribution of days of continuous
zeroes. You can see that usually there are on average
`r zero_run_summary(hai_bucaramanga)$mean[1]` continuous zeroes before a
case. However there have been runs of
`r zero_run_summary(hai_bucaramanga)$max[1]` continuous days with no
cases (continuous zeroes) until one case:

```{r}
zero_run_summary(hai_bucaramanga)
```

The composition block shows that there are way less females than males.
Which, again was already observed in the notes as the smaller stratum:

```{r summary-strata}
prop_strata(hai_bucaramanga)
```

Finally the cases per date shows how many cases on average occur per
date-axis. That is on average how many cases there are per event date or
per report date. In this case we can see that at least half of the dates
have zero cases:

```{r}
cases_per_date(hai_bucaramanga)
```

## Intermezzo: The first nowcast

Even with all this sparcity here we can nowcast using the
diseasenowcasting package:

```{r}
hai_fit <- hai_bucaramanga |> 
  run_nowcast(engine = engine_diseasenowcasting(temporal_effects = "none"))
```

Due to the sparcity, the nowcast predicts almost no cases at any time
with just maybe a second case at the now for males:

```{r, message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5, fig.alt="A sparse nowcast"}
autoplot(hai_fit) 
```

## 8. Changing to weekly data

Though nowcasting is possible with sparse data, oftentimes people prefer
to rescale data to a coarser time frame. Here we'll rescale the dataset
to a weekly reporting time frame with `aggregate_time_units()`:

```{r}
hai_bucaramanga <- hai_bucaramanga |> 
  aggregate_time_units(to = "weeks")
```

This aggregation of the epidemic process allows the nowcasting models to
infer more from it as you can see a more clear epidemic shape in the
autoplot:

```{r, fig.width=11, fig.height=13, message=FALSE, warning=FALSE, fig.alt="The now auto plot with weekly data"}
autoplot(hai_bucaramanga)
```

One should run again the `summary()` in the aggregated to draw
conclusions from this data. For us, however it will suffice to see that
the sparcity has reduced to usually just 1 week with zero cases before
seeing a week with cases so we can move on with the tutorial:

```{r}
zero_run_summary(hai_bucaramanga)
```

## 9. Is the reporting delay stable?

A nowcast might assume that the delay distribution is the same
throughout. Here we check those assumptions:

First, we just look at the delay aggregated through time

```{r drift-plot, fig.width=9, fig.height=5.5, message=FALSE, warning=FALSE, fig.alt="Fan chart of the reporting delay over time."}
plot_delay_drift(hai_bucaramanga)
```

The median moves from an extremely wide distribution in 2020 to a
shorter delay after 2021. We can confirm that split of two delay
behaviours with the following tests:

### Testing for the delay's drift

The `diagnose_drift()` function runs an autocorrelation-robust
Mann-Kendall test for a *monotonic* trend in the delay:

```{r drift-test, warning=FALSE}
diagnose_drift(hai_bucaramanga)
```

Here the median shows a reduction (`sens_slope` of
`r diagnose_drift(hai_bucaramanga)$sens_slope[1]`) which is consistent
with the plot. The `-1` can be interpreted as the decrease in the median
per unit of time. In this case, the median delay reduced 1 week per
week.

### Is there an abrupt shift?

A trend test cannot see a steep change like a jump. The
`diagnose_changepoint()` function uses Pettitt's test to look for
exactly that.

```{r changepoint-test, warning=FALSE}
diagnose_changepoint(hai_bucaramanga)
```

Here neither the median nor the spread have a changepoint. The change
was a drift in the trend as we saw in the plot and not an abrupt shift.
You can see the section on [delay
changes](https://rodrigozepeda.github.io/tbl.now/articles/diagnosing-a-tbl-now.html#part-3-what-needs-a-statistical-test-reporting-artefacts)
of the [Diagnosing a
tbl_now](https://rodrigozepeda.github.io/tbl.now/articles/diagnosing-a-tbl-now.html)
article for more information.

## 10. Are there any reporting batches?

A reporting batch consists on several reports that were all held for any
reason by the reporter (e.g. a laboratory) and then reported all
together as a group so that there is a giant signal in the reporting
process. To see them we can start by plotting against the reporting
axis:

```{r,  fig.width=9, fig.height=4.8}
plot_reporting_process(hai_bucaramanga)
```

We can see that there were several reports that dropped near the end of
2022. The reporting hexamap allows us to visualize the date of the
report at the same time as the day of the event and the corresponding
delay

```{r,  fig.width=9, fig.height=4.8}
plot_reporting_hexamap(hai_bucaramanga)
```


::: {.alert .alert-info}
The batch at `2022-07-17` was actually introduced by us by censoring all
the cases at that date in section 3.
:::

Here lies a figure with 3 axis. We can see for example our censoring at `2022-07-17` 
as the collection of points in the same axis. Values shared by the same
report axis with a high delay(i.e. far from the event-date axis might correspond
to reporting batches)

::: {.alert .alert-info}
You can also use

```{r, eval = FALSE}
plot_reporting_triangle(hai_bucaramanga)
```

to visualize the batches; however for this specific case they are easier
to visualize in the hexamap.
:::

Either way we can see that `2022-07-17` is a candidate for possible
batches. In this package, the `diagnose_batches()` function helps
identify potential dates with batches. We recommend using it with or
without the strata depending on whether the context warrants the
possibility that batching could differ by strata (e.g. if strata
correspond to healthcare facilities maybe yes; if they correspond to
gender of the patient maybe no).

Here we run it without the strata which we remove first. We'll keep only
those observations that have been identified as potential batches:

```{r}
hai_bucaramanga |> 
  remove_all_strata() |> 
  diagnose_batches() |> 
  filter(batch)
```

This identifies some dates that might be batches due to their high
reported numbers. A second batch method is the `diagnose_batches2()`
function. This one takes a specific date and checks whether the reports
that arrived on a date came from systematically older dates. We can
combine the findings from the previous run to test some of them and add
evidence to which are batches:

```{r}
hai_bucaramanga |> 
  remove_all_strata() |> 
  diagnose_batches2(at = ymd("2021-11-28"))
```

Where we further identify the date of `2021-11-28` as a potential batch
with its `p_value < 0.003`.


## 11. Nowcasting

Everything from sections 1 to 10 was about deciding *what* to fit.
Three of those findings change the call we are about to make:

-   The object is **weekly** now (section 8), so `now` is the week of
    `r format(get_now(hai_bucaramanga))` and every delay corresponds
    to weeks. 
-   The reporting delay **drifts down** across the series. So 
    we'll utilize the data starting on 2021 where the delay
    has settled. 
    
```{r}
hai_bucaramanga <- hai_bucaramanga |> 
  filter(year(specimen_date) > 2020)
```

The `run_nowcast()` function takes the `tbl_now` and an
**engine**. An engine is one modelling package plus every argument that
package needs:

```{r nowcast-weekly}
hai_fit_weekly <- hai_bucaramanga |>
  run_nowcast(engine = engine_diseasenowcasting())
```

The `autoplot()` shows the nowcast with  gray columns representing what
had been reported by `now`, and the green fan the nowcast's prediction intervals.

```{r nowcast-weekly-plot, message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5, fig.alt="Weekly nowcast for each sex: grey columns of reported cases with a green fan of predicted cases over the final weeks."}
autoplot(hai_fit_weekly)
```

Calling `tidy()` gives the nowcast as a table:

```{r, eval=FALSE}
tidy(hai_fit_weekly)
```

```{r nowcast-weekly-tidy, echo = FALSE}
tidy(hai_fit_weekly) |> arrange(event_date) |> tail()
```

## 12. Evaluate your nowcast

One way to evaluate the nowcast is to go back in time and 
see, what would the nowcast have produced with the information
available at a past date and contrast it with the information
known by now. You can use `nowcast_backtest()` for that purpose:
for every past date it truncates the data to the reports that were 
available *then*, refits the nowcast,  and scores the result against what
was eventually registered.

To best evaluate the nowcast we need something to compare against. Hence
we fit a second model:  `baselinenowcast`. 

One can fit it with the same call just a different engine

```{r bln-error, error = TRUE, warning=FALSE, eval=FALSE}
hai_fit_baseline <- hai_bucaramanga |>
  to_count(to = "count-incidence") |> 
  complete_zeroes() |> 
  run_nowcast(engine = engine_baselinenowcast())
```


Now the backtest. It takes the **same engines** you fitted with, so
there is no second place to keep the arguments in step, and it refits
every engine once per date — with three dates and two engines that is
six fits:

```{r backtest, warning=FALSE, message=FALSE}
hai_backtest <- nowcast_backtest(
  hai_counts,
  engine_diseasenowcasting(temporal_effects = "none"),
  engine_baselinenowcast(max_delay = 20),
  now_dates = get_now(hai_counts) - 7 * c(12, 8, 4),
  seed      = 20260904
)

tidy(hai_backtest)
```

::: {.alert .alert-info}
Pass `seed` to the backtest rather than calling `set.seed()` before it.
`nowcast_backtest()` seeds immediately before each fit, from the engine's
label and the date it is fitting, so a fit depends only on *which* fit it
is. One `set.seed()` at the top only pins anything if every method draws
the same random numbers in the same order — which stops being true the
moment you drop an engine or add a date.
:::

`tidy()` returns one row per (method, retrospective `now`, target), which
goes straight into `dplyr`:

```{r backtest-scores}
tidy(hai_backtest) |>
  group_by(method) |>
  summarise(
    wis         = mean(wis),
    ae_median   = mean(ae_median),
    coverage_50 = mean(coverage_50),
    coverage_90 = mean(coverage_90)
  )
```

`wis` is the weighted interval score (lower is better; it rewards a
sharp interval only if the interval is also right), `ae_median` is the
absolute error of the median, and the two coverage columns are the share
of targets that fell inside the 50% and 90% intervals.

Two things to read here. `diseasenowcasting` scores better on both
accuracy measures, which is what you would hope for from a model that
has an epidemic process in it when the delay is drifting. And **both**
methods cover far more than the nominal 50%: on counts this small a
discrete interval jumps from covering too little to covering everything,
so over-coverage here is a property of the data, not a fault of either
model.

That table is also flattering to both of them, because it scores *every*
event date — and most of those weeks were settled long before the
retrospective `now`, so they are free points that any method gets right.
The sharper question is how the methods do on the weeks that were still
open:

```{r backtest-open}
tidy(hai_backtest) |>
  filter(event_date > now - 7 * 8) |>
  group_by(method) |>
  summarise(
    n           = n(),
    wis         = mean(wis),
    coverage_50 = mean(coverage_50),
    coverage_90 = mean(coverage_90)
  )
```

The gap widens, and it is stable across all three retrospective dates
rather than resting on one lucky week:

```{r backtest-plot, fig.width=9, fig.height=4, fig.alt="Grouped bar chart of the mean weighted interval score for each method at each of the three retrospective now dates."}
tidy(hai_backtest) |>
  filter(event_date > now - 7 * 8) |>
  group_by(method, now) |>
  summarise(wis = mean(wis), .groups = "drop") |>
  ggplot(aes(x = factor(now), y = wis, fill = method)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  labs(x = "Retrospective `now`", y = "Mean WIS (last 8 weeks)", fill = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

::: {.alert .alert-warning}
Three retrospective dates is few. It is enough to see that neither model
is broken and to notice a consistent ordering, but it is not enough to
claim one package beats the other on this kind of data. A backtest costs
`length(engines) x length(now_dates)` model fits, so lengthening
`now_dates` is the first thing to do when the conclusion matters.
:::

## 13. Ensemble nowcast

The backtest above ranked the two models, but it did not say to *drop*
either of them: they were built on different assumptions, and averaging
models that disagree usually beats picking the winner in advance —
especially when you must commit before you know which weeks are hard.

Because every `run_nowcast()` result has the same shape, combining them
is one call. We use three members: the two fits from section 12, and a
third that is the **same package with a different epidemic process**.
An ensemble is not only a hedge across packages — two structurally
different models from one package disagree in their own way, and that
disagreement is worth combining too:

```{r ar1-fit, warning=FALSE, message=FALSE}
hai_fit_ar1 <- hai_counts |>
  run_nowcast(engine = engine_diseasenowcasting(
    temporal_effects = "none",
    model = diseasenowcasting::model(
      epidemic = diseasenowcasting::ar1_epidemic()
    ),
    seed  = 20260904,
    label = "dnc_ar1"
  ))

members <- list(
  diseasenowcasting = hai_fit_dnc,
  dnc_ar1           = hai_fit_ar1,
  baselinenowcast   = hai_fit_bln
)

hai_ensemble <- nowcast_ensemble(members)
hai_ensemble
```

Both `diseasenowcasting` members report the same `method`, so it is the
**names** that tell them apart: the names of the `members` list here,
and an engine's `label` in a backtest. We give the AR(1) engine the same
`label` its list entry has so that the two line up when we come to
weights below.

By default every member counts the same, and the members are combined
**quantile by quantile**: the ensemble's 90th percentile is the average
of the members' 90th percentiles. That always works, and it tends to
come out *narrower* than the members. Pooling the draws instead treats
the ensemble as a mixture, which is usually wider and better calibrated
— it needs every member to have kept draws, and errors rather than
silently dropping one that has not:

```{r ensemble-pool, warning=FALSE, message=FALSE}
nowcast_ensemble(members, type = "linear_pool")
```

The result is an ordinary `tbl_nowcast`, so it plots and tidies like any
single fit:

```{r ensemble-plot, message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5, fig.alt="The ensemble nowcast: grey reported columns with a green fan over the final weeks, for each sex."}
autoplot(hai_ensemble)
```

### Weighting the members

Equal weights are a defensible default, but the backtest already knows
which member did better. `nowcast_weights()` turns it into weights, and
`nowcast_ensemble()` takes them directly.

One catch: the weights are matched to the members **by name**, and a
member whose name the backtest has never seen falls back to its *method*
— so `dnc_ar1` would silently inherit the plain `diseasenowcasting`
weight. If you want the two epidemic processes weighted separately, the
labelled engine has to be in the backtest as well:

```{r ensemble-backtest, warning=FALSE, message=FALSE}
hai_backtest_all <- nowcast_backtest(
  hai_counts,
  engine_diseasenowcasting(temporal_effects = "none"),
  engine_diseasenowcasting(
    temporal_effects = "none",
    model = diseasenowcasting::model(
      epidemic = diseasenowcasting::ar1_epidemic()
    ),
    label = "dnc_ar1"
  ),
  engine_baselinenowcast(max_delay = 20),
  now_dates = get_now(hai_counts) - 7 * c(12, 8, 4),
  seed      = 20260904
)

nowcast_weights(hai_backtest_all, type = "inverse_score")
```

`"inverse_score"` makes each weight proportional to `1 / mean WIS`. It is
the safe rule: monotone in performance, and it never collapses onto a
single model. The alternative, `type = "optim"`, searches for the weights
that would have *minimised* the ensemble's WIS over the backtest window;
it is better in principle but it overfits happily on three retrospective
dates, and here it puts the entire weight on a single member.

Either rule plugs straight into the ensemble:

```{r ensemble-weighted, warning=FALSE, message=FALSE}
nowcast_ensemble(
  members,
  weights  = "inverse_score",
  backtest = hai_backtest_all
)
```

The two `diseasenowcasting` variants turn out to be nearly
indistinguishable on this data — an AR(1) epidemic and the default
process agree when there is this little signal to disagree about — so
the learned weights sit close to equal, and the weighted ensemble lands
almost exactly where the unweighted one did. That is a perfectly good
outcome to report: on a few hundred infections spread thinly over two
and a half years, the honest answer is a handful of cases with a wide
interval, and three models agreeing on it is more reassuring than any
one of them being confident.

```{r learning-more, echo=FALSE, results="asis"}
# The fragment is looked up by ABSOLUTE path via system.file(), never by a
# relative one. pkgdown/rmarkdown render into an intermediates directory under
# tempdir() and copy relative resources alongside; a path containing "../.."
# escapes that directory ("cannot create file '/tmp/RtmpXXX/../../man/...'"),
# which fails wherever tempdir() sits near the filesystem root -- as it does on
# CI. `system.file()` also works under pkgload::load_all(), which shims it to
# find inst/ in the source tree.
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```
