---
title: "The nowcasting workflow: hospital-acquired infections in Bucaramanga, Colombia"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{The nowcasting workflow: hospital-acquired infections in Bucaramanga, Colombia}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 150,
  fig.align = "center"
)
options(tibble.width = Inf, width = 160, pillar.width = 160)

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)

# epinowcast compiles a Stan model, which takes minutes. Flip to TRUE to fit it.
run_stan <- FALSE
```

This article is an end-to-end walk-through of the `tbl.now` workflow on
a real, **deliberately unpolished** dataset. We will:

1.  **Build** a `tbl_now` from a messy surveillance extract and show how
    to `diagnose()` possible data errors.
2.  **Clean** what it reported: duplicate records, missing dates, etc.
3.  **Look** at the data with `autoplot()`, and the `summary()`
    functions.
4.  **Describe** the behaviour of the reporting delay.
5.  **Attach temporal effects** to a model using the data.
6.  **Nowcast** with two different engines using everything we found.

We'll start the process by nowcasting with two dates (event and report
dates) and then we'll focus on nowcasting with three dates (event,
report, and validation dates).

Let's start by calling the libraries:

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(tbl.now)
```

## The data

`hai_bucaramanga` is a raw line list of **healthcare-associated
infections** (IAAS, *Infecciones Asociadas a la Atención en Salud*)
notified in the county of Bucaramanga, Sandander, Colombia, as published
by the county. Each row is one infection: a specimen taken from a
hospitalised patient, the laboratory result, and the microorganism
isolated.

```{r load-data}
data(hai_bucaramanga)
```

```{r}
hai_bucaramanga
```

For our example, six columns matter:

-   `id` — a unique patient identifier.
-   `specimen_date` — the **event date**, when the sample was taken.
-   `report_date` — the **report date**, when the laboratory issued the
    result.
-   `sex` - the sex of the individual.

For the tutorial, we will do perform a nowcast assuming we are standing
on July 18th 2022. For that purpose we filter our data to what it would
have looked like:

```{r}
hai_bucaramanga <- hai_bucaramanga |>
  filter((is.na(specimen_date) | specimen_date <= ymd("2022/07/18")), 
         (is.na(report_date)   | report_date   <= ymd("2022/07/18")))
```

::: {.alert .alert-warning}
For this tutorial we are keeping those events that have missing dates.
That is what the `is.na()` section on the left side of the filter is
doing as we keep either cases by July 18th 2022 or missing.
:::

## 1. Initial data cleaning

In this tutorial we show how to discover several anomalies in the data
as well as several modeling indicators. We suggest, however, that before
transforming the dataset into a `tbl_now` you use traditional data
analysis tools to clean the data first. Here we'll just select the
columns that matter to us and move onto the `tbl_now` capabilities.
However if you already have a pipeline for cleaning and preparing your
data this is where you would do it as we'll see the `tbl_now`
diagnostics don't cover everything.

```{r}
#We keep only the columns of interest
hai_bucaramanga <- hai_bucaramanga |> 
  select(id, specimen_date, report_date, sex)
```



## 2. Building the `tbl_now`

We'll create the `tbl_now()` object. For the beginning of the example,
we will declare two dates (`specimen_date` for the event and
`report_date`) for the report. We'll further specify that the data is in
line-list format (*i.e.* each row is a case), and stratify by `sex`.

```{r build-tbl-now-raw}
hai_bucaramanga <- hai_bucaramanga |>
  tbl_now(
    event_date   = specimen_date,
    report_date  = report_date,
    strata       = sex,
    data_type    = "linelist",
    event_units  = "days",
    report_units = "days"
  )
```

The `tbl.now()` automatically fires several warnings. Let's take a look!

## 3. What is wrong?

These same warnings can also be accessed with the `diagnose()` function
which returns the warnings as well as additional notes:

```{r diagnose-raw}
diagnose(hai_bucaramanga) 
```

Let's see what they identify.

### The warnings

In general the warnings will look for missing data within the dates and
the strata or for violations of the nowcasting hypotheses (for example
reports that "see the future" and identify an event before it happens!).
In our case, specifically we identify missing dates in both the report
and the event as well as some reports that have incorrect dates.

There is no solution that works in all cases for these problems and
oftentimes the ideal solution is to identify the reason for the
misingness. Here we posit some approaches for the warnings:

1.  **For simultaneosuly missing event and report dates** there is
    nothing one can do. In those cases we don't know when something
    happened (to add to the epidemic process) or when it was reported
    (to add to the delay process). Hence we remove those rows.

```{r}
#Remove those that have both dates missing
hai_bucaramanga <- hai_bucaramanga |> 
  filter(!(is.na(report_date) & is.na(specimen_date)))
```

2.  **For the rows with missing reports, we can censor them above**.
    This will tell the models in
    [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/)
    that the measurement is not exact; instead, that we know the report
    arrived before a certain date though we don't know that *exact* date
    in particular. We can do this because if we are seeing the reports
    **today** then we know that at most they took until **today** to
    arrive. This censoring will add a column `.is_censored_report` which
    will indicate that that report is censored.

When calling the function we specify that the ones we are censoring are
the missing values and that the date they should input is "2022/10/18"
which corresponds to the date of the nowcast in this example.

```{r, warning=FALSE, message=FALSE}
hai_bucaramanga <- hai_bucaramanga |> 
  censor_reports(is.na(report_date), to_report = as.Date("2022/07/18"))
```

3.  **For the delays before the reports we visualize the data**. Doing a
    quick analysis we see that its a date formatting issue. You can see
    in the first entries that the specimen date is in format
    `year-day-month` (read with `lubridate::ydm()`) while the
    `report_date` and `received_date` is in format `year-month-day`
    (read with `lubridate::ymd()`) .

```{r, eval=TRUE}
hai_bucaramanga |> filter(report_date < specimen_date) 
```

We can correct the date format to fix this issue:

```{r, warning=FALSE}
hai_bucaramanga <- hai_bucaramanga |> 
  mutate(specimen_date = if_else(specimen_date > report_date, 
                                 ydm(as.character(specimen_date)), 
                                 ymd(as.character(specimen_date))))
```

3.  Currently there is **nothing we can do for the cases were the event
    date is unknown**. Hence we just document them and we'll remove from
    the dataset given that they represent such a small percent
    (`r 100*(hai_bucaramanga |> filter(is.na(specimen_date)) |> nrow())/nrow(hai_bucaramanga)`%).

```{r, warning=FALSE}
hai_bucaramanga <- hai_bucaramanga |> filter(!is.na(specimen_date))
```

Re-running `diagnose()` confirms we solved this issues:

```{r}
diagnose(hai_bucaramanga)
```

### The notes

The notes convey additinal information that, at a glance, can tell you
whether something needs to be corrected in your data.

It helps you know that you have columns that are not strata or
covariates (in case you forgot to declare them!):

```         
declarations/undeclared: 3 columns "id", "received_date", and "microorganism" are not declared as strata or covariates.
```

It quantifies how much time has happened between the last observation
and the now for each strata:

```         
now/now_gap_event [Female]: The last event date is 32 days before now ("2022-07-18").
```

It also tells you how what is the smallest stratum (in case you have
strata with so few observations you need to collapse them into another
one):

```         
strata/size [Female]: The smallest stratum is "Female" with 195 cases, 37.7% of the total.
```

And how sparse your data is:

```         
strata/sparsity [Female]: The sparsest stratum is "Female": 777 of the 929 "days" between the minimum event (2020-01-02) and the now (2022-07-18) carry no cases at all (83.6%, against 64.5% pooled over every stratum).
```

This is particularly useful for triangle-based methods that require to
fill with zeroes the whole reporting triangle.

## 4. More cleaning

Besides the `diagnose()` one always has to do some additional cleaning
of datasets. One particular issue is for repeated rows; that is when an
individual is registered twice in the dataset for the same disease at
the same exact times.

Here we can see that the same individuals `id` are sometimes repeated:

```{r dupes}
hai_bucaramanga |>
  count(id, report_date, specimen_date) |>
  filter(n > 1) |> 
  as_tibble()
```

We can use `dplyr::distinct()` to remove them without losing
information:

```{r clean-distinct}
hai_bucaramanga <- hai_bucaramanga |> distinct()
```

## 5. Temporal effects

Temporal covariates can be added to a `tbl_now` so that we can describe
their effect or use them for nowcasting. They can be specified via the
`temporal_effects()` function. It can add holidays, weekends, day of the week 
effects and more. For the purpose of this tutorial we will add the weekends
and check whether there is a weekend effect or not:

```{r}
temporal_effects(weekend = TRUE)
```

You can add the effects to the `tbl_now()` with the `add_temporal_effects()` function:

```{r}
hai_bucaramanga <- hai_bucaramanga |> 
  add_temporal_effects(t_effects = temporal_effects(weekend = TRUE))

hai_bucaramanga
```

The temporal effect gets added lazily. You can use
`compute_temporal_effects()` to show it as a column in case you require it (see them at the end of the tbl_now). 

```{r}
hai_bucaramanga |> compute_temporal_effects()
```


## 6. Looking at the data

`autoplot()` draws a diagnostic grid with the panels in green describing
the **epidemic** process and the red ones, the **reporting** process.

```{r autoplot, fig.width=11, fig.height=13, message=FALSE, warning=FALSE, fig.alt="A six-panel diagnostic grid: empirical delay distribution, observed epidemic curve, month-of-year effects for cases and for reporting, and periodograms for both processes."}
autoplot(hai_bucaramanga)
```

Four things stand out, and each of them changes a modelling decision
later.

**The epidemic curve is extremely sparse.** There are only a handful of
cases a year with most of the times 0 or 1 cases being reported.

```{r delay-dist, fig.width=9, fig.height=4.5, fig.alt="Epidemic process."}
plot_epidemic_process(hai_bucaramanga)
```

> **Note** This is the same sparcity the `note` in the `diagnostic()`
> function gave us.

**The delay distribution has some extreme delays.** This is the most
consequential feature of the dataset, so it is worth its own plot:

```{r delay-dist2, fig.width=9, fig.height=4.5, fig.alt="Histogram of the reporting delay."}
plot_delay_distribution(hai_bucaramanga)
```

A lot of results are reported way after 500 days!

**There doesn't seem to be a day-of-the-week effect** In both reports
and events there is no aparent day of the week effect.

```{r delay-dist3, fig.width=9, fig.height=4.5, fig.alt="Effects in the week"}
plot_day_of_week_effects(hai_bucaramanga)
```

**One week had a lot of reports** Epiweek 29 is an anomaly in the amount
of reports it seems to have. This would warrant further investigation.

```{r delay-dist4, fig.width=9, fig.height=4.5, fig.alt="Effects on the report"}
plot_week_of_year_effects(hai_bucaramanga, type = "report")
```

**There doesn't seem to be a weekend effect either** In both reports
and events there is no aparent effect of a weekend.

```{r delay-dist5, fig.width=9, fig.height=4.5, fig.alt="Effects in the week"}
plot_weekend_effects(hai_bucaramanga)
```

## 7. Summarising the data

The `summary()` of a `tbl_now` quantifies what we see in the plots. You
can either call `summary()` all at once:

```{r summary-delay, eval = FALSE}
summary(hai_bucaramanga) 
```

or as we do here call each of the summaries we are going to discuss (see
`?nowcast_summary_components` for the whole list).

We can start with the delays. Here we can see that up until the 3rd
quartile (`q75`) delays are less than 100 days but then they explode by the 90th
quantile (`q90`):

```{r}
delay_summary(hai_bucaramanga)
```

The `zero_run_summary()` quantifies the sparcity by counting how many
days with continuous zeroes happened for either the event or the report
dates. That is, it measures the distribution of days of continuous
zeroes. You can see that usually there are on average
`r round(zero_run_summary(hai_bucaramanga)$mean[1],2)` continuous zeroes before a
case. However there have been runs of
`r zero_run_summary(hai_bucaramanga)$max[1]` continuous days with no
cases (continuous zeroes) until one case:

```{r}
zero_run_summary(hai_bucaramanga)
```

The composition block shows that there are way less females than males.
Which, again was already observed in the notes as the smaller stratum:

```{r summary-strata}
prop_strata(hai_bucaramanga)
```

Finally the cases per date shows how many cases on average occur per
date-axis. That is on average how many cases there are per event date or
per report date. In this case we can see that at least half of the dates
have zero cases:

```{r}
cases_per_date(hai_bucaramanga)
```

## Intermezzo: The first nowcast

Even with all this sparcity here we can nowcast using the
diseasenowcasting package:

```{r, eval = has("diseasenowcasting")}
hai_fit <- hai_bucaramanga |> 
  run_nowcast(engine = engine_diseasenowcasting(temporal_effects = "none"))
```

Due to the sparcity, the nowcast predicts almost no cases at any time
with just maybe a second case at the now for males:

```{r, message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5, fig.alt="A sparse nowcast"}
autoplot(hai_fit) 
```

## 8. Changing to weekly data

Though nowcasting is possible with sparse data, oftentimes people prefer
to rescale data to a coarser time frame (and some packages can't handle the sparcity either). 
Here we'll rescale the dataset to a weekly reporting time frame with `aggregate_time_units()`:

```{r}
hai_bucaramanga <- hai_bucaramanga |> 
  aggregate_time_units(to = "weeks")
```

This aggregation of the epidemic process allows the nowcasting models to
infer more from it as you can see a clearer epidemic-process shape:

```{r, fig.width=11, fig.height=13, message=FALSE, warning=FALSE, fig.alt="The now auto plot with weekly data"}
autoplot(hai_bucaramanga)
```

One should run again the `summary()` and `diagnose()` in the aggregated to draw
conclusions from this data. For us, however it will suffice to see that
the sparcity has reduced to usually just 1 week with zero cases before
having cases again:

```{r}
zero_run_summary(hai_bucaramanga)
```

Let's move on with the tutorial. 

## 9. Is the reporting delay stable?

A nowcast might assume that the delay distribution is the same
throughout. Here we check those assumptions:

First, we just look at the delay aggregated through time

```{r drift-plot, fig.width=9, fig.height=5.5, message=FALSE, warning=FALSE, fig.alt="Fan chart of the reporting delay over time."}
plot_delay_drift(hai_bucaramanga)
```

The median moves from an extremely wide distribution in 2020 to a
shorter delay after 2021. We can confirm that split of two delay
behaviours with the following tests:

### Testing for the delay's drift

The `diagnose_drift()` function runs an autocorrelation-robust
Mann-Kendall test for a *monotonic* trend in the delay:

```{r drift-test, warning=FALSE}
diagnose_drift(hai_bucaramanga)
```

Here the median shows a reduction (`sens_slope` of
`r diagnose_drift(hai_bucaramanga)$sens_slope[1]`) which is consistent
with the plot. The `-1` can be interpreted as the decrease in the median
per unit of time. In this case, the median delay reduced 1 week per
week.

### Is there an abrupt shift?

A trend test cannot see a steep change like a jump. The
`diagnose_changepoint()` function uses Pettitt's test to look for
exactly that.

```{r changepoint-test, warning=FALSE}
diagnose_changepoint(hai_bucaramanga)
```

Here neither the median nor the spread have a changepoint. The change
was a drift in the trend as we saw in the plot and not an abrupt shift.
You can see the section on [delay
changes](https://rodrigozepeda.github.io/tbl.now/articles/diagnosing-a-tbl-now.html#part-3-what-needs-a-statistical-test-reporting-artefacts)
of the [Diagnosing a
tbl_now](https://rodrigozepeda.github.io/tbl.now/articles/diagnosing-a-tbl-now.html)
article for more information.

## 10. Are there any reporting batches?

A reporting batch consists on several reports that were all held for any
reason by the reporter (e.g. a laboratory) and then reported all
together as a group so that there is a giant signal in the reporting
process. To see them we can start by plotting against the reporting
axis:

```{r,  fig.width=9, fig.height=4.8}
plot_reporting_process(hai_bucaramanga)
```

We can see that there were several reports that dropped near the end of
2022. The reporting hexamap allows us to visualize the date of the
report at the same time as the day of the event and the corresponding
delay

```{r,  fig.width=9, fig.height=4.8}
plot_reporting_hexamap(hai_bucaramanga)
```


::: {.alert .alert-info}
The batch at `2022-07-17` was actually introduced by us by censoring all
the cases at that date in section 3.
:::

Here lies a figure with 3 axis. We can see for example our censoring at `2022-07-17` 
as the collection of points in the same axis. Values shared by the same
report axis with a high delay might be batches (_i.e._ several reported at once
with some far from the event-date  axis). You can see for example a point with delay
around 32 weeks just before a report date of October 31st that corresponds to
an event somewhere between November 2020 and Match 2021. This has a high delay 
and several more points were reported at that report date. We'll come back to that. 

::: {.alert .alert-info}
You can also use

```{r, eval = FALSE}
plot_reporting_triangle(hai_bucaramanga)
```

to visualize the batches; however for this specific case they are easier
to visualize in the hexamap.
:::

In this package, the `diagnose_batches()` function helps
identify potential dates with batches. We recommend using it with or
without the strata depending on whether the context warrants the
possibility that batching could differ by strata (e.g. if strata
correspond to healthcare facilities maybe yes; if they correspond to
gender of the patient maybe no).

Here we run it without the strata which we remove first. We'll keep only
those observations that have been identified as potential batches:

```{r}
hai_bucaramanga |> 
  remove_all_strata() |> 
  diagnose_batches(drop_censored = TRUE) 
```

This identifies some dates that might be batches due to their high
reported numbers. A second batch method is the `diagnose_batches2()`
function. This one takes a specific date and checks whether the reports
that arrived on a date came from systematically older dates. We can
combine the findings from the previous run to test some of them and add
evidence to which are batches:

```{r}
hai_bucaramanga |> 
  remove_all_strata() |> 
  diagnose_batches2(at = ymd("2021-11-28"))
```

Where we further identify the date of `2021-11-28` as a potential batch
with its `p_value < 0.003`.

Given that the 


## 11. Nowcasting

Everything from sections 1 to 10 was about deciding *what* to fit.
Three of those findings change the call we are about to make:

-   The object is **weekly** now (section 8), so `now` is the week of
    `r format(get_now(hai_bucaramanga))` and every delay corresponds
    to weeks. 
-   The reporting delay **drifts down** across the series. So 
    we'll utilize the data starting on 2021 where the delay
    has settled. 
    
```{r}
hai_bucaramanga <- hai_bucaramanga |> 
  filter(year(specimen_date) > 2020)
```

The `run_nowcast()` function takes the `tbl_now` and an
**engine**. An engine contains a modelling package's nowcasting
function plus every argument that it needs:

```{r nowcast-weekly, eval= has("diseasenowcasting")}
hai_fit_weekly <- hai_bucaramanga |>
  run_nowcast(engine = engine_diseasenowcasting())
```

The `autoplot()` shows the nowcast with  gray columns representing what
had been reported by `now`, and the green fan the nowcast's prediction intervals.

```{r nowcast-weekly-plot,eval= has("diseasenowcasting"), message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5, fig.alt="Weekly nowcast for each sex: grey columns of reported cases with a green fan of predicted cases over the final weeks."}
autoplot(hai_fit_weekly)
```

Calling `tidy()` gives the nowcast as a table:

```{r, eval=FALSE}
tidy(hai_fit_weekly)
```

```{r nowcast-weekly-tidy, echo = FALSE, eval= has("diseasenowcasting")}
tidy(hai_fit_weekly) |> arrange(event_date) |> tail()
```

## 12. Evaluate your nowcast

One way to evaluate the nowcast is to go back in time and 
see, what would the nowcast have produced with the information
available at a previous date contrasting it with the information
known by now. You can use `nowcast_backtest()` for that purpose:
for every past date it truncates the data to the reports that were 
available *then*, refits the nowcast,  and scores the result against what
was eventually observed.

To best evaluate the nowcast we need something to compare against. Hence
we will fit a second model:  `baselinenowcast`. 

One can fit it with the same call just a different engine

```{r bln-error, error = TRUE, warning=FALSE, eval=has("baselinenowcast"), message=FALSE}
hai_fit_baseline <- hai_bucaramanga |>
  run_nowcast(engine = engine_baselinenowcast())
```

Any backtest uses the  **same engines** refitting 
every engine once per past date. Here we show the backtest with just 3 past dates.
On a real scenario we recommend using as many as possible to draw any conclusions:

```{r backtest, warning=FALSE, message=FALSE, eval=has("baselinenowcast") & has("diseasenowcasting")}
hai_backtest <- hai_bucaramanga |> 
  nowcast_backtest(
    engine_diseasenowcasting(),
    engine_baselinenowcast(),
    now_dates = c(ymd("2022-04-24"), ymd("2022-05-22"), ymd("2022-06-19")),
    seed      = 20260904
  )
```
Printing the object shows the results from the backtest which includes the individual results for the weighted interval score (`wis`), the median absolute error (`ae_median`), and the coverage of the 50 and 90% intervals (`coverage_*`):

```{r}
hai_backtest
```

In this scenario,  `diseasenowcasting` scores better on both
accuracy measures and **both** methods cover far more than the nominal 90.
We caution however that we cannot conclude anything from just 3 data points
and in a real scenario one should use more. 

Date-level indicators for the `wis`, `ae_median` and coverage can be obtained with tidy:


```{r, eval=FALSE}
tidy(hai_backtest)
```

```{r, eval=has("baselinenowcast") & has("diseasenowcasting")}
tidy(hai_backtest) |> arrange(event_date) |> tail()
```


## 13. Ensemble nowcast

An ensemble is a weighted average between models. It leverages the
fact that no model is perfect hence combining the predictions of each
might lead to a more robust nowcast. You can think of an ensemble as an average
of the models. 

To create an ensemble model one needs to pass the model
specifications (with a name) as well as the backtest:

```{r ensemble-weighted, warning=FALSE, message=FALSE, eval=has("baselinenowcast") & has("diseasenowcasting")}
hai_ensemble <- nowcast_ensemble(
  hai_fit_weekly,
  hai_fit_baseline,
  weights  = "inverse_score",
  backtest = hai_backtest
)
```

The same `tidy()` and `autoplot()` functions can be used here:

```{r, eval=has("baselinenowcast") & has("diseasenowcasting")}
autoplot(hai_ensemble)
```

One can fit ensembles with more than two models and different model specifications. Check out the article on [ensemble nowcasting](https://rodrigozepeda.github.io/tbl.now/articles/ensemble-nowcasting.html) for more information. 

## Summary

In this example we showed:
1. How to create a `tbl_now()` object.
2. How to `diagnose()` and summarise with `summary()` as well as visualize with `autoplot()` a `tbl_now()`.
3. How to add `temporal_effects()` to the object which can be displayed in the plots. 
4. How to aggregate the time units with `aggregate_time_units()` to pass from days to weeks.  
5. How to identify drifts and batches with `plot_delay_drift()`, `diagnose_drift()` and `diagnose_changepoint()`. 6. How to identify potential batches with `plot_reporting_hexamap()` in conjunction with `diagnose_batches()` (and `diagnose_batches2()`). 
7. How to perform a nowcast by setting an `engine()` and running with `run_nowcast()`.
8. How to clean with `tidy()` and `autoplot()` a nowcast.
9. How to backtest a nowcast with `nowcast_backtest()` and create an ensemble with `nowcast_ensemble()`.

::: {.alert .alert-tip}
If you have any questions regarding this article or comments please [open an issue](https://github.com/RodrigoZepeda/tbl.now/issues/new)
:::



```{r learning-more, echo=FALSE, results="asis"}
# The fragment is looked up by ABSOLUTE path via system.file(), never by a
# relative one. pkgdown/rmarkdown render into an intermediates directory under
# tempdir() and copy relative resources alongside; a path containing "../.."
# escapes that directory ("cannot create file '/tmp/RtmpXXX/../../man/...'"),
# which fails wherever tempdir() sits near the filesystem root -- as it does on
# CI. `system.file()` also works under pkgload::load_all(), which shims it to
# find inst/ in the source tree.
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```
