---
title: "A complete nowcasting workflow: hospital-acquired infections in Bucaramanga"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{A complete nowcasting workflow: hospital-acquired infections in Bucaramanga}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 150,
  fig.align = "center"
)
options(tibble.width = Inf, width = 160, pillar.width = 160)

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)

# epinowcast compiles a Stan model, which takes minutes. Flip to TRUE to fit it.
run_stan <- FALSE
```

This article is an end-to-end walk-through of the `tbl.now` workflow on a real,
**deliberately unpolished** dataset. We will:

1. **Build** a `tbl_now` from a messy surveillance extract — *before* cleaning
   it — and let `diagnose()` report what is wrong, rather than guessing.
2. **Clean** what it reported: duplicate records, missing dates, and reports
   that arrive *before* the event they describe. Then re-run the diagnosis to
   confirm the fix rather than assume it.
3. **Look** at the data with `autoplot()`, read the numbers off it with
   `summary()`, and use the standalone `plot_*()` diagnostics.
4. Ask whether the reporting delay is **stable** — does it drift, does it jump?
5. Attach **temporal effects** that the data actually justifies.
6. **Nowcast**, first with `diseasenowcasting` and then with five other engines,
   letting everything we found in steps 1–4 inform the model.

The thread running through all of it: *look before you model.* Every modelling
choice at the end is traceable to something we measured at the beginning.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr, quietly = TRUE)
library(lubridate)
library(ggplot2)
library(tbl.now)
```

## The data

`hai_bucaramanga` is a line list of **healthcare-associated infections** (IAAS,
*Infecciones Asociadas a la Atención en Salud*) notified in the municipality of
Bucaramanga, Colombia, published on the Colombian open-data portal. Each row is
one infection: a specimen taken from a hospitalised patient, the laboratory
result, and the microorganism isolated.

```{r load-data}
data(hai_bucaramanga)

glimpse(hai_bucaramanga)
```

For nowcasting, two columns matter:

* `specimen_date` — the **event date**, when the sample was taken.
* `report_date` — the **report date**, when the laboratory issued the result.

The delay between them is the specimen-to-result turnaround, and it is exactly
what a nowcast has to model. (`received_date` would split that into a transport
and a processing leg, but it is 84% missing and stops in 2019, so we leave it
alone.)

## 1. Choosing a time resolution

Before anything else, the data has to be put on a time grid, because the grid
constrains everything after it — including which nowcasting engines you can use
at all.

A `tbl_now` will happily take the daily dates, but daily is not usable here.
Counting needs rows that have both dates, so this one filter runs ahead of the
cleaning proper:

```{r resolution}
hai_bucaramanga |>
  filter(!is.na(specimen_date), !is.na(report_date)) |>
  summarise(
    cases            = n(),
    distinct_days    = n_distinct(specimen_date),
    cases_per_day    = round(n() / n_distinct(specimen_date), 2),
    distinct_weeks   = n_distinct(floor_date(specimen_date, "week", week_start = 1)),
    cases_per_week   = round(n() / n_distinct(floor_date(specimen_date, "week", week_start = 1)), 2)
  )
```

Roughly **1.5 cases per day** cannot support a delay distribution with any
resolution at all. Aggregating to weeks gives about three and a half cases a
week, which is still thin but workable.

Monthly would be denser still, and statistically it is the better fit for this
data — but it is worth knowing that **your time unit constrains your choice of
engine**. Of the five nowcasting packages used at the end of this article, only
`diseasenowcasting` and `baselinenowcast` accept monthly data: `NobBS` supports
only daily and weekly units, and `epinowcast` accepts only `"day"` and `"week"`
timesteps. Weeks are the finest unit every engine agrees on, so weeks it is.

```{r aggregate-weeks-raw}
hai_weekly_raw <- hai_bucaramanga |>
  mutate(
    event_week  = floor_date(specimen_date, "week", week_start = 1),
    report_week = floor_date(report_date,   "week", week_start = 1)
  )
```

## 2. Building the `tbl_now`

Now the data can be handed to `tbl_now()`. We declare the two dates, say the
data is a line list (one row per case), and stratify by the type of intensive
care unit.

This is deliberately done **before** cleaning. `?hai_bucaramanga` documents the
defects, but rather than trusting the help page — or discovering them one at a
time — we let the object tell us what is wrong with it. `tbl_now()` accepts
imperfect data and complains about it:

```{r build-tbl-now-raw}
hai_raw_now <- hai_weekly_raw |>
  tbl_now(
    event_date  = event_week,
    report_date = report_week,
    strata      = icu_type,
    data_type   = "linelist"
  )
```

Three warnings, before we have looked at a single number. `tbl_now()` also
inferred that both dates are **weekly**, computed `.delay` and the numeric
indices, and set the **now** to the most recent report week. From here on,
everything — the plots, the diagnostics, and every model converter — reads those
settings off the object instead of being told again.

## 3. What is wrong with it?

The warnings above are the loud findings. `diagnose()` returns all of them, and
the quiet ones too, as a tibble sorted worst first. `status` is an ordered
factor — `error` > `warning` > `note` > `ok` > `not_run` > `skipped` — so
filtering to what needs acting on is a comparison:

```{r diagnose-raw}
diagnose(hai_raw_now) |>
  filter(status <= "note") |>
  select(check, scope, stratum, status, n_affected, n_total, message)
```

That is the cleaning list, derived rather than assumed. Three of the findings
are defects to fix — missing event dates, missing report dates, and reports
that precede their event. The others are things the file cannot be blamed for
but you still need to know: twelve columns were never declared, so they silently
split every reporting-triangle cell; and the last event date sits well behind
`now`, because reporting on this extract stopped before the extract did.

Two statuses deserve attention, because neither means "fine":

```{r diagnose-raw-skipped}
diagnose(hai_raw_now) |>
  filter(status == "skipped") |>
  select(check, scope, message)
```

`skipped` is not `ok`. `ok` means a check ran and found nothing; `skipped` means
it could not run at all, and a check that cannot be performed is never reported
as a pass. Note the first row in particular: **`diagnose()` will not look for
duplicate records in a line list**, because one row *is* one case, so two
identical rows are two infections rather than a repeat. That check is the one
piece of the cleaning below that `diagnose()` cannot do for you, and §4 does it
by hand.

The `not_run` rows are different again — questions `diagnose()` deliberately
refuses to answer:

```{r diagnose-raw-signposts}
diagnose_signposts(hai_raw_now) |>
  select(scope, status, message)
```

`diagnose()` is **structural**: it never runs a statistical test, so it is fast
and its answer never depends on a random seed. Whether the delay drifts, and
whether reports arrive in batches, are statistical questions, and each signpost
names the call that answers it. §6 runs them.

## 4. Cleaning

Now the defects, one at a time. Nothing below is specific to this dataset —
these are the fixes worth knowing on *any* surveillance extract.

### Duplicate records

This is the check `diagnose()` skipped, and the reason it skipped it is exactly
why it needs doing here: the *object* cannot tell a repeated case from a real
second one, but the *source* can. `hai_bucaramanga`'s `id` column is documented
as a unique autonumber, so it should have no repeats:

```{r dupes}
hai_bucaramanga |>
  count(id) |>
  filter(n > 1) |>
  nrow()
```

It does. Are those rows genuinely identical, or do they differ somewhere and
merely share an id? That distinction decides whether it is safe to drop them:

```{r dupes-identical}
# Rows that are byte-identical to another row, across every column.
sum(duplicated(hai_bucaramanga))
```

The count of repeated ids and the count of exact duplicate rows agree, so every
repeat is a whole-record copy — an export artifact, not a real second infection.
`distinct()` removes them without losing information:

```{r clean-distinct}
hai <- distinct(hai_bucaramanga)

nrow(hai_bucaramanga) - nrow(hai)
```

### Missing dates

The source encodes missing dates as `1900-01-01`; that sentinel has already been
converted to `NA` when the dataset was built (see `?hai_bucaramanga`). A row
without both dates cannot contribute to a reporting triangle — which is what the
two `missing` findings above were saying.

That is a large fraction of the file, and it is worth pausing on rather than
filtering away reflexively. Missingness that depends on *when* a case occurred
would bias the nowcast; missingness spread evenly over time mostly costs
precision. Here it is heavily concentrated in the early years:

```{r missing-by-year}
hai |>
  filter(!is.na(specimen_date)) |>
  mutate(year = year(specimen_date)) |>
  summarise(
    cases        = n(),
    missing_rate = round(mean(is.na(report_date)), 2),
    .by          = year
  ) |>
  arrange(year)
```

```{r clean-dates}
hai <- hai |> filter(!is.na(specimen_date), !is.na(report_date))

nrow(hai)
```

### Reports that arrive before the event

This is the check that most often catches real data-entry problems, and it is
easy to forget: a result cannot be issued before the sample was taken, so the
delay must never be negative. `diagnose_ordering()` counted these at the weekly
resolution; at daily resolution the picture is:

```{r negative-delays}
hai |>
  mutate(delay_days = as.numeric(report_date - specimen_date)) |>
  summarise(
    negative = sum(delay_days < 0),
    worst    = min(delay_days),
    median   = median(delay_days),
    max      = max(delay_days)
  )
```

Eighty-one records have a report date *before* their specimen date, one by
almost a year. There is no way to tell from the file whether the event date, the
report date, or both are wrong, so the only defensible move is to drop them and
say so:

```{r clean-negative}
hai <- hai |> filter(report_date >= specimen_date)

nrow(hai)
```

### The cleaning cascade

It is worth writing the whole cascade down. Readers of your analysis — and you,
in six months — need to know how much data reached the model and why the rest
did not:

```{r cascade, echo=FALSE}
tibble::tibble(
  Step = c(
    "Raw records",
    "After removing exact duplicates",
    "After requiring both dates",
    "After requiring report >= specimen"
  ),
  Records = c(
    nrow(hai_bucaramanga),
    nrow(distinct(hai_bucaramanga)),
    nrow(distinct(hai_bucaramanga) |>
           filter(!is.na(specimen_date), !is.na(report_date))),
    nrow(hai)
  )
) |>
  mutate(`Lost at this step` = c(NA, -diff(Records))) |>
  knitr::kable(caption = "Cleaning cascade")
```

Just under half the file survives. That is uncomfortable, and it is the honest
starting point for everything that follows.

### Rebuilding, and checking that it worked

```{r aggregate-weeks}
hai_weekly <- hai |>
  mutate(
    event_week  = floor_date(specimen_date, "week", week_start = 1),
    report_week = floor_date(report_date,   "week", week_start = 1)
  )

hai_now <- hai_weekly |>
  tbl_now(
    event_date  = event_week,
    report_date = report_week,
    strata      = icu_type,
    data_type   = "linelist"
  )

hai_now
```

No warnings this time. Re-running the diagnosis confirms it rather than assuming
it, which is the point of having the check return data:

```{r diagnose-clean}
diagnose(hai_now) |>
  filter(status <= "note") |>
  select(check, scope, stratum, status, n_affected, message)
```

The three defects are gone. What remains are the two structural notes — the
undeclared columns, which are deliberate here (we are stratifying only by
`icu_type`), and the staleness and truncation of the tail, which are properties
of the extract rather than mistakes in it. A finding you have read and accepted
is a different thing from one you have not seen.

```{r getters}
c(
  now          = format(get_now(hai_now)),
  event_units  = get_event_units(hai_now),
  report_units = get_report_units(hai_now),
  strata       = paste(get_strata(hai_now), collapse = ", ")
)
```
## 5. Looking at the data

`autoplot()` draws a diagnostic grid: the case-count panels in green describe
the **epidemic** process, the red ones the **reporting** process.

```{r autoplot, fig.width=11, fig.height=13, message=FALSE, warning=FALSE, fig.alt="A six-panel diagnostic grid: empirical delay distribution, observed epidemic curve, month-of-year effects for cases and for reporting, and periodograms for both processes."}
autoplot(hai_now)
```

Four things stand out, and each of them changes a modelling decision later.

**The epidemic curve has a large 2021 wave.** Case counts sit in the low single
digits for most of the series and then rise sharply through 2021 — the period
when COVID-19 filled Colombian intensive care units. The dashed line marks where
the data become incomplete: everything to its right is still arriving.

**The delay distribution is not a single hump.** This is the most consequential
feature of the dataset, so it is worth its own plot:

```{r delay-dist, fig.width=9, fig.height=4.5, fig.alt="Histogram of the reporting delay in weeks, showing a large spike at zero and a long right tail."}
plot_delay_distribution(hai_now)
```

More than half of all results are reported in the **same week** the specimen was
taken, but the tail runs out past thirty weeks. Tabulating the cumulative share
makes the shape explicit:

```{r delay-quantiles}
hai_weekly |>
  mutate(delay_weeks = as.numeric(report_week - event_week) / 7) |>
  count(delay_weeks) |>
  mutate(cumulative = round(cumsum(n) / sum(n), 3)) |>
  filter(delay_weeks %in% c(0, 1, 4, 8, 9, 13, 17, 22, 26)) |>
  select(delay_weeks, n, cumulative)
```

This is **bimodal**. Roughly 74% of results arrive within a week, then reporting
almost stops — only four more percentage points accumulate between weeks 1 and
8 — and then a *second* wave of reports lands between weeks 9 and 18, taking the
total to 98%. That is the signature of two different reporting streams: routine
same-week laboratory results, plus a slower retrospective process that sweeps up
the rest months later.

**The reporting process has an annual cycle.** In the bottom-right periodogram
the reporting series peaks at a period of about 12 months, which the epidemic
series does not. An annual rhythm in *reporting* — not in infections — is
consistent with that second, retrospective stream being an annual audit or
reconciliation exercise.

### The same picture, as numbers

`autoplot()` is for looking; `summary()` is for reading off values, and returns a
tibble so you can do either. It confirms the bimodality above without needing the
plot:

```{r summary-delay}
summary(hai_now) |>
  filter(component == "delay") |>
  select(quantity, stratum, n, total, mean, sd, q50, q90, max)
```

A **mean of 3.5 weeks against a median of 0** is the whole story in two numbers:
most results arrive in the same week, and a long right tail drags the average out
to three and a half. A symmetric delay would have had them nearly equal. The
standard deviation, larger than the mean, says the same thing.

The composition block explains why the strata behave so differently downstream:

```{r summary-strata}
summary(hai_now) |>
  filter(component == "composition") |>
  select(quantity, total, prop)
```

Adult units carry 78% of the cases. That is the imbalance `diagnose_strata()`
flagged as a note, and it is worth remembering when a per-stratum nowcast for
`Neonatal` — 62 cases spread over 317 event weeks — comes back with very wide
intervals.

## 6. Is the reporting delay stable?

A nowcast fitted to the whole series assumes the delay distribution is the same
throughout. If it is not, the model averages over reporting regimes that are not
exchangeable, and the nowcast is biased. `tbl.now` gives you two tests for this,
which answer different questions.

First, look at it:

```{r drift-plot, fig.width=9, fig.height=5.5, message=FALSE, warning=FALSE, fig.alt="Fan chart of the reporting delay over time, showing the median as a solid line and the interquartile and 10-90 percent bands. The bands widen substantially from early 2020 onward."}
plot_delay_drift(hai_now, changepoint = TRUE)
```

The median delay bounces around zero throughout, but the spread of the
distribution clearly widens from 2020 onwards. The two tests confirm exactly
that split.

### Is there a gradual trend?

`diagnose_drift()` runs an autocorrelation-robust Mann-Kendall test for a
*monotonic* trend:

```{r drift-test, warning=FALSE}
diagnose_drift(hai_now)
```

No. Neither the median nor the spread shows a monotonic trend. Read `tau` before
`p_value`: both effect sizes are tiny, so this is a genuine null rather than an
underpowered one.

### Is there an abrupt shift?

A trend test cannot see a step change — a jump up and a jump down can even
cancel out to no trend at all. `diagnose_changepoint()` uses Pettitt's test to
look for exactly that:

```{r changepoint-test, warning=FALSE}
diagnose_changepoint(hai_now)
```

Here is the finding that matters. The **median** delay has no change point, but
the **spread** does, and a decisive one: the 10–90 percentile range of the delay
jumps from about 2 weeks to about 4.8 weeks, and the estimated change point is
**early March 2020** — the week COVID-19 reached Colombia and hospital
laboratories were abruptly redirected.

So the typical case was reported no more slowly after March 2020, but the
*variability* of reporting more than doubled. That is a real change in the
reporting process, and it has a direct modelling consequence, which we act on in
section 8.

### Does this differ by unit?

```{r drift-strata, warning=FALSE}
diagnose_drift(hai_now, by_strata = TRUE)
```

The neonatal unit shows a positive trend in the median delay
(`tau` = 0.14, p = 0.032). Resist the urge to make something of it: this is one
significant result out of six tests, and a Bonferroni threshold would be
0.05 / 6 = 0.008. It does not survive. With only 42 weeks contributing, the
honest reading is *suggestive, not established* — worth watching, not worth
modelling.

Note also that `sens_slope` is `0` everywhere. Delays are whole numbers of weeks
and more than half of them are zero, so the median pairwise slope really is
zero; that is a property of discrete, heavily-tied data, not a bug.

## 7. Temporal effects

`temporal_effects()` attaches calendar covariates a model can use. The
temptation is to add everything available; the discipline is to add only what
the data supports.

**Day-of-week effects are not justified here.** They are the usual first choice
for surveillance data — laboratories are quieter at weekends — but this dataset
does not behave that way:

```{r dow-check}
hai |>
  count(weekday = wday(specimen_date, label = TRUE, week_start = 1)) |>
  mutate(percent = round(100 * n / sum(n), 1))
```

Specimens are spread almost evenly across all seven days, weekends included. And
in any case we aggregated to weeks, so a day-of-week term has nothing to attach
to. Adding one would be modelling noise.

**An annual cycle is justified**, on the reporting side, by the periodogram in
section 5 and by the bimodal delay. Since the data are weekly, that is a Fourier
season of period 52:

```{r temporal-effects}
hai_effects <- temporal_effects(
  week_of_year = TRUE,   # position within the year
  seasons      = 52      # an annual Fourier cycle on weekly data
)

hai_effects
```

Attach them to the report date, because the annual rhythm we saw is a property
of *reporting*, not of infection:

```{r attach-effects}
hai_seasonal <- hai_now |>
  add_temporal_effects(hai_effects, date_type = "report_date")

hai_seasonal
```

Effects are attached **lazily** — the object records the specification but does
not materialise the columns until a model asks for them (or you call
`compute_temporal_effects()`):

```{r compute-effects}
hai_seasonal |>
  compute_temporal_effects() |>
  select(event_week, report_week, starts_with(".report_")) |>
  head(4)
```

## 8. Nowcasting with `diseasenowcasting`

Now we let the diagnostics decide the model. Three findings carry over:

* **The delay distribution changed in March 2020.** Fitting across that boundary
  would average two reporting regimes. We restrict to the post-change-point
  period.
* **The delay reaches 98% by about 20 weeks.** That sets the maximum delay: long
  enough to capture the second reporting stream, short enough not to carry a
  mostly-empty tail.
* **No monotonic drift within that period**, so a single delay distribution for
  the restricted window is defensible.

```{r informed-window}
changepoint_date <- as.Date("2020-03-09")

hai_model <- hai_weekly |>
  filter(event_week >= changepoint_date) |>
  tbl_now(
    event_date  = event_week,
    report_date = report_week,
    strata      = icu_type,
    data_type   = "linelist",
    verbose     = FALSE
  ) |>
  censor_delays_above(max_delay = 20)

hai_model
```

`censor_delays_above()` is doing real work: it caps the delay dimension at 20
weeks so the model is not asked to estimate probabilities for delays that occur
a handful of times in the whole series.

`diseasenowcasting` consumes a `tbl_now` **directly** — no converter, no
restating of which column is which:

```{r dnc-fit, eval=has("diseasenowcasting"), message=FALSE, warning=FALSE}
# `diseasenowcasting::` rather than `library()` -- it is not on CRAN, and
# attaching it would leave `R CMD check` with an undeclared dependency.
hai_fit <- diseasenowcasting::nowcast(hai_model, type = "one_stage", n_draws = 2000)

hai_fit
```

The posterior draws of the eventual counts come from `predict()`:

```{r dnc-predict, eval=has("diseasenowcasting"), message=FALSE, warning=FALSE}
hai_prediction <- predict(hai_fit)

draws       <- S7::prop(hai_prediction, "draws")        # draws x event weeks
event_weeks <- S7::prop(hai_prediction, "event_dates")
observed    <- S7::prop(hai_prediction, "observed_series")

nowcast_summary <- tibble::tibble(
  event_week = event_weeks,
  observed   = observed,
  median     = apply(draws, 2, median),
  lower      = apply(draws, 2, quantile, 0.025),
  upper      = apply(draws, 2, quantile, 0.975)
)

tail(nowcast_summary, 8)
```

```{r dnc-plot, eval=has("diseasenowcasting"), fig.width=9, fig.height=4.8, message=FALSE, warning=FALSE, fig.alt="Nowcast of weekly hospital-acquired infections. Grey bars show cases reported so far, a red line and ribbon show the nowcast median and 95 percent interval, which rises above the bars over the most recent weeks."}
recent <- nowcast_summary |> filter(event_week >= max(event_week) - 7 * 30)

ggplot(recent, aes(x = event_week)) +
  geom_col(aes(y = observed, fill = "Reported so far"), width = 5.5) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#e78b7f", alpha = 0.35) +
  geom_line(aes(y = median, colour = "Nowcast"), linewidth = 0.9) +
  scale_fill_manual(name = NULL, values = c("Reported so far" = "#dfe5e0")) +
  scale_colour_manual(name = NULL, values = c("Nowcast" = "#B85348")) +
  labs(
    x = "Event week (specimen taken)", y = "Infections",
    title = "Nowcast of hospital-acquired infections",
    subtitle = paste0("Now = ", format(get_now(hai_model)),
                      "; the gap at the right edge is what has not been reported yet")
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335"))
```

The bars are what had been reported by the `now`; the ribbon is the model's
estimate of where those weeks will end up.

Look closely at what the nowcast actually does here, because it is not the
textbook picture. The **median barely moves**: it exceeds the observed count in
only one of the 152 modelled weeks. What changes dramatically is the
**uncertainty** — the mean width of the 95% interval is about 0.1 cases over the
settled part of the series and about 2 cases over the final ten weeks, a
twenty-fold widening.

That is the correct behaviour for data this sparse, and it is worth
understanding rather than treating as a disappointment. We know from the delay
distribution that roughly a quarter of results arrive more than a week late, so
a week sitting five weeks from `now` really is missing about a quarter of its
eventual cases. But a quarter of *two cases* is half a case, and the median of a
posterior over counts is an integer. The correction is real; it is simply
smaller than the resolution of the answer. With one to three infections a week,
a nowcast's honest output is a statement about what it does not know, not a
large upward revision.

This is also why the point of the earlier diagnostics was never to produce a
better point estimate. It was to make sure the interval is trustworthy — that it
comes from one reporting regime rather than two averaged together.

### Carrying the strata and effects through

The enriched object needs no extra arguments: `diseasenowcasting` reads the
`icu_type` stratum and the seasonal columns off the object itself.

```{r dnc-seasonal, eval=run_stan && has("diseasenowcasting"), message=FALSE}
hai_model_seasonal <- hai_model |>
  add_temporal_effects(hai_effects, date_type = "report_date")

diseasenowcasting::nowcast(hai_model_seasonal, type = "one_stage", n_draws = 2000)
```

## 9. The same data, other engines

The value of describing the data once is that every other package is now one
call away. Each converter reads the event date, report date, strata and units
off `hai_model`.

### baselinenowcast — a reporting triangle

```{r bln, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
library(baselinenowcast)

hai_triangle <- tbl_now_to_baselinenowcast(hai_model, verbose = FALSE)

hai_triangle[1:4, 1:6]
```

```{r bln-fit, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
bln_samples <- baselinenowcast(hai_triangle, output_type = "samples", draws = 1000)

bln_samples |>
  as_tibble() |>
  summarise(median = median(pred_count), .by = reference_date) |>
  tail(4)
```

### surveillance — the classic line-list nowcast

```{r sur, eval=has("surveillance"), message=FALSE, warning=FALSE}
library(surveillance)

hai_sur <- tbl_now_to_surveillance(hai_model, verbose = FALSE)
sur_now <- get_now(hai_model)

sur_fit <- nowcast(
  now          = sur_now,
  when         = seq(sur_now - 7 * 8, sur_now, by = "1 week"),
  data         = hai_sur,
  dEventCol    = "dHospital",
  dReportCol   = "dReport",
  aggregate.by = "1 week",
  D            = 20,
  method       = "bayes.notrunc.bnb"
)

tibble::tibble(
  event_week = as.Date(epoch(sur_fit)),
  nowcast    = as.numeric(upperbound(sur_fit))
) |>
  tail(4)
```

### NobBS — Nowcasting by Bayesian Smoothing

`NobBS` takes a plain line list, which a `linelist` `tbl_now` already is — and
even its arguments come off the object.

```{r nobbs, eval=has("NobBS"), message=FALSE, warning=FALSE}
library(NobBS)

nobbs_fit <- NobBS(
  data        = as.data.frame(hai_model),
  now         = get_now(hai_model),
  units       = "1 week",
  onset_date  = get_event_date(hai_model),
  report_date = get_report_date(hai_model),
  max_D       = 20
)

nobbs_fit$estimates |> tail(4)
```

### epinowcast — a Bayesian model with delay and reference modules

`epinowcast` compiles a Stan model, so the fit below is not run when this page
is built; the conversion is.

```{r enw-convert, eval=has("epinowcast"), message=FALSE, warning=FALSE}
library(epinowcast)

hai_enw <- tbl_now_to_epinowcast(hai_model, verbose = FALSE, quiet = TRUE)

hai_enw
```

```{r enw-fit, eval=run_stan && has("epinowcast"), message=FALSE}
epinowcast(
  hai_enw,
  fit = enw_fit_opts(pp = TRUE, chains = 2, iter_sampling = 500, iter_warmup = 500)
)
```

For a side-by-side comparison of all of these engines on one dataset, with a
held-back ground truth, see the
[nowcasting-models article](https://rodrigozepeda.github.io/tbl.now/articles/nowcasting-models.html).

## Summary

The workflow, start to finish:

```{r recap, eval=FALSE}
# 1. Choose a resolution the data (and your engine) can support.
hai_weekly_raw <- hai_bucaramanga |>
  mutate(event_week  = floor_date(specimen_date, "week", week_start = 1),
         report_week = floor_date(report_date,   "week", week_start = 1))

# 2. Describe it once -- before cleaning, so the object can tell you what is wrong.
hai_raw_now <- tbl_now(hai_weekly_raw, event_date = event_week,
                       report_date = report_week, strata = icu_type,
                       data_type = "linelist")

# 3. Ask what is wrong with it, instead of guessing.
diagnose(hai_raw_now) |> filter(status <= "note")

# 4. Fix what it reported, and confirm the fix.
hai <- hai_bucaramanga |>
  distinct() |>                                          # duplicates: diagnose() skips
  filter(!is.na(specimen_date), !is.na(report_date)) |>  #   these two it found
  filter(report_date >= specimen_date)                   #   and this one

hai_weekly <- hai |>
  mutate(event_week  = floor_date(specimen_date, "week", week_start = 1),
         report_week = floor_date(report_date,   "week", week_start = 1))

hai_now <- tbl_now(hai_weekly, event_date = event_week, report_date = report_week,
                   strata = icu_type, data_type = "linelist")

diagnose(hai_now) |> filter(status <= "note")

# 5. Look at it, and read the numbers off it.
autoplot(hai_now)
summary(hai_now)

# 6. Run the statistical tests diagnose() signposted but would not run itself.
diagnose_drift(hai_now)
diagnose_changepoint(hai_now)

# 7. Let those findings set the model window and the maximum delay.
hai_model <- hai_now |> censor_delays_above(max_delay = 20)

# 8. Hand it to whichever engine you like.
diseasenowcasting::nowcast(hai_model)         # diseasenowcasting
tbl_now_to_baselinenowcast(hai_model)         # baselinenowcast
tbl_now_to_epinowcast(hai_model)              # epinowcast
tbl_now_to_surveillance(hai_model)            # surveillance
as.data.frame(hai_model)                      # NobBS
```

Two closing points worth carrying to your own data.

**The diagnostics were not decoration.** The change-point test found a real
break in March 2020 and that set the model window; the delay quantiles set
`max_delay`; the day-of-week check *stopped* us adding a term the data did not
support. Had we skipped straight to `nowcast()`, all three would have been
guesses.

**Half of this dataset never reached the model.** Duplicate records, missing
dates and impossible delays removed 750 of 1,423 rows. A nowcast built on the
remainder is only as good as the assumption that what was dropped resembles what
was kept — which for the missing dates, concentrated in the early years, is
worth stating out loud rather than burying.

```{r learning-more, echo=FALSE, results="asis"}
# The fragment is looked up by ABSOLUTE path via system.file(), never by a
# relative one. pkgdown/rmarkdown render into an intermediates directory under
# tempdir() and copy relative resources alongside; a path containing "../.."
# escapes that directory ("cannot create file '/tmp/RtmpXXX/../../man/...'"),
# which fails wherever tempdir() sits near the filesystem root -- as it does on
# CI. `system.file()` also works under pkgload::load_all(), which shims it to
# find inst/ in the source tree.
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```
