---
title: "More on the tbl_now object"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{More on the tbl_now object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has <- function(pkg) requireNamespace(pkg, quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # `html_vignette` renders at 96 dpi, which for the wide multi-panel figures
  # below produces PNGs wider than the 700px the vignette CSS will ever show
  # and inflates the installed size of `doc/` for pixels nobody sees. 72 dpi
  # keeps every figure at essentially its displayed width.
  dpi = 72
)
```


## Why this vignette 

This article describes in depth a `tbl_now`: what it records,
every attribute it carries, how it goes through a `dplyr` pipeline, and the
additional functions that come with it. If you have not met the package yet, read the
[*Get started* vignette](https://rodrigozepeda.github.io/tbl.now/articles/tbl.now.html)
first.

## Introduction 

<!--The [tidyverse](https://tidyverse.org/) ecosystem [@tidyverse] provides a widely adopted framework for data analysis. Within this paradigm, data is structured such that rows represent individual observations and columns represent variables. This approach is commonly referred to as [tidy data](https://tidyr.tidyverse.org/articles/tidy-data.html) [@wickham2014tidy].

Several tidyverse extensions exist for working with time series including [tsibble](https://tsibble.tidyverts.org/), [tibbletime](https://business-science.github.io/tibbletime/), and [timetk](https://business-science.github.io/timetk/) [@wang2020new; @wang2019tidy; @timetk]. These packages provide time-aware tibbles and tools that integrate smoothly with the tidyverse. However, epidemiological nowcasting requires two time indices simultaneously: an **event time** and a **reporting time**. Classical time-series abstractions assume only a single index and therefore do not fully support this structure.

The `tbl.now` class and package addresses this gap. -->

The `tbl.now` extends a regular  [tibble()](https://tibble.tidyverse.org/) to explicitly encode epidemiological event and report dates, allowing consistent data transformation, delay computation, and integration with the [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) and other modeling workflows.

More concretely, `tbl.now` was designed to:

+ Standardize the data inputs required by nowcasting models (including [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/)).

+ Preserve [tidyverse](https://tidyverse.org/) compatibility so users can continue to apply familiar [dplyr](https://dplyr.tidyverse.org/) operations.

+ [Diagnose reporting artifacts](https://rodrigozepeda.github.io/tbl.now/articles/diagnosing-a-tbl-now.html) directly from the data (such as reporting-delay drift and change points) as well as batch (backlog) reporting.

+ Facilitate integration into iterative modeling workflows with [different nowcasting packages](https://rodrigozepeda.github.io/tbl.now/articles/nowcasting-models.html) [@gelman2020bayesian; @wickham2023r]:

+ Facility backtesting and ensembling [of nowcasts](https://rodrigozepeda.github.io/tbl.now/articles/ensemble-nowcasting.html)


We begin by loading the required packages:

```{r setup, message=FALSE}
library(dplyr, quietly = TRUE)
library(lubridate)
library(tbl.now)
```

## How `tbl.now` works 

In an epidemiological nowcast, we typically observe at least two key dates[^1]:

[^1]: More key dates are possible such as a `revision_date`. For example in the case of influenza one might consider the `event_date` = symptom onset, the `report_date` = when the patient was first diagnosed by a medical professional, and `revision_date` = when the positive test's results for influenza were recorded. We  come back to these multiple dates [in the revision vignette](https://rodrigozepeda.github.io/tbl.now/articles/example_revisions.html).  

+ `event_date`: when the underlying event occurred (e.g., symptom onset, exposure, sample collection).

+ `report_date`: when the event was recorded in the data system (e.g., lab result processed, clinical visit documented).

The nowcasting task is:

> To estimate, for each past `event_date`, how many events (e.g. cases) have already occurred
but have not yet been reported as of **now**. That is, the nowcast will predict how
many observations will eventually be observed for each (past or present) `event_date`.

Visually:

```{r nowcast-explainer, echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4.6, out.width = "100%", fig.alt = "Stacked bar chart of cases by event date. Green bars show cases already reported as of now; pale red bars stacked on top show the cases not yet reported, which grow rapidly over the most recent event dates. A black line traces the total that will eventually be reported, rising steadily, while the green observed bars turn downward near now."}
library(ggplot2)
library(dplyr)

now_date <- ymd("2024-03-31")
horizon <- 24

explain <- dplyr::tibble(event_date = now_date - (horizon:0))
explain$age <- as.numeric(now_date - explain$event_date)

# A rising epidemic: the eventual (final) count for each event date.
explain$eventual <- round(20 * exp(0.055 * (horizon - explain$age)))
# Reporting completeness grows with how old the event date is.
explain$completeness <- 1 - exp(-0.55 * (explain$age + 0.35))
explain$observed <- round(explain$eventual * explain$completeness)
explain$missing <- explain$eventual - explain$observed

lab_obs <- "Already reported as of now"
lab_mis <- "Not yet reported (what the nowcast estimates)"
lab_now <- "Nowcast: a model for what will eventually be reported"

bars <- tidyr::pivot_longer(
  explain[, c("event_date", "observed", "missing")],
  cols = c("observed", "missing"), names_to = "part", values_to = "cases"
)
bars$part <- factor(
  bars$part, levels = c("missing", "observed"), labels = c(lab_mis, lab_obs)
)

ggplot(bars, aes(x = event_date, y = cases)) +
  geom_col(aes(fill = part), width = 0.75) +
  geom_line(data = explain, aes(y = eventual, colour = lab_now), linewidth = 0.9) +
  geom_point(data = explain, aes(y = eventual, colour = lab_now), size = 1.5) +
  scale_fill_manual(
    name = NULL, breaks = c(lab_obs, lab_mis),
    values = stats::setNames(c("#5F7E62", "#e78b7f"), c(lab_obs, lab_mis))
  ) +
  scale_colour_manual(name = NULL, values = stats::setNames("#262626", lab_now)) +
  geom_vline(aes(xintercept = now_date), data = dplyr::tibble(now_date = now_date),
             linetype = "dashed", colour = "#262626", linewidth = 0.4) +
  annotate(
    "label", x = now_date, y = max(explain$eventual) * 1.10,
    label = "now", hjust = 0.5, vjust = 0, fontface = "bold",
    size = 3.4, colour = "#262626"
  ) +
  coord_cartesian(ylim = c(0, max(explain$eventual) * 1.15)) +
  labs(
    x = "Event date", y = "Cases",
    title = "What a nowcast estimates",
    subtitle = paste(
      "Recent event dates look like a downturn only because",
      "their reports have not arrived yet"
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    legend.box = "vertical",
    legend.margin = margin(b = -4),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", colour = "#262626")
  ) +
  guides(
    fill = guide_legend(order = 1, ncol = 2),
    colour = guide_legend(order = 2)
  )
```


>In the figure above, the green bars represent the number of cases (events)
that have been observed until **now**; the pale red segments are the 
reports still in transit. Because completeness decays sharply
over the most recent event dates, the observed counts bend *downwards* near the
right edge even though the epidemic is still growing. The **nowcast** is the red
line: an estimate of the height each bar will eventually reach.

A `tbl.now` object is a specialized [tibble()](https://tibble.tidyverse.org/) that:

+ Identifies the `event_date` and `report_date` columns.

+ Stores these as `attributes()` to enable consistent processing.

+ Automatically computes auxiliary fields such as delay, numerical indices, and frequency units.

+ Ensures the dataset is correctly formatted for multiple frameworks including [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/), epidist, epinow, NobBS, and more (see `list_nowcast_methods()`).

## Example: A simple `tbl.now`

Consider the following dataset:

<div class="small-table">
| symptom_onset | medical_visit |  n |
| ------------- | ------------- | -: |
| 2023-12-25    | 2023-12-26    | 10 |
| 2023-12-26    | 2023-12-26    |  2 |
| 2023-12-25    | 2023-12-27    |  5 |
| 2023-12-26    | 2023-12-27    | 11 |
</div>

Where:

+ `symptom_onset` is the `event_date`.

+ `medical_visit` is the `report_date`.

+ `n` is the number of reported cases for each event–report combination.

We can convert this into a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) by first creating a `data.frame` and then using the `tbl_now()` function:

```{r}
#Create a data.frame
df <- data.frame(
  symptom_onset = c(ymd("2023/12/25"), ymd("2023/12/26"), 
                    ymd("2023/12/25"), ymd("2023/12/26")),
  medical_visit = c(ymd("2023/12/26"), ymd("2023/12/26"), 
                    ymd("2023/12/27"), ymd("2023/12/27")),
  n = c(10, 2, 5, 11)
)

#Convert to tbl.now
df |>
  tbl_now(event_date = symptom_onset, report_date = medical_visit, case_count = n)
```


This performs several operations automatically:

+ Detects the **data type** (`count-incidence` in this example). See [below](#data-types) for all the data types available. 

+ Infers the **frequency units** of event and report dates (daily).

+ Tags the correct columns as `event_date`, `report_date`, and `case_count.`

+ Computes `.event_num`, `.report_num`, and `.delay` columns the numerical versions (indexed at 0) of event, report and `.delay = report_date - event_date` columns.

+ Identifies the appropriate **now** date (the most recent report date).

The remaining sections describe these features and the broader `tbl.now` toolkit.


## Attributes of a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) 

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) stores information about its structure using object attributes, ensuring consistent behavior across `dplyr` transformations. The primary attributes are:

<table>
<thead>
<tr>
<th align="center">&nbsp;</th>
<th align="left">Argument</th>
<th align="left">What it records</th>
</tr>
</thead>
<tbody>
<tr>
<td align="center"><img src="man/figures/event_date.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="event_date"></td>
<td align="left"><code>event_date</code></td>
<td align="left">The column storing <strong>event dates</strong>; i.e. when the epidemiological phenomenon of interest happened (symptom onset, hospitalisation, death, &hellip;). <strong>Required.</strong></td>
</tr>
<tr>
<td align="center"><img src="man/figures/report_date.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="report_date"></td>
<td align="left"><code>report_date</code></td>
<td align="left">The column storing <strong>report dates</strong>; i.e. when that event became known to the surveillance system. <strong>Required</strong>, unless it is reconstructed from <code>delay</code>.</td>
</tr>
<tr>
<td align="center"><img src="man/figures/revision_date.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="revision"></td>
<td align="left"><code>revision_date</code></td>
<td align="left">An optional third date indicating when the report was resolved (see <code>revision_type</code>). <em>Optional</em>.</td>
</tr>
<tr>
<td align="center"><img src="man/figures/revision_type.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="revision"></td>
<td align="left"><code>revision_type</code>, <code>revision_levels</code></td>
<td align="left">What the revision date resolved to. Only <code>confirmed</code>, <code>retracted</code>, <code>pending</code> or <code>NA</code> are ever stored; set <code>revision_levels</code> as a named dictionary mapping the data's labels into those four ( e.g. <code>c(positive = "confirmed")</code>). <em>Optional</em>.</td>
</tr>
<tr>
<td align="center"><img src="man/figures/now.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="now"></td>
<td align="left"><code>now</code></td>
<td align="left">The date the nowcast is anchored to &mdash; &ldquo;today&rdquo; from the model's point of view. <em>Optional</em>; defaults to the latest date.</td>
</tr>
<tr>
<td align="center"><img src="man/figures/strata.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="strata"></td>
<td align="left"><code>strata</code></td>
<td align="left">Columns you want a separate nowcast for (e.g. gender, region). <em>Optional</em>.</td>
</tr>
<tr>
<td align="center"><img src="man/figures/covariates.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="covariates"></td>
<td align="left"><code>covariates</code></td>
<td align="left">Columns that inform the nowcast but that you do <em>not</em> want it broken down by (e.g. temperature or precipitation). <em>Optional</em>.</td>
</tr>
<tr>
<td align="center"><img src="man/figures/case_count.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="case_count"></td>
<td align="left"><code>case_count</code></td>
<td align="left">The column holding the counts when the data is given as aggregated (rather than line-list). <em>Optional</em>.</td>
</tr>
<tr>
<td align="center"><img src="man/figures/datatype.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="data_type"></td>
<td align="left"><code>data_type</code></td>
<td align="left">Whether the data represents a <code>linelist</code> (each row is a case), <code>count-incidence</code>(each row is a collection of cases per event-report date) or <code>count-cumulative</code>(each row is the cummulative number cases for that event accumulating in the report axis). <em>Optional</em>; inferred by default.</td>
</tr>
<tr>
<td align="center"><img src="man/figures/units.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="units"></td>
<td align="left"><code>event_units</code>, <code>report_units</code>, <code>revision_units</code></td>
<td align="left">The time grid each date lives on: <code>days</code>, <code>weeks</code>, <code>months</code>, <code>years</code> or <code>numeric</code>. <em>Optional</em>; inferred (<code>"auto"</code>) by default.</td>
</tr>
<tr>
<td align="center"><img src="man/figures/censoring.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="is_censored_report"></td>
<td align="left"><code>is_censored_report</code>,<code>is_censored_revision</code></td>
<td align="left">Flags dates from either the report or the revision axis that are only an upper bound, i.e. the true report happened <i>before</i> the date given in the database.  <em>Optional</em>.</td>
</tr>
<tr>
<td align="center"><img src="man/figures/temporal_effects.svg" height="80" style="height:80px;width:auto;max-width:80px;" alt="temporal_effects"></td>
<td align="left"><code>t_effects</code></td>
<td align="left">Columns holding temporal effects (day of week, holidays, &hellip;) that some models can use. <em>Optional</em>.</td>
</tr>
</tbody>
</table>

On top of these, a `tbl_now` keeps the **generated** columns it computes for
you: `.event_num`, `.report_num` and `.delay`, plus `.revision_num` and
`.revision_delay` when there is a revision date. They are protected: removing
or renaming one demotes the object back to an ordinary tibble, because the rest
of the package can no longer trust it.

You can access any attribute using the corresponding [getter](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html), e.g. [get_event_date()](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html) or [get_strata()](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html).


Below we provide more information on some of the attributes.

### Data types 

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)  can represent one of three data structures:


1. **Linelist**: Each row corresponds to a single reported observation.

<div class="small-table">
```{r, echo=FALSE, cache=FALSE}
df_linelist <- data.frame(
  patient = 1:6,
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/13"), 3)
  ),
  report_date = c(
    ymd("2020/09/12"),
    ymd("2020/09/12"),
    ymd("2020/09/13"),
    ymd("2020/09/13"),
    ymd("2020/09/13"),
    ymd("2020/09/13")
  )
)

knitr::kable(df_linelist, caption = "Linelist data")
```
</div>

2. **Count-incidence**: Each row summarizes how many events with a given `event_date` were reported **exactly** on that `report_date.`

<div class="small-table">
```{r, echo=FALSE, cache=FALSE}
df_count_inc <- data.frame(
  n = c(7, 1, 9, 5, 0, 2),
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/13"), 3)
  ),
  report_date = c(
    ymd("2020/09/12"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/15")
  )
)

knitr::kable(df_count_inc, caption = "Count-incidence data")
```
</div>

3. **Count-cumulative** Each row summarizes how many events with a given `event_date` had been reported up to and including that `report_date`. The distinction is crucial for nowcasting models that operate either on daily increments or cumulative totals.

<div class="small-table">
```{r, echo=FALSE, cache=FALSE}
df_count_cum <- data.frame(
  n = c(1, 5, 8, 2, 2, 4),
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/13"), 3)
  ),
  report_date = c(
    ymd("2020/09/12"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/13"),
    ymd("2020/09/14"),
    ymd("2020/09/15")
  )
)

knitr::kable(df_count_cum, caption = "Count-cumulative data")
```
</div>

The `to_count()` function allows you to convert between different data-types as we see below:

### Converting Between Data Types

The [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html) function supports structured transformations. Here we start with linelist data as an example:

```{r}
#The original data.frame has one row per patient
df_linelist <- data.frame(
  patient     = 1:6,
  event_date  = c(rep(ymd("2020/09/12"), 3), rep(ymd("2020/09/13"), 3)),
  report_date = c(rep(ymd("2020/09/12"), 2), rep(ymd("2020/09/13"), 4))
)

#We can convert it to a tbl.now
df_linelist <- df_linelist |> 
  tbl_now(event_date = event_date, report_date = report_date, 
          data_type = "linelist")

#This is what it looks like
df_linelist
```

+ **Linelist → Count-Incidence**: Aggregates by event–report date, counting only cases reported on that date.

```{r}
df_linelist |>
  to_count(to = "count-incidence")
```

+ **Linelist → Count-Cumulative**: Aggregates by event–report date, producing cumulative counts up to each report date.

```{r}
df_linelist |>
  to_count(to = "count-cumulative")
```

> **Note** In the previous example the `n` counts `3` as it is aggregating the `1` observed at `report_date = 2020-09-13` and the `2` observed at `report_date = 2020-09-12`. This is the difference between the **count-incidence** that specifies the ones observed **exactly** on that date and the **count-cumulative** that specifies the ones observed up **until and including** that date. 

+ **Count-Incidence → Count-Cumulative**: Computes cumulative sums for each event date across report dates.

```{r}
df_count_inc <- df_linelist |>
  to_count(to = "count-incidence")

#This is count incidence:
df_count_inc

#Turns to count cumulative:
df_count_inc |>
  to_count(to = "count-cumulative")
```

+ **Aggregation within the same type**: The [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html) may also be used to re-aggregate datasets that contain duplicate event–report pairs. This is useful when raw surveillance feeds contain repeated entries such as in this case:

```{r}
tbl_example <- data.frame(
  n = c(8, 11, 0, 1, 1, 5, 2, 4, 1, 10, 9, 11, 3, 1),
  sex = c(rep("M", 3), rep("F", 4), rep("M", 2), rep("F", 5)),
  event_date = c(
    rep(ymd("2020/09/12"), 3),
    rep(ymd("2020/09/12"), 4),
    rep(ymd("2020/09/13"), 2),
    rep(ymd("2020/09/13"), 5)
  ),
  report_date = c(
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/15"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17")
  )) |>
  tbl_now(
    event_date = event_date, report_date = report_date,
    data_type = "count-incidence", case_count = n, verbose = FALSE,
    warn_non_uniqueness = FALSE
  )

tbl_example
```

This dataset intentionally contains repeated event_date–report_date pairs for each `sex`. You can  aggregate redundant rows with the [to_count()](https://rodrigozepeda.github.io/tbl.now/reference/to_count.html)  function that collapses duplicates by summing the `case_count` column.


```{r}
tbl_example |>
  to_count(to = "count-incidence")
```

The function ensures that:

+ Rows are grouped by `event_date`, `report_date`, and any `strata`, and `is_censored_report`.

+ The `case_count` column is summed within each group.

+ Attributes are preserved so the resulting object remains a valid `tbl_now`.


### Changing the time unit

`to_count()` changes *if there is a count and what the count column is*; it does not change *how long a period is*. Sometimes, surveillance data is too sparse to nowcast and the usual fix is to work in weeks.
The [aggregate_time_units()](https://rodrigozepeda.github.io/tbl.now/reference/aggregate_time_units.html) function  moves every date onto a coarser time-grid. For example passing from daily data:

```{r}
daily <- data.frame(
  onset    = ymd("2024/01/01") + c(0, 1, 3, 8, 9, 15),
  reported = ymd("2024/01/01") + c(2, 2, 5, 9, 12, 16),
  sex      = c("F", "M", "F", "M", "F", "M")
) |>
  tbl_now(
    event_date = onset, report_date = reported, strata = sex,
    data_type = "linelist", units = "days", verbose = FALSE
  )

daily
```

to weekly:

```{r}
weekly <- daily |> 
  aggregate_time_units(to = "weeks", verbose = FALSE)

weekly
```

### Censoring the dates that are not really dates

Surveillance data arrives with missing report dates or stand-ins such as
`2222-02-22` for "never reported". Deleting those rows throws away
real cases; believing them makes the nowcast think reporting is far slower than
it is. The
[censoring](https://rodrigozepeda.github.io/tbl.now/reference/censoring.html)
 keep the case and record its delay as an upper *bound*:

```{r, warning=FALSE}
messy <- data.frame(
  onset    = ymd("2020/01/01") + 0:3,
  reported = ymd(c("2020/01/03", NA, "2222/02/22", "2020/01/06"))
  ) |>
  tbl_now(
    event_date = onset, report_date = reported, data_type = "linelist",
    units = "days", now = ymd("2020/01/10"), verbose = FALSE
  ) 

#Here we use a condition to tell what reports should be upper bounded
fixed <- messy |>
  censor_reports(
    is.na(reported) | reported > ymd("2100/01/01"), 
    to_report = ymd("2020/01/10"), verbose = FALSE
  ) 

fixed
```

The main idea is that reports are always upper bounded (if you are seeing a report today then the most recent date it could have arrived is today!). Some nowcasting models work with these censored frameworks. 

### Temporal effects

Oftentimes, temporal covariates improve nowcasting performance by helping to adjust systematic changes within the calendar cycle (e.g., day-of-week effects, seasonal effects, or other reporting artefacts). The [temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/temporal_effects.html) function creates a *specification* (recipe) of the features to compute:

```{r, eval=has("almanac")}
library(almanac)

t_eff <- temporal_effects(
  day_of_week  = TRUE,
  week_of_year = TRUE,
  holidays     = cal_us_federal()
)

t_eff
```

```{r, eval=!has("almanac"), echo =FALSE}
t_eff <- temporal_effects(
  day_of_week  = TRUE,
  week_of_year = TRUE
)

t_eff
```

> Note that the holidays calendar is an [rcalendar](https://davisvaughan.github.io/almanac/reference/rcalendar.html) object from the [almanac](https://davisvaughan.github.io/almanac/articles/almanac.html) package.

#### How do they work?

Temporal effects in `tbl.now` follow a **lazy evaluation** pattern:

1. **Add** with [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html) (or via the `t_effects` argument of `tbl_now()`). This records *what* should be computed but adds **no columns** yet.

2. **Materialise the columns** with [compute_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/compute_temporal_effects.html) when you are ready to use them in a model.

```{r}
data("denguedat")

# Step 1 — create the tbl_now and attach the spec (no columns added yet)
df_now <- denguedat |>
  tbl_now(
    event_date = onset_week, report_date = report_week,
    verbose = FALSE, strata = gender
  )

df_now <- df_now |>
  add_temporal_effects(t_eff)

# The footer shows "T. effects (lazy): ..." — spec is recorded but not computed
df_now
```

The footer now shows the spec with `(lazy)` to signal that the columns have not been computed yet. No new columns appear in the tibble at this point. When we compute, they appear (scroll to the right):

```{r}
# Step 2 — materialise the columns when needed
df_computed <- compute_temporal_effects(df_now)

# Columns are now present and annotated [t_effect]
df_computed
```

After `compute_temporal_effects()`:

+ The effect columns (`.event_day_of_week`, `.event_week_of_year`, `.event_holiday`) are added.
+ The function `get_temporal_effect_cols()` lists the column names while
+ The original call remains accessible via `get_temporal_effects()`, so you always know which effects were requested even after further dplyr operations.

```{r, eval=TRUE}
get_temporal_effects(df_computed)     # The spec (list of configs)
get_temporal_effect_cols(df_computed) # The computed column names
```

> **Holidays on a grid coarser than days.** `denguedat` is weekly, and a week is
> not a holiday -- it *contains* holidays. On daily data `.event_holiday` is the
> usual `0`/`1` indicator; on weekly, monthly or yearly data it is the **share of
> the period's days**. A week containing Christmas Day for example would have a
> scores `1/7`. A week with no holiday a `0`.

#### Around-holiday and around-weekend effects

Reporting often *rebounds* on the first working day(s) after a holiday or a
weekend. To capture that, `temporal_effects()` has `holiday_lags` and
`weekend_lags`: each takes a depth `N` and creates indicator columns
`..._holiday_lag_1 … ..._holiday_lag_N` (and `..._weekend_lag_k`) that flag dates
falling exactly `k` **working days** after a holiday / weekend. Working days skip
weekends and other holidays, so the effect lands on the first day back at work.

```{r, eval=has("almanac")}
# Flag the two working days after a holiday, and the working day after a weekend
after_eff <- temporal_effects(
  holidays     = cal_us_federal(),
  holiday_lags = 2,
  weekend_lags = 1
)
after_eff
```

```{r, eval=!has("almanac"), echo=FALSE}
# Flag the two working days after a holiday, and the working day after a weekend
after_eff <- temporal_effects(
  weekend_lags = 1
)
after_eff
```

The inverse,  a slowdown in the days *leading up to* a break is a negative
depth. `..._holiday_lead_k` / `..._weekend_lead_k`:

```{r, eval=has("almanac")}
# Flag Christmas Eve (and the eve of every other holiday), plus 
# Thursday and Friday before each weekend
before_eff <- temporal_effects(
  holidays     = cal_us_federal(),
  holiday_lags = -1,
  weekend_lags = -2 #Flag thursday and Friday
)
before_eff
```

```{r, eval=!has("almanac"), echo=FALSE}
before_eff <- temporal_effects(
  weekend_lags = -1
)
before_eff
```

To model both sides of the same break, attach one specification per direction:

```{r, eval=FALSE}
df_now |>
  add_temporal_effects(after_eff) |> 
  add_temporal_effects(before_eff) 
```

#### Event-, report-, and revision-date effects

By default effects are derived from the **event date** . Pass
`date_type = "report_date"` to `add_temporal_effects()` to derive them from the
**report date** instead:

```{r, eval=TRUE}
df_now |>
  add_temporal_effects(temporal_effects(week_of_year = TRUE), date_type = "event_date") |>
  add_temporal_effects(temporal_effects(day_of_week = TRUE),  date_type = "report_date")
```

### The revision process

Some surveillance systems have a **third** date. A case can not only
be reported but once its reported it is later *resolved*: a laboratory issues 
the result that confirms it, or rules it out for example. 
Influenza is the standard picture: symptoms begin (the event), the
patient visits a doctor (the report), and days later a swab comes back positive
(the **confirmation**) or negative (a *retraction*). 

A `tbl_now` can carry this with `revision_date`, an optional
`revision_type`, and its own `revision_units`. 

The timeline it assumes is:

$$\text{event date} \to \text{report date} \to \text{revision date}$$

We will use `covid_us`, the CDC's COVID-19 case surveillance data for 2020,
which records all three dates: when symptoms began, when the first positive
specimen was collected, and when the case was registered at CDC with a status.

```{r, message=FALSE}
data("covid_us")

covid_now <- covid_us |>
  filter(onset_dt >= as.Date("2020-09-01")) |>
  tbl_now(
    event_date        = onset_dt,      # symptoms began
    report_date       = pos_spec_dt,   # the first positive specimen
    revision_date   = cdc_report_dt, # the case was registered at CDC
    revision_type   = current_status,
    revision_levels = c(
      "Laboratory-confirmed case" = "confirmed",
      "Probable Case"             = "pending"
    ),
    case_count = n,
    strata     = sex,
    data_type  = "count-incidence",
    verbose    = FALSE
  )

covid_now
```

#### `revision_levels`: the revision dictionary

CDC does not say `"confirmed"`; it says `"Laboratory-confirmed case"`. The
`revision_levels` exist to translate between the data and what `tbl.now` expects. It
is a named vector mapping the data to `tbl.now`'s valid levels.  

```{r, eval=FALSE}
c("Laboratory-confirmed case" = "confirmed", "Probable Case" = "pending")
```


### Getting, removing and changing attributes

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)’s attributes can be modified using the functions in [add_*](https://rodrigozepeda.github.io/tbl.now/reference/add.html), [change_*](https://rodrigozepeda.github.io/tbl.now/reference/change.html), or [remove_*](https://rodrigozepeda.github.io/tbl.now/reference/remove.html). These functions share a consistent interface that allows users to incrementally manipulate strata, covariates, and temporal effects.

The example below demonstrates how to create a  [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html), add strata and temporal effects, later modify the strata, and finally remove the temporal effects.

```{r, message=FALSE}
data("mpoxdat")

df_now <- mpoxdat |>
  tbl_now(
    event_date = dx_date, report_date = dx_report_date,
    case_count = n, verbose = FALSE, strata = race
  )

df_now
```

You can see that the strata is `race` with the corresponding [get_*](https://rodrigozepeda.github.io/tbl.now/reference/nowcast_data_getters.html):

```{r}
get_strata(df_now)
```

Strata can be modified with the [change_*](https://rodrigozepeda.github.io/tbl.now/reference/change.html) family of functions. The following example adds a new column containing an uppercase version of the existing race variable and sets it as the new strata:

```{r}
df_now <- df_now |>
  mutate(RACE_UPPER = toupper(race)) |>
  change_strata(RACE_UPPER)

get_strata(df_now)
```

To attach a lazy temporal-effects spec, use [add_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html), then materialise with [compute_temporal_effects()](https://rodrigozepeda.github.io/tbl.now/reference/compute_temporal_effects.html):

```{r, R.options=list(pillar.width = 250)}
df_now <- df_now |>
  add_temporal_effects(temporal_effects(week_of_year = TRUE))

# Spec is stored (lazy):
get_temporal_effects(df_now)

# Compute to add columns:
df_now <- compute_temporal_effects(df_now)
get_temporal_effect_cols(df_now)
```

Attributes can be removed using the corresponding [remove_*](https://rodrigozepeda.github.io/tbl.now/reference/remove.html) functions. `remove_temporal_effects()` drops both the spec and any computed columns:

```{r}
df_now <- df_now |>
  remove_temporal_effects() |>
  remove_all_strata()

df_now
```

```{r}
get_temporal_effects(df_now) # Empty list — no spec
get_temporal_effect_cols(df_now) # character(0) — no computed cols
get_strata(df_now)
```

## Modifying a tbl_now() with dplyr

[tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) objects extend [tibble()](https://tibble.tidyverse.org/), and therefore support standard [dplyr](https://dplyr.tidyverse.org/)
verbs. The class attempts to preserve and adapt its internal attributes when operations are performed. For example, renaming a strata column will automatically update the stored strata attribute.

```{r}
library(dplyr, quietly = TRUE)

data(denguedat)

df_now <- tbl_now(denguedat,
  event_date = onset_week,
  report_date = report_week, strata = gender,
  verbose = FALSE
)

# Current strata
get_strata(df_now)
```


After renaming the column, the strata attribute updates accordingly:

```{r}
df_now <- df_now |>
  rename(male_or_female = gender)

get_strata(df_now)
```

Certain operations may cause a [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) object to drop back to a standard tibble. This occurs when the operation removes necessary structure—for example, collapsing all data into a single row with [summarise()](https://dplyr.tidyverse.org/reference/summarise.html):

```{r}
df_now |>
  summarise(number_males = sum(male_or_female == "Male"))
```
 
 
Other operations like mutate and select work as long as the original columns and the required for the attributes are kept:

```{r}
df_now <- df_now |>
  mutate(GENDER = toupper(male_or_female)) |> 
  select(GENDER, everything())

df_now
```

 
### Updating a tbl_now()

A [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) object can be updated using the [update()](https://rodrigozepeda.github.io/tbl.now/reference/update.tbl_now.html) method with another data.frame, tibble, or [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html) as input. When the column structure is compatible, the update process retains the strata, covariate, and temporal-effect attributes from the original object, and recalculates "now" estimates using the combined data.

Below is an example using an initial dataset:

```{r}
df <- data.frame(
  patient = 1:6,
  event_date = c(rep(ymd("2020/09/12"), 3), rep(ymd("2020/09/13"), 3)),
  report_date = c(
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/13"), ymd("2020/09/14"), ymd("2020/09/15")
  )
)

df_now <- tbl_now(df,
  event_date = event_date,
  report_date = report_date, verbose = FALSE
)
```

And a follow-up dataset containing newly reported cases:

```{r}
df_new <- data.frame(
  patient = 7:13,
  event_date = c(
    ymd("2020/09/13"),
    rep(ymd("2020/09/14"), 3),
    rep(ymd("2020/09/15"), 3)
  ),
  report_date = c(
    ymd("2020/09/14"), ymd("2020/09/14"), ymd("2020/09/15"),
    ymd("2020/09/16"), ymd("2020/09/15"), ymd("2020/09/16"),
    ymd("2020/09/17")
  )
)
```

We can update the original object by incorporating the new data:

```{r}
df_updated <- update(df_now, new_data = df_new)

df_updated
```

## Other functions (utilities)

### Convert epidemiological weeks to dates

The function [week_2_date()](https://rodrigozepeda.github.io/tbl.now/reference/week_2_date.html) converts epidemiological week/year combinations into a calendar date aligned on the first day of the week (Sunday).

```{r}
df <- data.frame(
  epidemiological_week = 1:5,
  epidemiological_year = rep(2024, 5)
)

df |>
  week_2_date(
    week_col = epidemiological_week,
    year_col = epidemiological_year
  )
```
 
### Reports

The functions [get_initial_reported_cases() and get_latest_reported_cases()](https://rodrigozepeda.github.io/tbl.now/reference/get_latest_first.html) extract the number of cases first reported for each event date and the most recently reported totals, respectively. These utilities allow users to quantify revisions between initial and final reports.
 
 
```{r}
df_reports <- data.frame(
  n = c(10, 1, 1, 0, 0, 3),
  event_date = rep(ymd("2020/09/12"), 6),
  report_date = c(
    ymd("2020/09/12"), ymd("2020/09/13"), ymd("2020/09/14"),
    ymd("2020/09/15"), ymd("2020/09/16"), ymd("2020/09/17")
  )
)

tbl_reports <- df_reports |>
  tbl_now(
    event_date = event_date, report_date = report_date,
    verbose = FALSE, case_count = n, report_units = "days",
    event_units = "days"
  )

tbl_reports
```


The initial reported cases:

```{r}
get_initial_reported_cases(tbl_reports)
```

and the latest totals:

```{r}
get_latest_reported_cases(tbl_reports)
```

For scenarios with revisions check also `get_nth_revised_cases()`, `get_initial_revised_cases()` and 
`get_final_revised_cases()`.

### Week alignment

The [align_weeks()](https://rodrigozepeda.github.io/tbl.now/reference/align_weeks.html) 
function standardizes dates within the same epidemiological week to a single reference day
(for example, the start of the week). This is helpful when computing differences across weekly reporting periods, avoiding fractional time intervals.

```{r}
df <- data.frame(
  date = c(ymd("2022-10-31"), ymd("2022-11-07"), ymd("2022-11-13")),
  epiweek = c(44, 45, 46)
)

# Align to Sundays
df_aligned <- align_weeks(df, date_col = date)
df_aligned
```

You can verify the resulting weekday using the [wday() function from the lubridate package](https://lubridate.tidyverse.org/reference/day.html):

```{r}
df_aligned |>
  mutate(day_label = wday(date_aligned, label = TRUE, abbr = FALSE))
```

### Complete zeroes

The [complete_zeroes()](https://rodrigozepeda.github.io/tbl.now/reference/complete_zeroes.html) function fills with zeroes those cases where the `event` or `report` weeks have not been observed. 

Consider for example the following data with just two observations per date:

```{r}
ndata <- tibble(
  event_date = c(as.Date("2021/01/12"), as.Date("2021/01/14"), as.Date("2021/01/14")),
  report_date = c(as.Date("2021/01/13"), as.Date("2021/01/15"), as.Date("2021/01/18")),
  case_count = c(10, 5, 1)
)

ndata <- tbl_now(ndata, event_date, report_date,
  verbose = FALSE, case_count = case_count, data_type = "count-incidence"
)

ndata
```

Notice that there are no observations for `2021/01/13`. Furthermore, if we assume that the maximum possible observed delay is of `4`,  we can fill the unobserved cases with:

```{r}
complete_zeroes(ndata, max_delay = 4)
```

Which looks at all the possible report dates and event dates and sets the counts to zero if they have not been observed.

### Censoring extreme delays

The function `censor_reporting_delays_above()` marks all delays above a threshold value (`max_delay`) as censored. This is useful to indicate extreme delays in some nowcast models:

```{r}
df <- data.frame(
  onset = as.Date("2020-01-01") + c(0, 0, 1, 2),
  reported = as.Date("2020-01-01") + c(1, 5, 2, 300)
)
tn <- tbl_now(df,
  event_date = onset, report_date = reported,
  data_type = "linelist", verbose = FALSE
)

# the 300-day report becomes censored (an upper bound on its delay)
censor_reporting_delays_above(tn, max_delay = 60)
```

## Converting to data formats from other packages

`tbl.now` ships converters that move data between a `tbl_now` and the data structures used by other nowcasting and delay-estimation packages. They all follow the same naming convention:

+ `tbl_now_from_*()` builds a `tbl_now` (it wraps [as_tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/as_tbl_now.html), so `...` is forwarded to [tbl_now()](https://rodrigozepeda.github.io/tbl.now/reference/tbl_now.html)).

+ `tbl_now_to_*()` converts a `tbl_now` into that package's native object.

Each function accepts a `verbose` argument that reports the choices it made, such as the inferred `now`, data type, strata, and conversion format. For example, here we can convert to `tsibble`:

```{r converters-roundtrip, eval=has("tsibble"), message=FALSE}
library(tsibble)

dengue_now <- tbl_now(denguedat,
  event_date = "onset_week",
  report_date = "report_week", strata = "gender",
  verbose = FALSE
)

# tbl_now -> tsibble -> tbl_now
dengue_ts <- tbl_now_to_tsibble(dengue_now, verbose = FALSE)

#This returns a tsibble
dengue_ts

#Which can be converted back to tbl.now
as_tbl_now(dengue_ts, report_date = "report_week", verbose = FALSE)
```

See the [vignette on using different models](https://rodrigozepeda.github.io/tbl.now/articles/nowcasting-models.html) to see all conversion formats. 


```{r learning-more, echo=FALSE, results="asis"}
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
```

## References
