---
title: "One dataset, many nowcasts: using tbl.now with five modelling packages"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{One dataset, many nowcasts: using tbl.now with five modelling packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The heavy model fits (Stan / JAGS / MCMC) are shown as illustrative code but
# not run when the vignette is built. Flip `run_models` to TRUE (and make sure
# the corresponding packages are installed) to actually fit them.
run_models <- FALSE

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)
```

## Why this vignette?

If you work in public health surveillance you have almost certainly hit the
same wall: **reported case counts for the most recent days or weeks are
incomplete.** A case that happened last week may not be reported until next
week, so the freshest numbers always look artificially low. *Nowcasting* is the
family of methods that corrects for this reporting delay and estimates what the
recent counts will look like once reporting is complete.

The good news is that the `R` ecosystem has many excellent nowcasting packages.
The bad news is that **each one wants its data in a different shape**: a
reporting triangle here, a cumulative long table there, a line list somewhere
else. Preparing the same outbreak data five different ways is tedious and
error-prone.

This is exactly the gap [`tbl.now`](https://rodrigozepeda.github.io/tbl.now/)
fills. You describe your data **once** — which column is the event date, which
is the report date, what the counts mean — and `tbl.now` hands it to each
modelling package in the format that package expects. In this vignette we take a
single dengue line list and, from that one object, drive five different
nowcasting / delay-estimation tools:

| Package | What it does | `tbl.now` bridge |
|---|---|---|
| [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) | Bayesian structural time-series nowcast | consumes a `tbl_now` **directly** |
| [baselinenowcast](https://baselinenowcast.epinowcast.org/) | fast, assumption-light baseline nowcast | `tbl_now_to_baselinenowcast()` |
| [epinowcast](https://package.epinowcast.org/) | flexible Bayesian nowcast (delay + reference modules) | `tbl_now_to_epinowcast()` |
| [epidist](https://epidist.epinowcast.org/) | estimates the reporting **delay distribution** | `tbl_now_to_epidist()` |
| [NobBS](https://cran.r-project.org/package=NobBS) | Nowcasting by Bayesian Smoothing | a plain line list (`as.data.frame()`) |

You do not need to be an expert in any of them to follow along — the point is
how little changes on *your* side when you switch models.

## The data: one `tbl_now` to rule them all

`tbl.now` ships with `denguedat`, a weekly dengue line list (one row per case)
with the week of symptom **onset** and the week the case was **reported**.

```{r setup, message=FALSE}
library(tbl.now)

data(denguedat)
head(denguedat)
```

We build a single `tbl_now`. We only have to say which column is the event date
(symptom onset) and which is the report date; `tbl.now` infers the rest (the
data are a line list, measured in weeks, and the "now" is the last report).

```{r build-tbl-now}
# A compact recent window keeps the vignette fast to build.
dengue <- denguedat[denguedat$onset_week >= as.Date("2009-01-01"), ]

dengue_now <- tbl_now(
  dengue,
  event_date  = onset_week,
  report_date = report_week,
  data_type   = "linelist"
)

dengue_now
```

That is the *only* data-wrangling step in this whole vignette. Everything below
starts from `dengue_now`.

## 1. diseasenowcasting — no conversion at all

[`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) was
designed hand-in-hand with `tbl.now`, so it takes a `tbl_now` **directly**. You
just hand it the object:

```{r dnc, eval=run_models && has("diseasenowcasting")}
library(diseasenowcasting)

dnc_fit <- nowcast(dengue_now)

dnc_fit
```

`diseasenowcasting` will even reach into the `tbl_now` for any temporal effects
you attached (see the [temporal effects](https://rodrigozepeda.github.io/tbl.now/articles/tbl.now.html#temporal-effects)
section of the main vignette). For example, to let the model account for a
seasonal cycle and a post-holiday reporting bump:

```{r dnc-teff, eval=FALSE}
library(almanac)

dengue_seasonal <- dengue_now |>
  add_temporal_effects(
    temporal_effects(
      seasons      = 52,                          # annual seasonality (weekly data)
      holidays     = rcalendar(hol_christmas()),  # Christmas holiday effect
      holiday_lags = 2                            # + the two weeks "after" it
    )
  )

nowcast(dengue_seasonal)
```

## 2. baselinenowcast — a reporting triangle

[`baselinenowcast`](https://baselinenowcast.epinowcast.org/) is a deliberately
simple, fast baseline. It works from a **reporting triangle**: a matrix with one
row per reference (onset) week and one column per reporting delay, where the
lower-right corner is the not-yet-observed part we want to fill in.

`tbl_now_to_baselinenowcast()` builds that triangle for you (aggregating the
line list to weekly incidence on the way):

```{r bln-convert, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
library(baselinenowcast)

# `delays_unit` defaults to NULL and is inferred from the object's weekly units;
# pass it explicitly (e.g. delays_unit = "weeks") if the units are ambiguous.
dengue_triangle <- tbl_now_to_baselinenowcast(dengue_now, verbose = FALSE)

# rows = onset weeks, columns = delay in weeks
dengue_triangle[1:5, 1:6]
```

Because `as_reporting_triangle()` is registered as a method for `tbl_now`, you
can also write the idiomatic `baselinenowcast` call and pass the `tbl_now`
straight in:

```{r bln-method, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
dengue_triangle2 <- as_reporting_triangle(dengue_now)
identical(dim(dengue_triangle), dim(dengue_triangle2))
```

From here you follow `baselinenowcast`'s own workflow — estimate the delay from
the triangle and sample nowcasts:

```{r bln-fit, eval=run_models && has("baselinenowcast")}
# Estimate the delay from the triangle, apply it, and draw nowcast samples
# (see the baselinenowcast documentation for the full workflow).
point_nowcast <- dengue_triangle |>
  estimate_and_apply_delays()

nowcast_samples <- point_nowcast |>
  sample_nowcasts()
```

## 3. epinowcast — a preprocessed Bayesian model object

[`epinowcast`](https://package.epinowcast.org/) fits a flexible Bayesian model
with separate modules for the reporting delay and the reference (epidemic)
process. It expects **cumulative** counts and a preprocessed object built by
`enw_preprocess_data()`.

`tbl_now_to_epinowcast()` handles the cumulation and the preprocessing, returning
an object you can pass straight to `epinowcast::epinowcast()`:

```{r enw-convert, eval=run_models && has("epinowcast"), message=FALSE, warning=FALSE}
library(epinowcast)

dengue_enw <- tbl_now_to_epinowcast(dengue_now, verbose = FALSE, quiet = TRUE)

dengue_enw
```

```{r enw-fit, eval=FALSE}
# A minimal epinowcast fit from the preprocessed object
fit <- epinowcast(
  dengue_enw,
  fit = enw_fit_opts(pp = TRUE, chains = 2, iter_sampling = 500)
)
```

If you attached temporal effects to the `tbl_now`, they are carried into the
`metareference` / `metareport` tables as covariate columns, ready for use in
epinowcast's reference and report modules.

## 4. epidist — the reporting delay distribution

Sometimes the quantity you actually want is the **delay distribution** itself —
how long, on average, between onset and report, and how variable is it?
[`epidist`](https://epidist.epinowcast.org/) estimates exactly that, treating
each case as an interval-censored onset/report pair.

`tbl_now_to_epidist()` converts the `tbl_now` into the censored line list
`epidist` expects. Because our data are weekly, each event is censored over its
whole week automatically:

```{r epidist-convert, eval=has("epidist"), message=FALSE, warning=FALSE}
library(epidist)

dengue_epidist <- tbl_now_to_epidist(dengue_now, verbose = FALSE)

head(dengue_epidist)
```

```{r epidist-fit, eval=FALSE}
# Fit the delay distribution (see the epidist documentation for model choices)
delay_model <- dengue_epidist |>
  as_epidist_marginal_model() |>
  epidist()
```

The fitted delay distribution can then feed back into a nowcast — for example as
a prior in `epinowcast` or `EpiNow2`.

## 5. NobBS — Nowcasting by Bayesian Smoothing

[`NobBS`](https://cran.r-project.org/package=NobBS) works from a **line list**
with an onset-date column and a report-date column — precisely what a `linelist`
`tbl_now` already is. Drop the `tbl.now` classes with `as.data.frame()` and hand
the columns to `NobBS()`:

```{r nobbs-convert, eval=has("NobBS")}
dengue_linelist <- as.data.frame(dengue_now)

names(dengue_linelist)
```

```{r nobbs-fit, eval=run_models && has("NobBS")}
library(NobBS)

nobbs_fit <- NobBS(
  data        = dengue_linelist,
  now         = get_now(dengue_now),
  units       = "1 week",
  onset_date  = get_event_date(dengue_now),
  report_date = get_report_date(dengue_now)
)

nobbs_fit$estimates |> head()
```

Notice that even the arguments come from the `tbl_now`: `get_now()`,
`get_event_date()` and `get_report_date()` tell `NobBS` what `tbl.now` already
figured out.

## Carrying delay effects into each model

Reporting delays are rarely constant — they wax and wane with the day of the
week, the season, and around holidays. `tbl.now` lets you attach these
**temporal (delay) effects** once, with
[`add_temporal_effects()`](https://rodrigozepeda.github.io/tbl.now/reference/add_temporal_effects.html),
and then carries the resulting covariate columns into whichever model format can
hold them. (Use [`autoplot()`](https://rodrigozepeda.github.io/tbl.now/reference/autoplot.tbl_now.html)
with `panels = "delay_calendar"` or `panels = "delay_seasonality"` first to *see*
whether such effects exist in your data before modelling them.)

```{r attach-effects, eval=has("almanac"), message=FALSE}
library(almanac)

dengue_eff <- dengue_now |>
  add_temporal_effects(
    temporal_effects(
      week_of_year = TRUE,                        # week-of-year delay effect
      holidays     = rcalendar(hol_christmas()),  # holiday indicator
      holiday_lags = 1                            # + the first week after a holiday
    )
  )

# The spec is lazy; materialise it to see the columns it produces
compute_temporal_effects(dengue_eff) |>
  get_temporal_effect_cols()
```

These columns then ride along automatically when you convert. For example, into
the `baselinenowcast` long format and into `epinowcast`'s metadata tables:

```{r effects-baselinenowcast, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
bln_long <- tbl_now_to_baselinenowcast(dengue_eff, format = "long", verbose = FALSE)
grep("holiday|week_of_year", names(bln_long), value = TRUE)
```

```{r effects-epinowcast, eval=run_models && has("epinowcast"), message=FALSE, warning=FALSE}
enw <- tbl_now_to_epinowcast(dengue_eff, verbose = FALSE, quiet = TRUE)
# The effect columns appear in the reference and report metadata tables,
# ready to be used in epinowcast's reference / report module formulas.
grep("holiday|week_of_year", names(enw$metareference[[1]]), value = TRUE)
```

**Where do the columns end up, and does the model use them?** Carrying a column
and *using* it are two different things — `tbl.now` does the former; whether a
model consumes it is up to that model:

| Target | Delay-effect columns | Generic `covariates` | How the model can use them |
|---|---|---|---|
| `diseasenowcasting` | consumed automatically | — | `nowcast()` reads the effect columns straight from the `tbl_now` and adds them to the model |
| `epinowcast` | in `metareference` **and** `metareport` | dropped (fixed schema) | reference / report module formulas (e.g. `~ 1 + .event_holiday`) |
| `baselinenowcast` (long) | extra columns | extra columns | carried for you; the core triangle estimator ignores them |
| `baselinenowcast` (matrix) | dropped | dropped | the matrix holds only counts |
| `epidist` | extra linelist columns | extra linelist columns | available to formula-based delay models |
| `data.table` / `tsibble` | kept | kept | yours to use however you like |
| `NobBS` | not used | not used | `NobBS` models only the onset/report dates |

Two practical notes for practitioners:

* **Delay effects are date-derived**, so they aggregate cleanly from a line list
  to counts and survive every conversion above. A **generic per-case covariate**
  (say a per-patient lab value) does *not* aggregate to a reference-date ×
  report-date cell, so it is only meaningful for the line-list / long formats —
  and `epinowcast`'s cumulative schema will reject it.
* The `baselinenowcast` **matrix** and any purely count-based format cannot carry
  covariates at all; use the **long** format when you want the effects to travel.

## The pay-off

Look back at what changed between the five models: **almost nothing.** We
described the dengue data once as a `tbl_now`, and then a single converter call
(or, for `diseasenowcasting`, no call at all) handed it to each package in the
shape it needed:

```{r recap, eval=FALSE}
dengue_now <- tbl_now(dengue, event_date = onset_week, report_date = report_week,
                      data_type = "linelist")

dengue_now                              # diseasenowcasting
tbl_now_to_baselinenowcast(dengue_now)  # baselinenowcast
tbl_now_to_epinowcast(dengue_now)       # epinowcast
tbl_now_to_epidist(dengue_now)          # epidist
as.data.frame(dengue_now)               # NobBS or others
```

For a public-health team this means you can **benchmark several nowcasting
methods on your own surveillance data without re-plumbing it each time** —
making it practical to compare models, cross-check results, and pick the tool
that works best for your disease and reporting system.

## See also

* The main [`tbl.now` vignette](https://rodrigozepeda.github.io/tbl.now/articles/tbl.now.html)
  for the full anatomy of a `tbl_now`, data types, and temporal effects.
* [`autoplot()`](https://rodrigozepeda.github.io/tbl.now/reference/autoplot.tbl_now.html)
  for a quick diagnostic overview (delay distribution, epidemic curve, calendar
  effects, seasonality) of any `tbl_now`.
