---
title: "One dataset, many nowcasts: using tbl.now with different modelling packages"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{One dataset, many nowcasts: using tbl.now with different modelling packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The heavy model fits (Stan / JAGS / MCMC) are shown as illustrative code but
# not run when the vignette is built. Flip `run_models` to TRUE (and make sure
# the corresponding packages are installed) to actually fit them.
run_models <- FALSE

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)
```

## Why this vignette?

Preparing the same data different ways for the different nowcasting models can be tedious and
error-prone. This is exactly what [`tbl.now`](https://rodrigozepeda.github.io/tbl.now/)
helps with. You can describe your data once by specifying which column is the event date, which
is the report date, whether your data is linelist or has counts (and what _those_ counts mean!) and `tbl.now` will hands it to each modelling package in the format that package expects. 

> In this vignette we take a single dataset (`denguedat`), and from that one object, use several different nowcasting / delay-estimation tools:

| Package | What it does | Additional requirements | `tbl.now` converter |
|---|---|---|---|
| [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) | flexible Bayesian nowcast (delay + epidemic processes) | none (uses `RTMB`) | consumes a `tbl_now` **directly** |
| [baselinenowcast](https://baselinenowcast.epinowcast.org/) | fast, assumption-light baseline nowcast | none | `tbl_now_to_baselinenowcast()` |
| [epinowcast](https://package.epinowcast.org/) | flexible Bayesian nowcast (delay + reference modules) | **Stan** | `tbl_now_to_epinowcast()` |
| [epidist](https://epidist.epinowcast.org/) | estimates only the reporting **delay distribution** | **Stan** | `tbl_now_to_epidist()` |
| [NobBS](https://cran.r-project.org/package=NobBS) | Nowcasting by Bayesian Smoothing | **JAGS** | a plain line-list (`as.data.frame()`) |
| [surveillance](https://cran.r-project.org/package=surveillance) | the classic Höhle & an der Heiden nowcast | none for the method used here (**JAGS** for `bayes.trunc.ddcp`) | `tbl_now_to_surveillance()` |
| [nowcaster](https://covid19br.github.io/nowcaster/) | Bayesian nowcast fitted with INLA | **R-INLA** | `tbl_now_to_nowcaster()` |

You do not need to be an expert in any of them to follow along, the point is to show
how little changes on *your* side when you switch models.

::: {.alert .alert-warning}
**Each of these is a separate package, and `tbl.now` does not install any of
them.** It only knows how to *talk* to them. Install whichever you actually want
to use — and note that some also need software outside R (the *Additional
requirements* column above): **Stan** via `cmdstanr`, **JAGS** as a standalone
program, and **R-INLA**, which is not on CRAN.

```r
# CRAN packges
install.packages(c("baselinenowcast", "NobBS", "surveillance"))

# Not on CRAN:
install.packages("cmdstanr",   repos = c('https://stan-dev.r-universe.dev', getOption("repos")))
install.packages("epidist",    repos = "https://epinowcast.r-universe.dev")
install.packages("epinowcast", repos = "https://epinowcast.r-universe.dev")
install.packages("INLA",       repos = "https://inla.r-inla-download.org/R/stable")

# Packages from Github:
install.packages("pak") 
pak::pkg_install("covid19br/nowcaster")
pak::pkg_install("RodrigoZepeda/diseasenowcasting")
```

Every code block below is guarded, so the article still builds when a package is
missing — but nothing will run on *your* machine until you install it.
:::

## The data

`tbl.now` ships with `denguedat`, a weekly dengue line-list (one row per case)
from Puerto Rico with the week of symptom **onset** and the week the case was **reported**.

```{r setup, message=FALSE}
library(dplyr)
library(tbl.now)

data(denguedat)
```

```{r, echo = FALSE}
head(denguedat)
```

We build a single `tbl_now`. We only have to say which column is the event date
(`onset_week`) and which is the report date (`report_week`); `tbl.now` infers the rest (the
data is line-list, measured in weeks, and the "now" is the last report, `2010-12-20`).

We cut **both** dates in mid-July 2002. The example will assume we are
back in July 2002 and try to nowcast from them using only the information available
at that time (`now = "2002-07-22"`). 

```{r build-tbl-now}
cutoff <- as.Date("2002-07-22")

dengue <- denguedat |>
  filter(onset_week < cutoff, report_week < cutoff)

dengue_now <- tbl_now(
  dengue,
  event_date  = onset_week,
  report_date = report_week,
  data_type   = "linelist"
  )

dengue_now
```

That is `r format(nrow(dengue_now), big.mark = ",")` cases over
`r dplyr::n_distinct(dengue$onset_week)` weeks, from 1990 to
`r format(get_now(dengue_now))`. Everything in this article — every fit, every
figure — is built from this one object, so following along step by step
reproduces exactly what you see.

::: {.alert .alert-info}
**A note on series length.** Every example below runs on the full object, and
none takes more than a few seconds. The *fits* split into two groups:

* `diseasenowcasting` and `baselinenowcast` are given the **whole** series. Nothing is trimmed.
* `NobBS`, `nowcaster` and `surveillance` need the series shortened or the time
they take to run becomes impractical. Each package has an argument for that, and we
show how to use **it**: `moving_window` in `NobBS`, `wdw` in
  `nowcaster`, `when` in `surveillance`. 
* `epinowcast` has no such argument, because the reporting triangle is already
  built by the time you hold a preprocessed object. It is trimmed one step
  earlier, in its own section. We'll explain this below. 
  
:::

### A more complex object: strata and temporal effects

Reporting delays are rarely constant and many series are reported for several **strata** (here, `gender`) that you may want to nowcast separately. Here we build a second
object, `dengue_seasonal`, that adds a `gender` stratum and a week-of-year plus
annual-Fourier **temporal (delay) effects** on top of `dengue_now`:

```{r build-seasonal, message=FALSE}
dengue_seasonal <- dengue_now |>
  add_strata(gender) |>      # nowcast Male/Female separately
  add_temporal_effects(
    temporal_effects(
      week_of_year = TRUE,   # a separate level per epidemiological week
      seasons      = 52      # an annual Fourier cycle (period = 52 weeks)
    )
  )
```

For each package below we show the **bare** conversion of `dengue_now` first, then
how the same call carries the strata and the temporal effects when handed
`dengue_seasonal`. 

::: {.alert .alert-info}
**TIP**: Run `autoplot()`
with `panels = "delay_seasonality"` or `"delay_calendar"` first to *see* whether
such effects are present before modelling them.
:::

> **Checking each nowcast**
Every section below ends with the same figure: three panels — the **total**
series and the two **gender** strata — each showing what had been reported by
`now` (grey bars), what those weeks eventually reached (the dark **truth** line),
and that package's nowcast, with its uncertainty interval where the package
provides one.


```{r comparison-load, echo=FALSE, message=FALSE}
library(ggplot2)

comparison <- readRDS("nowcast-comparison.rds")
nowcasts   <- comparison$nowcasts
truth      <- comparison$truth
meta       <- comparison$meta

strata_levels  <- c("Total", "Female", "Male")

# The fits are precomputed (see `data-raw/nowcast_comparison.R`), so the article
# cannot call `print(fit)` or `tidy(fit)` live. It shows the REAL output of those
# calls instead: the precompute captured them and they are replayed here, so the
# code you see above each result is exactly the code that produced it.
show_fit <- function(package) {
  captured <- comparison$display[[package]]$printed
  if (is.null(captured)) return(invisible(NULL))
  cat(captured, sep = "\n")
}

show_tidy <- function(package, n = 5) {
  tidied <- comparison$display[[package]]$tidy
  if (is.null(tidied)) return(invisible(NULL))
  head(tidied, n)
}
engine_colours <- c(
  diseasenowcasting = "#0072B2",
  baselinenowcast   = "#D55E00",
  epinowcast        = "#009E73",
  NobBS             = "#CC79A7",
  surveillance      = "#E69F00",
  nowcaster         = "#56B4E9"
)
```


```{r comparison-helper, message=FALSE, echo=FALSE}
# One figure per package: Total, Female and Male side by side.
nowcast_panels <- function(package, subtitle = NULL) {
  window <- meta$horizon

  tru <- truth[truth$event_date %in% window, ]
  est <- nowcasts[
    nowcasts$package == package & nowcasts$event_date %in% window,
  ]
  tru$stratum <- factor(tru$stratum, levels = strata_levels)
  est$stratum <- factor(est$stratum, levels = strata_levels)

  colour <- unname(engine_colours[package])

  ggplot(mapping = aes(x = event_date)) +
    # What had been reported by `now` -- the incomplete data the engine saw.
    geom_col(
      data = tru, aes(y = observed, fill = "Reported by now"), width = 5.5
    ) +
    # The interval, where the engine gives one.
    geom_ribbon(
      data = est[!is.na(est$lower), ],
      aes(ymin = lower, ymax = upper), fill = colour, alpha = 0.25
    ) +
    # The nowcast...
    geom_line(data = est, aes(y = estimate, colour = "Nowcast"), linewidth = 0.9) +
    geom_point(data = est, aes(y = estimate, colour = "Nowcast"), size = 1.1) +
    # ...then the truth ON TOP, so it stays visible where the two coincide
    # (over the settled weeks they lie exactly on each other).
    geom_line(
      data = tru, aes(y = truth, linetype = "Truth"),
      linewidth = 0.6, colour = "#262626"
    ) +
    facet_wrap(~stratum, nrow = 1, scales = "free_y") +
    scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
    scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
    scale_colour_manual(name = NULL, values = c("Nowcast" = colour)) +
    labs(
      x = "Event week (symptom onset)", y = "Cases",
      title = paste0(package, " nowcast vs. what actually happened"),
      subtitle = subtitle %||% paste0(
        "Now = ", format(meta$now),
        "; last ", length(window), " weeks"
      )
    ) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position  = "top",
      legend.box       = "horizontal",
      legend.margin    = margin(b = -4),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold", colour = "#334335")
    )
}
```


## diseasenowcasting

[`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) is
designed hand-in-hand with `tbl.now`, so it takes a `tbl_now` **directly**. You
just hand it the object:

### Simple nowcast

```{r dnc, eval=run_models && has("diseasenowcasting")}
library(diseasenowcasting)

# The whole series, with the package's own defaults -- no trimming.
dnc_fit <- nowcast(dengue_now)

dnc_fit
```

```{r dnc-out, echo=FALSE, eval=!run_models}
show_fit("diseasenowcasting")
```


### With strata and effects.

`diseasenowcasting` reaches into the `tbl_now` itself, so the enriched object needs no extra arguments: it picks up the `gender` stratum and the week-of-year / seasonal effect columns automatically.

```{r dnc-teff, eval=run_models && has("diseasenowcasting")}
nowcast(dengue_seasonal)   # the strata and temporal effects are used automatically
```

```{r dnc-teff-out, echo=FALSE, eval=!run_models}
show_fit("diseasenowcasting_seasonal")
```

Predictions can be obtained via `tidy`:

```{r dnc-teff-tidy-1, eval=FALSE}
tidy(nowcast(dengue_seasonal))
```

```{r dnc-teff-tidy, echo=FALSE}
show_tidy("diseasenowcasting_seasonal")
```

In both stratified and unstratified cases:

```{r dnc-tidy-1, eval=FALSE}
tidy(dnc_fit)
```

```{r dnc-tidy, echo=FALSE}
show_tidy("diseasenowcasting")
```

::: {.alert .alert-info}
**NOTE**: this needs **diseasenowcasting >= 2.1.0**. Earlier versions declared
their own `tidy()` generic rather than re-exporting the shared one, so a bare
`tidy(dnc_fit)` returned that package's *model-parameter* table (`term`,
`std.error`, ...) instead of a nowcast — no error, just a different table. On
2.1.0 and later, `tidy()` returns the nowcast and the parameter table moved to
`model_parameters()`. On an older version, use
`tbl.now::tidy(predict(dnc_fit))`.

A related hazard remains with \pkg{broom}: loading it overwrites `tbl.now`'s
`tidy.list()` method, which is what \pkg{NobBS} and \pkg{nowcaster} fits
dispatch on. Qualify the call as `tbl.now::tidy()` there if \pkg{broom} is
attached.
:::

```{r dnc-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the diseasenowcasting package", fig.alt="Three panels (Total, Female, Male) comparing the diseasenowcasting nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("diseasenowcasting")
```

## baselinenowcast

[`baselinenowcast`](https://baselinenowcast.epinowcast.org/) is a 
simple, fast baseline. It works from a **reporting triangle**, a matrix with one
row per event (reference) week and one column per reporting delay. The
lower-right corner of the matrix corresponds to the not-yet-observed part the nowcast will fill in.

`tbl_now_to_baselinenowcast()` builds that triangle for you (aggregating the
line-list to weekly incidence on the way):

### Simple nowcast

```{r bln-convert, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
library(baselinenowcast)

# The converter transforms into the required reporting triangle.
dengue_triangle <- tbl_now_to_baselinenowcast(dengue_now, verbose = FALSE)

# You can see some entries of the triangle where
# rows = onset weeks, columns = delay in weeks
dengue_triangle[1:5, 1:6]
```

From here you can follow `baselinenowcast`'s own workflow. For example calling
`baselinenowcast()` to estimate the delay from the triangle, apply it, and
draw nowcast samples:

::: {.alert .alert-warning}
**Trim the rows that are forecasts, not nowcasts.** `baselinenowcast` divides
each observed row by the share of the delay distribution that should have
arrived by now. In `denguedat` almost nothing is reported in the same week —
`P(delay = 0)` is about **0.05** — so the newest event week, observable only at
delay 0, gets divided by 0.05 and explodes: a single case became an estimate of
**257 with an upper bound of 1584**, against a truth of 15.

This is not a week you can choose your way out of: completing the triangle to
the `now` *always* leaves a final row seen only at delay 0. `baselinenowcast`'s
own error text names the remedy — truncate "to an earlier reference time to
ensure a nowcast, not a forecast, is being produced" — so drop trailing rows
whose expected observed share is too small:

```{r bln-trim, eval=has("baselinenowcast")}
# Drop trailing reference times whose expected observed share is below 10%:
# those are forecasts, not nowcasts.
trim_forecast_rows <- function(triangle, min_observed_share = 0.10) {
  pmf <- as.numeric(estimate_delay(triangle))
  repeat {
    n_rows <- nrow(triangle)
    if (n_rows < 10) break
    observed_delays <- sum(!is.na(triangle[n_rows, ]))
    share <- sum(head(pmf, max(observed_delays, 1L)))
    if (share >= min_observed_share) break
    triangle <- triangle[seq_len(n_rows - 1L), , drop = FALSE]
  }
  triangle
}
```
:::

```{r bln-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One-call workflow: estimate the delay, apply it, and draw nowcast samples.
nowcast_samples <- baselinenowcast(
  trim_forecast_rows(dengue_triangle),
  output_type = "samples",
  draws       = 1000
)

```

### With strata and effects.

A reporting-triangle *matrix* is a single table with no room for strata or
covariates, so `format = "matrix"` pools the strata (with a warning) and drops
the effect columns. To nowcast each stratum we need to do it separately via a
`format = "triangle_list"`. This creates one reporting-triangle per stratum.

```{r bln-strata, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
# One reporting triangle per stratum, straight from the object.
triangles_by_stratum <- tbl_now_to_baselinenowcast(
  dengue_seasonal,
  format  = "triangle_list",
  verbose = FALSE
)
```

```{r bln-strata-print, eval=has("baselinenowcast"), warning=FALSE}
triangles_by_stratum
```

This can be used to nowcast each stratum via `lapply`:

```{r bln-strata-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One nowcast per triangle, each trimmed the same way.
nowcasts_by_stratum <- triangles_by_stratum |>
  lapply(\(tri) baselinenowcast(
    trim_forecast_rows(tri),
    output_type = "samples",
    draws       = 1000
  ))
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r bln-tidy-1, eval=FALSE}
tidy(nowcast_samples)
```

```{r bln-tidy, echo=FALSE}
show_tidy("baselinenowcast")
```

```{r bln-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the baselinenowcast package", fig.alt="Three panels (Total, Female, Male) comparing the baselinenowcast nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("baselinenowcast")
```

## epinowcast

[`epinowcast`](https://package.epinowcast.org/) fits a flexible Bayesian model
with separate modules for the reporting delay and the reference (epidemic)
process. It expects a preprocessed object built by `enw_preprocess_data()`.

`tbl_now_to_epinowcast()` handles the preprocessing, returning
an object you can pass straight to `epinowcast::epinowcast()`:

**Series length: trimmed, before preprocessing.** `epinowcast` is the one engine
here with no argument for shortening the series, because by the time you hold a
preprocessed object the reporting triangle is already built. Its filters work one
step earlier, on the *observations* — and `tbl_now_to_epinowcast()` will hand you
those instead of the finished object if you ask with `preprocess = FALSE`:

::: {.alert .alert-warning}
**Warning: requires Stan.** This package fits its model with **Stan**, so it needs a working `cmdstanr` (or `rstan`) toolchain installed before any of the code below will run.
:::

### Simple nowcast

```{r enw-convert, eval=has("epinowcast"), warning=FALSE, message=FALSE}
library(epinowcast)

# Trim FIRST, then convert: two years of history is plenty and keeps the Stan
# fit tractable.
dengue_enw_recent <- dengue_now |>
  filter(onset_week >= cutoff - 7 * 104) |>
  tbl_now_to_epinowcast(max_delay = 10, verbose = FALSE, quiet = TRUE)

dengue_enw_recent
```

Two years preprocesses in well under a second, and it is `dengue_enw_recent` —
not the full series — that goes to the Stan fit.

This can then be passed to `epinowcast()`: 

```{r enw-fit, eval=run_models && has("epinowcast"), message = FALSE, warning = FALSE}
# A minimal epinowcast fit from the preprocessed object
fit <- epinowcast(
  dengue_enw_recent,   # the trimmed object, not the full series
  fit = enw_fit_opts(
    pp = TRUE, chains = 2, iter_sampling = 500, iter_warmup = 500
  )
)
```

`max_delay` above is a **modelling choice, not a detail**: left unset, the
converter infers it from the longest delay present (15 weeks here), and the
nowcast then carries one reference date per delay. Setting it to 10 matches the
maximum delay every other engine on this page is given, so the comparison at the
end is like for like.

### With strata and effects.

Handing `tbl_now_to_epinowcast()` the enriched object
does two things automatically: the `gender` stratum becomes epinowcast's grouping
(`by`), so the model fits a delay per sex, and the temporal-effect columns land in
the `metareference` / `metareport` tables so that they can be used in a module formula.

```{r effects-enw-fit, eval=run_models && has("epinowcast")}
# Same trim, from the enriched object.
enw_seasonal <- dengue_seasonal |>
  filter(onset_week >= cutoff - 7 * 104) |>
  tbl_now_to_epinowcast(max_delay = 10, verbose = FALSE, quiet = TRUE)

# Drop the seasonal terms into a reference-module formula and fit as usual: the
# covariates now enter the reference model.
epinowcast(
  enw_seasonal,
  reference = enw_reference(
    parametric   = ~ 1 + .event_season_52_sin + .event_season_52_cos,
    distribution = "lognormal",
    data         = enw_seasonal
  ),
  fit = enw_fit_opts(
    pp = TRUE, chains = 2, iter_sampling = 500, iter_warmup = 500
  )
)
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r enw-tidy-1, eval=FALSE}
tidy(fit)
```

```{r enw-tidy, echo=FALSE}
show_tidy("epinowcast")
```

```{r enw-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the epinowcast package", fig.alt="Three panels (Total, Female, Male) comparing the epinowcast nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("epinowcast")
```

## epidist

Sometimes the quantity you actually want is the **delay distribution** itself —
how long, on average, between onset and report, and how variable is it?
[`epidist`](https://epidist.epinowcast.org/) estimates exactly that, treating
each case as an interval-censored onset/report pair.

`tbl_now_to_epidist()` converts the `tbl_now` into the censored line-list
`epidist` expects. Because our data are weekly, each event is censored over its
whole week automatically:

::: {.alert .alert-warning}
**Warning: requires Stan.** This package fits its model with **Stan**, so it needs a working `cmdstanr` (or `rstan`) toolchain installed before any of the code below will run.
:::

### Simple delay fit

```{r epidist-convert, eval=has("epidist"), message=FALSE, warning=FALSE}
library(epidist)

dengue_epidist <- tbl_now_to_epidist(dengue_now, verbose = FALSE)

```

`epidist` offers several model types. The **latent** model is used here; it
carries one latent variable per observation, so it is given the last two years
rather than the whole twelve-year series — a delay distribution does not need
more than that to be pinned down.

```{r epidist-fit, eval=FALSE}
# Fit the delay distribution (see the epidist documentation for model choices)
delay_model <- dengue_now |>
  filter(onset_week >= cutoff - 7 * 104) |>
  tbl_now_to_epidist(verbose = FALSE) |>
  as_epidist_latent_model() |>
  epidist()
```

::: {.alert .alert-warning}
**The `marginal` model does not currently fit.** With **epidist 0.4.0** and
**primarycensored 1.5.1**, `as_epidist_marginal_model()` fails at Stan
compilation: the generated code calls `primarycensored_lpmf()` with 8 arguments
against a 9-argument signature (primarycensored gained an `L` lower-truncation
argument that the generated call does not supply). The
[latent model](https://epidist.epinowcast.org/) is unaffected and estimates the
same quantity, so it is what this article uses. Reported upstream; check the
[epidist issue tracker](https://github.com/epinowcast/epidist/issues/) for the
current status before reaching for `as_epidist_marginal_model()`.
:::

The fitted delay distribution can then feed back into a nowcast for example as
a data-informed prior in `epinowcast`.

### With strata and effects.

`epidist` has no separate grouping argument, so the strata (`gender`) and the
temporal-effect columns become extra columns that can be used within the  model's 
formula. For example:

```{r epidist-strata, eval=has("epidist"), message=FALSE, warning=FALSE}
dengue_epidist_eff <- tbl_now_to_epidist(dengue_seasonal, verbose = FALSE)
```

```{r epidist-strata-fit, eval=FALSE}
# A sex-varying mean delay.
delay_by_sex <- dengue_seasonal |>
  filter(onset_week >= cutoff - 7 * 104) |>
  tbl_now_to_epidist(verbose = FALSE) |>
  as_epidist_latent_model() |>
  epidist(formula = mu ~ 1 + gender)
```

`tidy()` works here too, but it returns a **different table**, because
`epidist` estimates a different thing. There are no per-date case estimates to
report, so instead of one row per event date you get one row per parameter of
the fitted delay distribution:

```{r epidist-tidy, eval=FALSE}
tidy(delay_model)
```

| column | meaning |
|---|---|
| `term` | a parameter of the fitted delay distribution |
| `estimate` | posterior median |
| `conf.low`, `conf.high` | interval bounds, using \pkg{broom}'s names |
| `level` | the width of that interval |
| `engine` | always `"epidist"` |

The `term` values are the distribution's own parameters (`mu`, `sigma`, or
whichever the `family` has) **plus** the derived `mean` and `sd` — the delay
summaries those parameters imply, which are usually the numbers you actually
want to quote. The fit keeps its draws, so `probs` works as it does elsewhere:
`tidy(delay_model, probs = c(0.05, 0.95))` adds `q5` and `q95` columns.

::: {.alert .alert-warning}
**Watch the dispatch.** `epidist()` returns an object whose class is
`c("brmsfit", "epidist_fit")`, in that order. If \pkg{broom.mixed} is loaded,
its `tidy.brmsfit()` method matches **first** and you get raw \pkg{brms}
parameters instead of the delay table above. Call `tidy.epidist_fit(fit)`
explicitly when you cannot be sure which method wins.
:::

The natural use of the result here is as a **data-informed delay prior** for one
of the other engines.

### Does the fitted delay match the observed one?

Every other section ends with the nowcast plotted against the truth. There is no
nowcast here, but there is an equivalent check: the delay distribution `epidist`
fitted, against the delays actually seen in the data.

The comparison needs care, and getting it wrong is easy. `epidist` estimates a
**continuous delay in days**. What the data record is a **week-index
difference**: `week(report) - week(onset)`. Those are not the same quantity —
`denguedat` is weekly, so a case whose onset fell late in its week and was
reported early the next week has a true delay of a day or two but a week
difference of 1.

So the fitted distribution has to be pushed through that same mapping before
being compared. With the onset time uniform inside its week, a case with true
delay $D$ lands in week $\lfloor (p + D)/7 \rfloor$ for $p \sim U(0, 7)$:

```{r epidist-delay-code, eval=FALSE}
draws <- predict_delay_parameters(delay_model)

# P(week difference = w), integrating over where in its week the onset fell.
week_probs <- function(mu, sigma, weeks, n_grid = 101) {
  onset_grid <- seq(0, 7, length.out = n_grid)
  vapply(weeks, function(w) {
    mean(vapply(onset_grid, function(p) {
      hi <- ifelse(7 * (w + 1) - p > 0, plnorm(7 * (w + 1) - p, mu, sigma), 0)
      lo <- ifelse(7 * w - p > 0, plnorm(7 * w - p, mu, sigma), 0)
      mean(hi - lo)
    }, numeric(1)))
  }, numeric(1))
}

fitted_bins <- week_probs(median(draws$mu), median(draws$sigma), 0:8)

# Observed: the share of cases at each week difference.
observed_bins <- dengue_now |>
  mutate(delay = as.numeric(report_week - onset_week) / 7) |>
  filter(delay >= 0, delay <= 8) |>
  count(delay) |>
  mutate(observed = n / sum(n))
```

Skipping that step is not a rounding error. Comparing the fitted lognormal
directly against the week counts credits week 0 with *every* delay under seven
days, when only a case whose onset fell early in the week can actually land
there — which inflates that bar roughly sixfold and makes a perfectly good fit
look broken.

```{r epidist-delay, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="epidist's fitted reporting-delay distribution against the delays observed in the data, pooled and by sex", fig.alt="Three panels (Total, Female, Male). Pale red bars show the observed share of cases at each delay in weeks; a darker red line with a shaded credible interval shows the delay distribution epidist fitted."}
ed <- comparison$display[["epidist"]]
if (!is.null(ed)) {
  strata_order <- c("Total", setdiff(unique(ed$fitted$stratum), "Total"))
  plot_data <- ed$fitted |>
    dplyr::left_join(ed$observed, by = c("stratum", "delay")) |>
    dplyr::mutate(stratum = factor(stratum, levels = strata_order))

  # Both series describe the REPORTING process, which the package palette draws
  # in red: `light_red` for the observed bars, `accent_red` for the fitted curve.
  # Distinguishing them by shape rather than by hue keeps that grammar intact.
  light_red <- "#e78b7f"
  accent_red <- "#B85348"

  ggplot(plot_data, aes(x = delay)) +
    geom_col(
      aes(y = observed, fill = "Observed"),
      width = 0.7, alpha = 0.85
    ) +
    geom_ribbon(
      aes(ymin = lower, ymax = upper, fill = "epidist (95% CrI)"),
      alpha = 0.25
    ) +
    geom_line(
      aes(y = fitted, colour = "epidist (fitted)"),
      linewidth = 0.9
    ) +
    geom_point(
      aes(y = fitted, colour = "epidist (fitted)"),
      size = 1.6
    ) +
    facet_wrap(~stratum, nrow = 1) +
    scale_fill_manual(
      NULL,
      values = c("Observed" = light_red, "epidist (95% CrI)" = accent_red)
    ) +
    scale_colour_manual(
      NULL, values = c("epidist (fitted)" = accent_red)
    ) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(
      x = "Week difference (report week - onset week)",
      y = "Share of cases",
      title = "Fitted delay distribution vs. observed delays"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "top", panel.grid.minor = element_blank())
}
```

The bars are what happened; the line is the same quantity implied by what
`epidist` fitted. Three things are worth reading off it:

* **The mode and the bulk should line up.** They do here: the fit puts the mass
  in the same place the data do. If the curve missed the observed mode, the
  family (`lognormal` by default) would be the first thing to reconsider.
* **Do not expect the bars to be matched exactly.** A parametric delay
  distribution is a smooth two-parameter summary being compared against a coarse
  weekly grid; small shoulder-by-shoulder differences are the price of that
  smoothing, not evidence of a bad fit.
* **The tail needs the numbers, not the picture.** Beyond five weeks both series
  are flat against the axis, but they are not equal: at seven weeks the fitted
  mass is about **0.05%** against an observed **0.31%**, and the fit runs two to
  eight times light across weeks 6–8. That is the more striking because the
  observed shares come from reports that had arrived by the `now`, so long delays
  are if anything **under**-represented in them (right-truncation, the bias
  nowcasting exists to correct). A tail that still outruns the fit despite that
  hints the `family` deserves attention — `epidist` also offers `Gamma()` and
  `weibull()` — if long delays are what you care about. For the bulk of the
  distribution, which is what a nowcast leans on, the lognormal is fine here.

The credible band is drawn but barely visible: 3,239 cases pin two parameters
down tightly, so the posterior for this curve is narrow. That is a statement
about parameter uncertainty only — it says nothing about how well the *shape*
matches, which is what the bars are for.

## NobBS

[`NobBS`](https://cran.r-project.org/package=NobBS) works from a **line-list**
with an onset-date column and a report-date column — precisely what a `linelist`
`tbl_now` already is. Drop the `tbl.now` classes with `as.data.frame()` and hand
the columns to `NobBS()`:

::: {.alert .alert-warning}
**Warning: requires JAGS.** This package fits its model with **JAGS**, which is a separate program: install JAGS itself, not just the R package, before running the code below.
:::

### Simple nowcast

```{r nobbs-convert, eval=has("NobBS")}
dengue_linelist <- as.data.frame(dengue_now)
```


```{r nobbs-fit, eval=run_models && has("NobBS")}
library(NobBS)

nobbs_fit <- NobBS(
  data          = dengue_linelist,
  now           = get_now(dengue_now),
  units         = "1 week",
  onset_date    = get_event_date(dengue_now),
  report_date   = get_report_date(dengue_now),
  max_D         = 10,   # delays beyond 10 weeks are negligible here
  moving_window = 104   # fit to the last two years; otherwise far too slow
)

```

The predictions come out with `tidy()`, in the same shape as every
other engine here:

```{r nobbs-tidy-1, eval=FALSE}
tidy(nobbs_fit)
```

```{r nobbs-tidy, echo=FALSE}
show_tidy("NobBS")
```


Notice that even the arguments come from the `tbl_now`: `get_now()`,
`get_event_date()` and `get_report_date()` tell `NobBS` what `tbl.now` already
figured out.

### With strata and effects.

Call `NobBS.strat()` and pass the column name to the strata option:

```{r nobbs-strata, eval=run_models && has("NobBS"), warning=FALSE, message=FALSE}
dengue_nobbs     <- as.data.frame(dengue_seasonal)
stratified_nobbs <- NobBS.strat(dengue_nobbs,
                                strata        = "gender",
                                now           = get_now(dengue_seasonal),
                                units         = "1 week",
                                onset_date    = get_event_date(dengue_seasonal),
                                report_date   = get_report_date(dengue_seasonal),
                                max_D         = 10,
                                moving_window = 104
                                )
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r nobbs-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the NobBS package", fig.alt="Three panels (Total, Female, Male) comparing the NobBS nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("NobBS")
```

## surveillance

[`surveillance`](https://cran.r-project.org/package=surveillance) is the
long-standing R package for outbreak detection and nowcasting, and implements
the Höhle & an der Heiden (2014) approach. `surveillance::nowcast()` works from
an individual-level **line list** with one column for the event date and one for
the report date, named by its `dEventCol` / `dReportCol` arguments.

`tbl_now_to_surveillance()` builds that data frame and renames the two dates to
`surveillance`'s own defaults:

### Simple nowcast

```{r sur-convert, eval=has("surveillance"), message=FALSE}
library(surveillance)

dengue_sur <- tbl_now_to_surveillance(dengue_now, verbose = FALSE)

head(dengue_sur)
```

The nowcast itself needs a `now`, the weeks you want estimated (`when`), and a
maximum delay `D`. All of it can come from the `tbl_now`:

```{r sur-fit, eval=run_models && has("surveillance"), message=FALSE, warning=FALSE}
sur_now  <- get_now(dengue_now)
sur_when <- seq(sur_now - 7 * 20, sur_now, by = "1 week")   # weeks to estimate

# `dRange` gives surveillance the grid explicitly, so it reaches the `now` even
# though the last week has no cases at all: a line list cannot express a zero
# week (no rows), so zero-completion is no help here.
sur_range <- seq(min(dengue_sur$dHospital), sur_now, by = "1 week")

sur_fit <- nowcast(
  now          = sur_now,
  when         = sur_when,
  data         = dengue_sur,
  dEventCol    = "dHospital",
  dReportCol   = "dReport",
  aggregate.by = "1 week",
  D            = 10,
  method       = "bayes.notrunc.bnb",
  control      = list(dRange = sur_range, N.tInf.max = 1000, nSamples = 1000)
)

```

The predictions come out with `tidy()`, in the same shape as every
other engine here:

```{r sur-tidy-1, eval=FALSE}
tidy(sur_fit)
```

```{r sur-tidy, echo=FALSE}
show_tidy("surveillance")
```

::: {.alert .alert-info}
**TIP**: the interval is already there. `surveillance::nowcast()` stores a
prediction interval in the returned object's `pi` slot, at whatever width
`control$alpha` names (`0.05`, so 95%, by default), and `tidy()` reads it into
`conf.low` / `conf.high`. The JAGS-backed `bayes.trunc` and `bayes.trunc.ddcp`
methods are worth reaching for when you want the delay distribution modelled
differently, **not** merely to get uncertainty out of the fit.
:::


### With strata and effects.

`surveillance::nowcast()` has **no strata argument** as it only models one series. The
strata ride along as ordinary columns, so fit one nowcast per stratum in a loop:

```{r sur-strata, eval=has("surveillance"), message=FALSE, warning=FALSE}
dengue_sur_eff <- tbl_now_to_surveillance(dengue_seasonal, verbose = FALSE)

# `gender` is kept as a column, so it can be split on.
head(dengue_sur_eff)
```

```{r sur-strata-fit, eval=run_models && has("surveillance"), message=FALSE}
sur_by_stratum <- dengue_sur_eff |>
  split(dengue_sur_eff$gender) |>
  lapply(\(df) nowcast(
    now          = sur_now,
    when         = sur_when,
    data         = df,
    dEventCol    = "dHospital",
    dReportCol   = "dReport",
    aggregate.by = "1 week",
    D            = 10,
    method       = "bayes.notrunc.bnb",
    control      = list(
      dRange     = seq(min(df$dHospital), sur_now, by = "1 week"),
      N.tInf.max = 1000,
      nSamples   = 1000
    )
  ))
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r sur-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the surveillance package", fig.alt="Three panels (Total, Female, Male) comparing the surveillance nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("surveillance")
```

## nowcaster

[`nowcaster`](https://covid19br.github.io/nowcaster/) fits a Bayesian
nowcast with [INLA](https://www.r-inla.org/), and like `surveillance` it starts
from a linelist.

`tbl_now_to_nowcaster()` renames the dates into its vocabulary:

::: {.alert .alert-warning}
**Warning: requires R-INLA.** This package fits its model with **R-INLA**, which is not on CRAN. Install it from <https://www.r-inla.org/download-install> before running the code below.
:::

### Simple nowcast

```{r nowcaster-convert, eval=has("nowcaster"), message=FALSE}
library(nowcaster)

dengue_nowcaster <- tbl_now_to_nowcaster(dengue_now, verbose = FALSE)

head(dengue_nowcaster)
```

```{r nowcaster-fit, eval=run_models && has("nowcaster") && has("INLA"), message=FALSE}
nowcaster_fit <- nowcasting_inla(
  dataset      = dengue_nowcaster,
  date_onset   = date_onset,
  date_report  = date_report,
  data.by.week = TRUE,
  use.epiweek  = FALSE,  # weeks END on the last record's weekday
  Dmax         = 20,     # how many PAST WEEKS to nowcast, not a max delay
  wdw          = 104,    # fitting window
  silent       = TRUE
)

```

The predictions come out with `tidy()`, in the same shape as every
other engine here:

```{r nowcaster-tidy-1, eval=FALSE}
tidy(nowcaster_fit)
```

```{r nowcaster-tidy, echo=FALSE}
show_tidy("nowcaster")
```


Two of those arguments are easy to misread, so they are worth spelling out:

* **`Dmax` is the nowcast horizon**, not a maximum delay: "till how many past
  weeks the nowcasting will estimate". Setting it to 10 gives you ten estimated
  weeks, not a ten-week delay cap.
* **`use.epiweek` decides where the weeks start.** With `FALSE` (the default) a
  week *ends* on the weekday of the last record; with `TRUE` it uses the CDC
  epiweek, which starts on Sunday. Since `denguedat`'s weeks are Mondays and our
  last report is a Monday, `FALSE` puts each of our weeks in its own bin —
  `TRUE` would shift them onto Sunday boundaries that do not line up.

::: {.alert .alert-warning}
**Make sure your last event week has a same-week report.** This is the one
`nowcaster` gotcha worth internalising, because it is silent and it looks like a
broken model.

`nowcasting_inla()` takes its last observable week from the last **onset** in the
data, not from the last report:

```r
Tmax <- max(pull(data_w, date_onset))
Y <- ifelse(Time + delay > Tmax.id, NA, Y)
```

Every cell whose report falls after that onset week is marked *unobservable* —
even when the report is sitting right there in your data. So if your final event
week happens to have no same-week report, `max(onset) < max(report)` and
`nowcaster` throws away a whole diagonal of real reports. The symptom is a
nowcast that sits **below** the counts you have already observed, which is
otherwise impossible.

Our cutoff is chosen so the last event week (`2002-07-15`) does contain a
same-week report. With it, `nowcaster` sees exactly what every other engine sees
— all 105 weeks match the observed counts, and it never estimates below them.
Shift the cutoff one week earlier and it loses a diagonal and six weeks come out
low.

To check your own data: `max(event_date) == max(report_date)` after aggregating
to your time unit. If they differ, either move the cutoff or accept that
`nowcaster` is working from one period less than the rest.
:::

### With strata and effects.

`nowcaster` *can* stratify, in a single fit — but only through `age_col`, and
only on a **numeric** column it can `cut()`. Its help calls that a "stratum
column", yet handing it a character one errors inside `cut()`, and a character
`bins_age` trips over an `if (bins_age == "SI-PNI")` comparison on a vector.

`tbl_now_to_nowcaster()` does the encoding for you: it pools the strata into one
label, codes it `1..n` as `stratum_code`, and attaches the breaks that put each
level in its own bin.

```{r nowcaster-strata, eval=has("nowcaster"), message=FALSE}
dengue_nowcaster_eff <- tbl_now_to_nowcaster(dengue_seasonal, verbose = FALSE)

# `as_tibble()` only for printing: the temporal effects ride along as extra
# columns, and a bare data.frame would wrap them onto a second block.
head(dengue_nowcaster_eff) |> as_tibble()

# The breaks `bins_age` needs come from the line list itself.
get_nowcaster_strata(dengue_nowcaster_eff)
```

One fit then covers every stratum: `$total` holds the pooled nowcast and `$age`
one row per stratum per week.

```{r nowcaster-strata-fit, eval=run_models && has("nowcaster") && has("INLA"), message=FALSE}
nowcaster_stratified <- nowcasting_inla(
  dataset      = dengue_nowcaster_eff,
  date_onset   = date_onset,
  date_report  = date_report,
  age_col      = stratum_code,
  bins_age     = get_nowcaster_strata(dengue_nowcaster_eff),
  data.by.week = TRUE,
  use.epiweek  = FALSE,
  Dmax         = 20,
  wdw          = 104,
  silent       = TRUE
)

tail(nowcaster_stratified$age)
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r nowcaster-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the nowcaster package", fig.alt="Three panels (Total, Female, Male) comparing the nowcaster nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("nowcaster")
```

<!-- I am ignoring this part for now until I have decided how to communicate it
content is AI generated please review by a human  before uncommenting. 

## Does this work on *your* data?

Everything above uses one weekly line list. `tbl.now` ships six other datasets
with different shapes — daily and weekly, line-list and count-incidence and
count-cumulative — so the table below records what actually happens when every
converter, plus one representative nowcast, is run against each of them. It is
generated by `data-raw/converter_matrix.R`, and it reports failures as failures.

```{r matrix-load, echo=FALSE, message=FALSE}
matrix_results <- readRDS("converter-matrix.rds")
```

```{r matrix-table, echo=FALSE}
matrix_results |>
  mutate(mark = dplyr::case_when(
    status == "ok"      ~ "ok",
    status == "timeout" ~ "slow",
    TRUE                ~ "fails"
  )) |>
  select(dataset, data_type, step, mark) |>
  tidyr::pivot_wider(names_from = step, values_from = mark, values_fill = "-") |>
  knitr::kable(caption = "Every converter against every shipped dataset")
```

```{r matrix-problems, echo=FALSE}
problems <- matrix_results |> filter(status != "ok")
if (nrow(problems) > 0) {
  problems |>
    select(dataset, data_type, step, status, message) |>
    knitr::kable(caption = "What went wrong, and why")
}
```

Two of these datasets needed fixing before they would convert at all, and both
problems are ones you will meet in your own data:

* **`mpoxdat` has a `race` column that must be declared.** Left as an undeclared
  extra column, the `(event_date, report_date)` pair is not unique — there is one
  row per race — and every target that needs a unique cell rejects it: a
  reporting triangle reported *832 duplicate pairs*, and `tsibble` refused for
  the same reason. `strata = race` fixes both. If a column genuinely is not a
  stratum, aggregate it away with `to_count()` instead.
* **`covidat` has rows registered *before* symptom onset.** A negative delay is
  not a delay, and `epidist` rejects it outright rather than guess. Filter them
  before converting.

A third needed fixing too, and it is the one most likely to catch you out:

* **FluSight's reporting weekday is not constant.** `target_end_date` is always a
  Saturday, but `as_of` lands on Saturdays *and* Wednesdays — so the delay
  between them is `0.571` weeks, not a whole number, and no reporting triangle
  can be built at all. That is an **alignment** problem masquerading as a
  cumulative-data problem. `align_weeks = TRUE` in `tbl_now()` snaps the reports
  onto a consistent weekday and the delays become integers.

**`count-cumulative` data does convert**, once its delays are whole periods.
De-accumulating a cumulative series produces *negative* increments wherever a
total was revised downward, and `baselinenowcast` ships
`preprocess_negative_values()` for exactly that — it redistributes each negative
back into earlier delays. `tbl_now_to_baselinenowcast()` calls it for you and
says so; pass `negatives = "error"` if you would rather be stopped than have the
correction applied silently.

What is left is a modelling limit rather than a conversion one:

* **`epidist` needs a delay it can censor, and counts of at least one.** It
  models an interval-censored onset-to-report pair; a cumulative series has no
  individual delays to censor, and its zero cells fail epidist's `n >= 1` check.
* **Not every model converges on every dataset.** `diseasenowcasting` reports
  *"joint fit failed to converge for all init attempts"* on the cumulative
  FluSight series. The converter did its job; the model could not fit the shape
  it was given.

The honest summary: **every converter works on `linelist`, `count-incidence` and
`count-cumulative` data** — provided the delays are whole periods and any
undeclared strata are declared. Those two data problems, not the converters, are
what you will spend your time on.
--->

## The `tidy()` function

The converters normalize what goes *into* each package. `tidy()` normalizes what
comes out. Every engine above returns something different: a matrix of
draws, an `stsNC` object, a Stan fit, an INLA summary, a bare list. The `tidy()`
turns any of them into the same table:

```{r tidy-demo, message=FALSE, warning=FALSE}
# A short window so this runs quickly; any fit from any section works the same.
recent_triangle <- dengue_now |>
  filter(onset_week >= cutoff - 365 * 2) |>
  to_count(to = "count-incidence") |>
  complete_zeroes() |>
  tbl_now_to_baselinenowcast(verbose = FALSE)

bln_fit <- baselinenowcast(recent_triangle, output_type = "samples", draws = 500)

tidy(bln_fit) |> tail(4)
```

The columns are the same whatever produced the fit:

| column | meaning |
|---|---|
| `event_date` | event/reference date, **on the engine's own grid** |
| `stratum` | `"all"` when the fit is unstratified |
| `estimate` | point nowcast (posterior median where available) |
| `conf.low`, `conf.high` | interval bounds, using \pkg{broom}'s names |
| `level` | the width that interval **actually** has |
| `engine` | which package produced it |
| `qXX` | (optional) quantile columns of the prediction |

Pass `probs` for columns with other quantiles:

```{r tidy-probs, message=FALSE, warning=FALSE}
tidy(bln_fit, probs = c(0.05, 0.5, 0.95)) |> tail(3)
```

::: {.alert .alert-warning}
**Only engines that keep draws can answer an arbitrary `probs`.** That is
`diseasenowcasting`, `baselinenowcast` and `epinowcast`. `NobBS`, `nowcaster` and
`surveillance` report a fixed set of summaries, so asking them for a quantile
they never computed is an **error** rather than a silent approximation.
:::

### Asking the other engines for the quantiles you want

That error is about *when* the quantiles are computed, not about whether you can
have them. Several of these packages will happily compute any quantile — you
just have to **ask at fit time**, because once the fit returns, the draws are
gone and there is nothing left to re-summarise.

`NobBS` is the clearest example. Its `specs` argument takes a `quantiles`
element, and whatever you put there is computed alongside the default summaries:

```{r nobbs-quantiles-1, eval=FALSE}
nobbs_quantiles <- NobBS(
  data          = dengue_linelist,
  now           = get_now(dengue_now),
  units         = "1 week",
  onset_date    = get_event_date(dengue_now),
  report_date   = get_report_date(dengue_now),
  max_D         = 10,
  moving_window = 104,
  specs         = list(quantiles = c(0.1, 0.5, 0.9))   # <- ask here
)

tail(nobbs_quantiles$estimates, 5)
```

```{r nobbs-quantiles, echo=FALSE}
comparison$display[["NobBS_quantiles"]]$estimates
```

The requested quantiles arrive as `q_0.1`, `q_0.5` and `q_0.9` columns on
`$estimates`, next to the usual `estimate` / `lower` / `upper`. `tidy()` still
returns the standard six columns for such a fit — it does not invent a schema
per engine — so join the extra columns on by date when you need them:

```{r nobbs-quantiles-join-1, eval=FALSE}
tidy(nobbs_quantiles) |>
  left_join(
    nobbs_quantiles$estimates |>
      select(event_date = onset_date, q_0.1, q_0.5, q_0.9),
    by = "event_date"
  ) |>
  tail(5)
```

```{r nobbs-quantiles-join, echo=FALSE}
qs <- comparison$display[["NobBS_quantiles"]]
if (!is.null(qs$tidy)) {
  qs$tidy |>
    dplyr::left_join(
      qs$estimates |>
        dplyr::select(event_date = onset_date, q_0.1, q_0.5, q_0.9) |>
        dplyr::mutate(event_date = as.Date(event_date)),
      by = "event_date"
    ) |>
    tail(5)
}
```

The same idea applies elsewhere: `surveillance::nowcast()` takes the interval
width through `control$alpha`, and `nowcaster::nowcasting_inla()` through its own
INLA quantile settings. Decide what you need **before** you fit.


## Summary


```{r comparison-all-code, eval=FALSE, echo=FALSE}
window <- meta$horizon
tru <- truth[truth$event_date %in% window, ]
est <- nowcasts[nowcasts$event_date %in% window, ]
tru$stratum <- factor(tru$stratum, levels = strata_levels)
est$stratum <- factor(est$stratum, levels = strata_levels)
est$package <- factor(est$package, levels = names(engine_colours))

ggplot(mapping = aes(x = event_date)) +
  geom_col(data = tru, aes(y = observed, fill = "Reported by now"), width = 5.5) +
  geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.7) +
  # Truth last, so it stays visible where the engines lie on top of it.
  geom_line(
    data = tru, aes(y = truth, linetype = "Truth"),
    linewidth = 0.7, colour = "#262626"
  ) +
  facet_wrap(~stratum, nrow = 1, scales = "free_y") +
  scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
  scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
  scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
  labs(x = "Event week (symptom onset)", y = "Cases",
       title = "Every engine on the same data",
       subtitle = paste0("Now = ", format(meta$now))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.box = "horizontal",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335")) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE, order = 1))
```


```{r comparison-all, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4.4, out.width="100%", fig.alt="Three panels (Total, Female, Male). Grey bars show the counts reported by now, a dark line the counts those weeks eventually reached, and one coloured line per package its nowcast."}
window <- meta$horizon
tru <- truth[truth$event_date %in% window, ]
est <- nowcasts[nowcasts$event_date %in% window, ]
tru$stratum <- factor(tru$stratum, levels = strata_levels)
est$stratum <- factor(est$stratum, levels = strata_levels)
est$package <- factor(est$package, levels = names(engine_colours))

ggplot(mapping = aes(x = event_date)) +
  geom_col(data = tru, aes(y = observed, fill = "Reported by now"), width = 5.5) +
  geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.7) +
  # Truth last, so it stays visible where the engines lie on top of it.
  geom_line(
    data = tru, aes(y = truth, linetype = "Truth"),
    linewidth = 0.7, colour = "#262626"
  ) +
  facet_wrap(~stratum, nrow = 1, scales = "free_y") +
  scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
  scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
  scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
  labs(x = "Event week (symptom onset)", y = "Cases",
       title = "Every engine on the same data",
       subtitle = paste0("Now = ", format(meta$now))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.box = "horizontal",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335")) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE, order = 1))
```


We described the dengue data once as a `tbl_now`, and then a single converter call
(or, for `diseasenowcasting`, no call at all) handed it to each package in the
shape it needed:

```{r recap, eval=FALSE}
dengue_now <- tbl_now(dengue, event_date = onset_week, report_date = report_week,
                      data_type = "linelist")

dengue_now                              # diseasenowcasting
tbl_now_to_baselinenowcast(dengue_now)  # baselinenowcast
tbl_now_to_epinowcast(dengue_now)       # epinowcast
tbl_now_to_epidist(dengue_now)          # epidist
tbl_now_to_surveillance(dengue_now)     # surveillance
tbl_now_to_nowcaster(dengue_now)        # nowcaster
as.data.frame(dengue_now)               # NobBS or others
```

Attaching **strata** and **temporal effects** once (`dengue_seasonal`) uses the
same converters: each package receives them in whatever way it can use
a grouping in `epinowcast`, covariate columns in `epidist` and the
`baselinenowcast` long format, one triangle/series per stratum where the model
takes a single series, and automatically in `diseasenowcasting`. 

```{r learning-more, echo=FALSE, results="asis"}
# The fragment is looked up by ABSOLUTE path via system.file(), never by a
# relative one. pkgdown/rmarkdown render into an intermediates directory under
# tempdir() and copy relative resources alongside; a path containing "../.."
# escapes that directory ("cannot create file '/tmp/RtmpXXX/../../man/...'"),
# which fails wherever tempdir() sits near the filesystem root -- as it does on
# CI. `system.file()` also works under pkgload::load_all(), which shims it to
# find inst/ in the source tree.
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
``` 
