---
title: "Using tbl.now with different modelling packages"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Using tbl.now with different modelling packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The heavy model fits (Stan / JAGS / MCMC) are shown as illustrative code but
# not run when the vignette is built: the results are read back from
# `nowcast-comparison.rds`, which is produced by running THIS FILE's own code
# (`data-raw/nowcast_models_precompute.R` purls the chunks below and runs them).
#
# Set TBL_NOW_RUN_MODELS=true to fit everything live instead. That is what the
# precompute does, which is why the two can no longer drift apart.

options(tibble.width = Inf, width = 160, pillar.width = 160)
options(pillar.print_max = 5, pillar.print_min = 5, 
        tibble.print_min = 5, dplyr.print_min = 5,
        tibble.print_max = 5, dplyr.print_max = 5)

run_models <- isTRUE(as.logical(Sys.getenv("TBL_NOW_RUN_MODELS", "false")))

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)
```

## Why this vignette?

Preparing the same data different ways for the different nowcasting models can be tedious and
error-prone. This is exactly what [`tbl.now`](https://rodrigozepeda.github.io/tbl.now/)
helps with. You can **describe your data once** by specifying which column is the `event_date`, which is the `report_date`, whether your data is linelist or has counts (and what _those_ counts mean!) and `tbl.now` will hand it to each modelling package in the format that package expects. 

In this vignette we take a single dataset (`covid_colombia`), and from that one object, use several different nowcasting / delay-estimation tools:

<div class="small-table">
| Package | What it does | Additional requirements | `tbl.now` converter | `run_nowcast()` method |
|---|---|---|---|---|
| [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) | flexible Bayesian nowcast (delay + epidemic processes) | none (uses `RTMB`) | consumes a `tbl_now` **directly** | `"diseasenowcasting"` |
| [baselinenowcast](https://baselinenowcast.epinowcast.org/) | fast, assumption-light baseline nowcast | none | `tbl_now_to_baselinenowcast()` | `"baselinenowcast"` |
| [epinowcast](https://package.epinowcast.org/) | flexible Bayesian nowcast (delay + reference modules) | **Stan** | `tbl_now_to_epinowcast()` | `"epinowcast"` |
| [epidist](https://epidist.epinowcast.org/) | estimates only the reporting **delay distribution** | **Stan** | `tbl_now_to_epidist()` | — (not a nowcast) |
| [NobBS](https://cran.r-project.org/package=NobBS) | Nowcasting by Bayesian Smoothing | **JAGS** | `tbl_now_to_nobbs()` | `"NobBS"` |
| [surveillance](https://cran.r-project.org/package=surveillance) | the classic Höhle & an der Heiden nowcast | none for the method used here (**JAGS** for `bayes.trunc.ddcp`) | `tbl_now_to_surveillance()` | `"surveillance"` |
| [EpiNow2](https://epiforecasts.io/EpiNow2/) | renewal-equation $R_t$, reporting truncation, and delay distributions | **Stan** (`cmdstanr`) | `tbl_now_to_EpiNow2()` | `"EpiNow2"` |
</div> 

You do not need to be an expert in any of these packages. The point of this vignette is to show you how to use any of them for nowcasting. 

::: {.alert .alert-info}
This article shows you to to use each package individually. The `tbl.now` package contains the `run_nowcast()` function which allows you run nowcast models without requiring knowledge of each of the package's frameworks. We recommend reading the article  [**One call, many models**](ensemble-nowcasting.html) first if your priority is to do nowcasts. Read this one if your prioirity is understanding how to use `tbl.now` for a **specific** package. In that case, read the description on [The data](#the-data) and then go to your favourite package's section. 
:::

::: {.alert .alert-warning}
This article solely aims to show how to use `tbl.now` within each package's own framework. We are purposefully not using or searching for the most optimal models from each 
package. Please **DO NOT CONCLUDE WHICH PACKAGE IS BEST BASED ON  THE RESULTS FROM THIS TUTORIAL**. 
:::



## The data

`tbl.now` includes `covid_colombia`, daily COVID-19 case counts from
Colombia's national surveillance system (INS) from 2020 to 2023. Each row is a
`(notification_date, diagnosis_date, sex)` combination with a case count `n`. It includes
notification date as the date the case was first notified (`event_date`), and the date the laboratory diagnosis was registered (`report_date`). The gap between the two is the reporting delay.  

> **OUR GOAL IS TO NOWCAST THE CASES AS THEY HAPPEN ACCORDING TO `notification_date`**.

```{r setup, message=FALSE}
library(dplyr)
library(lubridate)
library(tbl.now)

data(covid_colombia)
```

```{r, echo = FALSE, purl=FALSE}
head(covid_colombia)
```

We  will build a single `tbl_now`. To do so, we state which column is the event date
(`notification_date`), which is the report date (`diagnosis_date`), and which
 the counts (`n`); `tbl.now` infers the rest: that the grid is daily, and
that the "now" is the last report date.

For this example we will cut **both** dates at the start of April 2021 (Colombia's third wave). The example will assume we are back on that date, with only the information available until then (`now = "2021-04-01"`). No data observed after that date is kept: 

```{r build-tbl-now}
#Filter to simulate being back on April 2021
covid <- covid_colombia |>
  filter(notification_date < as.Date("2021-04-01") & 
           diagnosis_date < as.Date("2021-04-01"))

#Create the tbl_now object
covid_now <- covid |>
  tbl_now(
    event_date  = notification_date,
    report_date = diagnosis_date,
    case_count  = n,
    data_type   = "count-incidence"
  )

#We can see the tbl_now
covid_now
```

::: {.alert .alert-info}
**About the warning.** `sex` is in the data but was not declared as 
a covariate or strata, so each
`(notification_date, diagnosis_date)` cell has two rows (one per sex). 
One can correct it by summing the `(notification_date, diagnosis_date)` combinations by sex with
`to_count()`:

```{r, eval = FALSE}
covid_now |> 
  to_count(to = "count-incidence")
```

Though for the purpose of this tutorial we will **ignore the warning.** That is because
we will also create a second `tbl.now` from this one. The new one will be stratified by sex 
to show how the related packages work both in the stratified and not stratified case. 
:::

Everything in this article is built from either this dataset or a more complete one
with strata and temporal effects. 

### A more complex object: strata and temporal effects

Reporting delays are rarely constant and many series are reported for several
**strata** (here, `sex`) that you may want to nowcast separately. Here we build a
second object, `covid_seasonal`, that declares `sex` as a stratum and adds
day-of-week plus annual-Fourier **temporal (delay) effects**. We show
how one would use them after converting to each package's format: 

```{r build-seasonal, message=FALSE}
covid_seasonal <- covid_now |> 
  add_strata(sex) |> 
  add_temporal_effects(
    temporal_effects(
      day_of_week = TRUE,    # a separate level per weekday
      seasons     = 365      # an annual Fourier cycle (period = 365 days)
    )
  )
```


::: {.alert .alert-info}
**A note on series length.** This dataset carries `r round(sum(covid_now$n)/1000000, 2)` million cases. And not all packages will be able to handle that amount of data. Specifically we will see that:

* `diseasenowcasting` can take the **whole** series.
* `baselinenowcast` can keep every event day, but the **delays** will be capped with
  `max_delay = 30` (the real maximum delay in the data corresponds to `r max(covid_now$.delay)` days).
* The remaining packages (`NobBS`, `surveillance`, `epinowcast` and `epidist`)
  will be  trimmed both in the number of events and delays as the time it takes  
  to run those models is prohibitively expensive. 
:::

```{r comparison-load, echo=FALSE, message=FALSE, purl=FALSE}
library(ggplot2)

comparison <- readRDS("nowcast-comparison.rds")
nowcasts   <- comparison$nowcasts
truth      <- comparison$truth
meta       <- comparison$meta

strata_levels  <- c("Total", "Female", "Male")

# The fits are precomputed (see `data-raw/nowcast_models_precompute.R`, which
# purls and runs THIS file), so the article
# cannot call `print(fit)` or `tidy(fit)` live. It shows the REAL output of those
# calls instead: the precompute captured them and they are replayed here, so the
# code you see above each result is exactly the code that produced it.
show_fit <- function(package) {
  captured <- comparison$display[[package]]$printed
  if (is.null(captured)) return(invisible(NULL))
  cat(captured, sep = "\n")
}

show_tidy <- function(package, n = 5) {
  tidied <- comparison$display[[package]]$tidy
  if (is.null(tidied)) return(invisible(NULL))
  tail(tidied, n)
}

# The interval an engine reported, quoted in prose. COMPUTED from the same
# cached tables the section prints, never typed: a hand-written number next to a
# rebuilt figure is the drift this file's precompute exists to prevent.
#
# `width` is the last day's band; `typical` the median over the whole plotted
# window, which is the one that matters for "can you see it?" -- the band widens
# sharply at the `now` edge, so the last day flatters a narrow engine.
last_interval <- function(package) {
  tidied <- comparison$display[[package]]$tidy
  row    <- tidied[nrow(tidied), ]
  num    <- function(v) format(round(v), big.mark = ",", trim = TRUE)
  pct    <- function(v) sprintf("%.1f%%", 100 * v)

  shown <- nowcasts[
    nowcasts$package == package &
      nowcasts$stratum == "Total" &
      nowcasts$event_date %in% meta$horizon,
  ]

  list(
    date    = format(row$event_date),
    point   = num(row$estimate),
    lower   = num(row$conf.low),
    upper   = num(row$conf.high),
    width   = pct((row$conf.high - row$conf.low) / row$estimate),
    typical = pct(stats::median(
      (shown$upper - shown$lower) / shown$estimate, na.rm = TRUE
    ))
  )
}
# EVERY package that appears in `nowcasts` needs an entry. The Summary figure
# turns `package` into a factor with these levels, so a package missing here
# becomes an `NA` level: a grey unnamed line in the plot and an "NA" key in the
# legend, with no error to say why.
engine_colours <- c(
  diseasenowcasting = "#0072B2",
  baselinenowcast   = "#D55E00",
  epinowcast        = "#009E73",
  NobBS             = "#CC79A7",
  surveillance      = "#E69F00",
  EpiNow2           = "#56B4E9"
)
```


```{r comparison-helper, message=FALSE, echo=FALSE, purl=FALSE}
# One figure per package: Total, Female and Male side by side.
nowcast_panels <- function(package, subtitle = NULL) {
  window <- meta$horizon

  tru <- truth[truth$event_date %in% window, ]
  est <- nowcasts[
    nowcasts$package == package & nowcasts$event_date %in% window,
  ]
  tru$stratum <- factor(tru$stratum, levels = strata_levels)
  est$stratum <- factor(est$stratum, levels = strata_levels)

  colour <- unname(engine_colours[package])

  ggplot(mapping = aes(x = event_date)) +
    # What had been reported by `now` -- the incomplete data the engine saw.
    # width = 0.8 (one day). A wider bar on a DAILY series overlaps its
    # neighbours, and ggplot2 then stacks the overlaps -- so the grey bars would
    # show sums of several days rather than each day's count.
    geom_col(
      data = tru, aes(y = observed, fill = "Reported by now"), width = 0.8
    ) +
    # The interval, where the engine gives one.
    geom_ribbon(
      data = est[!is.na(est$lower), ],
      aes(ymin = lower, ymax = upper), fill = colour, alpha = 0.25
    ) +
    # The nowcast...
    geom_line(data = est, aes(y = estimate, colour = "Nowcast"), linewidth = 0.9) +
    geom_point(data = est, aes(y = estimate, colour = "Nowcast"), size = 1.1) +
    # ...then the truth ON TOP, so it stays visible where the two coincide
    # (over the settled days they lie exactly on each other).
    geom_line(
      data = tru, aes(y = truth, linetype = "Truth"),
      linewidth = 0.6, colour = "#262626"
    ) +
    facet_wrap(~stratum, nrow = 1, scales = "free_y") +
    scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
    scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
    scale_colour_manual(name = NULL, values = c("Nowcast" = colour)) +
    labs(
      x = "Event date (notification)", y = "Cases",
      title = paste0(package, " nowcast vs. what actually happened"),
      subtitle = subtitle %||% paste0(
        "Now = ", format(meta$now),
        "; last ", length(window), " days"
      )
    ) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position  = "top",
      legend.box       = "horizontal",
      legend.margin    = margin(b = -4),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold", colour = "#334335")
    )
}
```


## diseasenowcasting

::: {.alert .alert-info}
Install with: 
```r
install.packages("diseasenowcasting", repos = c("https://rodrigozepeda.r-universe.dev", getOption("repos")))
```
:::

[`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) is
designed hand-in-hand with `tbl.now`, so it takes a `tbl_now` **directly**. You
just hand it the object:

### Simple nowcast

```{r dnc, eval=run_models && has("diseasenowcasting")}
library(diseasenowcasting)
dnc_fit <- nowcast(covid_now)

dnc_fit
```

```{r dnc-out, echo=FALSE, purl=FALSE, eval=!run_models}
show_fit("diseasenowcasting")
```

Predictions can be obtained via `tidy`:

```{r dnc-tidy-1, eval=FALSE}
tidy(dnc_fit)
```

```{r dnc-tidy, echo=FALSE, purl=FALSE}
show_tidy("diseasenowcasting")
```

### With strata and effects.

`diseasenowcasting` reaches into the `tbl_now` itself, so the enriched object needs no extra arguments: it picks up the `sex` stratum and the day-of-week / seasonal effect columns automatically.

```{r dnc-teff, eval=run_models && has("diseasenowcasting")}
dnc_seasonal <- nowcast(covid_seasonal)   # strata and effects used automatically
```

```{r dnc-teff-out, echo=FALSE, purl=FALSE, eval=!run_models}
show_fit("diseasenowcasting_seasonal")
```

`tidy()` can also be used in stratified cases: 

```{r dnc-teff-tidy-1, eval=FALSE}
tidy(dnc_seasonal)
```

```{r dnc-teff-tidy, echo=FALSE, purl=FALSE}
show_tidy("diseasenowcasting_seasonal")
```


```{r dnc-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the diseasenowcasting package", fig.alt="Three panels (Total, Female, Male) comparing the diseasenowcasting nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("diseasenowcasting")
```

## baselinenowcast

::: {.alert .alert-info}
Install with: 

```r
install.packages("baselinenowcast")
```

:::

[`baselinenowcast`](https://baselinenowcast.epinowcast.org/) is a 
simple, fast baseline. It works from a **reporting triangle**, a matrix with one
row per event (reference) date and one column per reporting delay. The
lower-right corner of the matrix corresponds to the not-yet-observed part the nowcast will fill in.

### Simple nowcast

For unstratified data, `tbl_now_to_baselinenowcast()` builds that triangle directly from the
count-incidence data:

```{r bln-convert, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
library(baselinenowcast)

# Cap the delay axis: a single 185-day straggler gives the
# triangle 186 columns, almost all of them empty.
covid_triangle <- covid_now |>
  tbl_now_to_baselinenowcast(max_delay = 30, verbose = FALSE)

# rows = notification dates, columns = delay in days
covid_triangle[1:5, 1:6]
```

::: {.alert .alert-info}
**Why cap the delays?**  The reporting
triangle keeps every one of the `r nrow(covid_triangle)` event dates. Because the maximum delay in our data is `r max(covid_now$.delay)`  days it would mean a matrix of `r nrow(covid_triangle)` $\times$  `r max(covid_now$.delay)` $=$ `r scales::comma(nrow(covid_triangle)*max(covid_now$.delay))` entries. This makes the fit extremely slow.
:::

From here you can follow `baselinenowcast`'s own workflow. For example calling
`baselinenowcast()` to estimate the delay from the triangle, apply it, and
draw nowcast samples:


```{r bln-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One-call workflow: estimate the delay, apply it, and draw nowcast samples.
nowcast_samples <- baselinenowcast(
  covid_triangle,
  output_type = "samples",
  draws       = 1000
)

```

### With strata and effects.

`baselinenowcast` can also take a long tidy `data.frame` and nowcass every
stratum in one call, via its `strata_cols` argument. Use `format = "auto"`
(the default) or `format = "long"` on `tbl_now_to_baselinenowcast()` to get
that shape from a stratified `tbl_now`:

```{r bln-strata, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
# Long tidy data frame with the strata columns present; `format = "auto"`
# is what `tbl_now_to_baselinenowcast()` returns when the object has strata.
by_stratum <- tbl_now_to_baselinenowcast(
  covid_seasonal,
  max_delay = 30,
  verbose   = FALSE
)
```

```{r bln-strata-print, eval=has("baselinenowcast"), warning=FALSE}
head(by_stratum)
```

Hand it to `baselinenowcast(strata_cols = ...)` for one fit per stratum, or
opt into `strata_sharing = "delay"` / `"uncertainty"` to share the delay PMF
or the uncertainty parameters across strata:

```{r bln-strata-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One `baselinenowcast_df` covering every stratum.
nowcasts_by_stratum <- baselinenowcast(
  by_stratum,
  strata_cols = get_strata(covid_seasonal),
  output_type = "samples", draws = 1000
)
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r bln-tidy-1, eval=FALSE}
tidy(nowcast_samples)
```

```{r bln-tidy, echo=FALSE, purl=FALSE}
show_tidy("baselinenowcast")
```

```{r bln-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the baselinenowcast package", fig.alt="Three panels (Total, Female, Male) comparing the baselinenowcast nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("baselinenowcast")
```

## epinowcast

::: {.alert .alert-info}
Install with: 

```r
install.packages("epinowcast", repos = c("https://epinowcast.r-universe.dev", getOption("repos")))

# Also requires installation of STAN:
install.packages("cmdstanr", repos = c('https://stan-dev.r-universe.dev', getOption("repos")))
```

:::

[`epinowcast`](https://package.epinowcast.org/) fits a flexible Bayesian model
with separate modules for the reporting delay and the reference (epidemic)
process. It expects a preprocessed object built by `enw_preprocess_data()`.

`tbl_now_to_epinowcast()` handles the preprocessing, returning
an object you can pass straight to `epinowcast::epinowcast()`:


### Simple nowcast

`epinowcast` becomes very slow with the 2.3M cases reporting triangle. So we need to filter earlier, 
on the *observations* keeping only the most recent 90 days before applying `tbl_now_to_epinowcast()`.

```{r enw-convert, eval=has("epinowcast"), warning=FALSE, message=FALSE}
library(epinowcast)

# Trim, then convert. 90 days otherwise the model is too slow
covid_enw_recent <- covid_now |>
  filter(notification_date >= as.Date("2021-04-01") - 90) |>
  tbl_now_to_epinowcast(max_delay = 30, verbose = FALSE, quiet = TRUE)

covid_enw_recent
```

This can then be passed to `epinowcast()`: 

```{r enw-fit, eval=run_models && has("epinowcast"), message = FALSE, warning = FALSE}
# A minimal epinowcast fit from the preprocessed object
enw_fit <- epinowcast(
  covid_enw_recent,
  fit = enw_fit_opts(
    pp = TRUE, chains = 1, iter_sampling = 250, iter_warmup = 250,
    seed = 20260824
  )
)
```

Again, the `max_delay` is a **modelling choice, not a detail**: left unset, the
converter infers it from the longest delay present, and the
nowcast then carries one reference date per delay becoming extremely slow. 


### With strata and effects.

Handing `tbl_now_to_epinowcast()` the enriched object
does two things automatically: the `sex` stratum becomes epinowcast's grouping
(`by`), so the model fits a delay per sex, and the temporal-effect columns land in
the `metareference` / `metareport` tables so that they can be used in a module formula.

```{r effects-enw-fit, eval=run_models && has("epinowcast")}
# Same trim, from the enriched object.
enw_seasonal <- covid_seasonal |>
  filter(notification_date >= as.Date("2021-04-01") - 90) |>
  tbl_now_to_epinowcast(max_delay = 30, verbose = FALSE, quiet = TRUE)

# The temporal columns carry the same names as  in `covid_seasonal` after 
# `compute_temporal_effects()`. It can be used in any parametric model
enw_seasonal_fit <- epinowcast(
  enw_seasonal,
  reference = enw_reference(
    parametric   = ~ 1 + .event_day_of_week, 
    distribution = "lognormal",
    data         = enw_seasonal
  ),
  fit = enw_fit_opts(
    pp = TRUE, chains = 1, iter_sampling = 250, iter_warmup = 250,
    seed = 20260824
  )
)
```


::: {.alert .alert-warning}
**These fits are deliberately small.** Two chains, 250 warmup and 250 sampling
iterations on a short window, because this page is a tutorial and not a final nowcast. 
:::

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r enw-tidy-1, eval=FALSE}
tidy(enw_fit)
```

```{r enw-tidy, echo=FALSE, purl=FALSE}
show_tidy("epinowcast")
```

```{r enw-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the epinowcast package", fig.alt="Three panels (Total, Female, Male) comparing the epinowcast nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("epinowcast")
```

## NobBS

::: {.alert .alert-info}
Install with: 

```r
install.packages("NobBS")
```

:::

[`NobBS`](https://cran.r-project.org/package=NobBS) works from a **linelist**
with an onset-date column and a report-date column, and it counts **rows**. 
Each row is one case. Our data are counts, so they have to be expanded first.
`tbl_now_to_nobbs()` does that and names the columns what `NobBS()` expects:

### Simple nowcast

```{r nobbs-convert, eval=has("NobBS"), warning=FALSE}
# Trim, then convert. 90 days otherwise the model is too slow
covid_linelist <- covid_now |>
  filter(notification_date >= as.Date("2021-04-01") - 90) |>
  tbl_now_to_nobbs(verbose = FALSE)

nrow(covid_linelist)   # one row per case
```

We can then nowcast with the built data frame:

```{r nobbs-fit, eval=run_models && has("NobBS")}
library(NobBS)

nobbs_fit <- NobBS(
  data          = covid_linelist,
  now           = get_now(covid_now),
  units         = "1 day",
  onset_date    = "onset_date",
  report_date   = "report_date",
  max_D         = 15,   # delays beyond 15 days are negligible here
  moving_window = 30 
)

```

The predictions come out with `tidy()`, in the same shape as every
other engine here:

```{r nobbs-tidy-1, eval=FALSE}
tidy(nobbs_fit)
```

```{r nobbs-tidy, echo=FALSE, purl=FALSE}
show_tidy("NobBS")
```


Notice that even the arguments come from the `tbl_now`: `get_now()`,
`get_event_date()` and `get_report_date()` tell `NobBS` what `tbl.now` already
figured out.

### With strata and effects.

`NobBS.strat()` fits one nowcast per stratum, and its `strata` argument names
**one** column. `tbl_now_to_nobbs()` therefore adds a `strata` column holding
every declared stratum pasted together under the name `strata`. Use
`"strata"`  as the name for `NobBS.strat`:

```{r nobbs-strata, eval=run_models && has("NobBS"), warning=FALSE, message=FALSE}
covid_linelist_sex <- covid_seasonal |>
  filter(notification_date >= as.Date("2021-04-01") - 60) |>
  tbl_now_to_nobbs(verbose = FALSE) 

stratified_nobbs <- NobBS.strat(covid_linelist_sex,
                                strata        = "strata",
                                now           = get_now(covid_seasonal),
                                units         = "1 day",
                                onset_date    = "onset_date",
                                report_date   = "report_date",
                                max_D         = 15,
                                moving_window = 30
                                )
```

::: {.alert .alert-warning}
**These fits are deliberately small.** A moving window of 30 and a maximum delay of 15 is probably too short for this data. This is done for speed as this page is a tutorial and not a final nowcast. 
:::


In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r nobbs-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the NobBS package", fig.alt="Three panels (Total, Female, Male) comparing the NobBS nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("NobBS")
```

```{r nobbs-interval, echo=FALSE, purl=FALSE}
nobbs_ci <- last_interval("NobBS")
```


## surveillance

::: {.alert .alert-info}
Install with: 

```r
install.packages("surveillance")
```

:::

[`surveillance`](https://cran.r-project.org/package=surveillance) is the
long-standing R package for outbreak detection and nowcasting, and implements
the Höhle & an der Heiden (2014) approach. `surveillance::nowcast()` works from
an individual-level **line list** with one column for the event date and one for
the report date, named by its `dEventCol` / `dReportCol` arguments.

`tbl_now_to_surveillance()` builds that data frame and renames the two dates to
`surveillance`'s own defaults:

### Simple nowcast

```{r sur-convert, eval=has("surveillance"), message=FALSE, warning=FALSE}
library(surveillance)

# Trim, then convert. 90 days otherwise the model is too slow
covid_sur_now <- covid_now |>
  filter(notification_date >= as.Date("2021-04-01") - 60)

covid_sur <- tbl_now_to_surveillance(covid_sur_now, verbose = FALSE)

head(covid_sur)
```

`surveillance::nowcast()` requires you to pass two date **grids**. Use `get_surveillance_when()` to get the dates you want estimated, and `get_surveillance_range()` for the whole axis the model is laid on. The nowcast itself needs a `now`, the dates you want estimated (`when`), and a
maximum delay `D`. All of it can come from the `tbl_now`:

```{r sur-fit, eval=run_models && has("surveillance"), message=FALSE, warning=FALSE}
#Note that in the call we use our get_now(), get_surveillance_when()
#and get_surveillance_range() functions:
sur_fit <- nowcast(
  now          = get_now(covid_sur_now),
  when         = get_surveillance_when(covid_sur_now, length = 30),
  data         = covid_sur,
  dEventCol    = "dHospital",
  dReportCol   = "dReport",
  aggregate.by = "1 day",
  D            = 15,
  method       = "bayes.notrunc.bnb",
  control      = list(dRange = get_surveillance_range(covid_sur_now), 
                      N.tInf.max = 100000, nSamples = 1000)
)

```

The predictions come out with `tidy()`, in the same shape as every
other engine here:

```{r sur-tidy-1, eval=FALSE}
tidy(sur_fit)
```

```{r sur-tidy, echo=FALSE, purl=FALSE}
show_tidy("surveillance")
```


### With strata and effects.

`surveillance::nowcast()` has **no strata argument** as it only models one
series, so a stratified analysis means one fit per stratum. Ask the converter
for `format = "linelist_list"` and it does the splitting: one line list per stratum,
in a plain list you can use `lapply()` over.

```{r sur-strata, eval=has("surveillance"), message=FALSE, warning=FALSE}
# Trim first, as it can't handle the 2.3M cases of the whole window.
covid_sur_seasonal <- covid_seasonal |>
  filter(notification_date >= as.Date("2021-04-01") - 90)

covid_sur_eff <- tbl_now_to_surveillance(
  covid_sur_seasonal,
  format  = "linelist_list",
  verbose = FALSE
)
```

```{r sur-strata-print, eval=has("surveillance"), warning=FALSE}
covid_sur_eff
```

With no strata declared this is still a list, of length one named `"all"`, so the
`lapply()` below does not have to know which case it is in. The default
`format = "linelist"` gives the same information as one frame with a pasted
`strata` column, if you would rather split it yourself.

```{r sur-strata-fit, eval=run_models && has("surveillance"), message=FALSE}
sur_by_stratum <- covid_sur_eff |>
  lapply(\(df) nowcast(
    now          = get_now(covid_sur_seasonal),
    when         = get_surveillance_when(covid_sur_seasonal, length = 30),
    data         = df,
    dEventCol    = "dHospital",
    dReportCol   = "dReport",
    aggregate.by = "1 day",
    D            = 15,
    method       = "bayes.notrunc.bnb",
    control      = list(
      dRange     = get_surveillance_range(covid_sur_seasonal),
      N.tInf.max = 100000,
      nSamples   = 1000
    )
  ))
```

`tidy()` recognises the list and labels each block with its own stratum, so
`sur_by_stratum` tidies into one table exactly like the natively stratified
engines.

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r sur-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the surveillance package", fig.alt="Three panels (Total, Female, Male) comparing the surveillance nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("surveillance")
```

```{r sur-interval, echo=FALSE, purl=FALSE}
sur_ci <- last_interval("surveillance")
```


## EpiNow2

::: {.alert .alert-info}
Install with: 

```r
install.packages("EpiNow2")
```

:::

[`EpiNow2`](https://epiforecasts.io/EpiNow2/) is not
a single nowcasting framework but a more general tool that allows you to estimate different distributions related to your epidemic process. The `tbl_now_to_EpiNow2()` function therefore takes a `target` argument, named for the function the result is passed to, so whatever it gives you can be handed over unchanged.


| `target` | you get | for |
|---|---|---|
| `"estimate_infections"` (default) | `date` / `confirm` | `estimate_infections()`, `epinow()` |
| `"regional_epinow"` | the same plus `region` | `regional_epinow()` |
| `"estimate_truncation"` | a list of snapshots | `estimate_truncation()` |
| `"estimate_dist"` | interval-censored date columns | `estimate_dist()` |

::: {.alert .alert-warning}
**EpiNow2 models a *daily* process and has no `timestep`.** To pass weekly data
we utilize the  `accumulate` column. Units coarser than a week are refused rather than approximated.
:::

### Simple nowcast

`estimate_infections()` wants the series as known at the `now` — one row per day,
with the filler days marked:

```{r epinow2-convert, eval=has("EpiNow2")}
covid_en2 <- covid_now |>
  filter(notification_date >= as.Date("2021-04-01") - 60) |>
  tbl_now_to_EpiNow2(verbose = FALSE, quiet = TRUE)

head(covid_en2)
```

EpiNow2 needs **two** things fitted before it can nowcast:

* **`delays`** The infection-to-onset delay which we take from EpiNow2's shipped
  example.
* **`truncation`** is *the nowcast itself*. That **is** what the report 
  dimension of a `tbl_now` measures.

This is a two-step fit.

```{r epinow2-fit, eval=run_models && has("EpiNow2")}
library(EpiNow2)

# STEP 1 --- fit the truncation from the report dimension.
covid_snaps <- covid_now |>
  filter(notification_date >= as.Date("2021-04-01") - 90) |>
  tbl_now_to_EpiNow2(
    target = "estimate_truncation", snapshots = 5, verbose = FALSE, quiet = TRUE
  )

truncation_fit <- estimate_truncation(
  covid_snaps,
  stan = stan_opts(samples = 250, warmup = 250, chains = 1, seed = 20260824)
)

# `$dist` is defunct; the accessor is `get_parameters()`.
fitted_truncation <- get_parameters(truncation_fit)[["truncation"]]

# STEP 2 --- nowcast with it.
epinow2_fit <- estimate_infections(
  covid_en2,
  generation_time = gt_opts(example_generation_time),
  delays          = delay_opts(example_incubation_period),
  truncation      = trunc_opts(fitted_truncation),
  rt              = rt_opts(prior = LogNormal(mean = 2, sd = 0.1), rw = 7),
  gp              = NULL,
  stan            = stan_opts(samples = 250, warmup = 250, chains = 1, 
                              seed = 20260824)
)
```


The predictions come out with `tidy()`, in the same shape as every other engine
here:

```{r epinow2-tidy-1, eval=FALSE}
tidy(epinow2_fit)
```

```{r epinow2-tidy, echo=FALSE, purl=FALSE}
show_tidy("EpiNow2")
```

### With strata and effects.

`regional_epinow()` takes a single `region` column, so the object's strata are
folded into one label:

```{r epinow2-strata, eval=has("EpiNow2"), warning=FALSE}
covid_regional <- covid_seasonal |>
  filter(notification_date >= as.Date("2021-04-01") - 90) |>
  tbl_now_to_EpiNow2(target = "regional_epinow", verbose = FALSE, quiet = TRUE)

head(covid_regional)
```

```{r epinow2-strata-fit, eval=run_models && has("EpiNow2")}
regional_fit <- regional_epinow(
  covid_regional,
  generation_time = gt_opts(example_generation_time),
  delays          = delay_opts(example_incubation_period),
  truncation      = trunc_opts(fitted_truncation),
  rt              = rt_opts(prior = LogNormal(mean = 2, sd = 0.1), rw = 7),
  gp              = NULL,
  stan            = stan_opts(samples = 250, warmup = 250, chains = 1,
                              seed = 20260824)
)
```
::: {.alert .alert-warning}
**These fits are deliberately small.** Two chains, 250 warmup and 250 sampling
iterations on a short window, because this page is a tutorial and not a final nowcast. 
:::

```{r epinow2-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the EpiNow2 package", fig.alt="Three panels (Total, Female, Male) comparing the EpiNow2 nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("EpiNow2")
```


## epidist

::: {.alert .alert-info}
Install with: 
```r
install.packages("epidist", repos = c('https://epinowcast.r-universe.dev', getOption("repos")))
```

:::

Sometimes the quantity you actually want is the **delay distribution** itself.
[`epidist`](https://epidist.epinowcast.org/) estimates exactly that, treating
each case as an interval-censored onset/report pair.

`tbl_now_to_epidist()` converts the `tbl_now` into the censored form `epidist`
expects. Our data are daily counts, so the converter produces
`epidist_aggregate_data` with one row per distinct `(delay, observation time)`
combination, and `n` as that row's weight. `sex` is still undeclared on
`covid_now`, and `epidist` never sees it, so the two rows it splits each cell
into are pooled back together here -- declaring it (as `covid_seasonal` does
just below) is how you ask for it to reach the model instead.


### Simple delay fit

```{r epidist-convert, eval=has("epidist") & has("cmdstanr"), message=FALSE, warning=FALSE}
library(epidist)

# Fit the delay distribution (see the epidist documentation for model choices)
delay_model <- covid_now |>
  filter(notification_date >= as.Date("2021-04-01") - 90) |>
  tbl_now_to_epidist(verbose = FALSE) |>
  as_epidist_marginal_model() |>
  epidist(chains = 1, iter = 250, warmup = 250, backend = "cmdstanr")
```

The fitted delay distribution can then feed back into a nowcast.

### With strata and effects.

`epidist` has no separate grouping argument, so the strata (`sex`) and the
temporal-effect columns become extra columns that can be used within the  model's 
formula. For example:

```{r epidist-strata-fit, eval=run_models && has("epidist") & has("cmdstanr")}
# A sex-varying mean delay.
delay_by_sex <- covid_seasonal |>
  filter(notification_date >= as.Date("2021-04-01") - 90) |>
  tbl_now_to_epidist(verbose = FALSE) |>
  as_epidist_marginal_model() |>
  epidist(formula = mu ~ 1 + sex + .event_season_365_sin + .event_season_365_cos,
          chains = 1, iter = 250, warmup = 250, backend = "cmdstanr")
```

::: {.alert .alert-warning}
**These fits are deliberately small.** Two chains, 250 warmup and 250 sampling
iterations on a short window, because this page is a tutorial and not a final nowcast. 
:::

`tidy()` works here too, but it returns a slightly **different table** because
`epidist` estimates a different thing. In this case you get one row per parameter of
the fitted delay distribution:

```{r epidist-tidy, eval=FALSE}
tidy(delay_model)
```


```{r epidist-caveat, echo=FALSE, purl=FALSE}
show_tidy("epidist")
```




## The `tidy()` function

The converters normalize what goes *into* each package. `tidy()` normalizes what
comes out. Every engine above returns something different: a matrix of
draws, an `stsNC` object, a Stan fit, an INLA summary, a bare list. The `tidy()`
turns any of them into the same table:

```{r tidy-demo, message=FALSE, warning=FALSE, eval=FALSE}
tidy(nowcast_samples)
```

```{r tidy-demo-out, echo=FALSE, purl=FALSE}
show_tidy("baselinenowcast")
```


The columns are the same regardless of the package that produced the fit:

| column | meaning |
|---|---|
| `event_date` | event/reference date, **on the engine's own grid** |
| `stratum` | `"all"` when the fit is unstratified |
| `estimate` | point nowcast (posterior median where available) |
| `conf.low`, `conf.high` | interval bounds, using \pkg{broom}'s names |
| `level` | the width that interval **actually** has |
| `engine` | which package produced it |
| `qXX` | (optional) quantile columns of the prediction |

Pass `probs` for columns with other quantiles:

```{r tidy-probs, message=FALSE, warning=FALSE, eval=FALSE}
tidy(nowcast_samples, probs = c(0.05, 0.5, 0.95))
```

```{r tidy-probs-out, echo=FALSE, purl=FALSE}
show_tidy("baselinenowcast_probs")
```

::: {.alert .alert-warning}
**Only engines that keep draws can answer an arbitrary `probs`.** That is
`diseasenowcasting`, `baselinenowcast` and `epinowcast`. `NobBS` and
`surveillance` report a fixed set of summaries, so asking them for a quantile
they never computed is an **error** . To be
able to do so you need to specify at **fit time** the quantiles you want. For example with
`NobBS` you can use the `specs` to set `quantiles`:

```{r nobbs-quantiles-1, eval=run_models && has("NobBS")}
nobbs_quantiles <- NobBS(
  data          = covid_linelist,
  now           = get_now(covid_now),
  units         = "1 day",
  onset_date    = "onset_date",
  report_date   = "report_date",
  max_D         = 10,
  moving_window = 104,
  specs         = list(quantiles = c(0.1, 0.5, 0.9))   # <- ask here
)
```

so that they can be called with `tidy`:

```{r nobbs-probs, message=FALSE, warning=FALSE, eval=FALSE}
tidy(nobbs_quantiles, probs = c(0.1, 0.5, 0.9))
``` 

```{r nobbs-probs-out, echo=FALSE, purl=FALSE}
show_tidy("NobBS_quantiles")
```
:::


## Summary


```{r comparison-all-code, eval=FALSE, echo=FALSE, purl=FALSE}
window <- meta$horizon
tru <- truth[truth$event_date %in% window, ]
est <- nowcasts[nowcasts$event_date %in% window, ]
tru$stratum <- factor(tru$stratum, levels = strata_levels)
est$stratum <- factor(est$stratum, levels = strata_levels)
est$package <- factor(est$package, levels = names(engine_colours))

ggplot(mapping = aes(x = event_date)) +
  geom_col(data = tru, aes(y = observed, fill = "Reported by now"), width = 0.8) +
  geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.7) +
  # Truth last, so it stays visible where the engines lie on top of it.
  geom_line(
    data = tru, aes(y = truth, linetype = "Truth"),
    linewidth = 0.7, colour = "#262626"
  ) +
  facet_wrap(~stratum, nrow = 1, scales = "free_y") +
  scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
  scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
  scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
  labs(x = "Event date (notification)", y = "Cases",
       title = "Every engine on the same data",
       subtitle = paste0("Now = ", format(meta$now))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.box = "horizontal",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335")) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE, order = 1))
```


```{r comparison-all, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4.4, out.width="100%", fig.alt="Three panels (Total, Female, Male). Grey bars show the counts reported by now, a dark line the counts those dates eventually reached, and one coloured line per package its nowcast."}
window <- meta$horizon
tru <- truth[truth$event_date %in% window, ]
est <- nowcasts[nowcasts$event_date %in% window, ]
tru$stratum <- factor(tru$stratum, levels = strata_levels)
est$stratum <- factor(est$stratum, levels = strata_levels)
est$package <- factor(est$package, levels = names(engine_colours))

ggplot(mapping = aes(x = event_date)) +
  geom_col(data = tru, aes(y = observed, fill = "Reported by now"), width = 0.8) +
  geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.7) +
  # Truth last, so it stays visible where the engines lie on top of it.
  geom_line(
    data = tru, aes(y = truth, linetype = "Truth"),
    linewidth = 0.7, colour = "#262626"
  ) +
  facet_wrap(~stratum, nrow = 1, scales = "free_y") +
  scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
  scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
  scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
  labs(x = "Event date (notification)", y = "Cases",
       title = "Every engine on the same data",
       subtitle = paste0("Now = ", format(meta$now))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.box = "horizontal",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335")) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE, order = 1))
```


We described the data once as a `tbl_now`, and then a single converter call
(or, for `diseasenowcasting`, no call at all) handed it to each package in the
shape it needed:

```{r recap, eval=FALSE, purl=FALSE}
covid_now <- tbl_now(covid_colombia,
                     event_date  = notification_date,
                     report_date = diagnosis_date,
                     case_count  = n,
                     data_type   = "count-incidence")

covid_now                              # diseasenowcasting
tbl_now_to_baselinenowcast(covid_now)  # baselinenowcast
tbl_now_to_epinowcast(covid_now)       # epinowcast
tbl_now_to_epidist(covid_now)          # epidist
tbl_now_to_surveillance(covid_now)     # surveillance
tbl_now_to_nobbs(covid_now)            # NobBS
tbl_now_to_EpiNow2(covid_now)          # EpiNow2
as.data.frame(covid_now)               # others
```

Attaching **strata** and **temporal effects** once (`covid_seasonal`) uses the
same converters: each package receives them in whatever way it can use. 

```{r learning-more, echo=FALSE, results="asis", purl=FALSE}
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
``` 
