---
title: "One dataset, many nowcasts: using tbl.now with different modelling packages"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{One dataset, many nowcasts: using tbl.now with different modelling packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The heavy model fits (Stan / JAGS / MCMC) are shown as illustrative code but
# not run when the vignette is built. Flip `run_models` to TRUE (and make sure
# the corresponding packages are installed) to actually fit them.
run_models <- FALSE

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)
```

## Why this vignette?

Preparing the same data different ways for the different nowcasting models can be tedious and
error-prone. This is exactly what [`tbl.now`](https://rodrigozepeda.github.io/tbl.now/)
helps with. You can describe your data once by specifying which column is the event date, which
is the report date, whether your data is linelist or has counts (and what _those_ counts mean!) and `tbl.now` will hands it to each modelling package in the format that package expects. 

> In this vignette we take a single dataset (`covid_colombia`), and from that one object, use several different nowcasting / delay-estimation tools:

| Package | What it does | Additional requirements | `tbl.now` converter |
|---|---|---|---|
| [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) | flexible Bayesian nowcast (delay + epidemic processes) | none (uses `RTMB`) | consumes a `tbl_now` **directly** |
| [baselinenowcast](https://baselinenowcast.epinowcast.org/) | fast, assumption-light baseline nowcast | none | `tbl_now_to_baselinenowcast()` |
| [epinowcast](https://package.epinowcast.org/) | flexible Bayesian nowcast (delay + reference modules) | **Stan** | `tbl_now_to_epinowcast()` |
| [epidist](https://epidist.epinowcast.org/) | estimates only the reporting **delay distribution** | **Stan** | `tbl_now_to_epidist()` |
| [NobBS](https://cran.r-project.org/package=NobBS) | Nowcasting by Bayesian Smoothing | **JAGS** | `tbl_now_to_nobbs()` |
| [surveillance](https://cran.r-project.org/package=surveillance) | the classic Höhle & an der Heiden nowcast | none for the method used here (**JAGS** for `bayes.trunc.ddcp`) | `tbl_now_to_surveillance()` |
| [EpiNow2](https://epiforecasts.io/EpiNow2/) | renewal-equation $R_t$, reporting truncation, and delay distributions | **Stan** (`cmdstanr`) | `tbl_now_to_EpiNow2()` |

You do not need to be an expert in any of them to follow along, the point is to show
how little changes on *your* side when you switch models.

::: {.alert .alert-warning}
**Each of these is a separate package, and `tbl.now` does not install any of
them.** It only knows how to *talk* to them. Install whichever you actually want
to use — and note that some also need software outside R (the *Additional
requirements* column above): **Stan** via `cmdstanr` (needed by \pkg{epinowcast}, \pkg{epidist} and
\pkg{EpiNow2}) and **JAGS** as a standalone program (needed by \pkg{NobBS}).

```r
# CRAN packges
install.packages(c("baselinenowcast", "EpiNow2", "NobBS", "surveillance"))

# Not on CRAN:
install.packages("cmdstanr",   repos = c('https://stan-dev.r-universe.dev', getOption("repos")))
install.packages("epidist",    repos = "https://epinowcast.r-universe.dev")
install.packages("epinowcast", repos = "https://epinowcast.r-universe.dev")

# Packages from Github:
install.packages("pak") 
pak::pkg_install("RodrigoZepeda/diseasenowcasting")
```

Every code block below is guarded, so the article still builds when a package is
missing — but nothing will run on *your* machine until you install it.
:::

## The data

`tbl.now` ships with `covid_colombia`, **daily** COVID-19 case counts from
Colombia's national surveillance system (INS), 2020&ndash;2023. Each row is a
`(notification_date, diagnosis_date, sex)` combination with a case count `n`:
the date the case was notified, and the date the laboratory diagnosis was
registered. The gap between the two is the reporting delay.

```{r setup, message=FALSE}
library(dplyr)
library(tbl.now)

data(covid_colombia)
```

```{r, echo = FALSE}
head(covid_colombia)
```

We build a single `tbl_now`. We say which column is the event date
(`notification_date`), which is the report date (`diagnosis_date`) and which
holds the counts (`n`); `tbl.now` infers the rest — that the grid is daily, and
that the "now" is the last report date.

We cut **both** dates at the start of March 2023. The example assumes we are
back on that date, nowcasting with only the information available then
(`now = "2023-03-02"`).

`covid_colombia` is **daily** and already **aggregated**: one row per
`(notification_date, diagnosis_date, sex)` with a case count `n`. That makes it
`count-incidence` data rather than a line list. It also means a *pooled* object
has to sum over `sex` first — leaving `sex` in the data without declaring it as a
stratum would put two rows in every `(event, report)` cell, and a reporting
triangle has one slot per cell.

```{r build-tbl-now}
cutoff <- as.Date("2023-03-03")

covid <- covid_colombia |>
  filter(notification_date < cutoff, diagnosis_date < cutoff)

covid_now <- covid |>
  group_by(notification_date, diagnosis_date) |>
  summarise(n = sum(n), .groups = "drop") |>
  tbl_now(
    event_date  = notification_date,
    report_date = diagnosis_date,
    case_count  = n,
    data_type   = "count-incidence"
  )

covid_now
```

That is `r format(sum(covid$n), big.mark = ",")` cases over
`r dplyr::n_distinct(covid$notification_date)` days, from March 2020 to
`r format(get_now(covid_now))`. Everything in this article — every fit, every
figure — is built from this one object, so following along step by step
reproduces exactly what you see.

::: {.alert .alert-info}
**A note on series length**, with the measured cost of each choice on this
1,094-day series:

* `diseasenowcasting` takes the **whole** series untrimmed (33s). Nothing is
  dropped.
* `baselinenowcast` also keeps every day, but the **delays** are capped at 30
  days (55s). That is a modelling assumption, not a workaround: 99% of reports
  arrive within 15 days, and a single 330-day straggler otherwise makes the
  triangle 1094 &times; 331 and the fit take 314s.
* `NobBS` and `surveillance` need the series shortened, and each has its own
  argument for it: `moving_window` and `when`. They also need a
  **line list**, so the counts are expanded to one row per case — after trimming,
  because the whole series is 6.1M cases.
* `epinowcast` has no such argument, because the reporting triangle is already
  built by the time you hold a preprocessed object. It is trimmed one step
  earlier, in its own section.
* `epidist` is the awkward one here and gets its own note in that section.
:::

### A more complex object: strata and temporal effects

Reporting delays are rarely constant and many series are reported for several
**strata** (here, `sex`) that you may want to nowcast separately. Here we build a
second object, `covid_seasonal`, that declares `sex` as a stratum and adds
day-of-week plus annual-Fourier **temporal (delay) effects**. Daily data makes
day-of-week the effect worth modelling: only 9.6% of these cases are notified on
a Sunday against 16.1% on a Tuesday.

```{r build-seasonal, message=FALSE}
covid_seasonal <- covid |>
  tbl_now(
    event_date  = notification_date,
    report_date = diagnosis_date,
    case_count  = n,
    strata      = sex,       # nowcast Male/Female separately
    data_type   = "count-incidence",
    verbose     = FALSE
  ) |>
  add_temporal_effects(
    temporal_effects(
      day_of_week = TRUE,    # a separate level per weekday
      seasons     = 365      # an annual Fourier cycle (period = 365 days)
    )
  )
```

For each package below we show the **bare** conversion of `covid_now` first, then
how the same call carries the strata and the temporal effects when handed
`covid_seasonal`. 

::: {.alert .alert-info}
**TIP**: Run `autoplot()`
with `panels = "delay_seasonality"` or `"delay_calendar"` first to *see* whether
such effects are present before modelling them.
:::

> **Checking each nowcast**
Every section below ends with the same figure: three panels — the **total**
series and the two **sex** strata — each showing what had been reported by
`now` (grey bars), what those dates eventually reached (the dark **truth** line),
and that package's nowcast, with its uncertainty interval where the package
provides one.


```{r comparison-load, echo=FALSE, message=FALSE}
library(ggplot2)

comparison <- readRDS("nowcast-comparison.rds")
nowcasts   <- comparison$nowcasts
truth      <- comparison$truth
meta       <- comparison$meta

strata_levels  <- c("Total", "Female", "Male")

# The fits are precomputed (see `data-raw/nowcast_comparison.R`), so the article
# cannot call `print(fit)` or `tidy(fit)` live. It shows the REAL output of those
# calls instead: the precompute captured them and they are replayed here, so the
# code you see above each result is exactly the code that produced it.
show_fit <- function(package) {
  captured <- comparison$display[[package]]$printed
  if (is.null(captured)) return(invisible(NULL))
  cat(captured, sep = "\n")
}

show_tidy <- function(package, n = 5) {
  tidied <- comparison$display[[package]]$tidy
  if (is.null(tidied)) return(invisible(NULL))
  head(tidied, n)
}
engine_colours <- c(
  diseasenowcasting = "#0072B2",
  baselinenowcast   = "#D55E00",
  epinowcast        = "#009E73",
  NobBS             = "#CC79A7",
  surveillance      = "#E69F00"
)
```


```{r comparison-helper, message=FALSE, echo=FALSE}
# One figure per package: Total, Female and Male side by side.
nowcast_panels <- function(package, subtitle = NULL) {
  window <- meta$horizon

  tru <- truth[truth$event_date %in% window, ]
  est <- nowcasts[
    nowcasts$package == package & nowcasts$event_date %in% window,
  ]
  tru$stratum <- factor(tru$stratum, levels = strata_levels)
  est$stratum <- factor(est$stratum, levels = strata_levels)

  colour <- unname(engine_colours[package])

  ggplot(mapping = aes(x = event_date)) +
    # What had been reported by `now` -- the incomplete data the engine saw.
    geom_col(
      data = tru, aes(y = observed, fill = "Reported by now"), width = 5.5
    ) +
    # The interval, where the engine gives one.
    geom_ribbon(
      data = est[!is.na(est$lower), ],
      aes(ymin = lower, ymax = upper), fill = colour, alpha = 0.25
    ) +
    # The nowcast...
    geom_line(data = est, aes(y = estimate, colour = "Nowcast"), linewidth = 0.9) +
    geom_point(data = est, aes(y = estimate, colour = "Nowcast"), size = 1.1) +
    # ...then the truth ON TOP, so it stays visible where the two coincide
    # (over the settled days they lie exactly on each other).
    geom_line(
      data = tru, aes(y = truth, linetype = "Truth"),
      linewidth = 0.6, colour = "#262626"
    ) +
    facet_wrap(~stratum, nrow = 1, scales = "free_y") +
    scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
    scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
    scale_colour_manual(name = NULL, values = c("Nowcast" = colour)) +
    labs(
      x = "Event date (notification)", y = "Cases",
      title = paste0(package, " nowcast vs. what actually happened"),
      subtitle = subtitle %||% paste0(
        "Now = ", format(meta$now),
        "; last ", length(window), " days"
      )
    ) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position  = "top",
      legend.box       = "horizontal",
      legend.margin    = margin(b = -4),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold", colour = "#334335")
    )
}
```


## diseasenowcasting

[`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) is
designed hand-in-hand with `tbl.now`, so it takes a `tbl_now` **directly**. You
just hand it the object:

### Simple nowcast

```{r dnc, eval=run_models && has("diseasenowcasting")}
# `diseasenowcasting::` rather than `library()`: it is not on CRAN, so attaching
# it in a vignette leaves `R CMD check` with an undeclared dependency. Qualifying
# the call also keeps this article's `tidy()` calls on the `generics` generic --
# versions before 2.1.0 declared their own, which masks every other package's
# methods with no error.
dnc_fit <- diseasenowcasting::nowcast(covid_now)

dnc_fit
```

```{r dnc-out, echo=FALSE, eval=!run_models}
show_fit("diseasenowcasting")
```


### With strata and effects.

`diseasenowcasting` reaches into the `tbl_now` itself, so the enriched object needs no extra arguments: it picks up the `sex` stratum and the day-of-week / seasonal effect columns automatically.

```{r dnc-teff, eval=run_models && has("diseasenowcasting")}
diseasenowcasting::nowcast(covid_seasonal)   # strata and effects used automatically
```

```{r dnc-teff-out, echo=FALSE, eval=!run_models}
show_fit("diseasenowcasting_seasonal")
```

Predictions can be obtained via `tidy`:

```{r dnc-teff-tidy-1, eval=FALSE}
tidy(diseasenowcasting::nowcast(covid_seasonal))
```

```{r dnc-teff-tidy, echo=FALSE}
show_tidy("diseasenowcasting_seasonal")
```

In both stratified and unstratified cases:

```{r dnc-tidy-1, eval=FALSE}
tidy(dnc_fit)
```

```{r dnc-tidy, echo=FALSE}
show_tidy("diseasenowcasting")
```

::: {.alert .alert-info}
**NOTE**: this needs **diseasenowcasting >= 2.1.0**. Earlier versions declared
their own `tidy()` generic rather than re-exporting the shared one, so a bare
`tidy(dnc_fit)` returned that package's *model-parameter* table (`term`,
`std.error`, ...) instead of a nowcast — no error, just a different table. On
2.1.0 and later, `tidy()` returns the nowcast and the parameter table moved to
`model_parameters()`. On an older version, use
`tbl.now::tidy(predict(dnc_fit))`.

A related hazard remains with \pkg{broom}: loading it overwrites `tbl.now`'s
`tidy.list()` method, which is what \pkg{NobBS} fits dispatch on. Qualify the
call as `tbl.now::tidy()` there if \pkg{broom} is attached.
:::

```{r dnc-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the diseasenowcasting package", fig.alt="Three panels (Total, Female, Male) comparing the diseasenowcasting nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("diseasenowcasting")
```

## baselinenowcast

[`baselinenowcast`](https://baselinenowcast.epinowcast.org/) is a 
simple, fast baseline. It works from a **reporting triangle**, a matrix with one
row per event (reference) date and one column per reporting delay. The
lower-right corner of the matrix corresponds to the not-yet-observed part the nowcast will fill in.

`tbl_now_to_baselinenowcast()` builds that triangle directly from the
count-incidence data:

### Simple nowcast

```{r bln-convert, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
library(baselinenowcast)

# Cap the delays first: a single 330-day straggler would otherwise give the
# triangle 331 columns. 99% of reports arrive within 15 days.
covid_triangle <- covid_now |>
  filter(.delay <= 30) |>
  tbl_now_to_baselinenowcast(verbose = FALSE)

# rows = notification dates, columns = delay in days
covid_triangle[1:5, 1:6]
```

::: {.alert .alert-info}
**Why cap the delays?** It is a modelling assumption, not a workaround. The
triangle keeps every one of the `r nrow(covid_triangle)` event dates; only the
delay axis is bounded. Left uncapped the fit takes **314s** against **55s**, for
a tail that carries well under 1% of cases. `tbl_now_to_epinowcast()` has a
`max_delay` argument for the same purpose; for `baselinenowcast` the cap is an
ordinary `filter()` on the delay column.
:::

From here you can follow `baselinenowcast`'s own workflow. For example calling
`baselinenowcast()` to estimate the delay from the triangle, apply it, and
draw nowcast samples:


```{r bln-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One-call workflow: estimate the delay, apply it, and draw nowcast samples.
nowcast_samples <- baselinenowcast(
  covid_triangle,
  output_type = "samples",
  draws       = 1000
)

```

### With strata and effects.

A reporting-triangle *matrix* is a single table with no room for strata or
covariates, so `format = "matrix"` pools the strata (with a warning) and drops
the effect columns. To nowcast each stratum we need to do it separately via a
`format = "triangle_list"`. This creates one reporting-triangle per stratum.

```{r bln-strata, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
# One reporting triangle per stratum, straight from the object.
triangles_by_stratum <- tbl_now_to_baselinenowcast(
  covid_seasonal,
  format  = "triangle_list",
  verbose = FALSE
)
```

```{r bln-strata-print, eval=has("baselinenowcast"), warning=FALSE}
triangles_by_stratum
```

This can be used to nowcast each stratum via `lapply`:

```{r bln-strata-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One nowcast per triangle, each trimmed the same way.
nowcasts_by_stratum <- triangles_by_stratum |>
  lapply(\(tri) baselinenowcast(tri, output_type = "samples", draws = 1000))
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r bln-tidy-1, eval=FALSE}
tidy(nowcast_samples)
```

```{r bln-tidy, echo=FALSE}
show_tidy("baselinenowcast")
```

```{r bln-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the baselinenowcast package", fig.alt="Three panels (Total, Female, Male) comparing the baselinenowcast nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("baselinenowcast")
```

## epinowcast

[`epinowcast`](https://package.epinowcast.org/) fits a flexible Bayesian model
with separate modules for the reporting delay and the reference (epidemic)
process. It expects a preprocessed object built by `enw_preprocess_data()`.

`tbl_now_to_epinowcast()` handles the preprocessing, returning
an object you can pass straight to `epinowcast::epinowcast()`:

**Series length: trimmed, before preprocessing.** `epinowcast` is the one engine
here with no argument for shortening the series, because by the time you hold a
preprocessed object the reporting triangle is already built. Its filters work one
step earlier, on the *observations* — and `tbl_now_to_epinowcast()` will hand you
those instead of the finished object if you ask with `preprocess = FALSE`:

::: {.alert .alert-warning}
**Warning: requires Stan.** This package fits its model with **Stan**, so it needs a working `cmdstanr` (or `rstan`) toolchain installed before any of the code below will run.
:::

### Simple nowcast

```{r enw-convert, eval=has("epinowcast"), warning=FALSE, message=FALSE}
library(epinowcast)

# Trim FIRST, then convert: two years of history is plenty and keeps the Stan
# fit tractable.
covid_enw_recent <- covid_now |>
  filter(notification_date >= cutoff - 7 * 104) |>
  tbl_now_to_epinowcast(max_delay = 10, verbose = FALSE, quiet = TRUE)

covid_enw_recent
```

Two years preprocesses in well under a second, and it is `covid_enw_recent` —
not the full series — that goes to the Stan fit.

This can then be passed to `epinowcast()`: 

```{r enw-fit, eval=run_models && has("epinowcast"), message = FALSE, warning = FALSE}
# A minimal epinowcast fit from the preprocessed object
fit <- epinowcast(
  covid_enw_recent,   # the trimmed object, not the full series
  fit = enw_fit_opts(
    pp = TRUE, chains = 2, iter_sampling = 500, iter_warmup = 500
  )
)
```

`max_delay` above is a **modelling choice, not a detail**: left unset, the
converter infers it from the longest delay present (330 days here), and the
nowcast then carries one reference date per delay. Setting it to 10 matches the
maximum delay every other engine on this page is given, so the comparison at the
end is like for like.

### With strata and effects.

Handing `tbl_now_to_epinowcast()` the enriched object
does two things automatically: the `sex` stratum becomes epinowcast's grouping
(`by`), so the model fits a delay per sex, and the temporal-effect columns land in
the `metareference` / `metareport` tables so that they can be used in a module formula.

```{r effects-enw-fit, eval=run_models && has("epinowcast")}
# Same trim, from the enriched object.
enw_seasonal <- covid_seasonal |>
  filter(notification_date >= cutoff - 7 * 104) |>
  tbl_now_to_epinowcast(max_delay = 10, verbose = FALSE, quiet = TRUE)

# Drop the seasonal terms into a reference-module formula and fit as usual: the
# covariates now enter the reference model.
epinowcast(
  enw_seasonal,
  reference = enw_reference(
    parametric   = ~ 1 + .event_season_52_sin + .event_season_52_cos,
    distribution = "lognormal",
    data         = enw_seasonal
  ),
  fit = enw_fit_opts(
    pp = TRUE, chains = 2, iter_sampling = 500, iter_warmup = 500
  )
)
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r enw-tidy-1, eval=FALSE}
tidy(fit)
```

```{r enw-tidy, echo=FALSE}
show_tidy("epinowcast")
```

```{r enw-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the epinowcast package", fig.alt="Three panels (Total, Female, Male) comparing the epinowcast nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("epinowcast")
```

## epidist

Sometimes the quantity you actually want is the **delay distribution** itself —
how long, on average, between onset and report, and how variable is it?
[`epidist`](https://epidist.epinowcast.org/) estimates exactly that, treating
each case as an interval-censored onset/report pair.

`tbl_now_to_epidist()` converts the `tbl_now` into the censored form `epidist`
expects. Our data are daily counts, so the converter produces
`epidist_aggregate_data` — one row per distinct `(delay, observation time)`
combination with a weight — rather than one row per case:

::: {.alert .alert-warning}
**Warning: requires Stan.** This package fits its model with **Stan**, so it needs a working `cmdstanr` (or `rstan`) toolchain installed before any of the code below will run.
:::

### Simple delay fit

```{r epidist-convert, eval=has("epidist"), message=FALSE, warning=FALSE}
library(epidist)

covid_epidist <- tbl_now_to_epidist(covid_now, verbose = FALSE)

```

`epidist` offers several model types, and on **count** data the choice matters
more than it looks. The **marginal** model is the one built for aggregated
counts — but it does not currently compile (see the warning below). The
**latent** model works, at a price: it carries one latent variable per *case*, so
handing it the whole series expands 6.1M cases into 6.1M rows and it never
finishes. It is therefore fitted on the last 30 days (2,705 cases, about seven
minutes). A delay distribution needs delays, not a long series.

```{r epidist-fit, eval=FALSE}
# Fit the delay distribution (see the epidist documentation for model choices)
delay_model <- covid_now |>
  filter(notification_date >= cutoff - 30) |>
  tbl_now_to_epidist(verbose = FALSE) |>
  as_epidist_latent_model() |>
  epidist()
```

::: {.alert .alert-warning}
**The `marginal` model does not currently fit, and it is the one you would want
here.** With **epidist 0.4.0** and **primarycensored 1.5.1**,
`as_epidist_marginal_model()` fails at Stan compilation: the generated code calls
`primarycensored_lpmf()` with 8 arguments against a 9-argument signature
(primarycensored gained an `L` lower-truncation argument that the generated call
does not supply). That matters more for count data than for a line list, because
the marginal model is what consumes aggregated counts efficiently — without it,
the latent model has to expand them back to one row per case. Reported upstream;
check the [epidist issue tracker](https://github.com/epinowcast/epidist/issues/)
before reaching for `as_epidist_marginal_model()`.
:::

The fitted delay distribution can then feed back into a nowcast for example as
a data-informed prior in `epinowcast`.

### With strata and effects.

`epidist` has no separate grouping argument, so the strata (`sex`) and the
temporal-effect columns become extra columns that can be used within the  model's 
formula. For example:

```{r epidist-strata, eval=has("epidist"), message=FALSE, warning=FALSE}
covid_epidist_eff <- tbl_now_to_epidist(covid_seasonal, verbose = FALSE)
```

```{r epidist-strata-fit, eval=FALSE}
# A sex-varying mean delay.
delay_by_sex <- covid_seasonal |>
  filter(notification_date >= cutoff - 30) |>
  tbl_now_to_epidist(verbose = FALSE) |>
  as_epidist_latent_model() |>
  epidist(formula = mu ~ 1 + sex)
```

`tidy()` works here too, but it returns a **different table**, because
`epidist` estimates a different thing. There are no per-date case estimates to
report, so instead of one row per event date you get one row per parameter of
the fitted delay distribution:

```{r epidist-tidy, eval=FALSE}
tidy(delay_model)
```

| column | meaning |
|---|---|
| `term` | a parameter of the fitted delay distribution |
| `estimate` | posterior median |
| `conf.low`, `conf.high` | interval bounds, using \pkg{broom}'s names |
| `level` | the width of that interval |
| `engine` | always `"epidist"` |

The `term` values are the distribution's own parameters (`mu`, `sigma`, or
whichever the `family` has) **plus** the derived `mean` and `sd` — the delay
summaries those parameters imply, which are usually the numbers you actually
want to quote. The fit keeps its draws, so `probs` works as it does elsewhere:
`tidy(delay_model, probs = c(0.05, 0.95))` adds `q5` and `q95` columns.

::: {.alert .alert-warning}
**Watch the dispatch.** `epidist()` returns an object whose class is
`c("brmsfit", "epidist_fit")`, in that order. If \pkg{broom.mixed} is loaded,
its `tidy.brmsfit()` method matches **first** and you get raw \pkg{brms}
parameters instead of the delay table above. Call `tidy.epidist_fit(fit)`
explicitly when you cannot be sure which method wins.
:::

The natural use of the result here is as a **data-informed delay prior** for one
of the other engines.

### A caveat on this dataset: same-day reporting

Every other section ends with the nowcast plotted against the truth. The
equivalent check here would be the fitted delay distribution against the delays
actually observed — and on `covid_colombia` that check **fails**, for a reason
worth understanding before you reach for a delay model on your own data.

**57.4%** of the cases in the fitting window carry a delay of exactly **zero**
days: notification and laboratory diagnosis are recorded on the same date.
Another 20% arrive the next day. A lognormal — `epidist`'s default family, and
the usual choice for a delay — has no way to represent a point mass at zero,
because `log(0)` is undefined. Asked to fit one anyway, it does not fail loudly;
it inflates the variance until the density piles up near zero, and reports:

| term | estimate |
|---|---|
| `mu` | 7.36 |
| `sigma` | **17.95** |
| `mean` | 1.5 &times; 10^73^ days |

A mean delay of 10^73^ days is not a convergence warning you can tune away. It is
the model telling you that the family cannot describe the data.

::: {.alert .alert-warning}
**What to do instead.** If same-day reporting dominates your series, a
continuous positive-support distribution is the wrong tool. The options are to
model the delay as **discrete** (a distribution that puts real mass on 0), to fit
a **zero-inflated** or hurdle form that separates "reported same day" from "how
long otherwise", or to keep the continuous fit for the *non-zero* delays only and
report the zero share separately. `epidist` also offers `Gamma()` and
`weibull()`, but neither solves this: both have zero density at zero too.

This is a property of the data, not of `tbl.now` or `epidist` — the converter
does its job, and the diagnosis above comes straight from `tidy()`.
:::

## NobBS

[`NobBS`](https://cran.r-project.org/package=NobBS) works from a **line list**
with an onset-date column and a report-date column, and it counts **rows** —
each row is one case. Our data are counts, so they have to be expanded first.
`tbl_now_to_nobbs()` does that and names the columns what `NobBS()` expects:

::: {.alert .alert-warning}
**Warning: requires JAGS.** This package fits its model with **JAGS**, which is a separate program: install JAGS itself, not just the R package, before running the code below.
:::

### Simple nowcast

```{r nobbs-convert, eval=has("NobBS")}
# Trim BEFORE converting: the expansion is one row per case, and `moving_window`
# below limits only what NobBS fits, not what it is handed.
covid_linelist <- covid_now |>
  filter(notification_date >= cutoff - 180) |>
  tbl_now_to_nobbs(verbose = FALSE)

nrow(covid_linelist)   # one row per case
```

::: {.alert .alert-warning}
**Do not hand `NobBS()` count data directly.** It counts rows, so a table of
1,174 rows carrying 50,160 cases is nowcast as **1,174 cases** — no error, just
an answer that is 40 times too small. The converter exists to make that
impossible.
:::


```{r nobbs-fit, eval=run_models && has("NobBS")}
library(NobBS)

nobbs_fit <- NobBS(
  data          = covid_linelist,
  now           = get_now(covid_now),
  units         = "1 day",
  onset_date    = "onset_date",
  report_date   = "report_date",
  max_D         = 30,   # delays beyond 30 days are negligible here
  moving_window = 180   # fit to the last six months
)

```

The predictions come out with `tidy()`, in the same shape as every
other engine here:

```{r nobbs-tidy-1, eval=FALSE}
tidy(nobbs_fit)
```

```{r nobbs-tidy, echo=FALSE}
show_tidy("NobBS")
```


Notice that even the arguments come from the `tbl_now`: `get_now()`,
`get_event_date()` and `get_report_date()` tell `NobBS` what `tbl.now` already
figured out.

### With strata and effects.

Call `NobBS.strat()` and pass the column name to the strata option:

```{r nobbs-strata, eval=run_models && has("NobBS"), warning=FALSE, message=FALSE}
dengue_nobbs     <- as.data.frame(covid_seasonal)
stratified_nobbs <- NobBS.strat(dengue_nobbs,
                                strata        = "sex",
                                now           = get_now(covid_seasonal),
                                units         = "1 day",
                                onset_date    = get_event_date(covid_seasonal),
                                report_date   = get_report_date(covid_seasonal),
                                max_D         = 10,
                                moving_window = 104
                                )
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r nobbs-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the NobBS package", fig.alt="Three panels (Total, Female, Male) comparing the NobBS nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("NobBS")
```

## surveillance

[`surveillance`](https://cran.r-project.org/package=surveillance) is the
long-standing R package for outbreak detection and nowcasting, and implements
the Höhle & an der Heiden (2014) approach. `surveillance::nowcast()` works from
an individual-level **line list** with one column for the event date and one for
the report date, named by its `dEventCol` / `dReportCol` arguments.

`tbl_now_to_surveillance()` builds that data frame and renames the two dates to
`surveillance`'s own defaults:

### Simple nowcast

```{r sur-convert, eval=has("surveillance"), message=FALSE}
library(surveillance)

# Trim before converting: the converter expands counts to one row per case.
covid_sur <- covid_now |>
  filter(notification_date >= cutoff - 180) |>
  tbl_now_to_surveillance(verbose = FALSE)

head(covid_sur)
```

The nowcast itself needs a `now`, the dates you want estimated (`when`), and a
maximum delay `D`. All of it can come from the `tbl_now`:

```{r sur-fit, eval=run_models && has("surveillance"), message=FALSE, warning=FALSE}
sur_now  <- get_now(covid_now)
sur_when <- seq(sur_now - 30, sur_now, by = "1 day")   # days to estimate

# `dRange` gives surveillance the grid explicitly, so it reaches the `now` even
# if the most recent days carry no reports yet: a line list cannot express a
# zero day (no rows), so zero-completion is no help here.
sur_range <- seq(min(covid_sur$dHospital), sur_now, by = "1 day")

sur_fit <- nowcast(
  now          = sur_now,
  when         = sur_when,
  data         = covid_sur,
  dEventCol    = "dHospital",
  dReportCol   = "dReport",
  aggregate.by = "1 day",
  D            = 30,
  method       = "bayes.notrunc.bnb",
  control      = list(dRange = sur_range, N.tInf.max = 100000, nSamples = 1000)
)

```

The predictions come out with `tidy()`, in the same shape as every
other engine here:

```{r sur-tidy-1, eval=FALSE}
tidy(sur_fit)
```

```{r sur-tidy, echo=FALSE}
show_tidy("surveillance")
```

::: {.alert .alert-info}
**TIP**: the interval is already there. `surveillance::nowcast()` stores a
prediction interval in the returned object's `pi` slot, at whatever width
`control$alpha` names (`0.05`, so 95%, by default), and `tidy()` reads it into
`conf.low` / `conf.high`. The JAGS-backed `bayes.trunc` and `bayes.trunc.ddcp`
methods are worth reaching for when you want the delay distribution modelled
differently, **not** merely to get uncertainty out of the fit.
:::


### With strata and effects.

`surveillance::nowcast()` has **no strata argument** as it only models one series. The
strata ride along as ordinary columns, so fit one nowcast per stratum in a loop:

```{r sur-strata, eval=has("surveillance"), message=FALSE, warning=FALSE}
covid_sur_eff <- tbl_now_to_surveillance(covid_seasonal, verbose = FALSE)

# `sex` is kept as a column, so it can be split on.
head(covid_sur_eff)
```

```{r sur-strata-fit, eval=run_models && has("surveillance"), message=FALSE}
sur_by_stratum <- covid_sur_eff |>
  split(covid_sur_eff$sex) |>
  lapply(\(df) nowcast(
    now          = sur_now,
    when         = sur_when,
    data         = df,
    dEventCol    = "dHospital",
    dReportCol   = "dReport",
    aggregate.by = "1 week",
    D            = 10,
    method       = "bayes.notrunc.bnb",
    control      = list(
      dRange     = seq(min(df$dHospital), sur_now, by = "1 week"),
      N.tInf.max = 1000,
      nSamples   = 1000
    )
  ))
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r sur-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the surveillance package", fig.alt="Three panels (Total, Female, Male) comparing the surveillance nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("surveillance")
```

## EpiNow2

[`EpiNow2`](https://epiforecasts.io/EpiNow2/) is the odd one out here: it is not
a single nowcast model but **four entry points**, each taking a different shape
of data. `tbl_now_to_EpiNow2()` therefore takes a `target` argument, named for
the function the result is passed to, so whatever it gives you can be handed over
unchanged.

::: {.alert .alert-warning}
**Warning: requires Stan.** These models fit through `cmdstanr`, so a working
CmdStan installation is needed before running the code below.
:::

| `target` | you get | for |
|---|---|---|
| `"estimate_infections"` (default) | `date` / `confirm` | `estimate_infections()`, `epinow()` |
| `"regional_epinow"` | the same plus `region` | `regional_epinow()` |
| `"estimate_truncation"` | a list of snapshots | `estimate_truncation()` |
| `"estimate_dist"` | interval-censored date columns | `estimate_dist()` |

::: {.alert .alert-warning}
**EpiNow2 models a *daily* process and has no `timestep`.** Handing it a weekly
series as one row per week is read as one row per **day** — no error, just an
epidemic seven times too fast. The converter lays non-daily data on EpiNow2's own
daily grid using its `accumulate` column, so you do not have to. Units coarser
than a week are refused rather than approximated.
:::

### Simple nowcast

`estimate_infections()` wants the series as known at the `now` — one row per day,
with the filler days marked:

```{r epinow2-convert, eval=has("EpiNow2")}
covid_en2 <- covid_now |>
  filter(notification_date >= cutoff - 180) |>
  tbl_now_to_EpiNow2(verbose = FALSE, quiet = TRUE)

head(covid_en2)
```

```{r epinow2-fit, eval=run_models && has("EpiNow2")}
library(EpiNow2)

epinow2_fit <- estimate_infections(
  covid_en2,
  generation_time = gt_opts(example_generation_time),
  delays          = delay_opts(example_incubation_period + example_reporting_delay),
  rt              = rt_opts(prior = LogNormal(mean = 2, sd = 0.1)),
  stan            = stan_opts(samples = 500, warmup = 250, chains = 2)
)
```

Two things about that call, so the numbers below are read for what they are.
**The delay distributions are EpiNow2's own shipped examples** — COVID-in-the-UK
priors — not distributions fitted to these Colombian data. That is the normal
footing for a comparison of machinery rather than of estimates, but it does mean
the EpiNow2 curve answers a slightly different question from the engines that
learn their delay from the reporting triangle in front of them. Fitting the delay
from the data instead is one `estimate_dist()` call away — see below.
**Sampling is also lighter than the default** (500 draws, 250 warmup, 2 chains),
because this is much the slowest engine here.

The predictions come out with `tidy()`, in the same shape as every other engine
here:

```{r epinow2-tidy-1, eval=FALSE}
tidy(epinow2_fit)
```

```{r epinow2-tidy, echo=FALSE}
show_tidy("EpiNow2")
```

`tidy()` reads the interval width off the fit rather than assuming it: EpiNow2's
`CrIs` is a user argument, so a model fitted with `CrIs = c(0.5, 0.95)` reports
`level = 0.95` and one fitted with the defaults reports `level = 0.9`.

### The report dimension: `estimate_truncation()`

This is the one EpiNow2 model that uses the **report** dimension a `tbl_now`
exists to carry. It takes a list of snapshots — the series as it looked at each
of several report dates — which is exactly what the object already knows:

```{r epinow2-trunc, eval=has("EpiNow2")}
covid_snapshots <- covid_now |>
  filter(notification_date >= cutoff - 180) |>
  tbl_now_to_EpiNow2(
    target = "estimate_truncation", snapshots = 5,
    verbose = FALSE, quiet = TRUE
  )

covid_snapshots
```

```{r epinow2-trunc-fit, eval=run_models && has("EpiNow2")}
truncation_fit <- estimate_truncation(covid_snapshots)
```

Because the snapshots carry their report dates, this is the one EpiNow2 shape
that can be turned **back** into a `tbl_now` — differencing consecutive snapshots
recovers the incidence:

```{r epinow2-back, eval=has("EpiNow2")}
as_tbl_now(covid_snapshots)
```

### The delay distribution: `estimate_dist()`

`estimate_dist()` (new in EpiNow2 1.9.0) estimates a reporting-delay
distribution, accounting for double interval censoring and right truncation. It
takes the same schema as \pkg{epidist}, so the two are directly comparable on
the same object:

```{r epinow2-dist, eval=has("EpiNow2")}
covid_dist <- covid_now |>
  filter(notification_date >= cutoff - 180) |>
  tbl_now_to_EpiNow2(target = "estimate_dist", verbose = FALSE, quiet = TRUE)

head(covid_dist)
```

```{r epinow2-dist-fit, eval=run_models && has("EpiNow2")}
dist_fit <- estimate_dist(covid_dist, dist = "lognormal")
```

Like `tidy.epidist_fit()`, this returns a **delay-shaped** table — one row per
parameter, plus the distribution's `mean` and `sd`:

```{r epinow2-dist-tidy-1, eval=FALSE}
tidy(dist_fit)
```

```{r epinow2-dist-tidy, echo=FALSE}
show_tidy("EpiNow2_dist")
```

### With strata and effects.

`regional_epinow()` takes a single `region` column, so the object's strata are
folded into one label:

```{r epinow2-strata, eval=has("EpiNow2")}
covid_regional <- covid_seasonal |>
  filter(notification_date >= cutoff - 180) |>
  tbl_now_to_EpiNow2(target = "regional_epinow", verbose = FALSE, quiet = TRUE)

head(covid_regional)
```

```{r epinow2-strata-fit, eval=run_models && has("EpiNow2")}
regional_fit <- regional_epinow(
  covid_regional,
  generation_time = gt_opts(example_generation_time),
  delays          = delay_opts(example_incubation_period + example_reporting_delay)
)
```

`tidy()` on the result gives one block per region, so `stratum` stays a unique
key alongside `event_date`.

::: {.alert .alert-info}
**Two shapes EpiNow2 has that `tbl.now` does not convert for.**
`estimate_secondary()` models *two* data streams against each other (cases and
deaths, say) and one `tbl_now` is one stream, so there is no honest mapping.
`estimate_delay()` takes a bare vector of delays; EpiNow2's own help now points
at `estimate_dist()` instead, and it discards the censoring a `tbl_now` carries.
If you want it anyway it is `x$.delay`.
:::

## Does this work on *your* data?

Everything above uses one daily, count-incidence dataset. `tbl.now` ships six
others with different shapes — daily and weekly, line-list and count-incidence
and count-cumulative — so the table below records what actually happens when
every converter, plus one representative nowcast, is run against each of them.
It is generated by `data-raw/converter_matrix.R`, and it reports failures as
failures.

```{r matrix-load, echo=FALSE, message=FALSE}
matrix_results <- readRDS("converter-matrix.rds")
```

```{r matrix-table, echo=FALSE}
matrix_results |>
  mutate(mark = dplyr::case_when(
    status == "ok"      ~ "ok",
    status == "timeout" ~ "slow",
    TRUE                ~ "fails"
  )) |>
  select(dataset, data_type, step, mark) |>
  tidyr::pivot_wider(names_from = step, values_from = mark, values_fill = "-") |>
  knitr::kable(caption = "Every converter against every shipped dataset")
```

```{r matrix-problems, echo=FALSE}
problems <- matrix_results |> filter(status != "ok")
if (nrow(problems) > 0) {
  problems |>
    select(dataset, data_type, step, status, message) |>
    knitr::kable(caption = "What went wrong, and why")
}
```

Two of these datasets needed fixing before they would convert at all, and both
problems are ones you will meet in your own data:

* **`mpoxdat` has a `race` column that must be declared.** Left as an undeclared
  extra column, the `(event_date, report_date)` pair is not unique — there is one
  row per race — and every target that needs a unique cell rejects it: a
  reporting triangle reported *832 duplicate pairs*, and `tsibble` refused for
  the same reason. `strata = race` fixes both. If a column genuinely is not a
  stratum, aggregate it away with `to_count()` instead.
* **`covidat` has rows registered *before* symptom onset.** A negative delay is
  not a delay, and `epidist` rejects it outright rather than guess. Filter them
  before converting.

A third needed fixing too, and it is the one most likely to catch you out:

* **FluSight's reporting weekday is not constant.** `target_end_date` is always a
  Saturday, but `as_of` lands on Saturdays *and* Wednesdays — so the delay
  between them is `0.571` weeks, not a whole number, and no reporting triangle
  can be built at all. That is an **alignment** problem masquerading as a
  cumulative-data problem. `align_weeks = TRUE` in `tbl_now()` snaps the reports
  onto a consistent weekday and the delays become integers.

**`count-cumulative` data does convert**, once its delays are whole periods.
De-accumulating a cumulative series produces *negative* increments wherever a
total was revised downward, and `baselinenowcast` ships
`preprocess_negative_values()` for exactly that — it redistributes each negative
back into earlier delays. `tbl_now_to_baselinenowcast()` calls it for you and
says so; pass `negatives = "error"` if you would rather be stopped than have the
correction applied silently.

What is left is a modelling limit rather than a conversion one:

* **`epidist` needs a delay it can censor, and counts of at least one.** It
  models an interval-censored onset-to-report pair; a cumulative series has no
  individual delays to censor, and its zero cells fail epidist's `n >= 1` check.
* **Not every model converges on every dataset.** `diseasenowcasting` reports
  *"joint fit failed to converge for all init attempts"* on the cumulative
  FluSight series. The converter did its job; the model could not fit the shape
  it was given.

The honest summary: **every converter works on `linelist`, `count-incidence` and
`count-cumulative` data** — provided the delays are whole periods and any
undeclared strata are declared. Those two data problems, not the converters, are
what you will spend your time on.
--->

## The `tidy()` function

The converters normalize what goes *into* each package. `tidy()` normalizes what
comes out. Every engine above returns something different: a matrix of
draws, an `stsNC` object, a Stan fit, an INLA summary, a bare list. The `tidy()`
turns any of them into the same table:

```{r tidy-demo, message=FALSE, warning=FALSE}
# A short window so this runs quickly; any fit from any section works the same.
recent_triangle <- covid_now |>
  filter(notification_date >= cutoff - 365 * 2) |>
  to_count(to = "count-incidence") |>
  complete_zeroes() |>
  tbl_now_to_baselinenowcast(verbose = FALSE)

bln_fit <- baselinenowcast(recent_triangle, output_type = "samples", draws = 500)

tidy(bln_fit) |> tail(4)
```

The columns are the same whatever produced the fit:

| column | meaning |
|---|---|
| `event_date` | event/reference date, **on the engine's own grid** |
| `stratum` | `"all"` when the fit is unstratified |
| `estimate` | point nowcast (posterior median where available) |
| `conf.low`, `conf.high` | interval bounds, using \pkg{broom}'s names |
| `level` | the width that interval **actually** has |
| `engine` | which package produced it |
| `qXX` | (optional) quantile columns of the prediction |

Pass `probs` for columns with other quantiles:

```{r tidy-probs, message=FALSE, warning=FALSE}
tidy(bln_fit, probs = c(0.05, 0.5, 0.95)) |> tail(3)
```

::: {.alert .alert-warning}
**Only engines that keep draws can answer an arbitrary `probs`.** That is
`diseasenowcasting`, `baselinenowcast` and `epinowcast`. `NobBS` and
`surveillance` report a fixed set of summaries, so asking them for a quantile
they never computed is an **error** rather than a silent approximation.
:::

### Asking the other engines for the quantiles you want

That error is about *when* the quantiles are computed, not about whether you can
have them. Several of these packages will happily compute any quantile — you
just have to **ask at fit time**, because once the fit returns, the draws are
gone and there is nothing left to re-summarise.

`NobBS` is the clearest example. Its `specs` argument takes a `quantiles`
element, and whatever you put there is computed alongside the default summaries:

```{r nobbs-quantiles-1, eval=FALSE}
nobbs_quantiles <- NobBS(
  data          = covid_linelist,
  now           = get_now(covid_now),
  units         = "1 day",
  onset_date    = "onset_date",
  report_date   = "report_date",
  max_D         = 10,
  moving_window = 104,
  specs         = list(quantiles = c(0.1, 0.5, 0.9))   # <- ask here
)

tail(nobbs_quantiles$estimates, 5)
```

```{r nobbs-quantiles, echo=FALSE}
comparison$display[["NobBS_quantiles"]]$estimates
```

The requested quantiles arrive as `q_0.1`, `q_0.5` and `q_0.9` columns on
`$estimates`, next to the usual `estimate` / `lower` / `upper`. `tidy()` still
returns the standard six columns for such a fit — it does not invent a schema
per engine — so join the extra columns on by date when you need them:

```{r nobbs-quantiles-join-1, eval=FALSE}
tidy(nobbs_quantiles) |>
  left_join(
    nobbs_quantiles$estimates |>
      select(event_date = onset_date, q_0.1, q_0.5, q_0.9),
    by = "event_date"
  ) |>
  tail(5)
```

```{r nobbs-quantiles-join, echo=FALSE}
qs <- comparison$display[["NobBS_quantiles"]]
if (!is.null(qs$tidy)) {
  qs$tidy |>
    dplyr::left_join(
      qs$estimates |>
        dplyr::select(event_date = onset_date, q_0.1, q_0.5, q_0.9) |>
        dplyr::mutate(event_date = as.Date(event_date)),
      by = "event_date"
    ) |>
    tail(5)
}
```

The same idea applies elsewhere: `surveillance::nowcast()` takes the interval
width through `control$alpha`. Decide what you need **before** you fit.


## Summary


```{r comparison-all-code, eval=FALSE, echo=FALSE}
window <- meta$horizon
tru <- truth[truth$event_date %in% window, ]
est <- nowcasts[nowcasts$event_date %in% window, ]
tru$stratum <- factor(tru$stratum, levels = strata_levels)
est$stratum <- factor(est$stratum, levels = strata_levels)
est$package <- factor(est$package, levels = names(engine_colours))

ggplot(mapping = aes(x = event_date)) +
  geom_col(data = tru, aes(y = observed, fill = "Reported by now"), width = 0.8) +
  geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.7) +
  # Truth last, so it stays visible where the engines lie on top of it.
  geom_line(
    data = tru, aes(y = truth, linetype = "Truth"),
    linewidth = 0.7, colour = "#262626"
  ) +
  facet_wrap(~stratum, nrow = 1, scales = "free_y") +
  scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
  scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
  scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
  labs(x = "Event date (notification)", y = "Cases",
       title = "Every engine on the same data",
       subtitle = paste0("Now = ", format(meta$now))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.box = "horizontal",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335")) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE, order = 1))
```


```{r comparison-all, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4.4, out.width="100%", fig.alt="Three panels (Total, Female, Male). Grey bars show the counts reported by now, a dark line the counts those dates eventually reached, and one coloured line per package its nowcast."}
window <- meta$horizon
tru <- truth[truth$event_date %in% window, ]
est <- nowcasts[nowcasts$event_date %in% window, ]
tru$stratum <- factor(tru$stratum, levels = strata_levels)
est$stratum <- factor(est$stratum, levels = strata_levels)
est$package <- factor(est$package, levels = names(engine_colours))

ggplot(mapping = aes(x = event_date)) +
  geom_col(data = tru, aes(y = observed, fill = "Reported by now"), width = 0.8) +
  geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.7) +
  # Truth last, so it stays visible where the engines lie on top of it.
  geom_line(
    data = tru, aes(y = truth, linetype = "Truth"),
    linewidth = 0.7, colour = "#262626"
  ) +
  facet_wrap(~stratum, nrow = 1, scales = "free_y") +
  scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
  scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
  scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
  labs(x = "Event date (notification)", y = "Cases",
       title = "Every engine on the same data",
       subtitle = paste0("Now = ", format(meta$now))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.box = "horizontal",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335")) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE, order = 1))
```


We described the dengue data once as a `tbl_now`, and then a single converter call
(or, for `diseasenowcasting`, no call at all) handed it to each package in the
shape it needed:

```{r recap, eval=FALSE}
covid_now <- tbl_now(dengue, event_date = notification_date, report_date = diagnosis_date,
                      data_type = "linelist")

covid_now                              # diseasenowcasting
tbl_now_to_baselinenowcast(covid_now)  # baselinenowcast
tbl_now_to_epinowcast(covid_now)       # epinowcast
tbl_now_to_epidist(covid_now)          # epidist
tbl_now_to_surveillance(covid_now)     # surveillance
as.data.frame(covid_now)               # NobBS or others
```

Attaching **strata** and **temporal effects** once (`covid_seasonal`) uses the
same converters: each package receives them in whatever way it can use
a grouping in `epinowcast`, covariate columns in `epidist` and the
`baselinenowcast` long format, one triangle/series per stratum where the model
takes a single series, and automatically in `diseasenowcasting`. 

```{r learning-more, echo=FALSE, results="asis"}
# The fragment is looked up by ABSOLUTE path via system.file(), never by a
# relative one. pkgdown/rmarkdown render into an intermediates directory under
# tempdir() and copy relative resources alongside; a path containing "../.."
# escapes that directory ("cannot create file '/tmp/RtmpXXX/../../man/...'"),
# which fails wherever tempdir() sits near the filesystem root -- as it does on
# CI. `system.file()` also works under pkgload::load_all(), which shims it to
# find inst/ in the source tree.
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
``` 
