---
title: "One dataset, many nowcasts: using tbl.now with seven modelling packages"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{One dataset, many nowcasts: using tbl.now with seven modelling packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The heavy model fits (Stan / JAGS / MCMC) are shown as illustrative code but
# not run when the vignette is built. Flip `run_models` to TRUE (and make sure
# the corresponding packages are installed) to actually fit them.
run_models <- FALSE

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)
```

## Why this vignette?

Preparing the same data different ways for the different nowcasting models is tedious and
error-prone. This is exactly what [`tbl.now`](https://rodrigozepeda.github.io/tbl.now/)
does for you. You can describe your data once by specifying which column is the event date, which
is the report date, what the counts mean and `tbl.now` hands it to each
modelling package in the format that package expects. In this vignette we take a
single dengue line-list and, from that one object, drive seven different
nowcasting / delay-estimation tools:

| Package | What it does | `tbl.now` bridge |
|---|---|---|
| [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) | Bayesian structural time-series nowcast | consumes a `tbl_now` **directly** |
| [baselinenowcast](https://baselinenowcast.epinowcast.org/) | fast, assumption-light baseline nowcast | `tbl_now_to_baselinenowcast()` |
| [epinowcast](https://package.epinowcast.org/) | flexible Bayesian nowcast (delay + reference modules) | `tbl_now_to_epinowcast()` |
| [epidist](https://epidist.epinowcast.org/) | estimates the reporting **delay distribution** | `tbl_now_to_epidist()` |
| [NobBS](https://cran.r-project.org/package=NobBS) | Nowcasting by Bayesian Smoothing | a plain line-list (`as.data.frame()`) |
| [surveillance](https://cran.r-project.org/package=surveillance) | the classic Höhle & an der Heiden nowcast | `tbl_now_to_surveillance()` |
| [nowcaster](https://covid19br.github.io/nowcaster/) | Bayesian nowcast fitted with INLA | `tbl_now_to_nowcaster()` |

You do not need to be an expert in any of them to follow along — the point is
how little changes on *your* side when you switch models.

## The data: one `tbl_now` to rule them all

`tbl.now` ships with `denguedat`, a weekly dengue line-list (one row per case)
with the week of symptom **onset** and the week the case was **reported**.

```{r setup, message=FALSE}
library(dplyr)
library(tbl.now)

data(denguedat)
```

```{r, echo = FALSE}
head(denguedat)
```

We build a single `tbl_now`. We only have to say which column is the event date
(`onset_week`) and which is the report date (`report_week`); `tbl.now` infers the rest (the
data is line-list, measured in weeks, and the "now" is the last report, `2010-12-20`).

```{r build-tbl-now}
# A compact recent window keeps the vignette fast to build.
dengue <- denguedat |>
  filter(onset_week >= as.Date("2009-01-01"))

dengue_now <- tbl_now(
  dengue,
  event_date  = onset_week,
  report_date = report_week,
  data_type   = "linelist"
  )

dengue_now
```

### A more complex object: strata and temporal effects

Reporting delays are rarely constant and many series are reported for several **strata** (here, the
patient's sex) that you may want to nowcast separately.
[`tbl.now`](https://rodrigozepeda.github.io/tbl.now/) lets you attach both **once**
and carries them into whichever format can hold them. Here we build a second
object, `dengue_seasonal`, that adds a `gender` stratum and a week-of-year plus
annual-Fourier **temporal (delay) effect** on top of `dengue_now`:

```{r build-seasonal, message=FALSE}
dengue_seasonal <- dengue_now |>
  add_strata(gender) |>      # nowcast Male/Female separately
  add_temporal_effects(
    temporal_effects(
      week_of_year = TRUE,   # a separate level per epidemiological week
      seasons      = 52      # an annual Fourier cycle (period = 52 weeks)
    )
  )
```

For each package below we show the **bare** conversion of `dengue_now` first, then
how the same call carries the strata and the effects when handed
`dengue_seasonal`. 

::: {.alert .alert-info}
**TIP**: Run `autoplot()`
with `panels = "delay_seasonality"` or `"delay_calendar"` first to *see* whether
such effects are present before modelling them.
:::

## 1. diseasenowcasting — no conversion at all

[`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) was
designed hand-in-hand with `tbl.now`, so it takes a `tbl_now` **directly**. You
just hand it the object:

```{r dnc, eval=run_models && has("diseasenowcasting")}
library(diseasenowcasting)

dnc_fit <- nowcast(dengue_now)

dnc_fit
```

### With strata and effects.

`diseasenowcasting` reaches into the `tbl_now` itself, so the enriched object needs no extra arguments: it picks up the `gender`
stratum and the week-of-year / seasonal effect columns automatically.

```{r dnc-teff, eval=run_models && has("diseasenowcasting")}
nowcast(dengue_seasonal)   # the strata and temporal effects are used automatically
```

## 2. baselinenowcast — a reporting triangle

[`baselinenowcast`](https://baselinenowcast.epinowcast.org/) is a 
simple, fast baseline. It works from a **reporting triangle**: a matrix with one
row per event (reference) week and one column per reporting delay. The
lower-right corner of the matrix corresponds to the not-yet-observed part the nowcast will fill in.

`tbl_now_to_baselinenowcast()` builds that triangle for you (aggregating the
line-list to weekly incidence on the way):

```{r bln-convert, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
library(baselinenowcast)

# `delays_unit` defaults to NULL and is inferred from the object's weekly units;
# pass it explicitly (e.g. delays_unit = "weeks") if the units are ambiguous.
dengue_triangle <- tbl_now_to_baselinenowcast(dengue_now, verbose = FALSE)

# rows = onset weeks, columns = delay in weeks
dengue_triangle[1:5, 1:6]
```

From here you can follow `baselinenowcast`'s own workflow. For example calling
`baselinenowcast()` to estimate the delay from the triangle, apply it, and
draw nowcast samples:

```{r bln-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One-call workflow: estimate the delay, apply it, and draw nowcast samples.
nowcast_samples <- baselinenowcast(
  dengue_triangle2,
  output_type = "samples",
  draws       = 1000
)

head(nowcast_samples)
```

### With strata and effects.

> **NOTE** This section still needs to be verified

A reporting-triangle *matrix* is a single table with
no room for strata or covariates, so `format = "matrix"` pools the strata (with a
warning) and drops the effect columns. To keep them and to nowcast each stratum
separately take the **long** format, which carries `gender` and the effect
columns, and build one triangle per stratum:

```{r bln-strata, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
dengue_long <- tbl_now_to_baselinenowcast(dengue_seasonal, format = "long", verbose = FALSE)

# List with one reporting triangle per stratum
triangles_by_stratum <- dengue_long |>
  split(dengue_long$gender) |> #
  lapply(\(df) as_reporting_triangle(df, delays_unit = "weeks"))
```

```{r bln-strata-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
#One nowcast per triangle
nowcasts_by_stratum <- triangles_by_stratum |>
  lapply(\(tri) baselinenowcast(tri, output_type = "samples", draws = 1000))
```

## 3. epinowcast — a preprocessed Bayesian model object

[`epinowcast`](https://package.epinowcast.org/) fits a flexible Bayesian model
with separate modules for the reporting delay and the reference (epidemic)
process. It expects a preprocessed object built by `enw_preprocess_data()`.

`tbl_now_to_epinowcast()` handles the preprocessing, returning
an object you can pass straight to `epinowcast::epinowcast()`:

```{r enw-convert, eval=run_models && has("epinowcast"), warning=FALSE}
library(epinowcast)

dengue_enw <- tbl_now_to_epinowcast(dengue_now, verbose = FALSE, quiet = TRUE)

dengue_enw
```

This can then be passed to `epinowcast()`: 

```{r enw-fit, eval=run_models && has("epinowcast"), message = FALSE, warning = FALSE}
# A minimal epinowcast fit from the preprocessed object
fit <- epinowcast(
  dengue_enw,
  fit = enw_fit_opts(pp = TRUE, chains = 2, iter_sampling = 500)
)
```

### With strata and effects.

Handing `tbl_now_to_epinowcast()` the enriched object
does two things automatically: the `gender` stratum becomes epinowcast's grouping
(`by`), so the model fits a delay per sex, and the temporal-effect columns land in
the `metareference` / `metareport` tables so that they can be used in a module formula.

```{r effects-enw-fit, eval=run_models && has("epinowcast")}
enw_seasonal <- tbl_now_to_epinowcast(dengue_seasonal, verbose = FALSE, quiet = TRUE)

# Drop the seasonal terms into a reference-module formula and fit as usual: the
# covariates now enter the reference model.
epinowcast(
  enw_seasonal,
  reference = enw_reference(
    parametric   = ~ 1 + .event_season_52_sin + .event_season_52_cos,
    distribution = "lognormal",
    data         = enw_seasonal
  ),
  fit = enw_fit_opts(pp = TRUE, chains = 2, iter_sampling = 500)
)
```

## 4. epidist — the reporting delay distribution

Sometimes the quantity you actually want is the **delay distribution** itself —
how long, on average, between onset and report, and how variable is it?
[`epidist`](https://epidist.epinowcast.org/) estimates exactly that, treating
each case as an interval-censored onset/report pair.

`tbl_now_to_epidist()` converts the `tbl_now` into the censored line-list
`epidist` expects. Because our data are weekly, each event is censored over its
whole week automatically:

```{r epidist-convert, eval=has("epidist"), message=FALSE, warning=FALSE}
library(epidist)

dengue_epidist <- tbl_now_to_epidist(dengue_now, verbose = FALSE)

```

```{r epidist-fit, eval=FALSE}
# Fit the delay distribution (see the epidist documentation for model choices)
delay_model <- dengue_epidist |>
  as_epidist_marginal_model() |>
  epidist()
```

The fitted delay distribution can then feed back into a nowcast for example as
a data-informed prior in `epinowcast`.

### With strata and effects.

`epidist` has no separate grouping argument, so the strata (`gender`) and the
temporal-effect columns become extra columns that can be used within the  model's 
formula. For example:

```{r epidist-strata, eval=has("epidist"), message=FALSE, warning=FALSE}
dengue_epidist_eff <- tbl_now_to_epidist(dengue_seasonal, verbose = FALSE)
```

```{r epidist-strata-fit, eval=FALSE}
# A sex-varying mean delay with the fourier terms.
delay_by_sex <- dengue_epidist_eff |>
  as_epidist_marginal_model() |>
  epidist(formula = mu ~ 1 + gender + .event_season_52_sin + .event_season_52_cos)
```

## 5. NobBS — Nowcasting by Bayesian Smoothing

[`NobBS`](https://cran.r-project.org/package=NobBS) works from a **line-list**
with an onset-date column and a report-date column — precisely what a `linelist`
`tbl_now` already is. Drop the `tbl.now` classes with `as.data.frame()` and hand
the columns to `NobBS()`:

```{r nobbs-convert, eval=has("NobBS")}
dengue_linelist <- as.data.frame(dengue_now)
```

```{r nobbs-fit, eval=run_models && has("NobBS")}
library(NobBS)

nobbs_fit <- NobBS(
  data        = dengue_linelist,
  now         = get_now(dengue_now),
  units       = "1 week",
  onset_date  = get_event_date(dengue_now),
  report_date = get_report_date(dengue_now)
)

nobbs_fit$estimates |> head()
```

Notice that even the arguments come from the `tbl_now`: `get_now()`,
`get_event_date()` and `get_report_date()` tell `NobBS` what `tbl.now` already
figured out.

### With strata and effects.

Call `NobBS.strat()` and pass the column name to the strata option:

```{r nobbs-strata, eval=run_models && has("NobBS"), warning=FALSE, message=FALSE}
dengue_nobbs     <- as.data.frame(dengue_seasonal)
stratified_nobbs <- NobBS.strat(dengue_nobbs, 
                                strata        = "gender", 
                                now         = get_now(dengue_seasonal), 
                                units       = "1 week",
                                onset_date  = get_event_date(dengue_seasonal),
                                report_date = get_report_date(dengue_seasonal),
                                #These two values are set small for the example
                                max_D         = 5, 
                                moving_window = 7
                                )
```

## 6. surveillance — the classic nowcasting line list

[`surveillance`](https://cran.r-project.org/package=surveillance) is the
long-standing R package for outbreak detection and nowcasting, and implements
the Höhle & an der Heiden (2014) approach. `surveillance::nowcast()` works from
an individual-level **line list** with one column for the event date and one for
the report date, named by its `dEventCol` / `dReportCol` arguments.

`tbl_now_to_surveillance()` builds that data frame and renames the two dates to
`surveillance`'s own defaults, so nothing has to be spelled out twice:

```{r sur-convert, eval=has("surveillance"), message=FALSE}
library(surveillance)

dengue_sur <- tbl_now_to_surveillance(dengue_now, verbose = FALSE)

head(dengue_sur)
```

The nowcast itself needs a `now`, the weeks you want estimated (`when`), and a
maximum delay `D`. All of the bookkeeping comes straight off the `tbl_now`:

```{r sur-fit, eval=run_models && has("surveillance"), message=FALSE, warning=FALSE}
sur_now  <- get_now(dengue_now)
sur_when <- seq(sur_now - 7 * 10, sur_now, by = "1 week")

sur_fit <- nowcast(
  now          = sur_now,
  when         = sur_when,
  data         = dengue_sur,
  dEventCol    = "dHospital",
  dReportCol   = "dReport",
  aggregate.by = "1 week",
  D            = 10,
  method       = "bayes.notrunc.bnb"
)

data.frame(
  date     = as.Date(epoch(sur_fit)),
  nowcast  = as.numeric(upperbound(sur_fit))
) |> tail()
```

`method` is worth a word. `"bayes.notrunc.bnb"` and `"lawless"` both run in
well under a second here. The flagship `"bayes.trunc.ddcp"` — the change-point
model from the original paper — needs **JAGS**, and `"unif"` is only a
sanity-check baseline (it assumes a uniform delay and badly over-predicts on
this data).

Passing `format = "sts"` returns the observed curve as a `surveillance` `sts`
object instead, which is what the package's plotting and outbreak-detection
verbs take:

```{r sur-sts, eval=has("surveillance"), message=FALSE}
dengue_sts <- tbl_now_to_surveillance(dengue_now, format = "sts", verbose = FALSE)
```

### With strata and effects.

`surveillance::nowcast()` has **no strata argument** — it models one series. The
strata ride along as ordinary columns, so fit one nowcast per stratum in a loop:

```{r sur-strata, eval=has("surveillance"), message=FALSE, warning=FALSE}
dengue_sur_eff <- tbl_now_to_surveillance(dengue_seasonal, verbose = FALSE)

# `gender` is kept as a column, so it can be split on.
head(dengue_sur_eff)
```

```{r sur-strata-fit, eval=run_models && has("surveillance"), message=FALSE}
sur_by_stratum <- dengue_sur_eff |>
  split(dengue_sur_eff$gender) |>
  lapply(\(df) nowcast(
    now = sur_now, when = sur_when, data = df,
    dEventCol = "dHospital", dReportCol = "dReport",
    aggregate.by = "1 week", D = 10, method = "bayes.notrunc.bnb"
  ))
```

## 7. nowcaster — an INLA nowcast

[`nowcaster`](https://covid19br.github.io/nowcaster/) fits a Bayesian
nowcast with [INLA](https://www.r-inla.org/), and like `surveillance` it starts
from a line list — here with columns it calls `date_onset` and `date_report`.

`tbl_now_to_nowcaster()` renames the dates into that vocabulary:

```{r nowcaster-convert, eval=has("nowcaster"), message=FALSE}
library(nowcaster)

dengue_nowcaster <- tbl_now_to_nowcaster(dengue_now, verbose = FALSE)

head(dengue_nowcaster)
```

One gotcha: `nowcasting_inla()` takes `date_onset` and `date_report` as **bare
column names** (tidy-evaluation), not strings — so they are passed unquoted:

```{r nowcaster-fit, eval=run_models && has("nowcaster") && has("INLA"), message=FALSE}
nowcaster_fit <- nowcasting_inla(
  dataset      = dengue_nowcaster,
  date_onset   = date_onset,
  date_report  = date_report,
  data.by.week = TRUE,   # the data is already weekly
  Dmax         = 10,
  wdw          = 30,
  silent       = TRUE
)

# `$total` holds the posterior median and 50% / 95% credible bounds.
tail(nowcaster_fit$total)
```

Note that `nowcaster` aggregates onto **its own** weekly grid, which can be
offset by a day or two from the event weeks in your data. Align on the week
before joining its output to anything else — the comparison section below does
exactly that.

### With strata and effects.

`nowcaster`'s only native grouping is by **age** (`age_col` plus `bins_age`), and
a `tbl_now` stratum is not necessarily an age band. `tbl_now_to_nowcaster()`
therefore keeps the strata as ordinary columns rather than forcing them into
`age_col`, and you fit one nowcast per stratum:

```{r nowcaster-strata, eval=has("nowcaster"), message=FALSE}
dengue_nowcaster_eff <- tbl_now_to_nowcaster(dengue_seasonal, verbose = FALSE)

head(dengue_nowcaster_eff)
```

```{r nowcaster-strata-fit, eval=run_models && has("nowcaster") && has("INLA"), message=FALSE}
nowcaster_by_stratum <- dengue_nowcaster_eff |>
  split(dengue_nowcaster_eff$gender) |>
  lapply(\(df) nowcasting_inla(
    dataset = df, date_onset = date_onset, date_report = date_report,
    data.by.week = TRUE, Dmax = 10, wdw = 30, silent = TRUE
  ))
```

## Putting it together: every nowcast on one plot

Each engine above returns its answer in its own shape — a matrix of samples, a
`stsNC` object, a Stan fit, an INLA summary. Below they are all drawn on the
same axes so you can see how they actually differ.

**About this comparison.** The `dengue_now` used through the rest of this
article runs to the very end of `denguedat`, so its most recent weeks are *still
incomplete* — there is nothing to check a nowcast against. The figures below
therefore use an earlier window of comparable length (104 weeks across
2002–2003). Reports are truncated at `now`, so every engine faces a real
nowcasting problem, while the full `denguedat` still records what those weeks
eventually reached. That gives us a genuine **truth** line.

The fits are slow (Stan, JAGS and INLA), so they are computed once by
`data-raw/nowcast_comparison.R` and read back here.

<details>
<summary>Show the plotting code</summary>

```{r comparison-setup, message=FALSE}
library(ggplot2)

comparison <- readRDS("nowcast-comparison.rds")
nowcasts   <- comparison$nowcasts
truth      <- comparison$truth
meta       <- comparison$meta

# One colour per engine (Okabe-Ito, colour-blind safe).
engine_colours <- c(
  diseasenowcasting = "#0072B2",
  baselinenowcast   = "#D55E00",
  epinowcast        = "#009E73",
  NobBS             = "#CC79A7",
  surveillance      = "#E69F00",
  nowcaster         = "#56B4E9"
)

nowcast_plot <- function(strata, title, subtitle) {
  window <- meta$horizon
  tru <- truth[truth$stratum %in% strata & truth$event_date %in% window, ]
  est <- nowcasts[nowcasts$stratum %in% strata &
                    nowcasts$event_date %in% window, ]
  est$package <- factor(est$package, levels = names(engine_colours))

  ggplot(mapping = aes(x = event_date)) +
    # What was actually on hand at `now`: the incomplete data every engine saw.
    geom_col(
      data = tru, aes(y = observed, fill = "Reported by now"),
      width = 5.5, alpha = 0.9
    ) +
    # What those weeks eventually reached.
    geom_line(
      data = tru, aes(y = truth, linetype = "Truth"),
      linewidth = 1, colour = "#262626"
    ) +
    geom_point(data = tru, aes(y = truth), size = 1.7, colour = "#262626") +
    # One coloured line per engine.
    geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.8) +
    geom_point(data = est, aes(y = estimate, colour = package), size = 1.3) +
    scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
    scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
    scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
    labs(x = "Event week (symptom onset)", y = "Cases",
         title = title, subtitle = subtitle) +
    theme_minimal(base_size = 11) +
    theme(
      legend.position = "top",
      legend.box = "horizontal",
      legend.margin = margin(b = -2),
      legend.key.width = unit(1.1, "lines"),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", colour = "#334335")
    ) +
    guides(
      colour   = guide_legend(order = 1, nrow = 2, byrow = TRUE),
      linetype = guide_legend(order = 2),
      fill     = guide_legend(order = 3)
    )
}
```

</details>

```{r comparison-plot-total, echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=4.6, out.width="100%", fig.alt="Line chart comparing each package's nowcast against the eventually-reported truth over the last eleven event weeks. Grey bars show what had been reported as of now, a dark line shows the true final counts, and one coloured line per package shows its nowcast. The lines agree on the fully-observed weeks and fan out at the right-hand edge."}
nowcast_plot(
  "Total",
  "Every engine nowcasting the same dengue data",
  paste0("Now = ", format(meta$now),
         "; grey bars are what had been reported by then")
)
```

The grey bars are the incomplete data every engine was handed; the dark line is
what those weeks eventually reached. The gap between them, at the right-hand
edge, is the whole nowcasting problem — and how well each coloured line closes
it is the only thing worth judging.

```{r comparison-plot-strata, echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=4.6, out.width="100%", fig.alt="The same nowcast comparison split into one panel per gender stratum, with one coloured line per package and a dark truth line in each panel."}
nowcast_plot(
  setdiff(unique(truth$stratum), "Total"),
  "The same comparison, stratified by gender",
  "Each engine fitted separately per stratum"
) + facet_wrap(~ stratum)
```

Two caveats on the stratified figure. `surveillance` and `nowcaster` have no
strata argument at all — as noted in their sections, they are fitted once per
stratum in a loop — and `baselinenowcast`'s matrix format pools strata, so it
too is given one triangle per stratum. Splitting the data also roughly halves
the counts in each panel, which is why the engines disagree more here than on
the total.

`epidist` does not appear in either figure: it estimates the reporting-**delay
distribution** rather than a case-count nowcast, so there is nothing to plot on
this axis.

Two honest wrinkles in how these curves were assembled, both worth knowing if
you build a comparison like this yourself:

* **Not every engine reports the same quantity.** `diseasenowcasting` exposes
  both a smoothed *latent* epidemic curve and a posterior prediction of the
  eventual counts; only the latter is comparable with the others, so
  `predict()` is used rather than `summary()$latent`. Plotting the latent curve
  instead would have shown it dipping *below* data that had already been
  reported.
* **Not every engine uses the same weeks.** `denguedat` labels a week by its
  Monday, while `nowcaster` re-bins onto its own Tuesday-start weeks, so its
  label `2003-12-02` holds the cases filed here under `2003-12-08`. Every
  engine's dates are therefore snapped onto one common grid before plotting.
  Even so, `nowcaster`'s weeks contain a slightly different set of cases than
  everyone else's, which is why it sits one or two cases below the observed
  count in a handful of weeks.

`data-raw/nowcast_comparison.R` checks for exactly this: a nowcast should never
fall below what was already reported, and any engine that does so across most
weeks is flagged as a probable alignment bug rather than quietly plotted.

## Summary

We described the dengue data once as a `tbl_now`, and then a single converter call
(or, for `diseasenowcasting`, no call at all) handed it to each package in the
shape it needed:

```{r recap, eval=FALSE}
dengue_now <- tbl_now(dengue, event_date = onset_week, report_date = report_week,
                      data_type = "linelist")

dengue_now                              # diseasenowcasting
tbl_now_to_baselinenowcast(dengue_now)  # baselinenowcast
tbl_now_to_epinowcast(dengue_now)       # epinowcast
tbl_now_to_epidist(dengue_now)          # epidist
tbl_now_to_surveillance(dengue_now)     # surveillance
tbl_now_to_nowcaster(dengue_now)        # nowcaster
as.data.frame(dengue_now)               # NobBS or others
```

Attaching **strata** and **temporal effects** once (`dengue_seasonal`) then rides
along the same converters: each package receives them in whatever way it can use —
a grouping in `epinowcast`, covariate columns in `epidist` and the
`baselinenowcast` long format, one triangle/series per stratum where the model
takes a single series, and automatically in `diseasenowcasting`. 

## See also

```{r learning-more, child = "../../man/fragments/learning-more.Rmd"}
``` 
