---
title: "One dataset, many nowcasts: using tbl.now with five modelling packages"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{One dataset, many nowcasts: using tbl.now with five modelling packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The heavy model fits (Stan / JAGS / MCMC) are shown as illustrative code but
# not run when the vignette is built. Flip `run_models` to TRUE (and make sure
# the corresponding packages are installed) to actually fit them.
run_models <- FALSE

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)
```

## Why this vignette?

Preparing the same outbreak data different ways can be tedious and
error-prone. This is exactly what [`tbl.now`](https://rodrigozepeda.github.io/tbl.now/)
does for you. You describe your data once by specifying which column is the event date, which
is the report date, what the counts mean and `tbl.now` hands it to each
modelling package in the format that package expects. In this vignette we take a
single dengue line-list and, from that one object, drive five different
nowcasting / delay-estimation tools:

| Package | What it does | `tbl.now` bridge |
|---|---|---|
| [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) | Bayesian structural time-series nowcast | consumes a `tbl_now` **directly** |
| [baselinenowcast](https://baselinenowcast.epinowcast.org/) | fast, assumption-light baseline nowcast | `tbl_now_to_baselinenowcast()` |
| [epinowcast](https://package.epinowcast.org/) | flexible Bayesian nowcast (delay + reference modules) | `tbl_now_to_epinowcast()` |
| [epidist](https://epidist.epinowcast.org/) | estimates the reporting **delay distribution** | `tbl_now_to_epidist()` |
| [NobBS](https://cran.r-project.org/package=NobBS) | Nowcasting by Bayesian Smoothing | a plain line-list (`as.data.frame()`) |

You do not need to be an expert in any of them to follow along — the point is
how little changes on *your* side when you switch models.

## The data: one `tbl_now` to rule them all

`tbl.now` ships with `denguedat`, a weekly dengue line-list (one row per case)
with the week of symptom **onset** and the week the case was **reported**.

```{r setup, message=FALSE}
library(dplyr)
library(tbl.now)

data(denguedat)
```

```{r, echo = FALSE}
head(denguedat)
```

We build a single `tbl_now`. We only have to say which column is the event date
(`onset_week`) and which is the report date (`report_week`); `tbl.now` infers the rest (the
data is line-list, measured in weeks, and the "now" is the last report, `2010-12-20`).

```{r build-tbl-now}
# A compact recent window keeps the vignette fast to build.
dengue <- denguedat |>
  filter(onset_week >= as.Date("2009-01-01"))

dengue_now <- tbl_now(
  dengue,
  event_date  = onset_week,
  report_date = report_week,
  data_type   = "linelist"
  )

dengue_now
```

### A more complex object: strata and temporal effects

Reporting delays are rarely constant and many series are reported for several **strata** (here, the
patient's sex) that you may want to nowcast separately.
[`tbl.now`](https://rodrigozepeda.github.io/tbl.now/) lets you attach both **once**
and carries them into whichever format can hold them. Here we build a second
object, `dengue_seasonal`, that adds a `gender` stratum and a week-of-year plus
annual-Fourier **temporal (delay) effect** on top of `dengue_now`:

```{r build-seasonal, message=FALSE}
dengue_seasonal <- dengue_now |>
  add_strata(gender) |>      # nowcast Male/Female separately
  add_temporal_effects(
    temporal_effects(
      week_of_year = TRUE,   # a separate level per epidemiological week
      seasons      = 52      # an annual Fourier cycle (period = 52 weeks)
    )
  )
```

For each package below we show the **bare** conversion of `dengue_now` first, then
how the same call carries the strata and the effects when handed
`dengue_seasonal`. (Tip: run
[`autoplot()`](https://rodrigozepeda.github.io/tbl.now/reference/autoplot.tbl_now.html)
with `panels = "delay_seasonality"` or `"delay_calendar"` first to *see* whether
such effects are present before modelling them.)

## 1. diseasenowcasting — no conversion at all

[`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) was
designed hand-in-hand with `tbl.now`, so it takes a `tbl_now` **directly**. You
just hand it the object:

```{r dnc, eval=run_models && has("diseasenowcasting"), message=FALSE}
library(diseasenowcasting)

dnc_fit <- nowcast(dengue_now)

dnc_fit
```

### With strata and effects.

`diseasenowcasting` reaches into the `tbl_now` itself, so the enriched object needs no extra arguments: it picks up the `gender`
stratum and the week-of-year / seasonal effect columns automatically.

```{r dnc-teff, eval=run_models && has("diseasenowcasting"), message=FALSE}
nowcast(dengue_seasonal)   # the strata and temporal effects are used automatically
```

## 2. baselinenowcast — a reporting triangle

[`baselinenowcast`](https://baselinenowcast.epinowcast.org/) is a 
simple, fast baseline. It works from a **reporting triangle**: a matrix with one
row per reference (onset) week and one column per reporting delay, where the
lower-right corner is the not-yet-observed part we want to fill in.

`tbl_now_to_baselinenowcast()` builds that triangle for you (aggregating the
line-list to weekly incidence on the way):

```{r bln-convert, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
library(baselinenowcast)

# `delays_unit` defaults to NULL and is inferred from the object's weekly units;
# pass it explicitly (e.g. delays_unit = "weeks") if the units are ambiguous.
dengue_triangle <- tbl_now_to_baselinenowcast(dengue_now, verbose = FALSE)

# rows = onset weeks, columns = delay in weeks
dengue_triangle[1:5, 1:6]
```

Because `as_reporting_triangle()` is registered as a method for `tbl_now`, alternatively you
can also write the idiomatic `baselinenowcast` call and pass the `tbl_now`
straight in:

```{r bln-method, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
dengue_triangle2 <- as_reporting_triangle(dengue_now)
```

From here you follow `baselinenowcast`'s own workflow. The one-call
`baselinenowcast()` wrapper estimates the delay from the triangle, applies it, and
draws nowcast samples in a single step:

```{r bln-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One-call workflow: estimate the delay, apply it, and draw nowcast samples.
nowcast_samples <- baselinenowcast(
  dengue_triangle2,
  output_type = "samples",
  draws       = 1000
)

head(nowcast_samples)
```

If you want the intermediate point nowcast (rather than samples), the
single-triangle helper is `estimate_and_apply_delay()`:

```{r bln-point, eval=run_models && has("baselinenowcast"), message=FALSE}
point_nowcast <- estimate_and_apply_delay(dengue_triangle2)
```

### With strata and effects.

A reporting-triangle *matrix* is a single table with
no room for strata or covariates, so `format = "matrix"` pools the strata (with a
warning) and drops the effect columns. To keep them and to nowcast each stratum
separately take the **long** format, which carries `gender` and the effect
columns, and build one triangle per stratum:

```{r bln-strata, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
dengue_long <- tbl_now_to_baselinenowcast(dengue_seasonal, format = "long", verbose = FALSE)

# List with one reporting triangle per stratum
triangles_by_stratum <- dengue_long |>
  split(dengue_long$gender) |> #
  lapply(\(df) as_reporting_triangle(df, delays_unit = "weeks"))
```

```{r bln-strata-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
#One nowcast per triangle
nowcasts_by_stratum <- triangles_by_stratum |>
  lapply(\(tri) baselinenowcast(tri, output_type = "samples", draws = 1000))
```

## 3. epinowcast — a preprocessed Bayesian model object

[`epinowcast`](https://package.epinowcast.org/) fits a flexible Bayesian model
with separate modules for the reporting delay and the reference (epidemic)
process. It expects a preprocessed object built by `enw_preprocess_data()`.

`tbl_now_to_epinowcast()` handles the preprocessing, returning
an object you can pass straight to `epinowcast::epinowcast()`:

```{r enw-convert, eval=run_models && has("epinowcast"), message=FALSE, warning=FALSE}
library(epinowcast)

dengue_enw <- tbl_now_to_epinowcast(dengue_now, verbose = FALSE, quiet = TRUE)

dengue_enw
```

```{r enw-fit, eval=run_models && has("epinowcast"), message = FALSE, warning = FALSE}
# A minimal epinowcast fit from the preprocessed object
fit <- epinowcast(
  dengue_enw,
  fit = enw_fit_opts(pp = TRUE, chains = 2, iter_sampling = 500)
)
```

### With strata and effects.

Handing `tbl_now_to_epinowcast()` the enriched object
does two things automatically: the `gender` stratum becomes epinowcast's grouping
(`by`), so the model fits a delay per sex, and the temporal-effect columns land in
the `metareference` / `metareport` tables on every completed reference date,
including the recent incomplete ones ready to be used in a module formula.

```{r effects-enw-fit, eval=run_models && has("epinowcast")}
enw_seasonal <- tbl_now_to_epinowcast(dengue_seasonal, verbose = FALSE, quiet = TRUE)

# One group per stratum:
length(unique(enw_seasonal$metareference[[1]]$.group))
#> [1] 2

# The effect columns are in the per-reference-date metadata, ready for a formula:
names(enw_seasonal$metareference[[1]])
#> [1] ... ".event_week_of_year"  ".event_season_52_cos" ".event_season_52_sin" ...

# Drop the seasonal terms into a reference-module formula and fit as usual: the
# covariates now enter the reference model.
epinowcast(
  enw_seasonal,
  reference = enw_reference(
    parametric   = ~ 1 + .event_season_52_sin + .event_season_52_cos,
    distribution = "lognormal",
    data         = enw_seasonal
  ),
  fit = enw_fit_opts(pp = TRUE, chains = 2, iter_sampling = 500)
)
```

## 4. epidist — the reporting delay distribution

Sometimes the quantity you actually want is the **delay distribution** itself —
how long, on average, between onset and report, and how variable is it?
[`epidist`](https://epidist.epinowcast.org/) estimates exactly that, treating
each case as an interval-censored onset/report pair.

`tbl_now_to_epidist()` converts the `tbl_now` into the censored line-list
`epidist` expects. Because our data are weekly, each event is censored over its
whole week automatically:

```{r epidist-convert, eval=has("epidist"), message=FALSE, warning=FALSE}
library(epidist)

dengue_epidist <- tbl_now_to_epidist(dengue_now, verbose = FALSE)

```

```{r epidist-fit, eval=FALSE}
# Fit the delay distribution (see the epidist documentation for model choices)
delay_model <- dengue_epidist |>
  as_epidist_marginal_model() |>
  epidist()
```

The fitted delay distribution can then feed back into a nowcast for example as
a data-informed prior in `epinowcast`.

### With strata and effects.

`epidist` has no separate grouping argument, so the `gender` stratum and the
temporal-effect columns simply ride along as extra columns on the censored
line-list — where an `epidist` model **formula** can use them. The mean of the
delay distribution (`mu`) can be made to vary by sex, or by season through the
Fourier terms:

```{r epidist-strata, eval=has("epidist"), message=FALSE, warning=FALSE}
dengue_epidist_eff <- tbl_now_to_epidist(dengue_seasonal, verbose = FALSE)

# the strata and effect columns are on the line-list, ready for a formula
names(dengue_epidist_eff)
```

```{r epidist-strata-fit, eval=FALSE}
# A sex-varying mean delay: `gender` enters the model for `mu`.
delay_by_sex <- dengue_epidist_eff |>
  as_epidist_marginal_model() |>
  epidist(formula = mu ~ 1 + gender)

# A season-varying mean delay uses the Fourier pair instead:
delay_seasonal <- dengue_epidist_eff |>
  as_epidist_marginal_model() |>
  epidist(formula = mu ~ 1 + .event_season_52_sin + .event_season_52_cos)
```

## 5. NobBS — Nowcasting by Bayesian Smoothing

[`NobBS`](https://cran.r-project.org/package=NobBS) works from a **line-list**
with an onset-date column and a report-date column — precisely what a `linelist`
`tbl_now` already is. Drop the `tbl.now` classes with `as.data.frame()` and hand
the columns to `NobBS()`:

```{r nobbs-convert, eval=has("NobBS")}
dengue_linelist <- as.data.frame(dengue_now)
```

```{r nobbs-fit, eval=run_models && has("NobBS")}
library(NobBS)

nobbs_fit <- NobBS(
  data        = dengue_linelist,
  now         = get_now(dengue_now),
  units       = "1 week",
  onset_date  = get_event_date(dengue_now),
  report_date = get_report_date(dengue_now)
)

nobbs_fit$estimates |> head()
```

Notice that even the arguments come from the `tbl_now`: `get_now()`,
`get_event_date()` and `get_report_date()` tell `NobBS` what `tbl.now` already
figured out.

### With strata and effects.

Call `NobBS.strat()` and pass the column name to the strata option:

```{r nobbs-strata, eval=run_models && has("NobBS"), warning=FALSE, message=FALSE}
dengue_nobbs     <- as.data.frame(dengue_seasonal)
stratified_nobbs <- NobBS.strat(dengue_nobbs, 
                                strata        = "gender", 
                                now         = get_now(dengue_seasonal), 
                                units       = "1 week",
                                onset_date  = get_event_date(dengue_seasonal),
                                report_date = get_report_date(dengue_seasonal),
                                #These two values are set small for the example
                                max_D         = 5, 
                                moving_window = 7
                                )
```

## Summary

We described the dengue data once as a `tbl_now`, and then a single converter call
(or, for `diseasenowcasting`, no call at all) handed it to each package in the
shape it needed:

```{r recap, eval=FALSE}
dengue_now <- tbl_now(dengue, event_date = onset_week, report_date = report_week,
                      data_type = "linelist")

dengue_now                              # diseasenowcasting
tbl_now_to_baselinenowcast(dengue_now)  # baselinenowcast
tbl_now_to_epinowcast(dengue_now)       # epinowcast
tbl_now_to_epidist(dengue_now)          # epidist
as.data.frame(dengue_now)               # NobBS or others
```

Attaching **strata** and **temporal effects** once (`dengue_seasonal`) then rides
along the same converters: each package receives them in whatever way it can use —
a grouping in `epinowcast`, covariate columns in `epidist` and the
`baselinenowcast` long format, one triangle/series per stratum where the model
takes a single series, and automatically in `diseasenowcasting`. 

## See also

* The main [`tbl.now` vignette](https://rodrigozepeda.github.io/tbl.now/articles/tbl.now.html)
  for the full anatomy of a `tbl_now`, data types, and temporal effects.
* [`autoplot()`](https://rodrigozepeda.github.io/tbl.now/reference/autoplot.tbl_now.html)
  for a quick diagnostic overview (delay distribution, epidemic curve, calendar
  effects, seasonality) of any `tbl_now`.

```{r learning-more, child = "../../man/fragments/learning-more.Rmd"}
``` 
