---
title: "One dataset, many nowcasts: using tbl.now with seven modelling packages"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{One dataset, many nowcasts: using tbl.now with seven modelling packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The heavy model fits (Stan / JAGS / MCMC) are shown as illustrative code but
# not run when the vignette is built. Flip `run_models` to TRUE (and make sure
# the corresponding packages are installed) to actually fit them.
run_models <- FALSE

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)
```

## Why this vignette?

Preparing the same data different ways for the different nowcasting models can be tedious and
error-prone. This is exactly what [`tbl.now`](https://rodrigozepeda.github.io/tbl.now/)
helps with. You can describe your data once by specifying which column is the event date, which
is the report date, what the counts mean and `tbl.now` hands it to each
modelling package in the format that package expects. 

> In this vignette we take a single dataset (`denguedat`), and from that one object, drive seven different nowcasting / delay-estimation tools:

| Package | What it does | Additional requirements | `tbl.now` converter |
|---|---|---|---|
| [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) | flexible Bayesian nowcast (delay + epidemic processes) | none (uses `RTMB`) | consumes a `tbl_now` **directly** |
| [baselinenowcast](https://baselinenowcast.epinowcast.org/) | fast, assumption-light baseline nowcast | none | `tbl_now_to_baselinenowcast()` |
| [epinowcast](https://package.epinowcast.org/) | flexible Bayesian nowcast (delay + reference modules) | **Stan** | `tbl_now_to_epinowcast()` |
| [epidist](https://epidist.epinowcast.org/) | estimates only the reporting **delay distribution** | **Stan** | `tbl_now_to_epidist()` |
| [NobBS](https://cran.r-project.org/package=NobBS) | Nowcasting by Bayesian Smoothing | **JAGS** | a plain line-list (`as.data.frame()`) |
| [surveillance](https://cran.r-project.org/package=surveillance) | the classic Höhle & an der Heiden nowcast | none for the method used here (**JAGS** for `bayes.trunc.ddcp`) | `tbl_now_to_surveillance()` |
| [nowcaster](https://covid19br.github.io/nowcaster/) | Bayesian nowcast fitted with INLA | **R-INLA** | `tbl_now_to_nowcaster()` |

You do not need to be an expert in any of them to follow along, the point is to show
how little changes on *your* side when you switch models.

## The data: one `tbl_now` to rule them all

`tbl.now` ships with `denguedat`, a weekly dengue line-list (one row per case)
from Puerto Rico with the week of symptom **onset** and the week the case was **reported**.

```{r setup, message=FALSE}
library(dplyr)
library(tbl.now)

data(denguedat)
```

```{r, echo = FALSE}
head(denguedat)
```

We build a single `tbl_now`. We only have to say which column is the event date
(`onset_week`) and which is the report date (`report_week`); `tbl.now` infers the rest (the
data is line-list, measured in weeks, and the "now" is the last report, `2010-12-20`).

We cut **both** dates in mid-July 2002. That is what turns a historical dataset
into a nowcasting problem: the object's `now` becomes the last
report we kept, its recent weeks are genuinely incomplete — and because the full
`denguedat` runs to 2010, we still know what those weeks eventually reached, so
every figure below can be checked against the truth.

Mid-year is a deliberate choice. Cutting at the turn of the year lands on the
holiday reporting slump, which every engine struggles with and which tells you
more about December than about the models.

The exact *week* matters too, for a reason worth knowing before you pick your
own: `nowcaster` needs the last event week to contain at least one same-week
report. Its section explains why.

```{r build-tbl-now}
cutoff <- as.Date("2002-07-22")

dengue <- denguedat |>
  filter(onset_week < cutoff, report_week < cutoff)

dengue_now <- tbl_now(
  dengue,
  event_date  = onset_week,
  report_date = report_week,
  data_type   = "linelist"
  )

dengue_now
```

That is `r format(nrow(dengue_now), big.mark = ",")` cases over
`r dplyr::n_distinct(dengue$onset_week)` weeks, from 1990 to
`r format(get_now(dengue_now))`. Everything in this article — every fit, every
figure — is built from this one object, so following along step by step
reproduces exactly what you see.

::: {.alert .alert-info}
**A note on series length.** Every conversion below runs on the full object, and
none takes more than a few seconds. The *fits* split into two groups:

* `diseasenowcasting` and `baselinenowcast` are given the **whole** series. They
  cope with it, so nothing is trimmed.
* `NobBS`, `nowcaster` and `surveillance` need the series shortened or they
  become impractical. Where the package has an argument for that we use **it**,
  rather than subsetting the data first: `moving_window` in `NobBS`, `wdw` in
  `nowcaster`, `when` in `surveillance`. The object you built stays the object
  you model.
* `epinowcast` has no such argument, because the reporting triangle is already
  built by the time you hold a preprocessed object. It is trimmed one step
  earlier, in its own section.

Each section states which it does.
:::

### A more complex object: strata and temporal effects

Reporting delays are rarely constant and many series are reported for several **strata** (here, `gender`) that you may want to nowcast separately. Here we build a second
object, `dengue_seasonal`, that adds a `gender` stratum and a week-of-year plus
annual-Fourier **temporal (delay) effect** on top of `dengue_now`:

```{r build-seasonal, message=FALSE}
dengue_seasonal <- dengue_now |>
  add_strata(gender) |>      # nowcast Male/Female separately
  add_temporal_effects(
    temporal_effects(
      week_of_year = TRUE,   # a separate level per epidemiological week
      seasons      = 52      # an annual Fourier cycle (period = 52 weeks)
    )
  )
```

For each package below we show the **bare** conversion of `dengue_now` first, then
how the same call carries the strata and the temporal effects when handed
`dengue_seasonal`. 

::: {.alert .alert-info}
**TIP**: Run `autoplot()`
with `panels = "delay_seasonality"` or `"delay_calendar"` first to *see* whether
such effects are present before modelling them.
:::

## Checking each nowcast

Every section below ends with the same figure: three panels — the **total**
series and the two **gender** strata — each showing what had been reported by
`now` (grey bars), what those weeks eventually reached (the dark **truth** line),
and that package's nowcast, with its uncertainty interval where the package
provides one.

The two stratum panels come from fitting each engine **separately on each
stratum**, which is the one approach every package here supports. Some offer a
shortcut for it — `NobBS.strat()` is shown in its own section — but the panels
are built the uniform way so the packages stay comparable.

Because the Stan, JAGS and INLA fits take minutes, they are run once by
`data-raw/nowcast_comparison.R` — which contains the same calls, with the same
arguments, that each section shows — and read back here. If you change a call in
one place, change it in the other.

```{r comparison-load, echo=FALSE, message=FALSE}
library(ggplot2)

comparison <- readRDS("nowcast-comparison.rds")
nowcasts   <- comparison$nowcasts
truth      <- comparison$truth
meta       <- comparison$meta

strata_levels  <- c("Total", "Female", "Male")
engine_colours <- c(
  diseasenowcasting = "#0072B2",
  baselinenowcast   = "#D55E00",
  epinowcast        = "#009E73",
  NobBS             = "#CC79A7",
  surveillance      = "#E69F00",
  nowcaster         = "#56B4E9"
)
```

<details>
<summary>Show the plotting code</summary>

```{r comparison-helper, message=FALSE}
# One figure per package: Total, Female and Male side by side.
nowcast_panels <- function(package, subtitle = NULL) {
  window <- meta$horizon

  tru <- truth[truth$event_date %in% window, ]
  est <- nowcasts[
    nowcasts$package == package & nowcasts$event_date %in% window,
  ]
  tru$stratum <- factor(tru$stratum, levels = strata_levels)
  est$stratum <- factor(est$stratum, levels = strata_levels)

  colour <- unname(engine_colours[package])

  ggplot(mapping = aes(x = event_date)) +
    # What had been reported by `now` -- the incomplete data the engine saw.
    geom_col(
      data = tru, aes(y = observed, fill = "Reported by now"), width = 5.5
    ) +
    # The interval, where the engine gives one.
    geom_ribbon(
      data = est[!is.na(est$lower), ],
      aes(ymin = lower, ymax = upper), fill = colour, alpha = 0.25
    ) +
    # The nowcast...
    geom_line(data = est, aes(y = estimate, colour = "Nowcast"), linewidth = 0.9) +
    geom_point(data = est, aes(y = estimate, colour = "Nowcast"), size = 1.1) +
    # ...then the truth ON TOP, so it stays visible where the two coincide
    # (over the settled weeks they lie exactly on each other).
    geom_line(
      data = tru, aes(y = truth, linetype = "Truth"),
      linewidth = 0.6, colour = "#262626"
    ) +
    facet_wrap(~stratum, nrow = 1, scales = "free_y") +
    scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
    scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
    scale_colour_manual(name = NULL, values = c("Nowcast" = colour)) +
    labs(
      x = "Event week (symptom onset)", y = "Cases",
      title = paste0(package, " nowcast vs. what actually happened"),
      subtitle = subtitle %||% paste0(
        "Now = ", format(meta$now),
        "; last ", length(window), " weeks"
      )
    ) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position  = "top",
      legend.box       = "horizontal",
      legend.margin    = margin(b = -4),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold", colour = "#334335")
    )
}
```

</details>

## diseasenowcasting

[`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) is
designed hand-in-hand with `tbl.now`, so it takes a `tbl_now` **directly**. You
just hand it the object:

### Simple nowcast

```{r dnc, eval=run_models && has("diseasenowcasting")}
library(diseasenowcasting)

# The whole series, with the package's own defaults -- no trimming.
dnc_fit <- nowcast(dengue_now)

dnc_fit
```


### With strata and effects.

`diseasenowcasting` reaches into the `tbl_now` itself, so the enriched object needs no extra arguments: it picks up the `gender` stratum and the week-of-year / seasonal effect columns automatically.

```{r dnc-teff, eval=run_models && has("diseasenowcasting")}
nowcast(dengue_seasonal)   # the strata and temporal effects are used automatically
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r dnc-tidy-1, eval=FALSE}
tidy(dnc_fit) 
```

```{r dnc-tidy, eval=run_models && has("diseasenowcasting"), echo=FALSE}
tidy(dnc_fit) |> head(5)
```

```{r dnc-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the diseasenowcasting package", fig.alt="Three panels (Total, Female, Male) comparing the diseasenowcasting nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("diseasenowcasting")
```

## baselinenowcast

[`baselinenowcast`](https://baselinenowcast.epinowcast.org/) is a 
simple, fast baseline. It works from a **reporting triangle**, a matrix with one
row per event (reference) week and one column per reporting delay. The
lower-right corner of the matrix corresponds to the not-yet-observed part the nowcast will fill in.

`tbl_now_to_baselinenowcast()` builds that triangle for you (aggregating the
line-list to weekly incidence on the way):

### Simple nowcast

```{r bln-convert, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
library(baselinenowcast)

# An event week with NO reports yet does not exist as a row, so a triangle built
# straight from the data stops short of the `now`. `complete_zeroes()` fills
# those weeks in with zeroes (it needs counts, not a line list).
dengue_counts <- dengue_now |>
  to_count(to = "count-incidence") |>
  complete_zeroes()

# `delays_unit` defaults to NULL and is inferred from the object's weekly units;
# pass it explicitly (e.g. delays_unit = "weeks") if the units are ambiguous.
dengue_triangle <- tbl_now_to_baselinenowcast(dengue_counts, verbose = FALSE)

# rows = onset weeks, columns = delay in weeks
dengue_triangle[1:5, 1:6]
```

From here you can follow `baselinenowcast`'s own workflow. For example calling
`baselinenowcast()` to estimate the delay from the triangle, apply it, and
draw nowcast samples:

```{r bln-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One-call workflow: estimate the delay, apply it, and draw nowcast samples.
nowcast_samples <- baselinenowcast(
  dengue_triangle,
  output_type = "samples",
  draws       = 1000
)

head(nowcast_samples)
```

### With strata and effects.


A reporting-triangle *matrix* is a single table with no room for strata or
covariates, so `format = "matrix"` pools the strata (with a warning) and drops
the effect columns. To nowcast each stratum separately we require a 
`format = "triangle_list"` instead which creates one reporting-triangle per stratum. 

```{r bln-strata, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
# One reporting triangle per stratum, straight from the object.
triangles_by_stratum <- tbl_now_to_baselinenowcast(
  dengue_seasonal,
  format  = "triangle_list",
  verbose = FALSE
)
```

```{r bln-strata-print, eval=has("baselinenowcast"), warning=FALSE}
triangles_by_stratum
```

This can be used as a list:

```{r bln-strata-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One nowcast per triangle
nowcasts_by_stratum <- triangles_by_stratum |>
  lapply(\(tri) baselinenowcast(tri, output_type = "samples", draws = 1000))
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r bln-tidy-1, eval=FALSE}
tidy(nowcast_samples)
```

```{r bln-tidy, eval=run_models && has("baselinenowcast"), echo=FALSE}
tidy(nowcast_samples) |> head(5)
```

```{r bln-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the baselinenowcast package", fig.alt="Three panels (Total, Female, Male) comparing the baselinenowcast nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("baselinenowcast")
```

## epinowcast

[`epinowcast`](https://package.epinowcast.org/) fits a flexible Bayesian model
with separate modules for the reporting delay and the reference (epidemic)
process. It expects a preprocessed object built by `enw_preprocess_data()`.

`tbl_now_to_epinowcast()` handles the preprocessing, returning
an object you can pass straight to `epinowcast::epinowcast()`:

**Series length: trimmed, before preprocessing.** `epinowcast` is the one engine
here with no argument for shortening the series, because by the time you hold a
preprocessed object the reporting triangle is already built. Its filters work one
step earlier, on the *observations* — and `tbl_now_to_epinowcast()` will hand you
those instead of the finished object if you ask with `preprocess = FALSE`:

::: {.alert .alert-warning}
**Warning: requires Stan.** This package fits its model with **Stan**, so it needs a working `cmdstanr` (or `rstan`) toolchain installed before any of the code below will run.
:::

### Simple nowcast

```{r enw-convert, eval=has("epinowcast"), warning=FALSE, message=FALSE}
library(epinowcast)

# 1. Convert, but stop before preprocessing.
dengue_obs <- tbl_now_to_epinowcast(
  dengue_now,
  preprocess = FALSE, verbose = FALSE, quiet = TRUE
)

# 2. Trim with epinowcast's own filter (include_days counts DAYS, not weeks).
dengue_obs_recent <- enw_filter_reference_dates(dengue_obs, include_days = 104 * 7)

# 3. Preprocess the trimmed observations.
dengue_enw_recent <- enw_preprocess_data(
  dengue_obs_recent,
  max_delay = 10,
  timestep  = "week"
)

dengue_enw_recent
```

Two years preprocesses in well under a second, and it is `dengue_enw_recent` —
not the full series — that goes to the Stan fit.

This can then be passed to `epinowcast()`: 

```{r enw-fit, eval=run_models && has("epinowcast"), message = FALSE, warning = FALSE}
# A minimal epinowcast fit from the preprocessed object
fit <- epinowcast(
  dengue_enw_recent,   # the trimmed object, not the full series
  fit = enw_fit_opts(pp = TRUE, chains = 2, iter_sampling = 500)
)
```

### With strata and effects.

Handing `tbl_now_to_epinowcast()` the enriched object
does two things automatically: the `gender` stratum becomes epinowcast's grouping
(`by`), so the model fits a delay per sex, and the temporal-effect columns land in
the `metareference` / `metareport` tables so that they can be used in a module formula.

```{r effects-enw-fit, eval=run_models && has("epinowcast")}

# Same three steps, from the enriched object.
enw_seasonal <- dengue_seasonal |>
  tbl_now_to_epinowcast(preprocess = FALSE, verbose = FALSE, quiet = TRUE) |>
  enw_filter_reference_dates(include_days = 104 * 7) |>
  enw_preprocess_data(max_delay = 10, timestep = "week")

# Drop the seasonal terms into a reference-module formula and fit as usual: the
# covariates now enter the reference model.
epinowcast(
  enw_seasonal,
  reference = enw_reference(
    parametric   = ~ 1 + .event_season_52_sin + .event_season_52_cos,
    distribution = "lognormal",
    data         = enw_seasonal
  ),
  fit = enw_fit_opts(pp = TRUE, chains = 2, iter_sampling = 500)
)
```
In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r enw-tidy-1, eval=FALSE}
tidy(fit)
```

```{r enw-tidy, eval=run_models && has("epinowcast"), echo=FALSE}
tidy(fit) |> head(5)
```

```{r enw-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the epinowcast package", fig.alt="Three panels (Total, Female, Male) comparing the epinowcast nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("epinowcast")
```

## epidist

Sometimes the quantity you actually want is the **delay distribution** itself —
how long, on average, between onset and report, and how variable is it?
[`epidist`](https://epidist.epinowcast.org/) estimates exactly that, treating
each case as an interval-censored onset/report pair.

`tbl_now_to_epidist()` converts the `tbl_now` into the censored line-list
`epidist` expects. Because our data are weekly, each event is censored over its
whole week automatically:

::: {.alert .alert-warning}
**Warning: requires Stan.** This package fits its model with **Stan**, so it needs a working `cmdstanr` (or `rstan`) toolchain installed before any of the code below will run.
:::

### Simple nowcast

```{r epidist-convert, eval=has("epidist"), message=FALSE, warning=FALSE}
library(epidist)

dengue_epidist <- tbl_now_to_epidist(dengue_now, verbose = FALSE)

```

```{r epidist-fit, eval=FALSE}
# Fit the delay distribution (see the epidist documentation for model choices)
delay_model <- dengue_epidist |>
  as_epidist_marginal_model() |>
  epidist()
```

The fitted delay distribution can then feed back into a nowcast for example as
a data-informed prior in `epinowcast`.

### With strata and effects.

`epidist` has no separate grouping argument, so the strata (`gender`) and the
temporal-effect columns become extra columns that can be used within the  model's 
formula. For example:

```{r epidist-strata, eval=has("epidist"), message=FALSE, warning=FALSE}
dengue_epidist_eff <- tbl_now_to_epidist(dengue_seasonal, verbose = FALSE)
```

```{r epidist-strata-fit, eval=FALSE}
# A sex-varying mean delay with the fourier terms.
delay_by_sex <- dengue_epidist_eff |>
  as_epidist_marginal_model() |>
  epidist(formula = mu ~ 1 + gender + .event_season_52_sin + .event_season_52_cos)
```

## NobBS

[`NobBS`](https://cran.r-project.org/package=NobBS) works from a **line-list**
with an onset-date column and a report-date column — precisely what a `linelist`
`tbl_now` already is. Drop the `tbl.now` classes with `as.data.frame()` and hand
the columns to `NobBS()`:

::: {.alert .alert-warning}
**Warning: requires JAGS.** This package fits its model with **JAGS**, which is a separate program: install JAGS itself, not just the R package, before running the code below.
:::

### Simple nowcast

```{r nobbs-convert, eval=has("NobBS")}
dengue_linelist <- as.data.frame(dengue_now)
```


```{r nobbs-fit, eval=run_models && has("NobBS")}
library(NobBS)

nobbs_fit <- NobBS(
  data          = dengue_linelist,
  now           = get_now(dengue_now),
  units         = "1 week",
  onset_date    = get_event_date(dengue_now),
  report_date   = get_report_date(dengue_now),
  max_D         = 10,   # delays beyond 10 weeks are negligible here
  moving_window = 104   # fit to the last two years; otherwise far too slow
)

nobbs_fit$estimates |> tail()
```

Notice that even the arguments come from the `tbl_now`: `get_now()`,
`get_event_date()` and `get_report_date()` tell `NobBS` what `tbl.now` already
figured out.

### With strata and effects.

Call `NobBS.strat()` and pass the column name to the strata option:

```{r nobbs-strata, eval=run_models && has("NobBS"), warning=FALSE, message=FALSE}
dengue_nobbs     <- as.data.frame(dengue_seasonal)
stratified_nobbs <- NobBS.strat(dengue_nobbs,
                                strata        = "gender",
                                now           = get_now(dengue_seasonal),
                                units         = "1 week",
                                onset_date    = get_event_date(dengue_seasonal),
                                report_date   = get_report_date(dengue_seasonal),
                                max_D         = 10,
                                moving_window = 104
                                )
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r nobbs-tidy-1, eval=FALSE}
tidy(nobbs_fit)
```

```{r nobbs-tidy, eval=run_models && has("NobBS"), echo=FALSE}
tidy(nobbs_fit) |> head(5)
```

```{r nobbs-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the NobBS package", fig.alt="Three panels (Total, Female, Male) comparing the NobBS nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("NobBS")
```

## surveillance

[`surveillance`](https://cran.r-project.org/package=surveillance) is the
long-standing R package for outbreak detection and nowcasting, and implements
the Höhle & an der Heiden (2014) approach. `surveillance::nowcast()` works from
an individual-level **line list** with one column for the event date and one for
the report date, named by its `dEventCol` / `dReportCol` arguments.

`tbl_now_to_surveillance()` builds that data frame and renames the two dates to
`surveillance`'s own defaults:

### Simple nowcast

```{r sur-convert, eval=has("surveillance"), message=FALSE}
library(surveillance)

dengue_sur <- tbl_now_to_surveillance(dengue_now, verbose = FALSE)

head(dengue_sur)
```

The nowcast itself needs a `now`, the weeks you want estimated (`when`), and a
maximum delay `D`. All of it can come from the `tbl_now`:

```{r sur-fit, eval=run_models && has("surveillance"), message=FALSE, warning=FALSE}
sur_now  <- get_now(dengue_now)
sur_when <- seq(sur_now - 7 * 20, sur_now, by = "1 week")   # weeks to estimate

# `dRange` gives surveillance the grid explicitly, so it reaches the `now` even
# though the last week has no cases at all: a line list cannot express a zero
# week (no rows), so zero-completion is no help here.
sur_range <- seq(min(dengue_sur$dHospital), sur_now, by = "1 week")

sur_fit <- nowcast(
  now          = sur_now,
  when         = sur_when,
  data         = dengue_sur,
  dEventCol    = "dHospital",
  dReportCol   = "dReport",
  aggregate.by = "1 week",
  D            = 10,
  method       = "bayes.notrunc.bnb",
  control      = list(dRange = sur_range, N.tInf.max = 1000, nSamples = 1000)
)

data.frame(
  date     = as.Date(epoch(sur_fit)),
  nowcast  = as.numeric(upperbound(sur_fit))
) |> tail()
```


### With strata and effects.

`surveillance::nowcast()` has **no strata argument** as it only models one series. The
strata ride along as ordinary columns, so fit one nowcast per stratum in a loop:

```{r sur-strata, eval=has("surveillance"), message=FALSE, warning=FALSE}
dengue_sur_eff <- tbl_now_to_surveillance(dengue_seasonal, verbose = FALSE)

# `gender` is kept as a column, so it can be split on.
head(dengue_sur_eff)
```

```{r sur-strata-fit, eval=run_models && has("surveillance"), message=FALSE}
sur_by_stratum <- dengue_sur_eff |>
  split(dengue_sur_eff$gender) |>
  lapply(\(df) nowcast(
    now = sur_now, when = sur_when, data = df,
    dEventCol = "dHospital", dReportCol = "dReport",
    aggregate.by = "1 week", D = 10, method = "bayes.notrunc.bnb",
    # Each stratum runs out earlier than the pooled series, so it needs the
    # grid spelled out too -- otherwise its curve stops before the `now`.
    control = list(
      dRange = seq(min(df$dHospital), sur_now, by = "1 week"),
      N.tInf.max = 1000, nSamples = 1000
    )
  ))
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r sur-tidy-1, eval=FALSE}
tidy(sur_fit)
```

```{r sur-tidy, eval=run_models && has("surveillance"), echo=FALSE}
tidy(sur_fit) |> head(5)
```

```{r sur-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the surveillance package", fig.alt="Three panels (Total, Female, Male) comparing the surveillance nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("surveillance")
```

## nowcaster

[`nowcaster`](https://covid19br.github.io/nowcaster/) fits a Bayesian
nowcast with [INLA](https://www.r-inla.org/), and like `surveillance` it starts
from a linelist.

`tbl_now_to_nowcaster()` renames the dates into its vocabulary:

::: {.alert .alert-warning}
**Warning: requires R-INLA.** This package fits its model with **R-INLA**, which is not on CRAN. Install it from <https://www.r-inla.org/download-install> before running the code below.
:::

### Simple nowcast

```{r nowcaster-convert, eval=has("nowcaster"), message=FALSE}
library(nowcaster)

dengue_nowcaster <- tbl_now_to_nowcaster(dengue_now, verbose = FALSE)

head(dengue_nowcaster)
```

```{r nowcaster-fit, eval=run_models && has("nowcaster") && has("INLA"), message=FALSE}
nowcaster_fit <- nowcasting_inla(
  dataset      = dengue_nowcaster,
  date_onset   = date_onset,
  date_report  = date_report,
  data.by.week = TRUE,
  use.epiweek  = FALSE,  # weeks END on the last record's weekday
  Dmax         = 20,     # how many PAST WEEKS to nowcast, not a max delay
  wdw          = 104,    # fitting window
  silent       = TRUE
)

# `$total` holds the posterior median and 50% / 95% credible bounds.
tail(nowcaster_fit$total)
```

Two of those arguments are easy to misread, so they are worth spelling out:

* **`Dmax` is the nowcast horizon**, not a maximum delay: "till how many past
  weeks the nowcasting will estimate". Setting it to 10 gives you ten estimated
  weeks, not a ten-week delay cap.
* **`use.epiweek` decides where the weeks start.** With `FALSE` (the default) a
  week *ends* on the weekday of the last record; with `TRUE` it uses the CDC
  epiweek, which starts on Sunday. Since `denguedat`'s weeks are Mondays and our
  last report is a Monday, `FALSE` puts each of our weeks in its own bin —
  `TRUE` would shift them onto Sunday boundaries that do not line up.

::: {.alert .alert-warning}
**Make sure your last event week has a same-week report.** This is the one
`nowcaster` gotcha worth internalising, because it is silent and it looks like a
broken model.

`nowcasting_inla()` takes its last observable week from the last **onset** in the
data, not from the last report:

```r
Tmax <- max(pull(data_w, date_onset))
Y <- ifelse(Time + delay > Tmax.id, NA, Y)
```

Every cell whose report falls after that onset week is marked *unobservable* —
even when the report is sitting right there in your data. So if your final event
week happens to have no same-week report, `max(onset) < max(report)` and
`nowcaster` throws away a whole diagonal of real reports. The symptom is a
nowcast that sits **below** the counts you have already observed, which is
otherwise impossible.

Our cutoff is chosen so the last event week (`2002-07-15`) does contain a
same-week report. With it, `nowcaster` sees exactly what every other engine sees
— all 105 weeks match the observed counts, and it never estimates below them.
Shift the cutoff one week earlier and it loses a diagonal and six weeks come out
low.

To check your own data: `max(event_date) == max(report_date)` after aggregating
to your time unit. If they differ, either move the cutoff or accept that
`nowcaster` is working from one period less than the rest.
:::

### With strata and effects.

`nowcaster` *can* stratify, in a single fit — but only through `age_col`, and
only on a **numeric** column it can `cut()`. Its help calls that a "stratum
column", yet handing it a character one errors inside `cut()`, and a character
`bins_age` trips over an `if (bins_age == "SI-PNI")` comparison on a vector.

`tbl_now_to_nowcaster()` does the encoding for you: it pools the strata into one
label, codes it `1..n` as `stratum_code`, and attaches the breaks that put each
level in its own bin.

```{r nowcaster-strata, eval=has("nowcaster"), message=FALSE}
dengue_nowcaster_eff <- tbl_now_to_nowcaster(dengue_seasonal, verbose = FALSE)

head(dengue_nowcaster_eff)

# The breaks and the labels come along as attributes.
attr(dengue_nowcaster_eff, "nowcaster_bins")
attr(dengue_nowcaster_eff, "nowcaster_levels")
```

One fit then covers every stratum: `$total` holds the pooled nowcast and `$age`
one row per stratum per week.

```{r nowcaster-strata-fit, eval=run_models && has("nowcaster") && has("INLA"), message=FALSE}
nowcaster_stratified <- nowcasting_inla(
  dataset      = dengue_nowcaster_eff,
  date_onset   = date_onset,
  date_report  = date_report,
  age_col      = stratum_code,
  bins_age     = attr(dengue_nowcaster_eff, "nowcaster_bins"),
  data.by.week = TRUE,
  use.epiweek  = FALSE,
  Dmax         = 20,
  wdw          = 104,
  silent       = TRUE
)

tail(nowcaster_stratified$age)
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r nowcaster-tidy-1, eval=FALSE}
tidy(nowcaster_stratified, strata_levels = attr(dengue_nowcaster_eff, "nowcaster_levels"))
```

```{r nowcaster-tidy, eval=run_models && has("nowcaster") && has("INLA"), echo=FALSE}
tidy(nowcaster_stratified, strata_levels = attr(dengue_nowcaster_eff, "nowcaster_levels")) |> head(5)
```

```{r nowcaster-panels, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the nowcaster package", fig.alt="Three panels (Total, Female, Male) comparing the nowcaster nowcast against the counts reported by now and the counts those weeks eventually reached."}
nowcast_panels("nowcaster")
```

## One shape of answer: `tidy()`

The converters normalise what goes *into* each package. `tidy()` normalises what
comes back out. Every engine above returns something different — a matrix of
draws, an `stsNC` object, a Stan fit, an INLA summary, a bare list — and `tidy()`
turns any of them into the same table:

```{r tidy-demo, message=FALSE, warning=FALSE}
# A short window so this runs quickly; any fit from any section works the same.
recent_triangle <- dengue_now |>
  filter(onset_week >= cutoff - 365 * 2) |>
  to_count(to = "count-incidence") |>
  complete_zeroes() |>
  tbl_now_to_baselinenowcast(verbose = FALSE)

bln_fit <- baselinenowcast(recent_triangle, output_type = "samples", draws = 500)

tidy(bln_fit) |> tail(4)
```

The columns are the same whatever produced the fit:

| column | meaning |
|---|---|
| `event_date` | event/reference date, **on the engine's own grid** |
| `stratum` | `"all"` when the fit is unstratified |
| `estimate` | point nowcast (posterior median where available) |
| `conf.low`, `conf.high` | interval bounds, using \pkg{broom}'s names |
| `level` | the width that interval **actually** has |
| `engine` | which package produced it |

`level` is not decoration. `epinowcast` reports a q5–q95 band by default, i.e.
**90%**, while the others report 95%; without recording it you would compare the
two as though they were the same thing.

Pass `probs` for extra quantile columns, named after the probability:

```{r tidy-probs, message=FALSE, warning=FALSE}
tidy(bln_fit, probs = c(0.05, 0.5, 0.95)) |> tail(3)
```

::: {.alert .alert-warning}
**Only engines that keep draws can answer an arbitrary `probs`.** That is
`diseasenowcasting`, `baselinenowcast` and `epinowcast`. `NobBS`, `nowcaster` and
`surveillance` report a fixed set of summaries, so asking them for a quantile
they never computed is an **error** rather than a silent approximation.
:::

Two of these packages (`NobBS` and `nowcaster`) return an *unclassed* list, so
`tidy()` identifies them by their structure. If it ever cannot tell, name the
engine yourself:

```{r tidy-engine, eval=FALSE}
tidy(nobbs_fit,     engine = "NobBS")
tidy(nowcaster_fit, engine = "nowcaster")
```

One deliberate non-goal: `tidy()` does **not** re-grid. Packages that bin onto
their own week starts keep them, because silently snapping the dates would hide a
real difference between engines — as the `nowcaster` note below explains.

## Putting it together: every nowcast on one plot

Each engine above returns its answer in its own shape — a matrix of samples, a
`stsNC` object, a Stan fit, an INLA summary. Here they are on one set of axes,
built from the very same fits the sections above plotted.

<details>
<summary>Show the plotting code</summary>

```{r comparison-all-code, eval=FALSE}
window <- meta$horizon
tru <- truth[truth$event_date %in% window, ]
est <- nowcasts[nowcasts$event_date %in% window, ]
tru$stratum <- factor(tru$stratum, levels = strata_levels)
est$stratum <- factor(est$stratum, levels = strata_levels)
est$package <- factor(est$package, levels = names(engine_colours))

ggplot(mapping = aes(x = event_date)) +
  geom_col(data = tru, aes(y = observed, fill = "Reported by now"), width = 5.5) +
  geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.7) +
  # Truth last, so it stays visible where the engines lie on top of it.
  geom_line(
    data = tru, aes(y = truth, linetype = "Truth"),
    linewidth = 0.7, colour = "#262626"
  ) +
  facet_wrap(~stratum, nrow = 1, scales = "free_y") +
  scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
  scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
  scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
  labs(x = "Event week (symptom onset)", y = "Cases",
       title = "Every engine on the same data",
       subtitle = paste0("Now = ", format(meta$now))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.box = "horizontal",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335")) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE, order = 1))
```

</details>

```{r comparison-all, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4.4, out.width="100%", fig.alt="Three panels (Total, Female, Male). Grey bars show the counts reported by now, a dark line the counts those weeks eventually reached, and one coloured line per package its nowcast."}
window <- meta$horizon
tru <- truth[truth$event_date %in% window, ]
est <- nowcasts[nowcasts$event_date %in% window, ]
tru$stratum <- factor(tru$stratum, levels = strata_levels)
est$stratum <- factor(est$stratum, levels = strata_levels)
est$package <- factor(est$package, levels = names(engine_colours))

ggplot(mapping = aes(x = event_date)) +
  geom_col(data = tru, aes(y = observed, fill = "Reported by now"), width = 5.5) +
  geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.7) +
  # Truth last, so it stays visible where the engines lie on top of it.
  geom_line(
    data = tru, aes(y = truth, linetype = "Truth"),
    linewidth = 0.7, colour = "#262626"
  ) +
  facet_wrap(~stratum, nrow = 1, scales = "free_y") +
  scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
  scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
  scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
  labs(x = "Event week (symptom onset)", y = "Cases",
       title = "Every engine on the same data",
       subtitle = paste0("Now = ", format(meta$now))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.box = "horizontal",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335")) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE, order = 1))
```

The gap between the grey bars and the dark line, at the right-hand edge, is the
whole nowcasting problem; how well each coloured line closes it is the only
thing worth judging. Note the panels use free y-axes — splitting by gender
roughly halves the counts, so the strata are harder for every engine.

Three caveats worth carrying to your own data:

* **`epidist` is absent on purpose.** It estimates the reporting-**delay
  distribution**, not a case-count nowcast, so it has nothing to draw here.
* **Not every engine reports the same quantity.** `diseasenowcasting` exposes
  both a smoothed *latent* epidemic curve and a posterior prediction of the
  eventual counts; only the latter is comparable, so `predict()` is used rather
  than `summary()$latent`.
* **Not every engine uses the same weeks.** `denguedat` labels a week by its
  Monday; `nowcaster` re-bins onto its own **Tuesday**-start weeks and offers no
  argument to change that (`use.epiweek = TRUE` moves it to Sunday, not Monday).
  Its estimates are snapped onto the common grid for these figures, but its
  weeks hold a slightly different set of cases from everyone else's, so its line
  can sit a few cases below the grey bars (at most three, in these figures). That is a binning artefact, not a
  bad nowcast: checked against **its own** aggregation, `nowcaster`'s median is
  never below what it actually observed.

`data-raw/nowcast_comparison.R` guards against the version of this that *would*
be a bug: a nowcast should not fall below what was already reported, and any
engine doing so across most weeks is flagged as a probable misalignment rather
than quietly plotted.

## Summary

We described the dengue data once as a `tbl_now`, and then a single converter call
(or, for `diseasenowcasting`, no call at all) handed it to each package in the
shape it needed:

```{r recap, eval=FALSE}
dengue_now <- tbl_now(dengue, event_date = onset_week, report_date = report_week,
                      data_type = "linelist")

dengue_now                              # diseasenowcasting
tbl_now_to_baselinenowcast(dengue_now)  # baselinenowcast
tbl_now_to_epinowcast(dengue_now)       # epinowcast
tbl_now_to_epidist(dengue_now)          # epidist
tbl_now_to_surveillance(dengue_now)     # surveillance
tbl_now_to_nowcaster(dengue_now)        # nowcaster
as.data.frame(dengue_now)               # NobBS or others
```

Attaching **strata** and **temporal effects** once (`dengue_seasonal`) then rides
along the same converters: each package receives them in whatever way it can use —
a grouping in `epinowcast`, covariate columns in `epidist` and the
`baselinenowcast` long format, one triangle/series per stratum where the model
takes a single series, and automatically in `diseasenowcasting`. 

```{r learning-more, child = "../../man/fragments/learning-more.Rmd"}
``` 
