---
title: "One dataset, many nowcasts: using tbl.now with different modelling packages"
output: rmarkdown::html_vignette
bibliography: ../references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{One dataset, many nowcasts: using tbl.now with different modelling packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# The heavy model fits (Stan / JAGS / MCMC) are shown as illustrative code but
# not run when the vignette is built: the results are read back from
# `nowcast-comparison.rds`, which is produced by running THIS FILE's own code
# (`data-raw/nowcast_models_precompute.R` purls the chunks below and runs them).
#
# Set TBL_NOW_RUN_MODELS=true to fit everything live instead. That is what the
# precompute does, which is why the two can no longer drift apart.
run_models <- isTRUE(as.logical(Sys.getenv("TBL_NOW_RUN_MODELS", "false")))

has <- function(pkg) requireNamespace(pkg, quietly = TRUE)
```

## Why this vignette?

Preparing the same data different ways for the different nowcasting models can be tedious and
error-prone. This is exactly what [`tbl.now`](https://rodrigozepeda.github.io/tbl.now/)
helps with. You can describe your data once by specifying which column is the event date, which
is the report date, whether your data is linelist or has counts (and what _those_ counts mean!) and `tbl.now` will hand it to each modelling package in the format that package expects. 

> In this vignette we take a single dataset (`covid_colombia`), and from that one object, use several different nowcasting / delay-estimation tools:

| Package | What it does | Additional requirements | `tbl.now` converter | `run_nowcast()` method |
|---|---|---|---|---|
| [diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/) | flexible Bayesian nowcast (delay + epidemic processes) | none (uses `RTMB`) | consumes a `tbl_now` **directly** | `"diseasenowcasting"` |
| [baselinenowcast](https://baselinenowcast.epinowcast.org/) | fast, assumption-light baseline nowcast | none | `tbl_now_to_baselinenowcast()` | `"baselinenowcast"` |
| [epinowcast](https://package.epinowcast.org/) | flexible Bayesian nowcast (delay + reference modules) | **Stan** | `tbl_now_to_epinowcast()` | `"epinowcast"` |
| [epidist](https://epidist.epinowcast.org/) | estimates only the reporting **delay distribution** | **Stan** | `tbl_now_to_epidist()` | — (not a nowcast) |
| [NobBS](https://cran.r-project.org/package=NobBS) | Nowcasting by Bayesian Smoothing | **JAGS** | `tbl_now_to_nobbs()` | `"NobBS"` |
| [surveillance](https://cran.r-project.org/package=surveillance) | the classic Höhle & an der Heiden nowcast | none for the method used here (**JAGS** for `bayes.trunc.ddcp`) | `tbl_now_to_surveillance()` | `"surveillance"` |
| [EpiNow2](https://epiforecasts.io/EpiNow2/) | renewal-equation $R_t$, reporting truncation, and delay distributions | **Stan** (`cmdstanr`) | `tbl_now_to_EpiNow2()` | `"EpiNow2"` |

You do not need to be an expert in any of them to follow along, the point is to show
how little changes on *your* side when you switch models.

::: {.alert .alert-info}
This article shows how to use each package "by hand": convert, fit, and evaluate with `tidy()`.
The `tbl.now` package also contains the experimental `run_nowcast()` function
which allow you to do the same three steps in one call. See the companion
article: [**One call, many models**](ensemble-nowcasting.html) to see
how to use the `run_nowcast` function for any
of the aforementioned methods and generate ensemble nowcasts. 
:::

::: {.alert .alert-warning}
**Note** This article aims only to show how to use each package for nowcasting.
We are purposefully not using the optimal models from each 
package for _this particular dataset_. Our goal is just to show
_how one uses them_. Please **DO NOT CONCLUDE WHICH PACKAGE IS BEST BASED ON 
THE RESULTS FROM THIS TUTORIAL**. 
:::

::: {.alert .alert-warning}
**Each of these is a separate package, and `tbl.now` does not install any of
them.** It only knows how to *talk* to them. Install whichever you actually want
to use. Note that some also need software outside R (the *Additional
requirements* column above) like **Stan** or **JAGS**. 

```r
# CRAN packges
install.packages(c("baselinenowcast", "EpiNow2", "NobBS", "surveillance"))

# Not on CRAN:
install.packages("cmdstanr",   repos = c('https://stan-dev.r-universe.dev', getOption("repos")))
install.packages("epidist",    repos = "https://epinowcast.r-universe.dev")
install.packages("epinowcast", repos = "https://epinowcast.r-universe.dev")

# Packages from Github:
install.packages("pak") 
pak::pkg_install("RodrigoZepeda/diseasenowcasting")
```

:::

## The data

`tbl.now` ships with `covid_colombia`, daily COVID-19 case counts from
Colombia's national surveillance system (INS) from 2020 to 2023. Each row is a
`(notification_date, diagnosis_date, sex)` combination with a case count `n`. It includes
notification date as the date the case was first notified (`event_date`), and the date the laboratory diagnosis was registered (`report_date`). The gap between the two is the reporting delay.  **OUR GOAL IS TO NOWCAST THE CASES AS THEY HAPPEN ACCORDING TO `notification_date`**.

```{r setup, message=FALSE}
library(dplyr)
library(lubridate)
library(tbl.now)

data(covid_colombia)
```

```{r, echo = FALSE, purl=FALSE}
head(covid_colombia)
```

We  will build a single `tbl_now`. To do so, we state which column is the event date
(`notification_date`), which is the report date (`diagnosis_date`), and which
 the counts (`n`); `tbl.now` infers the rest — that the grid is daily, and
that the "now" is the last report date.

For this example we will cut **both** dates at the start of April 2021 (Colombia's third wave). The example will assume we are back on that date, with only the information available until then (`now = "2021-04-01"`). No data observed after that date is kept: 

```{r build-tbl-now}
cutoff <- as.Date("2021-04-01")

#Filter to simulate being back on April 2021
covid <- covid_colombia |>
  filter(notification_date < cutoff, diagnosis_date < cutoff)

#Create the tbl_now object
covid_now <- covid |>
  tbl_now(
    event_date  = notification_date,
    report_date = diagnosis_date,
    case_count  = n,
    data_type   = "count-incidence"
  )

#We can see the tbl_now
covid_now
```

::: {.alert .alert-info}
**About the warning.** `sex` is in the data but was not declared as 
a covariate or strata, so each
`(notification_date, diagnosis_date)` cell has two rows (one per sex). 
**You can ignore the warning.** The object will be
valid, and every `tbl_now_to_*()` converter pools undeclared columns for you, so
this article's pooled fits are one series summed over `sex`, exactly as intended.
:::

That is `r format(sum(covid$n), big.mark = ",")` cases over
`r dplyr::n_distinct(covid$notification_date)` days, from March 2020 to
`r format(get_now(covid_now))`. Everything in this article — every fit, every
figure — is built from this dataset.

### A more complex object: strata and temporal effects

Reporting delays are rarely constant and many series are reported for several
**strata** (here, `sex`) that you may want to nowcast separately. Here we build a
second object, `covid_seasonal`, that declares `sex` as a stratum and adds
day-of-week plus annual-Fourier **temporal (delay) effects**. We show
how one would use them after converting to each package's format: 

```{r build-seasonal, message=FALSE}
covid_seasonal <- covid_now |> 
  add_strata(sex) |> 
  add_temporal_effects(
    temporal_effects(
      day_of_week = TRUE,    # a separate level per weekday
      seasons     = 365      # an annual Fourier cycle (period = 365 days)
    )
  )
```


::: {.alert .alert-info}
**A note on series length.** This dataset carries 2.3M cases, and what that costs
depends entirely on whether a package counts rows or reads counts:

* `diseasenowcasting` takes the **whole** series untrimmed. Nothing is dropped.
* `baselinenowcast` also keeps every day, but the **delays** are capped with
  `max_delay = 30`. That is a modelling assumption, not a workaround: the long
  tail carries under 1% of cases, and one 185-day straggler otherwise gives the
  triangle 186 columns instead of 30.
* The remaining packages (`NobBS`, `surveillance`, `epinowcast` and `epidist`)
  are trimmed before converting as the time they take to run the models is prohibitively expensive. 
:::

```{r comparison-load, echo=FALSE, message=FALSE, purl=FALSE}
library(ggplot2)

comparison <- readRDS("nowcast-comparison.rds")
nowcasts   <- comparison$nowcasts
truth      <- comparison$truth
meta       <- comparison$meta

strata_levels  <- c("Total", "Female", "Male")

# The fits are precomputed (see `data-raw/nowcast_models_precompute.R`, which
# purls and runs THIS file), so the article
# cannot call `print(fit)` or `tidy(fit)` live. It shows the REAL output of those
# calls instead: the precompute captured them and they are replayed here, so the
# code you see above each result is exactly the code that produced it.
show_fit <- function(package) {
  captured <- comparison$display[[package]]$printed
  if (is.null(captured)) return(invisible(NULL))
  cat(captured, sep = "\n")
}

show_tidy <- function(package, n = 5) {
  tidied <- comparison$display[[package]]$tidy
  if (is.null(tidied)) return(invisible(NULL))
  tail(tidied, n)
}

# The interval an engine reported, quoted in prose. COMPUTED from the same
# cached tables the section prints, never typed: a hand-written number next to a
# rebuilt figure is the drift this file's precompute exists to prevent.
#
# `width` is the last day's band; `typical` the median over the whole plotted
# window, which is the one that matters for "can you see it?" -- the band widens
# sharply at the `now` edge, so the last day flatters a narrow engine.
last_interval <- function(package) {
  tidied <- comparison$display[[package]]$tidy
  row    <- tidied[nrow(tidied), ]
  num    <- function(v) format(round(v), big.mark = ",", trim = TRUE)
  pct    <- function(v) sprintf("%.1f%%", 100 * v)

  shown <- nowcasts[
    nowcasts$package == package &
      nowcasts$stratum == "Total" &
      nowcasts$event_date %in% meta$horizon,
  ]

  list(
    date    = format(row$event_date),
    point   = num(row$estimate),
    lower   = num(row$conf.low),
    upper   = num(row$conf.high),
    width   = pct((row$conf.high - row$conf.low) / row$estimate),
    typical = pct(stats::median(
      (shown$upper - shown$lower) / shown$estimate, na.rm = TRUE
    ))
  )
}
# EVERY package that appears in `nowcasts` needs an entry. The Summary figure
# turns `package` into a factor with these levels, so a package missing here
# becomes an `NA` level: a grey unnamed line in the plot and an "NA" key in the
# legend, with no error to say why.
engine_colours <- c(
  diseasenowcasting = "#0072B2",
  baselinenowcast   = "#D55E00",
  epinowcast        = "#009E73",
  NobBS             = "#CC79A7",
  surveillance      = "#E69F00",
  EpiNow2           = "#56B4E9"
)
```


```{r comparison-helper, message=FALSE, echo=FALSE, purl=FALSE}
# One figure per package: Total, Female and Male side by side.
nowcast_panels <- function(package, subtitle = NULL) {
  window <- meta$horizon

  tru <- truth[truth$event_date %in% window, ]
  est <- nowcasts[
    nowcasts$package == package & nowcasts$event_date %in% window,
  ]
  tru$stratum <- factor(tru$stratum, levels = strata_levels)
  est$stratum <- factor(est$stratum, levels = strata_levels)

  colour <- unname(engine_colours[package])

  ggplot(mapping = aes(x = event_date)) +
    # What had been reported by `now` -- the incomplete data the engine saw.
    # width = 0.8 (one day). A wider bar on a DAILY series overlaps its
    # neighbours, and ggplot2 then stacks the overlaps -- so the grey bars would
    # show sums of several days rather than each day's count.
    geom_col(
      data = tru, aes(y = observed, fill = "Reported by now"), width = 0.8
    ) +
    # The interval, where the engine gives one.
    geom_ribbon(
      data = est[!is.na(est$lower), ],
      aes(ymin = lower, ymax = upper), fill = colour, alpha = 0.25
    ) +
    # The nowcast...
    geom_line(data = est, aes(y = estimate, colour = "Nowcast"), linewidth = 0.9) +
    geom_point(data = est, aes(y = estimate, colour = "Nowcast"), size = 1.1) +
    # ...then the truth ON TOP, so it stays visible where the two coincide
    # (over the settled days they lie exactly on each other).
    geom_line(
      data = tru, aes(y = truth, linetype = "Truth"),
      linewidth = 0.6, colour = "#262626"
    ) +
    facet_wrap(~stratum, nrow = 1, scales = "free_y") +
    scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
    scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
    scale_colour_manual(name = NULL, values = c("Nowcast" = colour)) +
    labs(
      x = "Event date (notification)", y = "Cases",
      title = paste0(package, " nowcast vs. what actually happened"),
      subtitle = subtitle %||% paste0(
        "Now = ", format(meta$now),
        "; last ", length(window), " days"
      )
    ) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position  = "top",
      legend.box       = "horizontal",
      legend.margin    = margin(b = -4),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold", colour = "#334335")
    )
}
```


## diseasenowcasting

[`diseasenowcasting`](https://rodrigozepeda.github.io/diseasenowcasting/) is
designed hand-in-hand with `tbl.now`, so it takes a `tbl_now` **directly**. You
just hand it the object:

### Simple nowcast

```{r dnc, eval=run_models && has("diseasenowcasting")}
# TODO: When the diseasenowcasting 2.3.0 ships without the tidy()
# remove the :: and use library
dnc_fit <- diseasenowcasting::nowcast(covid_now)

dnc_fit
```

```{r dnc-out, echo=FALSE, purl=FALSE, eval=!run_models}
show_fit("diseasenowcasting")
```

Predictions can be obtained via `tidy`:

```{r dnc-tidy-1, eval=FALSE}
tidy(dnc_fit)
```

```{r dnc-tidy, echo=FALSE, purl=FALSE}
show_tidy("diseasenowcasting")
```

### With strata and effects.

`diseasenowcasting` reaches into the `tbl_now` itself, so the enriched object needs no extra arguments: it picks up the `sex` stratum and the day-of-week / seasonal effect columns automatically.

```{r dnc-teff, eval=run_models && has("diseasenowcasting")}
dnc_seasonal <- diseasenowcasting::nowcast(covid_seasonal)   # strata and effects used automatically
```

```{r dnc-teff-out, echo=FALSE, purl=FALSE, eval=!run_models}
show_fit("diseasenowcasting_seasonal")
```

`tidy()` can also be used in stratified cases: 

```{r dnc-teff-tidy-1, eval=FALSE}
tidy(dnc_seasonal)
```

```{r dnc-teff-tidy, echo=FALSE, purl=FALSE}
show_tidy("diseasenowcasting_seasonal")
```


```{r dnc-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the diseasenowcasting package", fig.alt="Three panels (Total, Female, Male) comparing the diseasenowcasting nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("diseasenowcasting")
```

## baselinenowcast

[`baselinenowcast`](https://baselinenowcast.epinowcast.org/) is a 
simple, fast baseline. It works from a **reporting triangle**, a matrix with one
row per event (reference) date and one column per reporting delay. The
lower-right corner of the matrix corresponds to the not-yet-observed part the nowcast will fill in.

`tbl_now_to_baselinenowcast()` builds that triangle directly from the
count-incidence data:

### Simple nowcast

```{r bln-convert, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
library(baselinenowcast)

# Cap the delay axis: a single 185-day straggler gives the
# triangle 186 columns, almost all of them empty.
covid_triangle <- covid_now |>
  tbl_now_to_baselinenowcast(max_delay = 30, verbose = FALSE)

# rows = notification dates, columns = delay in days
covid_triangle[1:5, 1:6]
```

::: {.alert .alert-info}
**Why cap the delays?**  The reporting
triangle keeps every one of the `r nrow(covid_triangle)` event dates. 
This makes the fit very slow for a tail that carries well under 1% of cases. 
:::

From here you can follow `baselinenowcast`'s own workflow. For example calling
`baselinenowcast()` to estimate the delay from the triangle, apply it, and
draw nowcast samples:


```{r bln-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One-call workflow: estimate the delay, apply it, and draw nowcast samples.
nowcast_samples <- baselinenowcast(
  covid_triangle,
  output_type = "samples",
  draws       = 1000
)

```

### With strata and effects.

To nowcast each stratum we need to specify a
`format = "triangle_list"`. This creates one reporting-triangle per stratum and
returns them as a list.

```{r bln-strata, eval=has("baselinenowcast"), message=FALSE, warning=FALSE}
# One reporting triangle per stratum, straight from the object.
triangles_by_stratum <- tbl_now_to_baselinenowcast(
  covid_seasonal,
  max_delay = 30, 
  format    = "triangle_list",
  verbose   = FALSE
)
```

```{r bln-strata-print, eval=has("baselinenowcast"), warning=FALSE}
triangles_by_stratum
```

This can be used to nowcast each stratum via `lapply`:

```{r bln-strata-fit, eval=run_models && has("baselinenowcast"), message=FALSE}
# One nowcast per triangle, each trimmed the same way.
nowcasts_by_stratum <- triangles_by_stratum |>
  lapply(\(tri) baselinenowcast(tri, output_type = "samples", draws = 1000))
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r bln-tidy-1, eval=FALSE}
tidy(nowcast_samples)
```

```{r bln-tidy, echo=FALSE, purl=FALSE}
show_tidy("baselinenowcast")
```

```{r bln-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the baselinenowcast package", fig.alt="Three panels (Total, Female, Male) comparing the baselinenowcast nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("baselinenowcast")
```

## epinowcast

[`epinowcast`](https://package.epinowcast.org/) fits a flexible Bayesian model
with separate modules for the reporting delay and the reference (epidemic)
process. It expects a preprocessed object built by `enw_preprocess_data()`.

`tbl_now_to_epinowcast()` handles the preprocessing, returning
an object you can pass straight to `epinowcast::epinowcast()`:

::: {.alert .alert-warning}
**Warning: requires Stan.** This package fits its model with **Stan**, so it needs a working `cmdstanr` (or `rstan`) toolchain installed before any of the code below will run.
:::

### Simple nowcast

`epinowcast` becomes very slow with the 2.3M cases reporting triangle. So we need to filter earlier, 
on the *observations* before applying `tbl_now_to_epinowcast()`.

```{r enw-convert, eval=has("epinowcast"), warning=FALSE, message=FALSE}
library(epinowcast)

# Trim, then convert. 180 days, not two years otherwise it takes too long to fit.
covid_enw_recent <- covid_now |>
  filter(notification_date >= cutoff - 180) |>
  tbl_now_to_epinowcast(max_delay = 30, verbose = FALSE, quiet = TRUE)

covid_enw_recent
```

This can then be passed to `epinowcast()`: 

```{r enw-fit, eval=run_models && has("epinowcast"), message = FALSE, warning = FALSE}
# A minimal epinowcast fit from the preprocessed object
enw_fit <- epinowcast(
  covid_enw_recent,
  fit = enw_fit_opts(
    pp = TRUE, chains = 2, iter_sampling = 500, iter_warmup = 500,
    seed = 20260824
  )
)
```

Again, the `max_delay` is a **modelling choice, not a detail**: left unset, the
converter infers it from the longest delay present (330 days here), and the
nowcast then carries one reference date per delay. This also becomes extremely slow. 


### With strata and effects.

Handing `tbl_now_to_epinowcast()` the enriched object
does two things automatically: the `sex` stratum becomes epinowcast's grouping
(`by`), so the model fits a delay per sex, and the temporal-effect columns land in
the `metareference` / `metareport` tables so that they can be used in a module formula.

```{r effects-enw-fit, eval=run_models && has("epinowcast")}
# Same trim, from the enriched object.
enw_seasonal <- covid_seasonal |>
  filter(notification_date >= cutoff - 180) |>
  tbl_now_to_epinowcast(max_delay = 30, verbose = FALSE, quiet = TRUE)

# Drop the seasonal terms into a reference-module formula and fit as usual: the
# covariates now enter the reference model. Thet have the same names as in 
# `covid_seasonal` when using `compute_temporal_effects`. 
enw_seasonal_fit <- epinowcast(
  enw_seasonal,
  reference = enw_reference(
    parametric   = ~ 1 + .event_season_365_sin + .event_season_365_cos,
    distribution = "lognormal",
    data         = enw_seasonal
  ),
  fit = enw_fit_opts(
    pp = TRUE, chains = 2, iter_sampling = 500, iter_warmup = 500,
    seed = 20260824
  )
)
```

::: {.alert .alert-info}
The Fourier terms are weakly identified in this model, so read the seasonal
coefficients as a demonstration of *how* to put temporal effects into a module
formula rather than as a finding about COVID-19 in Colombia. 
:::

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r enw-tidy-1, eval=FALSE}
tidy(enw_fit)
```

```{r enw-tidy, echo=FALSE, purl=FALSE}
show_tidy("epinowcast")
```

```{r enw-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the epinowcast package", fig.alt="Three panels (Total, Female, Male) comparing the epinowcast nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("epinowcast")
```

## NobBS

[`NobBS`](https://cran.r-project.org/package=NobBS) works from a **linelist**
with an onset-date column and a report-date column, and it counts **rows**. 
Each row is one case. Our data are counts, so they have to be expanded first.
`tbl_now_to_nobbs()` does that and names the columns what `NobBS()` expects:

::: {.alert .alert-warning}
**Warning: requires JAGS.** This package fits its model with **JAGS**, which is a separate program: install JAGS itself, not just the R package, before running the code below.
:::

### Simple nowcast

```{r nobbs-convert, eval=has("NobBS")}
# Trim BEFORE converting: the expansion is one row per case, and `moving_window`
# below limits only what NobBS fits, not what it is handed.
covid_linelist <- covid_now |>
  filter(notification_date >= cutoff - 60) |>
  tbl_now_to_nobbs(verbose = FALSE)

nrow(covid_linelist)   # one row per case
```

::: {.alert .alert-warning}
**Do not hand `NobBS()` count data directly.** It counts rows, so a table of
1,174 rows carrying 50,160 cases is nowcast as **1,174 cases**!. 
The converter exists to make that impossible.
:::


```{r nobbs-fit, eval=run_models && has("NobBS")}
library(NobBS)

nobbs_fit <- NobBS(
  data          = covid_linelist,
  now           = get_now(covid_now),
  units         = "1 day",
  onset_date    = "onset_date",
  report_date   = "report_date",
  max_D         = 30,   # delays beyond 30 days are negligible here
  moving_window = 60    # ...and fit the last 60 days, which is all we handed it
)

```

The predictions come out with `tidy()`, in the same shape as every
other engine here:

```{r nobbs-tidy-1, eval=FALSE}
tidy(nobbs_fit)
```

```{r nobbs-tidy, echo=FALSE, purl=FALSE}
show_tidy("NobBS")
```


Notice that even the arguments come from the `tbl_now`: `get_now()`,
`get_event_date()` and `get_report_date()` tell `NobBS` what `tbl.now` already
figured out.

### With strata and effects.

`NobBS.strat()` fits one nowcast per stratum, and its `strata` argument names
**one** column. `tbl_now_to_nobbs()` therefore adds a `strata` column holding
every declared stratum pasted together under the name `strata`. Use
`"strata"`  as the name for `NobBS.strat`:

```{r nobbs-strata, eval=run_models && has("NobBS"), warning=FALSE, message=FALSE}
covid_linelist_sex <- covid_seasonal |>
  filter(notification_date >= cutoff - 60) |>
  tbl_now_to_nobbs(verbose = FALSE)   # `sex` rides along, plus a `strata` column

stratified_nobbs <- NobBS.strat(covid_linelist_sex,
                                strata        = "strata",
                                now           = get_now(covid_seasonal),
                                units         = "1 day",
                                onset_date    = "onset_date",
                                report_date   = "report_date",
                                max_D         = 30,
                                moving_window = 60
                                )
```

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r nobbs-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the NobBS package", fig.alt="Three panels (Total, Female, Male) comparing the NobBS nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("NobBS")
```

```{r nobbs-interval, echo=FALSE, purl=FALSE}
nobbs_ci <- last_interval("NobBS")
```

::: {.alert .alert-warning}
**There is a credible interval in that figure — it is just too narrow to see.**
Every panel in this article draws the engine's interval as a shaded band, and
`NobBS` does return one. The `tidy()` table above gives
**`r nobbs_ci$point` [`r nobbs_ci$lower`, `r nobbs_ci$upper`]** for
`r nobbs_ci$date` — a band `r nobbs_ci$width` of the estimate wide, and that is
its *widest* day. Across the thirty days plotted the median width is
**`r nobbs_ci$typical`**, which is thinner than the line drawn on top of it.

Nothing was lost in translation — the model is simply that confident. At roughly
ten thousand cases a day a Poisson-like posterior puts its interval near
$\pm\sqrt{n}$, and `NobBS` estimates the delay distribution from a great deal of
data. **Read these bounds off `tidy()`, not off the figure**, and compare them
with `epinowcast` above and `EpiNow2` below, whose delay carries far more
uncertainty and whose bands show it.
:::

## surveillance

[`surveillance`](https://cran.r-project.org/package=surveillance) is the
long-standing R package for outbreak detection and nowcasting, and implements
the Höhle & an der Heiden (2014) approach. `surveillance::nowcast()` works from
an individual-level **line list** with one column for the event date and one for
the report date, named by its `dEventCol` / `dReportCol` arguments.

`tbl_now_to_surveillance()` builds that data frame and renames the two dates to
`surveillance`'s own defaults:

### Simple nowcast

```{r sur-convert, eval=has("surveillance"), message=FALSE}
library(surveillance)

# Trim before converting: the converter expands counts to one row per case.
covid_sur_now <- covid_now |>
  filter(notification_date >= cutoff - 60)

covid_sur <- tbl_now_to_surveillance(covid_sur_now, verbose = FALSE)

head(covid_sur)
```

`surveillance::nowcast()` requires you to pass two date **grids**. Use `get_surveillance_when()` to get the dates you want estimated, and `get_surveillance_range()` for the whole axis the model is laid on.
The nowcast itself needs a `now`, the dates you want estimated (`when`), and a
maximum delay `D`. All of it can come from the `tbl_now`:

```{r sur-fit, eval=run_models && has("surveillance"), message=FALSE, warning=FALSE}
#Note that in the call we use our get_now(), get_surveillance_when()
#and get_surveillance_range() functions to get those variables from the 
#tbl.now
sur_fit <- nowcast(
  now          = get_now(covid_sur_now),
  when         = get_surveillance_when(covid_sur_now, length = 30),
  data         = covid_sur,
  dEventCol    = "dHospital",
  dReportCol   = "dReport",
  aggregate.by = "1 day",
  D            = 30,
  method       = "bayes.notrunc.bnb",
  control      = list(dRange = get_surveillance_range(covid_sur_now), 
                      N.tInf.max = 100000, nSamples = 1000)
)

```

The predictions come out with `tidy()`, in the same shape as every
other engine here:

```{r sur-tidy-1, eval=FALSE}
tidy(sur_fit)
```

```{r sur-tidy, echo=FALSE, purl=FALSE}
show_tidy("surveillance")
```


### With strata and effects.

`surveillance::nowcast()` has **no strata argument** as it only models one
series, so a stratified analysis means one fit per stratum. Ask the converter
for `format = "linelist_list"` and it does the splitting, exactly as
`format = "triangle_list"` does for `baselinenowcast`: one line list per stratum,
in a plain list you can `lapply()` over.

```{r sur-strata, eval=has("surveillance"), message=FALSE, warning=FALSE}
# Trim first, exactly as above: the whole window is 2.3M cases and the converter
# expands every one of them to a row.
covid_sur_seasonal <- covid_seasonal |>
  filter(notification_date >= cutoff - 60)

covid_sur_eff <- tbl_now_to_surveillance(
  covid_sur_seasonal,
  format  = "linelist_list",
  verbose = FALSE
)
```

```{r sur-strata-print, eval=has("surveillance"), warning=FALSE}
covid_sur_eff
```

With no strata declared this is still a list, of length one named `"all"`, so the
`lapply()` below does not have to know which case it is in. The default
`format = "linelist"` gives the same information as one frame with a pasted
`strata` column, if you would rather split it yourself.

```{r sur-strata-fit, eval=run_models && has("surveillance"), message=FALSE}
sur_by_stratum <- covid_sur_eff |>
  lapply(\(df) nowcast(
    now          = get_now(covid_sur_seasonal),
    when         = get_surveillance_when(covid_sur_seasonal, length = 30),
    data         = df,
    dEventCol    = "dHospital",
    dReportCol   = "dReport",
    aggregate.by = "1 day",
    D            = 30,
    method       = "bayes.notrunc.bnb",
    control      = list(
      # The grid comes from the object, not from the piece: every stratum has to
      # be laid on the SAME axis, or a stratum whose first case arrived late
      # starts its own time at a different day.
      dRange     = get_surveillance_range(covid_sur_seasonal),
      # `N.tInf.max` caps the support of the nowcast distribution, so it has to
      # sit comfortably above the largest daily count in the stratum -- here
      # about 4,000. Too low and the posterior is silently truncated.
      N.tInf.max = 100000,
      nSamples   = 1000
    )
  ))
```

`tidy()` recognises the list and labels each block with its own stratum, so
`sur_by_stratum` tidies into one table exactly like the natively stratified
engines.

In both stratified and unstratified cases the predictions can be recovered with `tidy`:

```{r sur-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the surveillance package", fig.alt="Three panels (Total, Female, Male) comparing the surveillance nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("surveillance")
```

```{r sur-interval, echo=FALSE, purl=FALSE}
sur_ci <- last_interval("surveillance")
```

::: {.alert .alert-warning}
**As with `NobBS`, the prediction interval is drawn but invisible.** `tidy()`
reads it from the `stsNC` object's `pi` slot at the width `control$alpha` sets
(95% by default), and for `r sur_ci$date` it is
**`r sur_ci$point` [`r sur_ci$lower`, `r sur_ci$upper`]** — `r sur_ci$width` of
the estimate, against a median of `r sur_ci$typical` over the thirty days
plotted. You do **not** need the JAGS-backed `bayes.trunc` methods to get
uncertainty out of `surveillance`; `bayes.notrunc.bnb` above reports it. (The
`lawless` and `unif` methods may leave the slot empty, and then the bounds come
back `NA`.)
:::

## EpiNow2

::: {.alert .alert-danger}
EpiNow2's interface from `tbl.now` is still experimental we haven't checked it yet
:::

[`EpiNow2`](https://epiforecasts.io/EpiNow2/) is the odd one out here: it is not
a single nowcast model but **four entry points**, each taking a different shape
of data. `tbl_now_to_EpiNow2()` therefore takes a `target` argument, named for
the function the result is passed to, so whatever it gives you can be handed over
unchanged.

::: {.alert .alert-warning}
**Warning: requires Stan.** These models fit through `cmdstanr`, so a working
CmdStan installation is needed before running the code below.
:::

| `target` | you get | for |
|---|---|---|
| `"estimate_infections"` (default) | `date` / `confirm` | `estimate_infections()`, `epinow()` |
| `"regional_epinow"` | the same plus `region` | `regional_epinow()` |
| `"estimate_truncation"` | a list of snapshots | `estimate_truncation()` |
| `"estimate_dist"` | interval-censored date columns | `estimate_dist()` |

::: {.alert .alert-warning}
**EpiNow2 models a *daily* process and has no `timestep`.** Handing it a weekly
series as one row per week is read as one row per **day** — no error, just an
epidemic seven times too fast. The converter lays non-daily data on EpiNow2's own
daily grid using its `accumulate` column, so you do not have to. Units coarser
than a week are refused rather than approximated.
:::

### Simple nowcast

`estimate_infections()` wants the series as known at the `now` — one row per day,
with the filler days marked:

```{r epinow2-convert, eval=has("EpiNow2")}
covid_en2 <- covid_now |>
  filter(notification_date >= cutoff - 60) |>
  tbl_now_to_EpiNow2(verbose = FALSE, quiet = TRUE)

head(covid_en2)
```

EpiNow2 needs **two** things fitted before it can nowcast, and they are not the
same thing:

* **`delays`** convolves infections into reports. The infection-to-onset part of
  it is not in our data at all, so the incubation period stays EpiNow2's shipped
  example.
* **`truncation`** is the right-truncation correction — *the nowcast itself*.
  That **is** what the report dimension of a `tbl_now` measures, and
  `estimate_truncation()` fits it from snapshots of the series as it looked at
  successive report dates. This is the one EpiNow2 model that uses the report
  axis, which is why `tbl_now_to_EpiNow2()` has a target for it.

Its default is `trunc_opts()` = `Fixed(0)`: **no truncation, so no nowcast**.
So this is a two-step fit.

```{r epinow2-fit, eval=run_models && has("EpiNow2")}
library(EpiNow2)

# STEP 1 --- fit the truncation from the report dimension.
covid_snaps <- covid_now |>
  filter(notification_date >= cutoff - 60) |>
  tbl_now_to_EpiNow2(
    target = "estimate_truncation", snapshots = 5, verbose = FALSE, quiet = TRUE
  )

truncation_fit <- estimate_truncation(
  covid_snaps,
  stan = stan_opts(samples = 500, warmup = 500, chains = 2)
)

# `$dist` is defunct; the accessor is `get_parameters()`.
fitted_truncation <- get_parameters(truncation_fit)[["truncation"]]

# STEP 2 --- nowcast with it.
epinow2_fit <- estimate_infections(
  covid_en2,
  generation_time = gt_opts(example_generation_time),
  delays          = delay_opts(example_incubation_period),
  truncation      = trunc_opts(fitted_truncation),
  rt              = rt_opts(prior = LogNormal(mean = 2, sd = 0.1)),
  stan            = stan_opts(samples = 500, warmup = 250, chains = 2)
)
```

::: {.alert .alert-warning}
**Without step 1, EpiNow2 does not nowcast at all.** Given only a reporting
delay — whether EpiNow2's shipped UK one or one fitted from these data with
`estimate_dist()` — its median stayed roughly flat over the last two weeks and
sat *below* what had already been reported. `delays` tells the model how
infections turn into reports; it does not tell it that the newest days are
**incomplete**. Only `truncation` does.

With it, over the last seven days — where these data are about **50% complete** —
the fit sits below the already-reported count on 4 of 21 stratum-days instead of
most of them, and the last day is nowcast at 8,090 against 4,261 reported and
12,521 eventual.

It still dips below the observed value on about half of the *older* days, and
that is not a nowcasting failure: those days are ~92% complete, and a model that
fits a smooth infection curve will fall below a noisy daily count roughly half
the time. That is EpiNow2 doing what it is for.
:::

**The generation time and incubation period are still EpiNow2's shipped
examples**, and deliberately so: both are properties of transmission rather than
of reporting, and no amount of reporting data identifies them. What the report
dimension *does* identify is the truncation, and that is what step 1 fits.
**Sampling is also lighter than the default** (500 draws, 250 warmup, 2 chains),
because this is much the slowest engine here.

The predictions come out with `tidy()`, in the same shape as every other engine
here:

```{r epinow2-tidy-1, eval=FALSE}
tidy(epinow2_fit)
```

```{r epinow2-tidy, echo=FALSE, purl=FALSE}
show_tidy("EpiNow2")
```

`tidy()` reads the interval width off the fit rather than assuming it: EpiNow2's
`CrIs` is a user argument, so a model fitted with `CrIs = c(0.5, 0.95)` reports
`level = 0.95` and one fitted with the defaults reports `level = 0.9`.

### The report dimension: `estimate_truncation()`

This is step 1 above, looked at on its own — the one EpiNow2 model that uses the
**report** dimension a `tbl_now` exists to carry. It takes a list of snapshots —
the series as it looked at each of several report dates — which is exactly what
the object already knows:

```{r epinow2-trunc, eval=has("EpiNow2")}
covid_snapshots <- covid_now |>
  filter(notification_date >= cutoff - 60) |>
  tbl_now_to_EpiNow2(
    target = "estimate_truncation", snapshots = 5,
    verbose = FALSE, quiet = TRUE
  )

covid_snapshots
```

Because the snapshots carry their report dates, this is the one EpiNow2 shape
that can be turned **back** into a `tbl_now` — differencing consecutive snapshots
recovers the incidence:

```{r epinow2-back, eval=has("EpiNow2")}
as_tbl_now(covid_snapshots)
```

### The delay distribution: `estimate_dist()`

`estimate_dist()` (new in EpiNow2 1.9.0) estimates a reporting-delay
distribution, accounting for double interval
censoring and right truncation. It takes the same schema as \pkg{epidist}, so
the two are directly comparable on the same object:

```{r epinow2-dist, eval=has("EpiNow2")}
covid_dist <- covid_now |>
  filter(notification_date >= cutoff - 60) |>
  tbl_now_to_EpiNow2(target = "estimate_dist", verbose = FALSE, quiet = TRUE)

head(covid_dist)
```

Unlike the other targets, this one is **not** a nowcast and it is **not** what
corrects the recent days — `estimate_truncation()` above does that. It answers a
different question: how long reporting takes. Like `tidy.epidist_fit()`, it
returns a **delay-shaped** table — one row per parameter, plus the
distribution's `mean` and `sd`:

```{r epinow2-dist-fit, eval=run_models && has("EpiNow2")}
dist_fit <- estimate_dist(
  covid_dist,
  stan = stan_opts(samples = 500, warmup = 500, chains = 2)
)
```

```{r epinow2-dist-tidy-1, eval=FALSE}
tidy(dist_fit)
```

```{r epinow2-dist-tidy, echo=FALSE, purl=FALSE}
show_tidy("EpiNow2_dist")
```

### With strata and effects.

`regional_epinow()` takes a single `region` column, so the object's strata are
folded into one label:

```{r epinow2-strata, eval=has("EpiNow2")}
covid_regional <- covid_seasonal |>
  filter(notification_date >= cutoff - 60) |>
  tbl_now_to_EpiNow2(target = "regional_epinow", verbose = FALSE, quiet = TRUE)

head(covid_regional)
```

```{r epinow2-strata-fit, eval=run_models && has("EpiNow2")}
regional_fit <- regional_epinow(
  covid_regional,
  generation_time = gt_opts(example_generation_time),
  delays          = delay_opts(example_incubation_period),
  # The same two-step logic as the pooled fit: without `truncation` this is not
  # a nowcast, just a smooth through the incomplete recent days. The truncation
  # fitted in step 1 above is reused here.
  truncation      = trunc_opts(fitted_truncation),
  rt              = rt_opts(prior = LogNormal(mean = 2, sd = 0.1)),
  stan            = stan_opts(samples = 500, warmup = 250, chains = 2)
)
```

`tidy()` on the result gives one block per region, so `stratum` stays a unique
key alongside `event_date`, and the pooled and stratified fits plot together
exactly as every other engine's do:

```{r epinow2-panels, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3.6, out.width="100%", fig.cap="Nowcast both stratified and total using the EpiNow2 package", fig.alt="Three panels (Total, Female, Male) comparing the EpiNow2 nowcast against the counts reported by now and the counts those dates eventually reached."}
nowcast_panels("EpiNow2")
```

::: {.alert .alert-info}
**EpiNow2's interval is much the widest here**, and that is the model rather
than the data: `estimate_infections()` fits a latent infection curve and
propagates the generation time, the incubation period *and* the fitted
truncation into every day, where the other engines model the reporting delay
alone. Read the width as the price of the extra structure.
:::

::: {.alert .alert-info}
**Two shapes EpiNow2 has that `tbl.now` does not convert for.**
`estimate_secondary()` models *two* data streams against each other (cases and
deaths, say) and one `tbl_now` is one stream, so there is no honest mapping.
`estimate_delay()` takes a bare vector of delays; EpiNow2's own help now points
at `estimate_dist()` instead, and it discards the censoring a `tbl_now` carries.
If you want it anyway it is `x$.delay`.
:::

## epidist

Sometimes the quantity you actually want is the **delay distribution** itself —
how long, on average, between onset and report, and how variable is it?
[`epidist`](https://epidist.epinowcast.org/) estimates exactly that, treating
each case as an interval-censored onset/report pair.

`tbl_now_to_epidist()` converts the `tbl_now` into the censored form `epidist`
expects. Our data are daily counts, so the converter produces
`epidist_aggregate_data` — one row per distinct `(delay, observation time)`
combination with a weight — rather than one row per case:

::: {.alert .alert-warning}
**Warning: requires Stan.** This package fits its model with **Stan**, so it needs a working `cmdstanr` (or `rstan`) toolchain installed before any of the code below will run.
:::

### Simple delay fit

```{r epidist-convert, eval=has("epidist"), message=FALSE, warning=FALSE}
library(epidist)

covid_epidist <- tbl_now_to_epidist(covid_now, verbose = FALSE)

```

`epidist` offers several model types, and on **count** data the choice matters
more than it looks. The **marginal** model is the one built for aggregated
counts: it works from the `(delay, observation time)` cells the converter
produces, so a month of cases costs a few hundred *weights* rather than a few
thousand rows.

```{r epidist-fit, eval=run_models && has("epidist")}
# Fit the delay distribution (see the epidist documentation for model choices)
delay_model <- covid_now |>
  filter(notification_date >= cutoff - 30) |>
  tbl_now_to_epidist(verbose = FALSE) |>
  as_epidist_marginal_model() |>
  epidist()
```

The fitted delay distribution can then feed back into a nowcast for example as
a data-informed prior in `epinowcast`.

### With strata and effects.

`epidist` has no separate grouping argument, so the strata (`sex`) and the
temporal-effect columns become extra columns that can be used within the  model's 
formula. For example:

```{r epidist-strata, eval=has("epidist"), message=FALSE, warning=FALSE}
covid_epidist_eff <- tbl_now_to_epidist(covid_seasonal, verbose = FALSE)
```

```{r epidist-strata-fit, eval=run_models && has("epidist")}
# A sex-varying mean delay.
delay_by_sex <- covid_seasonal |>
  filter(notification_date >= cutoff - 30) |>
  tbl_now_to_epidist(verbose = FALSE) |>
  as_epidist_marginal_model() |>
  epidist(formula = mu ~ 1 + sex + .event_season_365_sin + .event_season_365_cos)
```

`tidy()` works here too, but it returns a **different table**, because
`epidist` estimates a different thing. There are no per-date case estimates to
report, so instead of one row per event date you get one row per parameter of
the fitted delay distribution:

```{r epidist-tidy, eval=FALSE}
tidy(delay_model)
```


```{r epidist-caveat, echo=FALSE, purl=FALSE}
show_tidy("epidist")
```

::: {.alert .alert-warning}
The current model does not converge. Again this lies on how we specified the model
and in the amount of zero-day-delays we have (~34%). Read this as a demonstration of *how* 
to call the model from `tbl.now` rather than a demonstration of the model's effectiveness. 
:::


## The `tidy()` function

The converters normalize what goes *into* each package. `tidy()` normalizes what
comes out. Every engine above returns something different: a matrix of
draws, an `stsNC` object, a Stan fit, an INLA summary, a bare list. The `tidy()`
turns any of them into the same table:

```{r tidy-demo, message=FALSE, warning=FALSE, eval=FALSE}
tidy(nowcast_samples)
```

```{r tidy-demo-out, echo=FALSE, purl=FALSE}
show_tidy("baselinenowcast")
```


The columns are the same regardless of the package that produced the fit:

| column | meaning |
|---|---|
| `event_date` | event/reference date, **on the engine's own grid** |
| `stratum` | `"all"` when the fit is unstratified |
| `estimate` | point nowcast (posterior median where available) |
| `conf.low`, `conf.high` | interval bounds, using \pkg{broom}'s names |
| `level` | the width that interval **actually** has |
| `engine` | which package produced it |
| `qXX` | (optional) quantile columns of the prediction |

Pass `probs` for columns with other quantiles:

```{r tidy-probs, message=FALSE, warning=FALSE, eval=FALSE}
tidy(nowcast_samples, probs = c(0.05, 0.5, 0.95))
```

```{r tidy-probs-out, echo=FALSE, purl=FALSE}
show_tidy("baselinenowcast_probs")
```

::: {.alert .alert-warning}
**Only engines that keep draws can answer an arbitrary `probs`.** That is
`diseasenowcasting`, `baselinenowcast` and `epinowcast`. `NobBS` and
`surveillance` report a fixed set of summaries, so asking them for a quantile
they never computed is an **error** rather than a silent approximation. To be
able to do so you need to specify at **fit time** the quantiles you want. For example with
`NobBS` you can use the `specs` to set `quantiles`:

```{r nobbs-quantiles-1, eval=run_models && has("NobBS")}
nobbs_quantiles <- NobBS(
  data          = covid_linelist,
  now           = get_now(covid_now),
  units         = "1 day",
  onset_date    = "onset_date",
  report_date   = "report_date",
  max_D         = 10,
  moving_window = 104,
  specs         = list(quantiles = c(0.1, 0.5, 0.9))   # <- ask here
)
```

so that they can be called with `tidy`:

```{r nobbs-probs, message=FALSE, warning=FALSE, eval=FALSE}
tidy(nobbs_quantiles, probs = c(0.1, 0.5, 0.9))
``` 

```{r nobbs-probs-out, echo=FALSE, purl=FALSE}
show_tidy("NobBS_quantiles")
```
:::


## Summary


```{r comparison-all-code, eval=FALSE, echo=FALSE, purl=FALSE}
window <- meta$horizon
tru <- truth[truth$event_date %in% window, ]
est <- nowcasts[nowcasts$event_date %in% window, ]
tru$stratum <- factor(tru$stratum, levels = strata_levels)
est$stratum <- factor(est$stratum, levels = strata_levels)
est$package <- factor(est$package, levels = names(engine_colours))

ggplot(mapping = aes(x = event_date)) +
  geom_col(data = tru, aes(y = observed, fill = "Reported by now"), width = 0.8) +
  geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.7) +
  # Truth last, so it stays visible where the engines lie on top of it.
  geom_line(
    data = tru, aes(y = truth, linetype = "Truth"),
    linewidth = 0.7, colour = "#262626"
  ) +
  facet_wrap(~stratum, nrow = 1, scales = "free_y") +
  scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
  scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
  scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
  labs(x = "Event date (notification)", y = "Cases",
       title = "Every engine on the same data",
       subtitle = paste0("Now = ", format(meta$now))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.box = "horizontal",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335")) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE, order = 1))
```


```{r comparison-all, purl=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4.4, out.width="100%", fig.alt="Three panels (Total, Female, Male). Grey bars show the counts reported by now, a dark line the counts those dates eventually reached, and one coloured line per package its nowcast."}
window <- meta$horizon
tru <- truth[truth$event_date %in% window, ]
est <- nowcasts[nowcasts$event_date %in% window, ]
tru$stratum <- factor(tru$stratum, levels = strata_levels)
est$stratum <- factor(est$stratum, levels = strata_levels)
est$package <- factor(est$package, levels = names(engine_colours))

ggplot(mapping = aes(x = event_date)) +
  geom_col(data = tru, aes(y = observed, fill = "Reported by now"), width = 0.8) +
  geom_line(data = est, aes(y = estimate, colour = package), linewidth = 0.7) +
  # Truth last, so it stays visible where the engines lie on top of it.
  geom_line(
    data = tru, aes(y = truth, linetype = "Truth"),
    linewidth = 0.7, colour = "#262626"
  ) +
  facet_wrap(~stratum, nrow = 1, scales = "free_y") +
  scale_fill_manual(name = NULL, values = c("Reported by now" = "#dfe5e0")) +
  scale_linetype_manual(name = NULL, values = c("Truth" = "solid")) +
  scale_colour_manual(name = NULL, values = engine_colours, drop = FALSE) +
  labs(x = "Event date (notification)", y = "Cases",
       title = "Every engine on the same data",
       subtitle = paste0("Now = ", format(meta$now))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.box = "horizontal",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", colour = "#334335")) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE, order = 1))
```


We described the dengue data once as a `tbl_now`, and then a single converter call
(or, for `diseasenowcasting`, no call at all) handed it to each package in the
shape it needed:

```{r recap, eval=FALSE, purl=FALSE}
covid_now <- tbl_now(covid_colombia,
                     event_date  = notification_date,
                     report_date = diagnosis_date,
                     case_count  = n,
                     data_type   = "count-incidence")

covid_now                              # diseasenowcasting
tbl_now_to_baselinenowcast(covid_now)  # baselinenowcast
tbl_now_to_epinowcast(covid_now)       # epinowcast
tbl_now_to_epidist(covid_now)          # epidist
tbl_now_to_surveillance(covid_now)     # surveillance
tbl_now_to_nobbs(covid_now)            # NobBS
tbl_now_to_EpiNow2(covid_now)          # EpiNow2
as.data.frame(covid_now)               # others
```

Attaching **strata** and **temporal effects** once (`covid_seasonal`) uses the
same converters: each package receives them in whatever way it can use
a grouping in `epinowcast`, covariate columns in `epidist` and the
`baselinenowcast` long format, one triangle/series per stratum where the model
takes a single series, and automatically in `diseasenowcasting`. 

```{r learning-more, echo=FALSE, results="asis", purl=FALSE}
# The fragment is looked up by ABSOLUTE path via system.file(), never by a
# relative one. pkgdown/rmarkdown render into an intermediates directory under
# tempdir() and copy relative resources alongside; a path containing "../.."
# escapes that directory ("cannot create file '/tmp/RtmpXXX/../../man/...'"),
# which fails wherever tempdir() sits near the filesystem root -- as it does on
# CI. `system.file()` also works under pkgload::load_all(), which shims it to
# find inst/ in the source tree.
cat(
  knitr::knit_child(
    system.file("fragments", "learning-more.Rmd", package = "tbl.now"),
    quiet = TRUE
  ),
  sep = "\n"
)
``` 
