---
title: "Batch detection"
author: "tbl.now team"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Batch detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 3.6, fig.align = "center"
)
```

## What this vignette is about

Surveillance systems stall. A laboratory closes over a holiday, a data pipeline
breaks for a week, a reporting office falls behind and then catches up. When the
system restarts, a **backlog of reports arrives all at once**. We call that report
date a **batch**.

Batches matter because they look exactly like an epidemic taking off — a sudden
jump in reported cases — while being nothing of the sort. A nowcasting model that
assumes a stable reporting process will happily interpret a batch as a surge in
incidence.

This vignette shows how to find them. Everything here is **model-free**: no model
is fitted and nothing is estimated by a downstream nowcaster. You need only a
[`tbl_now`](https://rodrigozepeda.github.io/tbl.now/), which makes these tools
appropriate for exploratory data analysis *before* you commit to a model.

::: {.alert .alert-warning}
The batch-detection functions are **experimental**: the statistical behaviour and
the interface may still change, and each warns you when called. Treat a flagged
report date as a *potential* batch, not a confirmed one.
:::

The theory and its proofs live in the help pages — every function has a **"The
mathematics"** section (see `?batch_test`, `?batch_shape_test`,
`?simulate_batch`). This article is about *using* them.

```{r libraries, warning=FALSE, message=FALSE}
library(tbl.now)
library(dplyr)
library(ggplot2)
```

## The one idea you need

> **A batch moves reports. It does not create them.**

Suppose the desk was shut on Monday, Tuesday and Wednesday and reopened on
Thursday. Then Thursday reports its own cases *plus* three days of backlog. But if
you draw a box around Monday–Thursday and count everything inside, you get exactly
what you would have got with no closure at all. Nothing was invented; things merely
arrived on a different day.

A genuine epidemic surge is different: it *creates* cases, and the box count goes up.

`batch_test()` computes exactly these two quantities for every report date:

| quantity | meaning | what it detects |
|---|---|---|
| `deficit` | reports that went missing in the days before | **transport** (a batch) |
| `delta` | the box total minus what we expected | **creation** (a real surge) |

and reads the diagnosis off a 2×2 table:

| | `deficit` ≈ 0 | `deficit` ≫ 0 |
|---|---|---|
| `delta` ≈ 0 | nothing happened | **`"batch"`** |
| `delta` ≫ 0 | **`"surge"`** | `"batch_and_surge"` |
| `delta` < 0 | — | **`"hold_or_deletion"`** (the stall has not cleared) |

## Example 1 — Dengue with a planted batch

We start with real dengue data and *plant a batch ourselves*, so that we know the
right answer.

```{r dengue-data}
data(denguedat, package = "tbl.now")

dengue_tbl <- denguedat |>
  filter(report_week <= as.Date("1991-06-01")) |>
  tbl_now(
    event_date  = onset_week,
    report_date = report_week,
    data_type   = "linelist",
    verbose     = FALSE
  )
```

### Planting a batch

`simulate_batch()` closes the reporting desk on the dates you name and releases
everything on the next open date. It never changes an event date and never moves a
report *earlier* — it is a transport, exactly as the theory defines one.

```{r dengue-simulate}
closed_weeks <- as.Date(c("1990-06-04", "1990-06-11", "1990-06-18"))
release_week <- as.Date("1990-06-25")

dengue_batched <- simulate_batch(dengue_tbl, closed_dates = closed_weeks)
```

Here is what that did to the report axis. Note the three empty weeks and the spike
that pays for them.

```{r dengue-plot, echo=FALSE}
report_totals <- function(tbl_object, label) {
  as.data.frame(tbl_object) |>
    count(!!as.symbol(tbl.now::get_report_date(tbl_object)), name = "reported") |>
    rename(report_week = 1) |>
    mutate(series = label)
}

bind_rows(
  report_totals(dengue_tbl,     "original"),
  report_totals(dengue_batched, "with a planted batch")
) |>
  filter(report_week >= as.Date("1990-04-01"), report_week <= as.Date("1990-08-01")) |>
  ggplot(aes(x = report_week, y = reported, colour = series)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  labs(x = "Report week", y = "Reports that week", colour = NULL,
       title = "A batch: three weeks of silence, then a spike") +
  theme_minimal()
```

### Screening for it

```{r dengue-screen}
screened <- batch_test(dengue_batched, lookback = 3)
screened
```

`batch_test()` found the release week and nothing else. Look at what happened
around the closure:

```{r dengue-window}
screened |>
  filter(report_date >= as.Date("1990-06-04"), report_date <= as.Date("1990-07-02")) |>
  select(report_date, reported, baseline, deficit, delta, classification)
```

Three things to notice, and they are the whole method:

1. **During the closure** the verdict is `"hold_or_deletion"`. `delta` is strongly
   negative: reports have gone missing and no spike has yet arrived to pay for them.
   That is the correct read of a stall *in progress*.
2. **At the release week**, `reported` is far above `baseline` and `deficit` is large
   — but `delta` is small. The window total is essentially unchanged. Mass moved; it
   was not created. Verdict: `"batch"`.
3. **`deficit` is roughly the number of reports actually withheld.** It is not a
   p-value; it is an estimate of the size of the backlog.

### `delta` really is blind to the batch

The central claim of the theory is that `delta` cannot see a transport confined to
its window. This is not an approximation — it is a *pathwise* identity, and we can
check it exactly:

```{r dengue-pivot}
screened_clean <- batch_test(dengue_tbl, lookback = 3)

delta_clean   <- screened_clean$delta[screened_clean$report_date == release_week]
delta_batched <- screened$delta[screened$report_date == release_week]

c(clean = delta_clean, batched = delta_batched, identical = delta_clean == delta_batched)
```

Identical, to the last digit. Planting a batch changed the reported counts, the
baseline, and the deficit — and left `delta` exactly where it was. That is why
`delta` can be trusted to answer "was anything *created* here?" even in the presence
of a batch.

### Looking at *who* was reported, not just how many

`batch_test()` counts reports. `batch_shape_test()` asks a complementary question:
did the reports arriving on this date come from unusually **old** event dates? A
released backlog is old, so its delays should be inflated.

It compares the delays on the candidate date against those of nearby report dates,
using a permutation test. Set `guard` to skip the dates immediately around the
candidate — those are the deficit weeks, and including them would contaminate the
comparison.

```{r dengue-shape}
batch_shape_test(dengue_batched, at = release_week, guard = 3, seed = 1)
```

The mean delay at the release week is substantially longer than at its neighbours,
and the permutation test rejects. On the *original* data, the same date is
unremarkable:

```{r dengue-shape-clean}
batch_shape_test(dengue_tbl, at = release_week, guard = 3, seed = 1)
```

This test is remarkable for how little it assumes: it needs neither the reporting
delay distribution nor the epidemic curve. It only needs the epidemic to be
*locally log-linear*, which any smooth curve is. See the **"The mathematics"**
section of `?batch_shape_test` for why.

## Example 2 — Count-cumulative data (FluSight)

Some surveillance streams publish a running **cumulative** total for each week,
re-reported over time, which can be revised **down** as well as up. FluSight
hospitalisations are one such stream. Batch detection works there too — the theory
is about transport, and transport does not care whether the increments are signed.

```{r flusight-data}
data(flusight, package = "tbl.now")

flusight_tbl <- flusight |>
  filter(location_name == "California",
         as_of >= as.Date("2023-10-01"), as_of <= as.Date("2024-06-01"),
         observation >= 0) |>
  tbl_now(
    event_date  = target_end_date,
    report_date = as_of,
    case_count  = observation,
    data_type   = "count-cumulative",
    verbose     = FALSE
  )

get_data_type(flusight_tbl)
```

`batch_test()` de-accumulates the cumulative curve into the signed increment each
report announced, then proceeds exactly as before. Because the increments are signed,
the window total is a difference of counting processes rather than a count, so the
exact Poisson reference is replaced by a robust normal approximation. You do not have
to ask for this; it is chosen from the data type.

```{r flusight-screen}
flusight_batched <- simulate_batch(
  flusight_tbl,
  closed_dates = as.Date(c("2024-01-13", "2024-01-20"))
)

flusight_screened <- batch_test(flusight_batched, lookback = 2)
attr(flusight_screened, "null_model")

flusight_screened |>
  filter(report_date >= as.Date("2023-12-30"), report_date <= as.Date("2024-02-10")) |>
  select(report_date, reported, baseline, deficit, delta, classification)
```

The release date is flagged. But notice it is called `"batch_and_surge"`, and the
weeks around it are called `"surge"`. **That is correct, and worth understanding.**

Late January 2024 is the peak of the influenza season: cases really were being
created, fast. `delta` measures creation relative to a line extrapolated from outside
the window, and an exponential epidemic rise is not a straight line. So `"surge"`
fires — on the *original* data too:

```{r flusight-clean}
flusight_clean <- batch_test(flusight_tbl, lookback = 2)

peak_week <- as.Date("2024-01-27")
tibble(
  quantity = c("delta", "deficit"),
  clean    = c(flusight_clean$delta[flusight_clean$report_date == peak_week],
               flusight_clean$deficit[flusight_clean$report_date == peak_week]),
  batched  = c(flusight_screened$delta[flusight_screened$report_date == peak_week],
               flusight_screened$deficit[flusight_screened$report_date == peak_week])
)
```

Read that table carefully, because it is the method in miniature:

* **`delta` is identical** in the clean and batched data. The epidemic peak is real
  and `delta` sees it; the batch is invisible to `delta`.
* **`deficit` swings from negative to strongly positive.** The batch is entirely
  visible to `deficit`.

So `"surge"` came from the flu peak and `"batch"` came from our planted stall. The
classifier is telling the truth about both.

::: {.alert .alert-info}
**The surge verdict is the weaker one.** The transport test conditions on the window
total, so it does not care what the underlying intensity is or how good the baseline
is. The creation test compares against the baseline directly, and is therefore only
as good as it. On a steeply curving epidemic curve, `"surge"` will fire on ordinary
growth. If you want to detect genuine surges, fit a model. If you want to detect
batches, this screen is the right tool.
:::

## Example 3 — Weekends, holidays and other schedules

A reporting system that is *always* closed at weekends produces every batch symptom,
every single week. This is a **scheduled** transport, and it is not what we are
hunting. Left unmodelled it will flood the screen with false positives.

Here is daily data whose weekend reports are pushed to Monday.

```{r schedule-data}
set.seed(42)

origin_dates <- seq(as.Date("2021-01-04"), by = "day", length.out = 120)
report_rows  <- vector("list", length(origin_dates))

for (origin_index in seq_along(origin_dates)) {
  origin_date <- origin_dates[origin_index]
  n_cases     <- rpois(1, 30)
  if (n_cases == 0) next

  report_date <- origin_date + rgeom(n_cases, prob = 0.5)
  # The desk is shut at weekends: those reports slide to the following Monday.
  weekday     <- as.POSIXlt(report_date)$wday
  report_date <- report_date + ifelse(weekday == 6, 2, ifelse(weekday == 0, 1, 0))

  report_rows[[origin_index]] <- data.frame(onset = origin_date, report = report_date)
}

scheduled_tbl <- bind_rows(report_rows) |>
  filter(report <= as.Date("2021-05-01")) |>
  tbl_now(event_date = onset, report_date = report,
          data_type = "linelist", verbose = FALSE)
```

Screen it twice — once naively, once telling `batch_test()` that the cycle is seven
days long:

```{r schedule-compare}
naive    <- batch_test(scheduled_tbl, lookback = 3)
adjusted <- batch_test(scheduled_tbl, lookback = 3, period = 7)

c(
  "flagged without `period`" = sum(naive$batch,    na.rm = TRUE),
  "flagged with `period = 7`" = sum(adjusted$batch, na.rm = TRUE)
)
```

Every Monday is a phantom batch until the schedule is modelled; afterwards, none of
them are. `batch_test()` estimates the effect of each phase of the cycle by taking
**medians across cycles**, which recovers the schedule as long as irregular batches
hit any given phase in fewer than half of its cycles.

Now plant a *real*, irregular four-day stall on top of the weekly schedule and see
whether it can still be found:

```{r schedule-batch}
irregular_closure <- as.Date(c("2021-03-08", "2021-03-09", "2021-03-10", "2021-03-11"))

scheduled_batched <- simulate_batch(scheduled_tbl, closed_dates = irregular_closure,
                                    verbose = FALSE)

batch_test(scheduled_batched, lookback = 4, period = 7) |>
  filter(batch) |>
  select(report_date, reported, baseline, deficit, delta, classification)
```

The release date (12 March) is recovered. **Always pass `period` when your reporting
system has a fixed cycle.**

## Practical guidance

**Choosing `lookback`.** This is the width of the box. It must comfortably cover the
longest stall you think plausible: a three-day closure needs `lookback >= 3`. Too
small and the deficit falls outside the window; too large and you lose power, because
the box total gets noisier. When in doubt, try a few values — a real batch survives.

**Choosing `baseline_window`.** It must satisfy `baseline_window >= 2 * lookback + 3`,
which `batch_test()` enforces, so that a batch episode can never outvote the robust
baseline it is being measured against. The default is the smallest admissible value.

**Reading the output.**

* `batch` is the flag you should act on: it applies a Benjamini–Hochberg correction
  across every (report date, stratum) pair.
* `classification` uses raw p-values, so it is more liberal. It is there to tell you
  *which kind* of anomaly, not *whether*.
* Flags often cluster around an episode. One stall may light up its release date and a
  neighbour; that is expected, not a second batch.

**Strata.** A batch is a report date that is anomalous *for at least one stratum* — one
laboratory can stall while others do not. `batch_test()` returns one row per (report
date, stratum) and corrects across all of them.

**What this will not do.** If a stall has not cleared by the end of your data, you see
the lull and never the spike. Within the data you have, that is *indistinguishable*
from cases being deleted — an honest impossibility, not a failure of the method.
`batch_test()` labels it `"hold_or_deletion"` rather than guessing.

**A batch is not a single strange report.** It is tempting to look for one case
with a shockingly long delay. But a batch produces **many cases with mildly
inflated delays** — a whole backlog, each a few days late. Finding one huge
departure and finding many small ones are different statistical problems: the
former wants a maximum, the latter wants to *accumulate* the small pieces of
evidence, which is exactly what the volume/deficit contrast and the rank-sum shape
test do. A per-report outlier rule (useful in its own right for spotting a single
mis-recorded delay) is the wrong instrument here.

Use each tool for its question:

| question | tool |
|---|---|
| Did *this report date* release a backlog? | `batch_test()` |
| Did *this report date* draw on unusually old cases? | `batch_shape_test()` |
| Is *this individual report* implausibly delayed? | see `test_delay_drift()` and downstream model-based checks |

## See also

* `?batch_test`, `?batch_shape_test`, `?simulate_batch` — each help page has a
  **"The mathematics"** section deriving the method.
* [`test_delay_drift()`](../reference/test_delay_drift.html) and
  [`test_delay_changepoint()`](../reference/test_delay_changepoint.html) — the
  complementary question of whether the reporting *delay* is drifting or has a
  change point.
* The [Get started](tbl.now.html) article — building and manipulating a `tbl_now`.
